### 2020-04-14 manu_draft.Rmd
manuscript rmd template, started methods

---
title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
# bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = T, dev.args = list(family = 'serif'), dpi = 300, warning = F, cache.path = 'manu_draft_cache/',
  fig.process = function(x) {
  x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
  if (file.rename(x, x2)) x2 else x
  })
# libraries
library(Jabbrev)
library(tidyverse)
library(wqtrends)

# extract bib entries from online
# bib_scrp('manu_draft.Rmd', 'refs.bib')
```

`r paste('Compiled', Sys.time())`

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files('manu_draft.Rmd')
```

# Abstract

# Introduction

# Methods

## Study area and data sources

The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America and drains an area of approximately 200 thousand km$^2$ in the US state of California.  The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisan Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.  Major riverine inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.  The northern subembayments are river-dominanted (salinity ranging from 0 to 15 ppt) with seasonal freshwater pulses primarily occurring in the winter and spring during the rainy season and subsequent snowmelt in the upper watershed.  The southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on diurnal variation in the tidal cycle and effluent discharge from wastewater treatment plants.  South Bay is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.  Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.  Agricultural runoff from the upper watershed conctributes also contributes to nutrient loading in the SFE with annual nutrient export estimated as approximately 30 thousand kg dy$^-1$ of Nitrogen from the Delta.

Nitrogen and phospohorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.  However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.  The resistance of SFE to eutrophication has been attributed to several factors, including elevated susepended sediments that reduce light penetration in the water column, regular exchange with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.  Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where recent harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999.  These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with a purpose among others to devleop a framework for assessineg water quality condition.

* SF south Bay dataset, USGS monitoring data, show observed data summer/fall IQR plots (is the trend significant? when did it start or stop being significant?)

Analysis focused on near-surface (depth = 1-3m) chl-a data from Central Bay, South Bay, and Lower South Bay (stations 18-36, Figure 1), collected biweekly to monthly through long-term monitoring by USGS (Cloern and Schraga, 2016; Schraga et al.2017). For this initial work, we selected the time period 1990-2017 because it represented a suitable balance among three factors relevant to testing the statistical approaches: a. sufficient length of record; b. consistent biweekly-monthly sampling; and c. a diverse set of stations, across multiple subembayments, for which a and b were satisfied. While sampling frequency varied somewhat overtime or by station, all data were treated as individual date/station/concentration records within the statistical models (no spatial or temporal binning or averaging).

## GAM model format

* GAM general format and structures (gam0, gam1, gam2, gam6)

* Model application to SF time series and model evaluation, including transformation for response variables (why Box-Cox?)

*  Model predictions (complete time series, annual trend) and back-transformation estimates

## Long-term and seasonal trend analysis

* Secondary methods for trend analysis, estimate of seasonal values, hypothesis tests including mixed meta-analysis

## Sensitivity analysis

* Sensitivity analysis - do trend estimates or hypothesis test results change by model type?

# Results

# Discussion

# Conclusions

# Acknowledgments

# Figures

# Tables

# References

### 2020-04-14 manu_draft.Rmd
working on gams description, general

@@
 The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisan Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.
-Major riverine inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.
-The northern subembayments are river-dominanted (salinity ranging from 0 to 15 ppt) with seasonal freshwater pulses primarily occurring in the winter and spring during the rainy season and subsequent snowmelt in the upper watershed.
-The southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on diurnal variation in the tidal cycle and effluent discharge from wastewater treatment plants.
+Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.
+The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt) with seasonal freshwater pulses primarily occurring in the winter and spring during the rainy season and subsequent snowmelt in the upper watershed.
+The southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
 South Bay is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.
 Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.
-Agricultural runoff from the upper watershed conctributes also contributes to nutrient loading in the SFE with annual nutrient export estimated as approximately 30 thousand kg dy$^-1$ of Nitrogen from the Delta.
+Agricultural runoff from the upper watershed also contributes to nutrient loading in the SFE with the annual nutrient export estimated as approximately 30 thousand kg dy$^{-1}$ of nitrogen from the Delta.
 
@@
 However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.
-The resistance of SFE to eutrophication has been attributed to several factors, including elevated susepended sediments that reduce light penetration in the water column, regular exchange with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.
-Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where recent harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999.
-These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with a purpose among others to devleop a framework for assessineg water quality condition.
+The resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.
+Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999.
+These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with the purpose, among others, to develop a framework for assessing water quality condition.
+Critical questions that were to be addressed by the NMS included identifying water quality conditions to be considered healthy and protective of beneficial uses, how do current conditions compare to desirable water quality conditions, and what potential management actions are needed to achieve desired conditions if they currently do not persist.
+Although visual changes in observed data were apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed.
 
-* SF south Bay dataset, USGS monitoring data, show observed data summer/fall IQR plots (is the trend significant? when did it start or stop being significant?)
+The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 (Cloern and Schraga, 2016; Schraga et al.2017).
+This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.
+Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients.
+The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24 - 32), and Lower South Bay (stations 34-36).
+Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters.
+Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.
+While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).
 
-Analysis focused on near-surface (depth = 1-3m) chl-a data from Central Bay, South Bay, and Lower South Bay (stations 18-36, Figure 1), collected biweekly to monthly through long-term monitoring by USGS (Cloern and Schraga, 2016; Schraga et al.2017).
-For this initial work, we selected the time period 1990-2017 because it represented a suitable balance among three factors relevant to testing the statistical approaches: a.
-sufficient length of record; b.
-consistent biweekly-monthly sampling; and c.
-a diverse set of stations, across multiple subembayments, for which a and b were satisfied.
-While sampling frequency varied somewhat overtime or by station, all data were treated as individual date/station/concentration records within the statistical models (no spatial or temporal binning or averaging).
+## GAM application
 
-## GAM model format
+To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program, including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.
+These methods leverage functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods.
+In generals, GAMs are an extension of linear models that can incorporate additive smoothed tersm to model non-linear effects.
+The basic structural form is (Hastie and Tibshirani 1986):
+
+$$
+E\left(Y|X_1, X_2, ..., X_p\right) = s_0 + \sum_{j=1}^{p} s_j\left(X_j\right) 
+$$
+
+where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an interecept term $s_0$.
+The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.
+Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.
+
+The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.
+A spline is a piecewise function, such as a polynomial, that is connected at knots that smoothly join the separate pieces of the function.
+Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.
+There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to esitmating the smoother.
+The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).
+In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty.
+This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function.
+However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.
+The default is ten, and as described below, this default value was increased to prevent under-fitting the data.
+
+An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location.
+Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is a useful form for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.
+A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of component being modelled.
+This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.
+In others words, not using a cyclic cubic spline to modeal season would cause a non-continous overall function with step increases or decreases for the seasonal estimates between the years.
+This discrepancy can likely lead to inaccurate trend estimates when evaluating seasonal differences over time.
+
+Because GAM structures are flexible, the functional form for time can be described wih varying levels of complexity depending on the desired structural form.
+The method is useful for time series analysis because separate components can be explicitly modelled with differnet smoothing tersm.
+
+using some of the model structures they have documented and incorporated into their R package, baytrends.
+Table 1 summarizes the four primary model structures tested.
+The first three models (gam0, gam1, gam2) have the same structure as CBP’s base models.
+We added a fourth model, gam6.
+For all models, date is expressed as cyear, or centered decimal date, meaning that a date is turned into a decimal (i.e., 2002.41), and then a time series is centered so that the middle date in a record becomes zero.
+doy is the day of year as a numeric value from 1 to 366.
+All models include year as a linear effect.
+The functions s() model either year or doy as a smoothed, non-linear variable and ti() models the interaction between the two in the gam2 model.
+A central challenge to optimizing GAMs for applications such as this, where time series data is rather noisy with significant inner- and interannual seasonal variation, is estimating how smooth (vs.
+‘wiggly’) a spline should be to fit the data without over-fitting (i.e.
+following noise).
+In the mgcv package this is done by GCV.
+The fourth model, gam6, has the same structure as gam1 but permits a much higher degree in the s(year) by increasing the maximum allowable number of knots, k, and relies on mgcv’s internal model selection statistics to determine the appropriate degree of smoothness (using GCV).
 

### 2020-04-15 manu_draft.Rmd
more gams description

@@
 
-To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program, including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.
-These methods leverage functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods.
-In generals, GAMs are an extension of linear models that can incorporate additive smoothed tersm to model non-linear effects.
-The basic structural form is (Hastie and Tibshirani 1986):
+Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is (Hastie and Tibshirani 1986):
 
-$$
+\begin{equation}
 E\left(Y|X_1, X_2, ..., X_p\right) = s_0 + \sum_{j=1}^{p} s_j\left(X_j\right) 
-$$
+(\#eq:gengam)
+\end{equation}
 
@@
 Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.
+All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods (Wood, 2017).
 
 The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.
-A spline is a piecewise function, such as a polynomial, that is connected at knots that smoothly join the separate pieces of the function.
+A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.
 Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.
-There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to esitmating the smoother.
+There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother (Wood, 2003).
 The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).
@@
 This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.
-In others words, not using a cyclic cubic spline to modeal season would cause a non-continous overall function with step increases or decreases for the seasonal estimates between the years.
-This discrepancy can likely lead to inaccurate trend estimates when evaluating seasonal differences over time.
+In others words, not using a cyclic cubic spline to model season would cause a non-continous function across the time series with step increases or decreases for the seasonal estimates between the years.
+This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences over time.
 
-Because GAM structures are flexible, the functional form for time can be described wih varying levels of complexity depending on the desired structural form.
-The method is useful for time series analysis because separate components can be explicitly modelled with differnet smoothing tersm.
+To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program (Murphy et al.
+2019), including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.
+These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.
+This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.
+Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.
+We present four GAM structures, following a similar format as Murphy et al.
+(2019):
 
-using some of the model structures they have documented and incorporated into their R package, baytrends.
-Table 1 summarizes the four primary model structures tested.
-The first three models (gam0, gam1, gam2) have the same structure as CBP’s base models.
-We added a fourth model, gam6.
-For all models, date is expressed as cyear, or centered decimal date, meaning that a date is turned into a decimal (i.e., 2002.41), and then a time series is centered so that the middle date in a record becomes zero.
-doy is the day of year as a numeric value from 1 to 366.
-All models include year as a linear effect.
-The functions s() model either year or doy as a smoothed, non-linear variable and ti() models the interaction between the two in the gam2 model.
+\begin{equation}
+\boldsymbol{gam0:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + \epsilon 
+(\#eq:gamzero)
+\end{equation}
+
+\begin{equation}
+\boldsymbol{gam1:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
+(\#eq:gamone)
+\end{equation}
+
+\begin{equation}
+\boldsymbol{gam2:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
+(\#eq:gamtwo)
+\end{equation}
+
+\begin{equation}
+\boldsymbol{gam6:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2^*\left(year\right) + \epsilon
+(\#eq:gamsix)
+\end{equation}
+
+For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$).
+The $year$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).
+The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series.
+The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models.
+The $s_1()$ and $s_2()$ functions are different smoothers for $year$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.
+The $ti_1()$ function is a tensor product interaction between $year$ and $doy$ that allows the seasonal component to change between years.
+Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline.
+As with conventional linear models, the remaining variance in the response not captured by the functional form of the model is expressed as an error term $\epsilon$ that is normally-distributed having mean equal to zero and variance as $\sigma^2$.
+The residual variance is used to back-transform model estimates for trend testing, described below.
+
+The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.
+The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable.
+However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameter by better satisying the model assumptions.
+The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).
+These methods...
+A comparison of fit between the two methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.
+
 A central challenge to optimizing GAMs for applications such as this, where time series data is rather noisy with significant inner- and interannual seasonal variation, is estimating how smooth (vs.
@@
 following noise).
+The four models above, Table 1....
 In the mgcv package this is done by GCV.

### 2020-04-16 manu_draft.Rmd
general gam methods done

@@
 However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.
-The default is ten, and as described below, this default value was increased to prevent under-fitting the data.
 
@@
 We present four GAM structures, following a similar format as Murphy et al.
-(2019):
+(2019).
+The notation for gam3, gam4, and gam5 are not used because describe models used in Murphy et al.
+(2019) that were developed for purposes specific to Chesapeake Bay.
 
@@
 The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.
+A comparison of GAM perforamcne between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.
 The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable.
-However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameter by better satisying the model assumptions.
+However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.
 The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).
-These methods...
-A comparison of fit between the two methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.
+The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.
+The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package (Venables and Ripley 2002) to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
+Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation (Box and Cox 1964):
 
-A central challenge to optimizing GAMs for applications such as this, where time series data is rather noisy with significant inner- and interannual seasonal variation, is estimating how smooth (vs.
-‘wiggly’) a spline should be to fit the data without over-fitting (i.e.
-following noise).
-The four models above, Table 1....
-In the mgcv package this is done by GCV.
-The fourth model, gam6, has the same structure as gam1 but permits a much higher degree in the s(year) by increasing the maximum allowable number of knots, k, and relies on mgcv’s internal model selection statistics to determine the appropriate degree of smoothness (using GCV).
+\begin{equation}
+g\left(chl;\lambda\right) = \begin{cases} \frac{chl ^ {\lambda} - 1}{\lambda} & \lambda \ne 0 \\ \log\left(chl\right) & \lambda = 0 \end{cases}
+(\#eq:boxcox)
+\end{equation}
 
-* GAM general format and structures (gam0, gam1, gam2, gam6)
+where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value.
+Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.
+The *BoxCox* function from the forecast package (Hyndman et al.
+2020) in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.
+Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.
 
-* Model application to SF time series and model evaluation, including transformation for response variables (why Box-Cox?)
+A challenge to optimizing GAMs for time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
+The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency.
+However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.
+This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.
+Initial testing by Murphy et al.
+(2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
+We follow the same approach herein for gam1 and gam2.
+Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further.
+For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
+This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.
 
-* Model predictions (complete time series, annual trend) and back-transformation estimates
+All models were compared using standard summary statistics describing overall model fit to the observed data.
+In addition to the GCV score, models were comapred using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
+Comparisons were made between the different GAMs in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) and the transformation used to define the $g()$ function for chlorophyll at each station.
+Model predictions were based on the standard model output from the *predict()* function from the mgcv package.
+A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions.
+Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used.
 

### 2020-04-20 manu_draft.Rmd
cloern reminder for intro
getting through hypothesis tests for gams

@@
 # Introduction
+
+Cloern et al. 2020 [link](https://link.springer.com/article/10.1007/s12237-020-00737-w)
 
@@
 All models were compared using standard summary statistics describing overall model fit to the observed data.
-In addition to the GCV score, models were comapred using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
-Comparisons were made between the different GAMs in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) and the transformation used to define the $g()$ function for chlorophyll at each station.
+In addition to the GCV score, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
+Comparisons were made between the different GAMs in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) and for the transformations used to define the $g()$ function for chlorophyll at each station.
 Model predictions were based on the standard model output from the *predict()* function from the mgcv package.
@@
 Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used.
+For hypothesis tests that required estimates of central tendency and confidence intervals for mean values, back-transformations were based on an exponentiation or inverse function of the transformed response variable as before, with the addition of a correction factor.
+This bias-correction factor was simply the variance of the model residuals divided by two for variables in log-space and correction with a forecast variance estimate as specified in the *InvBoxCox()* function from the forecast package.
 
-## Long-term and seasonal trend analysis
+## Trend tests on GAM results
+
+Each of the GAM structures described in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) provide different approaches to quantify long-term variation in annual and seasonal changes.
+Although useful as a general approach to view functional forms for different time series, additional information regarding whether an observed change is statistically significant can be useful to inform environmental management decisions.
+We present two methods for performing trend tests on model results that address different changes that may be of management interest.
+First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized.
+This approach is similar to that presented in Murphy et al.
+(2019).
+Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.
+
+The ability to compare time periods of interest has importance for evaluating whether a significant change has occurred in relation to perceived trends in observed data.
+For example, the noticeable change in SFE water quality in South Bay starting from the 1990s and persisting through 2005 was of considerable interest by environmental managers given the apparent departure from long-term conditions to a state that suggested a trend towards more eutrophic conditions.
+However, statistical methods to evaluate this change that were able to quantify the temporal variation and to provide a statistically robust estimate of the change were not available.
+The use of GAMs in parallel with hypothesis testing can be used for this purpose.
+Methods proposed in Murphy et al.
+(2019) and similarly applied in Yang and Moyer (2020) were used and are briefly summarized herein.
+
+For two time periods of interest, a trend can be estimated as the difference in means from the GAM results between the periods, the percent change, and an appropriate p-value descibing the probability that the change is due to change alone.
+Murphy et al.
+(2019) defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.
+It was further suggested the the base and test periods include a minimum of two years if evaluating years at the beginning and end of a time series to account for instability of GAM estimates at the edges of the independent variable space.
+The differences are estimated as:
+
+\begin{equation}
+\text{Difference} = Z_d\hat{\beta}
+(\#eq:diffest)
+\end{equation}
+
+where $Z_d$ is a point estimate of the differences of the predictions between the base and test years and $\hat{\beta}$ is the estimated parameter vector from the fitted GAM.
+The standard error of this difference is then:
+
+\begin{equation}
+\text{se}_{\text{diff}} = \sqrt{Z_d\hat{\Sigma}_\beta Z_d^T}
+(\#eq:sediffeest)
+\end{equation}
+
+defined as the square root of the product of $Z_d$, the estimated variance-covariance matrix of parameter vector for the fitted GAM $\hat{\Sigma}_\beta$, and the transposition of $Z_d$.
+The trend is considered significant if the appropriate confidence interval at a specified $\alpha$ does not include zero.
+
+The second hypothesis test evaluates changes in seasonal estimates between years using results from a fitted GAM.
+This test provides both an estimate of a seasonal mean and its associated confidence interval and if the rate of change in the seasonal means has varied over a specific period interest.
+For example,
 

### 2020-04-29 manu_draft.Rmd
methods text draft done

@@
 Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients.
-The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24 - 32), and Lower South Bay (stations 34-36).
+The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36).
 Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters.
@@
 (2019).
-The notation for gam3, gam4, and gam5 are not used because describe models used in Murphy et al.
+The notation for gam3, gam4, and gam5 are not used because these describe models in Murphy et al.
 (2019) that were developed for purposes specific to Chesapeake Bay.
@@
 
-A challenge to optimizing GAMs for time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
+A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
 The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency.
@@
 The second hypothesis test evaluates changes in seasonal estimates between years using results from a fitted GAM.
-This test provides both an estimate of a seasonal mean and its associated confidence interval and if the rate of change in the seasonal means has varied over a specific period interest.
-For example,
+This test provides both an estimate of a seasonal mean, its associated confidence interval, and if the rate of change in the seasonal means has varied over a specific period of interest.
+For example, changes that occur during late fall or early spring can be evaluated between years to determine if the seasonal averages differ and if the direction and magnitude of change is significant.
+This has implications for understanding potential changes in seasonal bloom phenology resulting from shifting system responses to nutrient inputs or changes in the latter.
 
-* Secondary methods for trend analysis, estimate of seasonal values, hypothesis tests including mixed meta-analysis
+Seasonal averages and confidence intervals were estimated from fitted GAMs by first creating a matrix of the predictors (i.e., `year`, `doy`) covering the seasonal periods across the time series.
+For example, seasonal changes from late summer could be obtained from a prediction matrix covering all dates from Aug 1st to Sep.
+30th for the period from 2000 to 2010.
+The prediction matrix is then used to obtain fitted values from a GAM that describe the estimated values of chlorophyll that occurred during the periods of interest.
+Rather than the estimated values for the predictions, a linear predictor matrix is returned that provides a parameter vecotr at the supplied covariate values.
+This predictor matrix is then used to estimate the mean seasonal value within a year and associated variance using the sum of predictions from the matrix.
+As a result, each year has a mean value of chlorophyll with an associated confidence interval that describes the modelled expectation for the seasonal period.
 
-## Sensitivity analysis
+Although the seasonal averages obtained from GAMs describe the expected values for each year, describing the direction and magnitude of the seasonal change across years may also be of interest.
+Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected value of chlorophyll and the predictor is year as a numeric value.
+Regression analysis could be used to evaluate changes in the seasonal averages described above, although the fit from the regression would under-estimate the uncertainty in the model parameters by not accounting for the variance in the seasonal averages.
+To better describe the trends in the seasonal averages, we use mixed effects regression meta-analysis to evaluate seasonal changes and their uncertainty over a finite set of years.
+This analysis is an extension of conventional regression by accounting for different soures of uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:
 
-* Sensitivity analysis - do trend estimates or hypothesis test results change by model type?
+\begin{equation}
+chl_i = \beta_0 + \beta_1 year_i + b_i + \epsilon_i
+(\#eq:mixmet)
+\end{equation}
+
+where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and devation of the seasonal mean from the true mean ($S_i$).
+Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and vraiance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.
+The R package *mixmeta* was used to estimate trends in the seasonal averages over time (Sera et al.
+2019).
+Results appear similar to those from regression analysis (e.g., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.
+
+## Model application and interpretation
+
+Each of the four GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.
+For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoothers terms.
+Models were also evaluated using different transformations to determine which had improved fit for specific time series.
+Secondary analyses evaluating trends with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.
+Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.
 

### 2020-05-28 manu_draft.Rmd
started bib file, outline slight update

@@
     reference_docx: my_styles.docx
-# bibliography: refs.bib
+bibliography: refs.bib
 author: "Marcus W.
@@
 
-# extract bib entries from online
+# # extract bib entries from online
 # bib_scrp('manu_draft.Rmd', 'refs.bib')
@@
 The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America and drains an area of approximately 200 thousand km$^2$ in the US state of California.
-The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisan Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.
+The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisun Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.
 Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.
@@
 
-The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 (Cloern and Schraga, 2016; Schraga et al.2017).
+The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20].
 This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.
@@
 
-Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is (Hastie and Tibshirani 1986):
+Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is [@Hastie90]:
 
@@
 Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.
-All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods (Wood, 2017).
+All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods [@Wood17].
 
@@
 Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.
-There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother (Wood, 2003).
+There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother [@Wood03].
 The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).
@@
 
-To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program (Murphy et al.
-2019), including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.
+To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19], including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.
 These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.
@@
 Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.
-We present four GAM structures, following a similar format as Murphy et al.
-(2019).
-The notation for gam3, gam4, and gam5 are not used because these describe models in Murphy et al.
-(2019) that were developed for purposes specific to Chesapeake Bay.
+We present four GAM structures, following a similar format as @Murphy19.
+The notation for gam3, gam4, and gam5 are not used because these describe models in @Murphy19 that were developed for purposes specific to Chesapeake Bay.
 
@@
 The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.
-The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package (Venables and Ripley 2002) to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
-Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation (Box and Cox 1964):
+The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02 to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
+Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:
 
@@
 Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.
-The *BoxCox* function from the forecast package (Hyndman et al.
-2020) in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.
+The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.
 Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.
@@
 First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized.
-This approach is similar to that presented in Murphy et al.
-(2019).
+This approach is similar to that presented in @Murphy19.
 Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.
@@
 The use of GAMs in parallel with hypothesis testing can be used for this purpose.
-Methods proposed in Murphy et al.
-(2019) and similarly applied in Yang and Moyer (2020) were used and are briefly summarized herein.
+Methods proposed in @Murphy19 and similarly applied in @Yang20 were used and are briefly summarized herein.
 
 For two time periods of interest, a trend can be estimated as the difference in means from the GAM results between the periods, the percent change, and an appropriate p-value descibing the probability that the change is due to change alone.
-Murphy et al.
-(2019) defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.
+@Murphy19 defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.
 It was further suggested the the base and test periods include a minimum of two years if evaluating years at the beginning and end of a time series to account for instability of GAM estimates at the edges of the independent variable space.
@@
 where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and devation of the seasonal mean from the true mean ($S_i$).
-Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and vraiance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.
-The R package *mixmeta* was used to estimate trends in the seasonal averages over time (Sera et al.
-2019).
+Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.
+The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].
 Results appear similar to those from regression analysis (e.g., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.

### 2020-05-28 manu_draft.Rmd
added fig 1, table 1

@@
     reference_docx: my_styles.docx
+always_allow_html: true
 bibliography: refs.bib
@@
 
-For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$).
+For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
 The $year$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).
@@
 A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
-The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency.
+The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
 However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.
@@
 
+```{r modsumfig, fig.height = 4.5, fig.width = 6, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
+Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
+# get predictions for all four gams
+tomod <- rawdat %>%
+  filter(station %in% 22) %>%
+  filter(param %in% 'chl')
+
+# use previously fitted list of models
+trans <- 'boxcox'
+mods <- list(
+  gam0 = anlz_gam(tomod, mod = 'gam0', trans = trans),
+  gam1 = anlz_gam(tomod, mod = 'gam1', trans = trans),
+  gam2 = anlz_gam(tomod, mod = 'gam2', trans = trans),
+  gam6 = anlz_gam(tomod, mod = 'gam6', trans = trans)
+)
+
+p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 2, size = 0.8, alpha = 0.8) + 
+  theme(legend.position = 'right') + 
+  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
+  scale_colour_gradient(low = "white", high = "black") + 
+  theme_minimal() + 
+  theme(
+    legend.title = element_blank()
+  )
+
+p1
+```
+
 # Tables
 
+```{r modsumtab}
+tab <- tibble(
+  GAM = c('`gam0`', '`gam1`', '`gam2`', '`gam6`'),
+  Summary = c(
+    'Linear trend with seasonality', 
+    'Nonlinear trends seasonality (constrained knots)', 
+    'Nonlinear trends wiht seasonality, plus interaction',
+    'Nonelinear trends with seasonality (unconstrained knots)'
+  ),
+  Details = c(
+    'A linear trend in time over the course of the time series (cyear); and Within-year seasonal fluctuations that follow the same pattern every year - e.g.
+large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
+    'Treats cyear and doy same as gam0; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
+    'Treats cyear and doy same as gam1; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to gam6 (`ti(year, doy, bs = c("tp", "cc"))`)', 
+    'Variant of gam1 with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
+  )
+
+# table stuff
+cap.val <- 'Summary and details for each of the GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).'
+
+# table
+knitr::kable(tab, booktabs = T, caption = cap.val)
+```
+
 # References

### 2020-05-29 manu_draft.Rmd
table added for station info

@@
   })
+
 # libraries
@@
 library(wqtrends)
+library(flextable)
+library(officer)
+
+data(locs)
+
+locs <- locs %>% 
+  rename(station = Station)
 
@@
 Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients.
-The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36).
+The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab)).
 Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters.
@@
 
+```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salnity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.
+Rows are arranged from north to south.'}
+totab <- rawdat %>% 
+  filter(param %in% c('sal', 'din', 'chl')) %>% 
+  group_by(station, param) %>% 
+  summarise(
+    n = n(), 
+    medv = median(value, na.rm = T)
+  ) %>% 
+  ungroup() %>% 
+  gather('var', 'val', n, medv) %>% 
+  unite('var', param, var) %>% 
+  mutate(var = factor(var, levels = c('sal_n', 'sal_medv', 'din_n', 'din_medv', 'chl_n', 'chl_medv'))) %>% 
+  left_join(locs, by = 'station') %>% 
+  spread(var, val)
+
+tab <- flextable(totab) %>% 
+  set_header_labels(
+    station = 'Station', 
+    lat = 'Latitude', 
+    lon = 'Longitude', 
+    sal_n = 'n', 
+    sal_medv = 'Med.', 
+    din_n = 'n', 
+    din_medv = 'Med.', 
+    chl_n = 'n', 
+    chl_medv = 'Med.'
+  ) %>% 
+  add_header_row(values = c('', '', '', 'Salinity (psu)', '', 'DIN (mg/L)', '', 'Chl-a (ug/L)', '')) %>% 
+  merge_at(i = 1, j = 4:5, part = 'header') %>% 
+  merge_at(i = 1, j = 6:7, part = 'header') %>% 
+  merge_at(i = 1, j = 8:9, part = 'header') %>% 
+  colformat_num(j = c(5, 7, 9), digits = 1) %>%
+  colformat_num(j = c(2, 3), digits = 3) %>% 
+  align(align = 'right', part = 'header') %>% 
+  border_remove() %>% 
+  hline_top(border = fp_border(color = 'black')) %>% 
+  font(fontname = 'Times', part = 'all') %>%
+  fontsize(size = 12, part = 'all') %>% 
+  width(j = c(2, 3), width = 0.9) 
+tab
+```
+
 ```{r modsumtab}

### 2020-05-29 manu_draft.Rmd
table added for model performance

@@
 data(locs)
+data(modprf)
 
@@
 
+Table \@ref(tab:modprftab)
+
 # Discussion
@@
 
+```{r modprftab, tab.cap = "Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
+Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
+Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized."}
+totab <- modprf %>% 
+  group_by(station) %>% 
+  mutate(
+    AIC = AIC - max(AIC),
+    minaic = AIC == min(AIC), 
+    mingcv = GCV == min(GCV),
+    maxr2 = R2 == max(R2)
+  ) %>% 
+  ungroup
+
+boldaic <- as.numeric(rownames(totab)[totab$minaic])
+boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
+boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
+
+totab <- totab %>% 
+  select(-minaic, -mingcv, -maxr2) %>% 
+  mutate(
+    station = ifelse(duplicated(station), '', station)
+  )
+
+tab <- flextable(totab) %>% 
+  set_header_labels(
+    station = 'Station', 
+    model = 'GAM', 
+    AIC = 'Delta AIC',
+    R2 = 'R-squared'
+  ) %>% 
+  colformat_num(j = 3, digits = 1) %>%
+  colformat_num(j = c(4, 5), digits = 2) %>% 
+  bold(i = boldaic, j = 'AIC') %>% 
+  bold(i = boldgcv, j = 'GCV') %>% 
+  bold(i = boldr2, j = 'R2') %>% 
+  border_remove() %>% 
+  hline_top(border = fp_border(color = 'black')) %>% 
+  font(fontname = 'Times', part = 'all') %>%
+  font(j = 2, fontname = 'Consolas') %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  autofit()
+
+tab
+```
+
 # References

### 2020-05-29 R/dat_proc.R

@@ -1,0 +1,40 @@
+# libraries
+library(tidyverse)
+library(wqtrends)
+library(flextable)
+library(officer)
+
+# model performance summaries ---------------------------------------------
+
+modprf <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>% 
+  enframe() %>% 
+  group_by(value) %>% 
+  nest %>% 
+  mutate(
+    prf = purrr::map(value, function(x){
+      
+      load(file = x)
+      
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+      
+      dat <- get(nm) %>% 
+        select(model, modi) %>% 
+        deframe()
+      
+      prf <- anlz_fit(mods = dat)
+      
+      return(prf)
+      
+    })
+  ) %>% 
+  ungroup %>% 
+  select(-data) %>% 
+  unnest('prf') %>% 
+  select(-k) %>% 
+  mutate(
+    value = gsub('^data/mods\\_chl|\\.RData$', '', value)
+  ) %>% 
+  rename(station = value)
+
+save(modprf, file = 'data/modprf.RData', compress = 'xz')

### 2020-06-01 manu_draft.Rmd
p vals to mod summary table

@@
 
-```{r modprftab, tab.cap = "Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
+```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
 Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
-Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized."}
+Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
+Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
+The p-value for a row refers to the comparison of the model in the current row to the one preceeding.
+Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
 totab <- modprf %>% 

### 2020-06-01 R/funcs.R

@@ -1,0 +1,12 @@
+# function for formatting p-values in tables
+p_ast <- function(x){
+  
+  sig_cats <- c('**', '*', 'ns')
+  sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  
+  out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
+  out <- as.character(out)
+  
+  return(out)
+  
+}

### 2020-06-03 manu_draft.Rmd
added seastrnd data and table

@@
 data(modprf)
+data(seastrnd)
+
+source('R/funcs.R')
 
@@
 
-Table \@ref(tab:modprftab)
+Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab)
 
@@
     'Nonlinear trends wiht seasonality, plus interaction',
-    'Nonelinear trends with seasonality (unconstrained knots)'
+    'Nonlinear trends with seasonality (unconstrained knots)'
   ),
   Details = c(
-    'A linear trend in time over the course of the time series (cyear); and Within-year seasonal fluctuations that follow the same pattern every year - e.g.
-large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
-    'Treats cyear and doy same as gam0; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
-    'Treats cyear and doy same as gam1; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to gam6 (`ti(year, doy, bs = c("tp", "cc"))`)', 
-    'Variant of gam1 with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
+    'A linear trend in time over the course of the time series (year); and Within-year seasonal fluctuations that follow the same pattern every year, e.g., large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
+    'Treats year and doy same as `gam0`; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
+    'Treats cyear and doy same as `gam1`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `gam6` (`ti(year, doy, bs = c("tp", "cc"))`)', 
+    'Variant of `gam1` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
   )
@@
 Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
-The p-value for a row refers to the comparison of the model in the current row to the one preceeding.
+The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
+A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
 Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
@@
 
+```{r seastrndtab, tab.cap = 'Seasonal trends from `gam6` results by station and decade.
+Results show the annual slope estimate as chlorophyll change per year for each seasonal period across years (JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September, OND: October, November, December).
+Significant trends based on mixed-meta regression analysis are indicated with asterisks at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
+totab <- seastrnd %>% 
+  mutate(
+    pval = p_ast(pval),
+    pval = gsub('ns', '', pval), 
+    yrcoef = round(yrcoef, 2)
+    ) %>% 
+  unite('yrcoef', yrcoef, pval, sep = '') %>% 
+  spread(seas, yrcoef) %>% 
+  mutate(
+    station = ifelse(duplicated(station), '', station)
+  )
+
+boldjfm <- grep('\\*', totab$JFM)
+boldamj <- grep('\\*', totab$AMJ)
+boldjas <- grep('\\*', totab$JAS)
+boldond <- grep('\\*', totab$OND)
+
+tab <- flextable(totab) %>% 
+  set_header_labels(
+    station = 'Station', 
+    yrs = 'Years'
+  ) %>% 
+  add_header_row(values = c('', '', 'Chlorophyll change (ug/l) per year', '', '', '')) %>% 
+  merge_at(i = 1, j = 3:6, part = 'header') %>% 
+  bold(i = boldjfm, j = 'JFM') %>% 
+  bold(i = boldamj, j = 'AMJ') %>% 
+  bold(i = boldjas, j = 'JAS') %>% 
+  bold(i = boldond, j = 'OND') %>% 
+  border_remove() %>% 
+  hline_top(border = fp_border(color = 'black')) %>% 
+  font(fontname = 'Times', part = 'all') %>%
+  fontsize(size = 12, part = 'all') %>%   
+  width(j = c(2), width = 1) 
+tab
+```
+
 # References

### 2020-06-03 R/dat_proc.R

@@ -4,6 +4,8 @@ library(wqtrends)
 library(flextable)
 library(officer)
 
+source('R/funcs.R')
+
 # model performance summaries ---------------------------------------------
 
 modprf <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>% 
@@ -28,13 +30,91 @@ modprf <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
       
     })
   ) %>% 
-  ungroup %>% 
+  ungroup %>%
   select(-data) %>% 
   unnest('prf') %>% 
-  select(-k) %>% 
+  select(-k, -F) %>% 
   mutate(
-    value = gsub('^data/mods\\_chl|\\.RData$', '', value)
+    value = gsub('^data/mods\\_chl|\\.RData$', '', value), 
+    p.value = p_ast(p.value), 
+    p.value = ifelse(is.na(p.value), '-', p.value)
   ) %>% 
-  rename(station = value)
+  rename(
+    station = value, 
+    `P-value` = p.value
+    )
 
 save(modprf, file = 'data/modprf.RData', compress = 'xz')
+
+# seasonal trends by decade -----------------------------------------------
+
+seastrnd <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>% 
+  crossing(
+    fl = ., 
+    tibble(
+      doystr = c(1, 91, 182, 274), 
+      doyend = c(90, 181, 273, 364)
+    ), 
+    tibble(
+      yrstr = c(1990, 2000, 2010),
+      yrend = c(1999, 2009, 2017)
+    )
+  ) %>% 
+  group_by(fl) %>% 
+  nest() %>% 
+  mutate(
+    mod = purrr::map(fl, function(x){
+      
+      load(file = x)
+      
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+      
+      out <- get(nm) %>% 
+        filter(model == 'gam6') %>% 
+        select(model, modi) %>% 
+        deframe()
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest(c('data')) %>% 
+  group_by(doystr, doyend, mod, fl) %>% 
+  nest() %>% 
+  mutate(
+    avgseas = purrr::pmap(list(mods = mod, doystr, doyend), anlz_avgseason)
+  ) %>% 
[…]

### 2020-06-04 manu_draft.Rmd
station map added

@@
 library(officer)
+library(ggmap)
+library(sf)
+library(ggsn)
+library(USAboundaries)
 
@@
 
-The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20].
+The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20] (Figure \@ref(fig:sitemap)).
 This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.
 Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients.
-The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab)).
+The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
 Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters.
@@
 
+```{r, results = 'hide', message = F}
+# station locations
+sflocs <- locs %>% 
+  st_as_sf(coords = c('lon', 'lat'), crs = 4326)
+
+# for inset
+states <- us_states() %>% 
+  filter(!name %in% c('Alaska', 'Hawaii', 'Puerto Rico'))
+castate <- states %>% 
+  filter(name %in% 'California')
+locbuff <- st_buffer(sflocs, dist = 0.1)
+locbuff2 <- st_buffer(sflocs, dist = 0.07)
+insetbb <- st_buffer(sflocs, dist = 1.5) %>% 
+  st_bbox() %>% 
+  st_as_sfc(crs = 4326)
+
+# basemap
+dat_ext <- unname(st_bbox(locbuff))
+bsmap1 <- get_stamenmap(bbox = dat_ext, maptype = 'terrain-background', zoom = 11)
+
+# change opacity of basemap
+mapatt <- attributes(bsmap1)
+bsmap1_transparent <- matrix(adjustcolor(bsmap1, 
+                                                alpha.f = 0.75), 
+                                    nrow = nrow(bsmap1))
+attributes(bsmap1_transparent) <- mapatt
+
+# basemape plus stations
+p1 <- ggmap(bsmap1_transparent) +
+  geom_sf(data = sflocs, inherit.aes = F, col = 'black', pch = 21, fill = 'grey', size = 4) + 
+  geom_text(data = locs, aes(x = lon + 0.02, y = lat + 0.02, label = station), fontface = 'bold.italic') + 
+  theme(
+    axis.title = element_blank()
+  ) + 
+  north(locbuff, scale = 0.15, symbol = 10, location = 'bottomright') +
+  scalebar(locbuff2, dist = 10, dist_unit = 'km', location = 'bottomleft', transform = T, 
+           st.color = 'black', border.size = 0.5, st.dist = 0.02, st.size = 4)
+
+# inset
+p2 <- ggplot() + 
+  geom_sf(data = states, fill = 'white', colour = 'black') + 
+  geom_sf(data = castate, fill = 'grey', colour = 'black') + 
+  geom_sf(data = insetbb, fill = NA, color = 'blue', size = 2) +
+  theme_void()
+
+# final map
+p <- p1 + 
+  inset(ggplotGrob(p2), xmin = -122.3, xmax = -121.97, ymin = 37.75, ymax = 37.95)
+
+# save as png
+png('figs/sitemap.png', height = 1800, width = 1400, family = 'serif', res = 300)
+print(p)
+dev.off()
+```
+```{r sitemap, fig.cap = 'Station locations in the central and south San Francisco Estuary used for analysis.
+See Table \\@ref(tab:sumtab) for station descriptions.
+Full dataset described in @Schraga20.'}
+knitr::include_graphics('figs/sitemap.png')
+```
+
 ```{r modsumfig, fig.height = 4.5, fig.width = 6, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).

### 2020-06-07 manu_draft.Rmd
added obs summary plot

@@
 library(USAboundaries)
+library(hrbrthemes)
 
@@
 The resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.
-Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999.
+Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999 (Figure \@ref(fig:obsdat)).
 These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with the purpose, among others, to develop a framework for assessing water quality condition.
@@
 ```{r, results = 'hide', message = F}
+# treatment colors
+cls1 <- RColorBrewer::brewer.pal(9, 'Greys')
+cls2 <- RColorBrewer::brewer.pal(3, 'BrBG')
+
+toplo1 <- rawdat %>%
+  filter(param %in% 'chl') %>% 
+  filter(mo %in% c('Aug', 'Sep', 'Oct', 'Nov', 'Dec')) %>% 
+  # filter(station %in% c(24, 27, 30, 32)) %>% 
+  group_by(yr) %>% 
+  mutate(Median = median(value, na.rm = TRUE)) %>% 
+  ungroup()
+
+toplo2 <- rawdat %>%
+  filter(param %in% 'chl') %>% 
+  mutate(
+    yrcat = case_when(
+      yr < 2000  ~ '1990-2000', 
+      yr >= 2000 & yr < 2010 ~ '2000-2010', 
+      yr >= 2010 ~ '2010-2017'
+    ), 
+    yrcat = factor(yrcat, levels = c('1990-2000', '2000-2010', '2010-2017'))
+  )
+
+ylab <- expression(paste("Chlorophyll-a (", italic(mu), "g ", L^-1, ")"))
+
+p1 <- ggplot(toplo1, aes(x = yr, y = value, fill = Median, group = yr)) + 
+  geom_violin(draw_quantiles = 0.5)  + 
+  scale_fill_gradientn(colours = cls1) + 
+  scale_y_log10() +
+  theme_ipsum(base_family = 'serif') +
+  theme(
+    axis.title.x = element_blank()
+  ) +
+  labs(
+    y = ylab, 
+    subtitle = '(a) Annual summer/fall concentrations'
+  )
+
+p2 <- ggplot(toplo2, aes(x = mo, y = value, fill = yrcat)) + 
+  geom_violin(draw_quantiles = 0.5)  + 
+  scale_y_log10(ylab) +
+  theme_ipsum(base_family = 'serif') +
+  scale_fill_manual(values = cls2) +
+  theme(
+    legend.title = element_blank(), 
+    axis.title.x = element_blank()
+  ) +
+  labs(
+    y = ylab, 
+    subtitle = '(b) Monthly concentrations by decade'
+  )
+
+p <- p1 + p2 + plot_layout(ncol = 1)
+png('figs/obsdat.png', height = 8, width = 9, family = 'serif', units = 'in', res = 400)
+print(p)
+dev.off()
+```
+```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in Central and South SFE (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
+knitr::include_graphics('figs/obsdat.png')
+```
+
+```{r, results = 'hide', message = F}
 # station locations
@@
 # save as png
-png('figs/sitemap.png', height = 1800, width = 1400, family = 'serif', res = 300)
+png('figs/sitemap.png', height = 6.5, width = 5, units = 'in', family = 'serif', res = 300)
 print(p)
@@
 
-```{r modsumfig, fig.height = 4.5, fig.width = 6, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
-Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
+```{r, results = 'hide'}
 # get predictions for all four gams
@@
 
-p1
+png('figs/modsumfig.png', height = 4.5, width = 6, units = 'in', family = 'serif', res = 400)
+print(p1)
+dev.off()
+```
+```{r modsumfig, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
+Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
+knitr::include_graphics('figs/modsumfig.png')
 ```

### 2020-06-08 manu_draft.Rmd
model predictions results figure added

@@
 library(hrbrthemes)
+library(patchwork)
+library(lubridate)
+library(colorspace)
 
@@
 data(seastrnd)
+data(mods_chl18)
+data(mods_chl24)
+data(mods_chl27)
+data(mods_chl32)
 
@@
 
-Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab)
+Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat)
 
@@
   scale_y_log10() +
-  theme_ipsum(base_family = 'serif') +
+  theme_minimal(base_family = 'serif') +
   theme(
@@
   scale_y_log10(ylab) +
-  theme_ipsum(base_family = 'serif') +
+  theme_minimal(base_family = 'serif') +
   scale_fill_manual(values = cls2) +
@@
 p <- p1 + p2 + plot_layout(ncol = 1)
-png('figs/obsdat.png', height = 8, width = 9, family = 'serif', units = 'in', res = 400)
+png('figs/obsdat.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
 print(p)
@@
 ```
-```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in Central and South SFE (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
+```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
 knitr::include_graphics('figs/obsdat.png')
@@
 knitr::include_graphics('figs/modsumfig.png')
+```
+
+```{r, results = 'hide', message = F}
+# get predictions to plot
+prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
+  mutate(prddat = pmap(list(data, modi), function(data, modi){
+
+    prddat <- data.frame(
+        dec_time = seq(min(data$dec_time), max(data$dec_time), length = 1000)
+      ) %>%
+      mutate(
+        date = date_decimal(dec_time),
+        date = as.Date(date),
+        mo = month(date, label = TRUE),
+        doy = yday(date),
+        yr = year(date)
+      )
+
+    prd <- predict(modi, newdata = prddat) %>% 
+      as.numeric
+
+    annprd <- predict(modi, newdata = prddat, type = 'terms')
+    int <- attr(annprd, 'constant')
+    value <- rowSums(annprd) + int
+    annvalue <- rowSums(annprd[, !grepl('doy', colnames(annprd)), drop = FALSE]) + int
+    annvalue <- as.numeric(annvalue)
+
+    trans <- unique(modi$trans)
+
+    if(!inherits(trans, 'numeric'))
+      stop('log used')
+
+    prddat <- prddat %>% 
+      mutate(
+        chl = prd,
+        chl = forecast::InvBoxCox(chl, trans), 
+        annchl = forecast::InvBoxCox(annvalue, trans)
+        )
+
+    return(prddat)
+
+  })) %>% 
+  select(station, modi = model, prddat) %>% 
+  unnest(prddat)
+
+ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))
+
+toplo <- prdplo %>% 
+  filter(yr >= 1991)
+
+obsdat <- rawdat %>% 
+  filter(station %in% c(18, 24, 27, 32)) %>% 
+  filter(param == 'chl') %>% 
+  filter(yr >= 1991) %>% 
+  rename(chl = value)
+
+colpal <- "PuBuGn"
+
+p1 <- ggplot(toplo, aes(x = date, y = chl)) + 
+  geom_point(data = obsdat, color = 'grey', size = 0.15) + 
+  geom_line(size = 0.5) +
+  geom_line(aes(y = annchl), color = 'tomato1', size = 0.5, alpha = 0.9) +
+  theme_minimal(base_family = 'serif') + 
+  theme(
+    legend.position = 'right', 
+    legend.title = element_blank(), 
+    # axis.text.x= element_text(size = 8)
+    strip.background = element_blank()
+  ) + 
+  facet_grid(modi ~ station) +
+  scale_y_log10(limits = c(0.74, 83)) + 
+  labs(x = 'Date', subtitle = '(a) Model predictions with annual estimates', y = ylab)
+
+p2 <- ggplot(toplo, aes(x = doy, colour = yr, group = factor(yr))) + 
+  geom_line(aes(y = chl)) + 
+  theme_minimal(base_family = 'serif') + 
+  theme(
+    legend.position = 'right', 
+    legend.title = element_blank(), 
+    # axis.text.x= element_text(size = 8)
+    strip.background = element_blank()
+  ) + 
+  scale_colour_continuous_sequential(palette = colpal, rev = T, begin = 0.2) +
+  guides(colour = guide_colourbar(barheight = 10, barwidth = 1)) +
+  facet_grid(modi ~ station) +
+  scale_y_log10() + 
+  labs(x = 'Day of year', subtitle = '(b) Model predictions by day of year', y = ylab)
+
+p <- p1 + p2 + plot_layout(ncol = 1)
+
+png('figs/prddat.png', height = 10, width = 9, family = 'serif', units = 'in', res = 400)
+print(p)
+dev.off()
+```
+```{r prddat, fig.cap = "GAM predictions for select stations from north to south (18, 24, 27, 32) for models in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
+The results show (a) predictions across the time series with annual estimates (red line) and (b) predictions by day of year with one smooth per year.
+Observed data in (a) are shown with the gray points.
+Station locations are in Figure \\@ref(fig:sitemap)."}
+knitr::include_graphics('figs/prddat.png')
 ```

### 2020-06-09 manu_draft.Rmd
trends example plot added

@@
 data(mods_chl32)
+data(mods_chl34)
 
@@
 
-Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat)
+Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat), Figure \@ref(fig:trnddat)
 
@@
 
+```{r, results = 'hide', message = F}
+toplo <- mods_chl34 %>% 
+  filter(model %in% 'gam6') %>% 
+  pull(modi)
+
+ylim <- c(-1, 19)
+ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))
+
+p1 <- show_perchg(mods = toplo, baseyr = 1991, testyr = 2000, ylab = ylab) +
+  labs(
+    subtitle = '(a) 1990, 2000, Base: 3.6, Test: 5.7,\nChange: 57.7%, ns', 
+    title = NULL
+  )
+p2 <- show_perchg(mods = toplo, baseyr = 2000, testyr = 2010, ylab = NULL) +
+  labs(
+    subtitle = '(b) 2000, 2010, Base: 5.7, Test: 10.1,\nChange: 77.5%, p < 0.001', 
+    title = NULL
+  )
+p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2017, ylab = NULL) +
+  labs(
+    subtitle = '(c) 2010, 2017, Base: 10.1, Test: 6.1,\nChange: -39.8%, p < 0.01', 
+    title = NULL
+  )
+p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
+  scale_y_continuous(limits = ylim) + 
+  labs(
+    subtitle = '(d) 1990-2000, Jan-Jul, slope 0.6, p < 0.01', 
+    title = NULL
+  )
+p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
+  scale_y_continuous(limits = ylim) + 
+  labs(
+    subtitle = '(e) 2000-2010, Jan-Jul, slope 0.5, ns', 
+    title = NULL
+  )
+p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2016, ylab = NULL) +
+  scale_y_continuous(limits = ylim) + 
+  labs(
+    subtitle = '(f) 2010-2017, Jan-Jul, slope -0.4, ns', 
+    title = NULL
+  )
+p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
+  scale_y_continuous(limits = ylim) + 
+  labs(
+    subtitle = '(g) 1990-2000, Aug-Dec, slope 0.26, ns', 
+    title = NULL
+  )
+p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
+  scale_y_continuous(limits = ylim) + 
+  labs(
+    subtitle = '(h) 2000-2010, Aug-Dec, slope 0.56, p < 0.05', 
+    title = NULL
+  )
+p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2017, ylab = NULL) +
+  scale_y_continuous(limits = ylim) + 
+  labs(
+    subtitle = '(i) 2010-2017, Aug-Dec, slope -0.52, ns', 
+    title = NULL
+  )
+
+p <- p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + p9 + plot_layout(nrow = 3) & 
+  theme_minimal(base_family = 'serif', base_size = 12) &
+  theme(
+    axis.title.x = element_blank(),
+    legend.position = 'none'
+    )
+
+png('figs/trnddat.png', height = 9, width = 11, family = 'serif', units = 'in', res = 400)
+print(p)
+dev.off()
+```
+```{r trnddat, fig.cap = "Examples of trend tests using results of GAM predictions for station 34 (`gam6`).
+Plots (a), (b), and (c) show estimates of percent change as the difference of means between base and test years for chlorophyll.
+Plots (d) through (i) show seasonal averages and trend estimates over time.
+Plot (d), (e), and (f) show trend estimates for January through July and (g), (h), and (i) show trend estimates for August through December.
+The trend lines in (d) through (i) estimate the rate of change of chlorophyll per year, reported as the slope in the plot title.
+ns: not significant"}
+knitr::include_graphics('figs/trnddat.png')
+```
+
 # Tables

### 2020-06-10 manu_draft.Rmd
trends map done

@@
 data(seastrnd)
+data(seastrnd2)
+data(chgtrnd)
 data(mods_chl18)
@@
 
-Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat), Figure \@ref(fig:trnddat)
+Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat), Figure \@ref(fig:trnddat), Figure \@ref(fig:trndmap)
 
@@
 
+```{r, results = 'hide', message = F}
+# change opacity of basemap
+mapatt <- attributes(bsmap1)
+bsmap1_transparent <- matrix(adjustcolor(bsmap1, 
+                                         alpha.f = 0.4), 
+                             nrow = nrow(bsmap1))
+attributes(bsmap1_transparent) <- mapatt
+
+leglab2 <- expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+
+pthm <- theme_bw(base_family = 'serif', base_size = 14) +
+  theme(
+    legend.position = 'top',
+    strip.background = element_blank(),
+    axis.title = element_blank(), 
+    axis.text = element_text(size = 7)
+  )
+
+toplo1 <- chgtrnd %>% 
+  mutate(
+    station = as.numeric(station),
+    pval = ifelse(pval == 'sig', '*', '')
+    ) %>% 
+  left_join(locs, by = 'station') %>% 
+  filter(model %in% 'gam6')
+
+toplo2 <- seastrnd2 %>%
+  mutate(
+    station = as.numeric(station),
+    pval = ifelse(pval < 0.05, '*', ''), 
+    coefsgn = sign(yrcoef), 
+    coefsgn = factor(coefsgn, levels = c('1', '-1'), labels = c('inc', 'dec'))
+  ) %>% 
+  left_join(locs, by = 'station')
+
+p1 <- ggmap(bsmap1_transparent) +
+  geom_point(data = toplo1, aes(x = lon, y = lat, size = abs(perchg), shape = persgn, fill = perchg, colour = pval)) + #, color = 'black') +
+  geom_text(data = toplo1, aes(x = lon, y = lat, label = pval), nudge_x = 0.04) + 
+  facet_grid( ~ yrs) + 
+  scale_fill_gradient2('Chl-a % change', low = 'green', mid = 'white',  high = 'tomato1', midpoint = 0) +
+  scale_colour_manual(values = c('black', 'black'), guide = F) +
+  coord_map() + 
+  scale_shape_manual('Trend', values = c(24, 25)) + 
+  pthm +
+  scale_size(range = c(1, 6), guide = F) +
+  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 6)) +
+  labs(
+    subtitle = '(a) Percent change between year pairs'
+  )
+
+p2 <- ggmap(bsmap1_transparent) +
+  geom_point(data = toplo2, aes(x = lon, y = lat, size = abs(yrcoef), shape = coefsgn, fill = yrcoef, colour = pval)) + #, color = 'black') +
+  geom_text(data = toplo2, aes(x = lon, y = lat, label = pval), nudge_x = 0.04) + 
+  facet_grid(seas ~ yrs) + 
+  scale_fill_gradient2(leglab2, low = 'green', mid = 'white',  high = 'tomato1', midpoint = 0) +
+  scale_colour_manual(values = c('black', 'black'), guide = F) +
+  coord_map() + 
+  scale_shape_manual('Trend', values = c(24, 25)) + 
+  pthm +
+  scale_size(range = c(1, 6), guide = F) +
+  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 6)) +
+  labs(
+    subtitle = '(b) Annual rates of change across seasons by decade'
+  )
+
+p <- p1 + p2 + plot_layout(ncol = 1, heights = c(0.5, 1))
+
+png('figs/trndmap.png', height = 10, width = 6.5, family = 'serif', units = 'in', res = 400)
+p
+dev.off()
+```
+```{r trndmap, fig.cap= "Trend estimates (a) between year pairs and (b) across seasons by decade for chlorophyll at each station.
+The top plots show percent change comparing the first and last year within a decade and the bottom plots show seasonal estimates of change per year for chlorophyll concentrations for each decade.
+Point type, shape, and color represent the direction and magnitude of an estimated trend.
+Trends with $p<0.05$ are marked with an asterisk.
+All results are from `gam6`."}
+knitr::include_graphics('figs/trndmap.png')
+```
+
 # Tables

### 2020-06-10 R/dat_proc.R

@@ -118,3 +118,126 @@ seastrnd <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
 
 save(seastrnd, file = 'data/seastrnd.RData', compress = 'xz')
 
+# seasonal trends by decade, jan-jul and aug-dec --------------------------
+
+seastrnd2 <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>% 
+  crossing(
+    fl = ., 
+    tibble(
+      doystr = c(41, 214), 
+      doyend = c(213, 338)
+    ), 
+    tibble(
+      yrstr = c(1991, 2000, 2010),
+      yrend = c(1999, 2009, 2016)
+    )
+  ) %>% 
+  group_by(fl) %>% 
+  nest() %>% 
+  mutate(
+    mod = purrr::map(fl, function(x){
+      
+      load(file = x)
+      
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+      
+      out <- get(nm) %>% 
+        filter(model == 'gam6') %>% 
+        select(model, modi) %>% 
+        deframe()
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest(c('data')) %>% 
+  group_by(doystr, doyend, mod, fl) %>% 
+  nest() %>% 
+  mutate(
+    avgseas = purrr::pmap(list(mods = mod, doystr, doyend), anlz_avgseason)
+  ) %>% 
+  unnest('data') %>% 
+  mutate(
+    yrstr = purrr::pmap(list(yrstr, avgseas), function(yrstr, avgseas){
+      max(yrstr, min(avgseas$yr))
+    }), 
+    yrend = purrr::pmap(list(yrend, avgseas), function(yrend, avgseas){
+      min(yrend, max(avgseas$yr))
+    }),
+    metatrnd = purrr::pmap(list(avgseas, yrstr, yrend), anlz_mixmeta)
+  ) %>% 
+  mutate(
+    yrcoef = purrr::map(metatrnd, function(x) x[[1]]$coefficients['yr']), 
+    pval = purrr::map(metatrnd, function(x){
+      coefficients(summary(x[[1]])) %>% data.frame %>% .[2, 4]
+    }
+    )
+  ) %>% 
+  ungroup %>% 
+  select(station = fl, seas = doystr, yrs = yrstr, yrcoef, pval) %>% 
+  mutate(
[…]

### 2020-06-15 manu_draft.Rmd
log chl models added from sfbaytrends, minor edits to methods

@@
 These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with the purpose, among others, to develop a framework for assessing water quality condition.
-Critical questions that were to be addressed by the NMS included identifying water quality conditions to be considered healthy and protective of beneficial uses, how do current conditions compare to desirable water quality conditions, and what potential management actions are needed to achieve desired conditions if they currently do not persist.
+Critical questions that were to be addressed by the NMS included identifying water quality conditions to be considered healthy and protective of beneficial uses, describing how current conditions compare to desirable water quality conditions, and identifying what potential management actions are needed to achieve desired conditions if they currently do not persist.
 Although visual changes in observed data were apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed.
 
-The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20] (Figure \@ref(fig:sitemap)).
+The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20].
 This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.
@@
 However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.
+The effect of this choose on model fit is described below.
 
 An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location.
-Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is a useful form for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.
-A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of component being modelled.
+Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is useful for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.
+A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of a component being modelled.
 This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.
 In others words, not using a cyclic cubic spline to model season would cause a non-continous function across the time series with step increases or decreases for the seasonal estimates between the years.
-This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences over time.
+This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences.
 
@@
 The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.
-A comparison of GAM perforamcne between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.
+A comparison of GAM performance between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.
 The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable.
@@
 The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.
-The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02 to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
+The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
 Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:
@@
 Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.
-The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlrophyll time series with the optimal $\lambda$ value.
+The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlorophyll time series with the optimal $\lambda$ value.
 Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.
@@
 The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
-However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require an explicit declaration of how much variation could be explained for the separate $year$ and $doy$ components.
-This requires user input on the potential upper limit for the *k*-values that could be used during model fitting.
+However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.
 Initial testing by Murphy et al.
-(2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable and specified the upper limit for the number of knots for the $s_2(year)$ smoother as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
+(2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.
+Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
 We follow the same approach herein for gam1 and gam2.
@@
 For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
-This approach potentially allows the smoother for $year$ to be completely unconstrained in how much variance is fit within and between years.
+This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.
 
@@
 # Results
+
+* start with description of model performance, which models had the best performance (gam0 v gam6, boxcox v log)
+* then describe figure 4
 

### 2020-06-15 R/dat_proc.R
modprf with log

@@ -8,7 +8,7 @@ source('R/funcs.R')
 
 # model performance summaries ---------------------------------------------
 
-modprf <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>% 
+modprf <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names = T) %>% 
   enframe() %>% 
   group_by(value) %>% 
   nest %>% 
@@ -16,7 +16,7 @@ modprf <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
     prf = purrr::map(value, function(x){
       
       load(file = x)
-      
+  
       nm <- basename(x)
       nm <- gsub('\\.RData', '', nm)
       
@@ -24,18 +24,26 @@ modprf <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
         select(model, modi) %>% 
         deframe()
       
+      trans <- lapply(dat, function(x) x$trans) %>% 
+        enframe('model', 'trans') %>%
+        unnest('trans')
+        
       prf <- anlz_fit(mods = dat)
+
+      out <- trans %>% 
+        left_join(prf, by = 'model') %>% 
+        mutate(trans = ifelse(is.numeric(trans), as.character(round(trans, 2)), trans))
       
-      return(prf)
+      return(out)
       
     })
   ) %>% 
-  ungroup %>%
+  ungroup %>% 
   select(-data) %>% 
   unnest('prf') %>% 
   select(-k, -F) %>% 
   mutate(
-    value = gsub('^data/mods\\_chl|\\.RData$', '', value), 
+    value = gsub('^data/mods\\_chl|^data/modslog\\_chl$|\\.RData$', '', value), 
     p.value = p_ast(p.value), 
     p.value = ifelse(is.na(p.value), '-', p.value)
   ) %>% 

### 2020-06-15 manu_draft.Rmd
new model summary table with log and boxcox

@@
 ```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
-Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
-Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
-Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
+Results are for models of chlorophyll-a using BoxCox and logarathmic transformations.
+Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`),  $\\Delta$ GCV, and R-squared values.
+Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.
 The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
 A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
-Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
+ns: not significant at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
+
 totab <- modprf %>% 
+  gather('stat', 'val', AIC, GCV, R2, `P-value`) %>% 
+  mutate(trans = case_when(
+    trans != 'log10' ~ 'BoxCox', 
+    T ~ trans)
+  ) %>% 
+  unite('stat', trans, stat) %>% 
+  mutate(
+    stat = factor(stat, levels = c('BoxCox_AIC', 'log10_AIC', 'BoxCox_GCV', 'log10_GCV', 'BoxCox_R2', 'log10_R2', 'BoxCox_P-value', 'log10_P-value'))
+  ) %>% 
+  spread(stat, val) %>% 
+  mutate_at(c('BoxCox_AIC', 'log10_AIC', 'BoxCox_GCV', 'log10_GCV', 'BoxCox_R2', 'log10_R2'), as.numeric) %>% 
   group_by(station) %>%
   mutate(
-    AIC = AIC - max(AIC),
-    minaic = AIC == min(AIC), 
-    mingcv = GCV == min(GCV),
-    maxr2 = R2 == max(R2)
+    BoxCox_AIC = BoxCox_AIC - max(BoxCox_AIC),
+    log10_AIC = log10_AIC - max(log10_AIC),
+    BoxCox_GCV = BoxCox_GCV - max(BoxCox_GCV),
+    log10_GCV = log10_GCV - max(log10_GCV)#,
+    # minaic = AIC == min(AIC), 
+    # mingcv = GCV == min(GCV),
+    # maxr2 = R2 == max(R2)
   ) %>% 
@@
 
-boldaic <- as.numeric(rownames(totab)[totab$minaic])
-boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
-boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
+# boldaic <- as.numeric(rownames(totab)[totab$minaic])
+# boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
+# boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
 
 totab <- totab %>% 
-  select(-minaic, -mingcv, -maxr2) %>% 
+  # select(-minaic, -mingcv, -maxr2) %>% 
   mutate(
@@
   set_header_labels(
-    station = 'Station', 
-    model = 'GAM', 
-    AIC = 'Delta AIC',
-    R2 = 'R-squared'
+    station = '', 
+    model = '', 
+    BoxCox_AIC = 'Delta AIC',
+    log10_AIC = 'Delta AIC',
+    BoxCox_GCV = 'Delta GCV',
+    log10_GCV = 'Delta GCV',
+    BoxCox_R2 = 'R-squared',
+    log10_R2 = 'R-squared',
+    `BoxCox_P-value` = 'P-val',
+    `log10_P-value` = 'P-val'
   ) %>% 
-  colformat_num(j = 3, digits = 1) %>%
-  colformat_num(j = c(4, 5), digits = 2) %>% 
-  bold(i = boldaic, j = 'AIC') %>% 
-  bold(i = boldgcv, j = 'GCV') %>% 
-  bold(i = boldr2, j = 'R2') %>% 
+  merge_at(i = 1, j = 3:4, part = "header") %>% 
+  merge_at(i = 1, j = 5:6, part = "header") %>% 
+  merge_at(i = 1, j = 7:8, part = "header") %>% 
+  merge_at(i = 1, j = 9:10, part = "header") %>% 
+  add_header_row(values = c('Station', 'GAM', 'BoxCox', 'Log', 'BoxCox', 'Log', 'BoxCox', 'Log', 'BoxCox', 'Log'), top = F) %>% 
+  align(j = c(3:10), align = 'center', part = 'all') %>% 
+  colformat_num(j = c(3, 4), digits = 1) %>%
+  colformat_num(j = c(5, 6), digits = 3) %>% 
+  colformat_num(j = c(7, 8), digits = 2) %>% 
+  # bold(i = boldaic, j = 'AIC') %>% 
+  # bold(i = boldgcv, j = 'GCV') %>% 
+  # bold(i = boldr2, j = 'R2') %>% 
   border_remove() %>% 
@@
   font(j = 2, fontname = 'Consolas') %>% 
-  fontsize(size = 12, part = 'all') %>% 
+  fontsize(size = 8, part = 'all') %>% 
   autofit()
@@
 
+# Supplement
+
+```{r modprfboxtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
+Results are for models of chlorophyll-a using BoxCox transformations.
+Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
+Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
+Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
+The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
+A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
+Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
+totab <- modprf %>% 
+  filter(trans != 'log10') %>% 
+  select(-trans) %>% 
+  group_by(station) %>% 
+  mutate(
+    AIC = AIC - max(AIC),
+    minaic = AIC == min(AIC), 
+    mingcv = GCV == min(GCV),
+    maxr2 = R2 == max(R2)
+  ) %>% 
+  ungroup
+
+boldaic <- as.numeric(rownames(totab)[totab$minaic])
+boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
+boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
+
+totab <- totab %>% 
+  select(-minaic, -mingcv, -maxr2) %>% 
+  mutate(
+    station = ifelse(duplicated(station), '', station)
+  )
+
+tab <- flextable(totab) %>% 
+  set_header_labels(
+    station = 'Station', 
+    model = 'GAM', 
+    AIC = 'Delta AIC',
+    R2 = 'R-squared'
+  ) %>% 
+  colformat_num(j = 3, digits = 1) %>%
+  colformat_num(j = c(4, 5), digits = 2) %>% 
+  bold(i = boldaic, j = 'AIC') %>% 
+  bold(i = boldgcv, j = 'GCV') %>% 
+  bold(i = boldr2, j = 'R2') %>% 
+  border_remove() %>% 
+  hline_top(border = fp_border(color = 'black')) %>% 
+  font(fontname = 'Times', part = 'all') %>%
+  font(j = 2, fontname = 'Consolas') %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  autofit()
+
+tab
+```
+
+```{r modprflogtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
+Results are for models of chlorophyll-a using logarithmic (base-10) ransformations.
+Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
+Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
+Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
+The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
+A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
+Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
+totab <- modprf %>% 
+  filter(trans == 'log10') %>% 
+  select(-trans) %>% 
+  group_by(station) %>% 
+  mutate(
+    AIC = AIC - max(AIC),
+    minaic = AIC == min(AIC), 
+    mingcv = GCV == min(GCV),
+    maxr2 = R2 == max(R2)
+  ) %>% 
+  ungroup
+
+boldaic <- as.numeric(rownames(totab)[totab$minaic])
+boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
+boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
+
+totab <- totab %>% 
+  select(-minaic, -mingcv, -maxr2) %>% 
+  mutate(
+    station = ifelse(duplicated(station), '', station)
+  )
+
+tab <- flextable(totab) %>% 
+  set_header_labels(
+    station = 'Station', 
+    model = 'GAM', 
+    AIC = 'Delta AIC',
+    R2 = 'R-squared'
+  ) %>% 
+  colformat_num(j = 3, digits = 1) %>%
+  colformat_num(j = c(4, 5), digits = 2) %>% 
+  bold(i = boldaic, j = 'AIC') %>% 
+  bold(i = boldgcv, j = 'GCV') %>% 
+  bold(i = boldr2, j = 'R2') %>% 
+  border_remove() %>% 
+  hline_top(border = fp_border(color = 'black')) %>% 
+  font(fontname = 'Times', part = 'all') %>%
+  font(j = 2, fontname = 'Consolas') %>% 
+  fontsize(size = 12, part = 'all') %>% 
+  autofit()
+
+tab
+```
+
 # References

### 2020-06-15 R/dat_proc.R

@@ -43,7 +43,7 @@ modprf <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names =
   unnest('prf') %>% 
   select(-k, -F) %>% 
   mutate(
-    value = gsub('^data/mods\\_chl|^data/modslog\\_chl$|\\.RData$', '', value), 
+    value = gsub('^data/mods\\_chl|^data/modslog\\_chl|\\.RData$', '', value), 
     p.value = p_ast(p.value), 
     p.value = ifelse(is.na(p.value), '-', p.value)
   ) %>% 

### 2020-08-04 manu_draft.Rmd
gam0 removed from manu, added doc links to readme

@@
 \begin{equation}
-\boldsymbol{gam0:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + \epsilon 
-(\#eq:gamzero)
-\end{equation}
-
-\begin{equation}
 \boldsymbol{gam1:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
@@
 
-A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam0, gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
+A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
 The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
@@
 
-* start with description of model performance, which models had the best performance (gam0 v gam6, boxcox v log)
+* start with description of model performance, which models had the best performance (gam1 v gam6, boxcox v log)
 * then describe figure 4
@@
 mods <- list(
-  gam0 = anlz_gam(tomod, mod = 'gam0', trans = trans),
   gam1 = anlz_gam(tomod, mod = 'gam1', trans = trans),
@@
 
-p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 2, size = 0.8, alpha = 0.8) + 
+p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8) + 
   theme(legend.position = 'right') + 
@@
 
-png('figs/modsumfig.png', height = 4.5, width = 6, units = 'in', family = 'serif', res = 400)
+png('figs/modsumfig.png', height = 2.5, width = 6, units = 'in', family = 'serif', res = 400)
 print(p1)
@@
 prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
+  filter(model != 'gam0') %>% 
   mutate(prddat = pmap(list(data, modi), function(data, modi){
@@
 tab <- tibble(
-  GAM = c('`gam0`', '`gam1`', '`gam2`', '`gam6`'),
+  GAM = c('`gam1`', '`gam2`', '`gam6`'),
   Summary = c(
-    'Linear trend with seasonality', 
     'Nonlinear trends seasonality (constrained knots)', 
@@
   Details = c(
-    'A linear trend in time over the course of the time series (year); and Within-year seasonal fluctuations that follow the same pattern every year, e.g., large spring bloom and small fall bloom (`s(doy, bs = "cc")`)',
-    'Treats year and doy same as `gam0`; and A smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
+    'Treats year and doy with separate smoothers with a smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
     'Treats cyear and doy same as `gam1`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `gam6` (`ti(year, doy, bs = c("tp", "cc"))`)', 
@@
 Results are for models of chlorophyll-a using BoxCox and logarathmic transformations.
-Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`),  $\\Delta$ GCV, and R-squared values.
+Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam1`),  $\\Delta$ GCV, and R-squared values.
 Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.
-The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
+The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam1` is the base model and does not have a p-value).
 A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.

### 2020-08-04 R/dat_proc.R

@@ -22,12 +22,14 @@ modprf <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names =
       
       dat <- get(nm) %>% 
         select(model, modi) %>% 
+        filter(model != 'gam0') %>% 
         deframe()
       
       trans <- lapply(dat, function(x) x$trans) %>% 
         enframe('model', 'trans') %>%
-        unnest('trans')
-        
+        unnest('trans') %>% 
+        filter(model != 'gam0')
+  
       prf <- anlz_fit(mods = dat)
 
       out <- trans %>% 
@@ -217,10 +219,11 @@ chgtrnd <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
       
       nm <- basename(x)
       nm <- gsub('\\.RData', '', nm)
-      
+
       out <- get(nm) %>%  
         select(model, modi) %>% 
-        deframe()
+        filter(model != 'gam0') %>% 
+        deframe() 
       
       return(out)
       

### 2020-08-04 manu_draft.Rmd
maybe renaming gams

@@
 
-To evaluate GAMs with SFE chloropyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19], including incorporation and adaptation of their hierarchical model structure to evaluate long-term time series.
-These hierarchical models vary the functional form for time in relation to chlorophyll by adding smoothers and linear predictors with increasing flexibility in the GAM functions.
-This is a powerful approach for time series analysis because separate components can be explicitly modelled with different smoothing terms and the results can be idendependently evaluated and further tested from the model output.
-Varying degrees of model complexity also provides a balance between descriptive power and computational efficiency, where simpler models may be preferred in routine reporting or more complex models may be preferred for hypothesis testing.
-We present four GAM structures, following a similar format as @Murphy19.
-The notation for gam3, gam4, and gam5 are not used because these describe models in @Murphy19 that were developed for purposes specific to Chesapeake Bay.
+To evaluate GAMs with SFE chlorophyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19].
+We specifically build our models using the `gam1` and `gam2` structures in @Murphy19, but present a different naming convention to better distinguish the functional components of the models.
+The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Additive" and "Interactive", respectively.
+Below, we describe the "Additive*" star model as a slight modification of the former.
+Each model has a similar functional form for time in relation to chlorophyll with slight differences in the smoothers that describe time.
 
 \begin{equation}
-\boldsymbol{gam1:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
+\boldsymbol{Additive:} g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
 (\#eq:gamone)
@@
 \begin{equation}
-\boldsymbol{gam2:}\ g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
+\boldsymbol{Interactive:} g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
 (\#eq:gamtwo)
-\end{equation}
-
-\begin{equation}
-\boldsymbol{gam6:}\  g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2^*\left(year\right) + \epsilon
-(\#eq:gamsix)
 \end{equation}
@@
 A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
-The four models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
+The three models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
 However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.
@@
 
-Each of the four GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.
+Each of the three GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.
 For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoothers terms.
@@
 ```{r, results = 'hide'}
-# get predictions for all four gams
+# get predictions for all three gams
 tomod <- rawdat %>%
@@
 ```
-```{r modsumfig, fig.cap = "General output for each of the four GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
+```{r modsumfig, fig.cap = "General output for each of the three GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
 Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
@@
 Results are for models of chlorophyll-a using BoxCox transformations.
-Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
+Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
 Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
@@
 Results are for models of chlorophyll-a using logarithmic (base-10) ransformations.
-Four GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
+Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
 Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.

### 2020-08-06 manu_draft.Rmd
wan et al. to lit, fixing spacing on equations in manu

@@
 source('R/funcs.R')
+
+# model levels and labels
+modlbs <- c('Const seas', 'Int seas', 'Const seas*')
+modbst <- modlbs[3]
 
@@
 We specifically build our models using the `gam1` and `gam2` structures in @Murphy19, but present a different naming convention to better distinguish the functional components of the models.
-The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Additive" and "Interactive", respectively.
-Below, we describe the "Additive*" star model as a slight modification of the former.
+The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Constant season" and "Interactive season", respectively.
+Below, we describe the "Constant season*" model as a slight modification of the former.
 Each model has a similar functional form for time in relation to chlorophyll with slight differences in the smoothers that describe time.
@@
 \begin{equation}
-\boldsymbol{Additive:} g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + \epsilon 
+\boldsymbol{Constant\ season:} g\left(chl\right) = s_0 + \beta_1 decYear + s_1\left(doy\right) + s_2\left(decYear\right) + \epsilon 
 (\#eq:gamone)
@@
 \begin{equation}
-\boldsymbol{Interactive:} g\left(chl\right) = s_0 + \beta_1 year + s_1\left(doy\right) + s_2\left(year\right) + ti_1(doy, year) + \epsilon
+\boldsymbol{Interactive\ season:} g\left(chl\right) = s_0 + \beta_1 decYear + s_1\left(doy\right) + s_2\left(decYear\right) + ti_1(doy, decYear) + \epsilon
 (\#eq:gamtwo)
@@
 
-For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $year$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
-The $year$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).
+For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $decYear$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
+The $decYear$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).
 The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series.
 The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models.
-The $s_1()$ and $s_2()$ functions are different smoothers for $year$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.
-The $ti_1()$ function is a tensor product interaction between $year$ and $doy$ that allows the seasonal component to change between years.
+The $s_1()$ and $s_2()$ functions are different smoothers for $decYear$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.
+The $ti_1()$ function is a tensor product interaction between $decYear$ and $doy$ that allows the seasonal component to change between years.
 Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline.
@@
 
-A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., gam1, gam2, or gam6) and determining how much variation could be explained by the smoothers in each function.
-The three models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
+A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., equation \@ref(eq:gamone) or \@ref(eq:gamtwo) and determining how much variation could be explained by the smoothers in each function.
+The two models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
 However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.
@@
 Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
-We follow the same approach herein for gam1 and gam2.
+We follow the same approach herein both models.
 Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further.
-For the application to SFE chlorophyll data, gam6 was also added as an extension of gam1, where the total number of knots for the $year$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
-This approach potentially allows the smoother for $year$ to be unconstrained in how much variance is fit within and between years.
+For the application to SFE chlorophyll data, an extension of the constant season model was also added as "Constant season*", where the total number of knots for the $decYear$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
+This approach potentially allows the smoother for $decYear$ to be unconstrained in how much variance is fit within and between years.
 
@@
 In addition to the GCV score, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
-Comparisons were made between the different GAMs in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) and for the transformations used to define the $g()$ function for chlorophyll at each station.
+Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station.
 Model predictions were based on the standard model output from the *predict()* function from the mgcv package.
@@
 
-Each of the GAM structures described in equations \@ref(eq:gamzero) to \@ref(eq:gamsix) provide different approaches to quantify long-term variation in annual and seasonal changes.
+Each of the GAM structures provide different approaches to quantify long-term variation in annual and seasonal changes.
 Although useful as a general approach to view functional forms for different time series, additional information regarding whether an observed change is statistically significant can be useful to inform environmental management decisions.
@@
 
-* start with description of model performance, which models had the best performance (gam1 v gam6, boxcox v log)
+* start with description of model performance, which models had the best performance (Const. seas., Int. seas., Const. seas*, boxcox v log)
 * then describe figure 4
@@
 mods <- list(
-  gam1 = anlz_gam(tomod, mod = 'gam1', trans = trans),
-  gam2 = anlz_gam(tomod, mod = 'gam2', trans = trans),
-  gam6 = anlz_gam(tomod, mod = 'gam6', trans = trans)
+  `Const seas` = anlz_gam(tomod, mod = 'gam1', trans = trans),
+  `Int seas` = anlz_gam(tomod, mod = 'gam2', trans = trans),
+  `Const seas*` = anlz_gam(tomod, mod = 'gam6', trans = trans)
 )
 
-p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8) + 
+p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8, faclev = modlbs) + 
   theme(legend.position = 'right') + 
@@
 ```
-```{r modsumfig, fig.cap = "General output for each of the three GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
+```{r modsumfig, fig.cap = "General output for each of the GAM structures for constant season, interactive season, and constant season*.
 Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
@@
 prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
-  filter(model != 'gam0') %>% 
   mutate(prddat = pmap(list(data, modi), function(data, modi){
@@
 ```
-```{r prddat, fig.cap = "GAM predictions for select stations from north to south (18, 24, 27, 32) for models in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).
+```{r prddat, fig.cap = "GAM predictions for select stations from north to south (18, 24, 27, 32) for each of the GAM structures for constant season, interactive season, and constant season*.
 The results show (a) predictions across the time series with annual estimates (red line) and (b) predictions by day of year with one smooth per year.
@@
 toplo <- mods_chl34 %>% 
-  filter(model %in% 'gam6') %>% 
+  filter(model %in% !!modbst) %>% 
   pull(modi)
@@
 ```
-```{r trnddat, fig.cap = "Examples of trend tests using results of GAM predictions for station 34 (`gam6`).
+```{r trnddat, fig.cap = "Examples of trend tests using results of GAM predictions for station 34 (`Const seas*`).
 Plots (a), (b), and (c) show estimates of percent change as the difference of means between base and test years for chlorophyll.
@@
   left_join(locs, by = 'station') %>% 
-  filter(model %in% 'gam6')
+  filter(model %in% !!modbst)
 
@@
 Trends with $p<0.05$ are marked with an asterisk.
-All results are from `gam6`."}
+All results are from `Const seas*`."}
 knitr::include_graphics('figs/trndmap.png')
@@
 tab <- tibble(
-  GAM = c('`gam1`', '`gam2`', '`gam6`'),
+  GAM = c('`Constant season`', '`Interactive season`', '`Constant season*`'),
   Summary = c(
@@
     'Treats year and doy with separate smoothers with a smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
-    'Treats cyear and doy same as `gam1`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `gam6` (`ti(year, doy, bs = c("tp", "cc"))`)', 
-    'Variant of `gam1` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
+    'Treats cyear and doy same as `Const seas`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `Const seas*` (`ti(decYear, doy, bs = c("tp", "cc"))`)',
+    'Variant of `Const seas` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
   )
@@
 # table stuff
-cap.val <- 'Summary and details for each of the GAM structures in equations \\@ref(eq:gamzero) to \\@ref(eq:gamsix).'
+cap.val <- 'Summary and details for each of the GAM structures.'
 
@@
 Results are for models of chlorophyll-a using BoxCox and logarathmic transformations.
-Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam1`),  $\\Delta$ GCV, and R-squared values.
+Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `Const seas`),  $\\Delta$ GCV, and R-squared values.
 Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.
-The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam1` is the base model and does not have a p-value).
+The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`Const seas` is the base model and does not have a p-value).
 A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
@@
 
-```{r seastrndtab, tab.cap = 'Seasonal trends from `gam6` results by station and decade.
+```{r seastrndtab, tab.cap = 'Seasonal trends from `Const seas*` results by station and decade.
 Results show the annual slope estimate as chlorophyll change per year for each seasonal period across years (JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September, OND: October, November, December).
@@
 Results are for models of chlorophyll-a using BoxCox transformations.
-Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
+Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC, GCV, and R-squared values.
 Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.

### 2020-08-06 R/dat_proc.R

@@ -6,6 +6,39 @@ library(officer)
 
 source('R/funcs.R')
 
+# model levels and labels
+modlvs <- c('gam0', 'gam1', 'gam2', 'gam6')
+modlbs <- c('gam0', 'Const seas', 'Int seas', 'Const seas*')
+names(modlvs) <- modlbs
+
+# best mod
+modbst <- modlbs[4]
+
+# # save all models with different naming convention ------------------------
+# 
+# mods <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names = T)
+# 
+# for(mod in mods){
+#   
+#   cat(mod, '\t')
+#   
+#   load(file = mod)
+#   nm <- gsub('\\.RData$', '', basename(mod))
+#   dat <- get(nm)
+#   
+#   dat <- dat %>% 
+#     filter(model != 'gam0') %>% 
+#     mutate(
+#       model = factor(model, levels = modlvs, labels = modlbs), 
+#       model = fct_drop(model)
+#     )
+#   
+#   assign(nm, dat)
+#   
+#   save(list = nm, file = mod, compress = 'xz')
+#     
+# }
+
 # model performance summaries ---------------------------------------------
 
 modprf <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names = T) %>% 
@@ -16,19 +49,17 @@ modprf <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names =
     prf = purrr::map(value, function(x){
       
       load(file = x)
-  
+
       nm <- basename(x)
       nm <- gsub('\\.RData', '', nm)
       
       dat <- get(nm) %>% 
         select(model, modi) %>% 
-        filter(model != 'gam0') %>% 
         deframe()
       
       trans <- lapply(dat, function(x) x$trans) %>% 
         enframe('model', 'trans') %>%
-        unnest('trans') %>% 
-        filter(model != 'gam0')
+        unnest('trans') 
   
       prf <- anlz_fit(mods = dat)
 
@@ -81,7 +112,7 @@ seastrnd <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
       nm <- gsub('\\.RData', '', nm)
       
       out <- get(nm) %>% 
-        filter(model == 'gam6') %>% 
+        filter(model == !!modbst) %>% 
         select(model, modi) %>% 
         deframe()
       
[…]

### 2020-08-07 manu_draft.Rmd
working on results text

@@
 
-Cloern et al. 2020 [link](https://link.springer.com/article/10.1007/s12237-020-00737-w)
-
 # Methods
@@
 The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America and drains an area of approximately 200 thousand km$^2$ in the US state of California.
-The estuary is subdivided into six sub-embayments having distinct geomorphological characteristics: Suisun Bay, San Pablo Bay, North Central Bay, Central Bay, South Bay,and Lower South Bay.
 Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.
-The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt) with seasonal freshwater pulses primarily occurring in the winter and spring during the rainy season and subsequent snowmelt in the upper watershed.
-The southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
-South Bay is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.
+The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt), whereas the southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
+The South Bay embayment is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.
 Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.
@@
 However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.
-The resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.
-Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations were observed beginning in 1999 (Figure \@ref(fig:obsdat)).
-These relatively recent occurrrences motivated the San Francisco Regional Water Quality Control Board and stakeholders to establish a Nutrient Management Strategy, with the purpose, among others, to develop a framework for assessing water quality condition.
-Critical questions that were to be addressed by the NMS included identifying water quality conditions to be considered healthy and protective of beneficial uses, describing how current conditions compare to desirable water quality conditions, and identifying what potential management actions are needed to achieve desired conditions if they currently do not persist.
-Although visual changes in observed data were apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed.
-
-The SFE Research Program of the US Geological Survey has been collecting water quality data at fixed locations in the bay since 1969 [@Cloern16;@Schraga20].
-This dataset is one of the longest continuous monitoring datasets available in a North American estuary and includes at least monthly sampling at 37 fixed sampling locations along a 145 km longitudinal transect from the lower Sacramento River in the Delta to lower South Bay.
-Water quality constituent measurements available for each station include salinity, temperature, light attenuation, chlorophyll-a, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients.
-The current analysis focused on near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
+Historical resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.
+Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations beginning in 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
+Although visual changes in observed data are apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed, particularly on a seasonal basis.
+
+The analysis evaluated near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
+Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
 Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters.
@@
 
-Generalized Additive Models are an extension of linear models that can incorporate additive smoothed terms to model non-linear effects. The basic structural form is [@Hastie90]:
+The basic structural form of a Generalized Additive Model is [@Hastie90]:
 
@@
 
-where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an interecept term $s_0$.
+where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an intercept term $s_0$.
 The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.
@@
 
-The flexibility of the smoothing functions is a powerful component of GAMs that uses localized, non-linear spline functions to approximate the relationship of $Y$ with predictor $X_j$.
-A spline is a piecewise function, such as a polynomial, that is connected at knots (or *k*-values) that smoothly join the separate pieces of the function.
-Increasing the number of knots creates a smoother fit of the response variable against the predictor at the risk of overfitting, whereas choosing a small number of knots creates a more rigid function that may underfit the data.
-There are several smoothing functions available in the *mgcv* package and the default method uses a thin-plate spline as a computationally efficient approach to estimating the smoother [@Wood03].
-The thin-plate spline can provide a balance between over- and under-fitting to the bivariate smoother by identifying the optimal number of knots through generalized cross validation (GCV).
-In addition to providing an overall summary statistic of model performance, the GCV optimization informs the number of knots for the spline by balancing the fit subject to a smoothing penalty.
-This frees the user from manually selecting knots, and therefore, the degree of smoothness modelled by the function.
-However, the *mgcv* package requires the user to specify an upper limit as a reasonable expection for the number of knots applied to each spline.
-The effect of this choose on model fit is described below.
-
-An additional smoother provided by *mgcv* is the cubic spline, which provides similar functionality as the thin-plate spline, but has an explicit form as a cubic polynomial function that is continuous to the second derivative at each knot location.
-Although visually, smooth fits for different spline types may appear similar if the number of knots are sufficiently high, the cubic spline is useful for modelling periodic variation, such as seasonal changes on the annual scale or diurnal variation at shorter scales.
-A slight modification of the cubic spline is the *cyclic* cubic spline that has the added constrainst of values being equal at the beginning and end of a component being modelled.
-This allows the seasonal component to not only be modelled as a separate component of an overall time series, but the seasonal estimates are joined continuously across years.
-In others words, not using a cyclic cubic spline to model season would cause a non-continous function across the time series with step increases or decreases for the seasonal estimates between the years.
-This discrepancy can likely lead to inaccurate or ambiguous trend estimates when evaluating seasonal differences.
-
 To evaluate GAMs with SFE chlorophyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19].
@@
 The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Constant season" and "Interactive season", respectively.
-Below, we describe the "Constant season*" model as a slight modification of the former.
+We also describe the "Constant season*" model below as a slight modification of the former.
 Each model has a similar functional form for time in relation to chlorophyll with slight differences in the smoothers that describe time.
@@
 
-For each GAM, chlorophyll is modelled using a link function $g()$ that relates the response variable to the independent variables $decYear$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
+For each GAM, chlorophyll was modelled using a link function $g()$ that relates the response variable to the independent variables $decYear$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
 The $decYear$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).
@@
 The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models.
-The $s_1()$ and $s_2()$ functions are different smoothers for $decYear$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component.
+The $s_1()$ and $s_2()$ functions are different smoothers for $decYear$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component [@Wood03].
 The $ti_1()$ function is a tensor product interaction between $decYear$ and $doy$ that allows the seasonal component to change between years.
@@
 
+A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., equation \@ref(eq:gamone) or \@ref(eq:gamtwo)) and determining how much variation could be explained by the smoothers in each function.
+The two models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
+However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, and $ti_1()$) require user input on the potential upper limit for how much variation could be explained by the predictors during model fitting.
+Each smoother is approximated by multiple, localized spline functions that are connected by knots (*k*-values) to create a continuous function.
+Increasing the number of knots creates a smoother fit of the response variable against the predictor, whereas choosing a small number of knots creates a more rigid function that may underfit the data.
+Initial testing by @Murphy19 showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in chlorophyll data from Chesapeake Bay.
+The upper limit for the number of knots for the $s_2(decYear)$ smoother was modified from the default value to the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
+We follow the same approach herein both models.
+Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further.
+For the application to SFE chlorophyll data, an extension of the constant season model was also added as "Constant season*", where the total number of knots for the $decYear$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
+This approach potentially allows the smoother for $decYear$ to be unconstrained in how much variance is fit within and between years.
+
 The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.
@@
 However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.
-The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation (Box and Cox 1964).
+The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation [@Box64].
 The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.
-The profile vector for each chlorophyll time series was obtained using the *boxcox* funtion from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
+The profile vector for each chlorophyll time series was obtained using the *boxcox* function from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
 Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:
@@
 
-A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., equation \@ref(eq:gamone) or \@ref(eq:gamtwo) and determining how much variation could be explained by the smoothers in each function.
-The two models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
-However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, $s_2^*()$ and $ti_1()$) require user input on the potential upper limit for the *k*-values that could be used during model fitting.
-Initial testing by Murphy et al.
-(2019) showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in the response variable.
-Instead, the upper limit for the number of knots for the $s_2(year)$ smoother was specified as the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
-We follow the same approach herein both models.
-Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further.
-For the application to SFE chlorophyll data, an extension of the constant season model was also added as "Constant season*", where the total number of knots for the $decYear$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
-This approach potentially allows the smoother for $decYear$ to be unconstrained in how much variance is fit within and between years.
-
 All models were compared using standard summary statistics describing overall model fit to the observed data.
-In addition to the GCV score, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
+In addition to the Generalized Cross-Validatoin score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
 Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station.
@@
 
-## Trend tests on GAM results
-
-Each of the GAM structures provide different approaches to quantify long-term variation in annual and seasonal changes.
-Although useful as a general approach to view functional forms for different time series, additional information regarding whether an observed change is statistically significant can be useful to inform environmental management decisions.
+## Using GAMs to evaluate trends
+
+Each of the GAM structures provided different approaches to quantify long-term variation in annual and seasonal changes.
+Although useful as a general approach to view functional forms for different time series, additional information on whether an observed change is statistically significant can be useful to inform environmental management decisions.
 We present two methods for performing trend tests on model results that address different changes that may be of management interest.
-First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized.
-This approach is similar to that presented in @Murphy19.
+First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized [@Murphy19].
 Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.
 
-The ability to compare time periods of interest has importance for evaluating whether a significant change has occurred in relation to perceived trends in observed data.
-For example, the noticeable change in SFE water quality in South Bay starting from the 1990s and persisting through 2005 was of considerable interest by environmental managers given the apparent departure from long-term conditions to a state that suggested a trend towards more eutrophic conditions.
-However, statistical methods to evaluate this change that were able to quantify the temporal variation and to provide a statistically robust estimate of the change were not available.
-The use of GAMs in parallel with hypothesis testing can be used for this purpose.
-Methods proposed in @Murphy19 and similarly applied in @Yang20 were used and are briefly summarized herein.
-
-For two time periods of interest, a trend can be estimated as the difference in means from the GAM results between the periods, the percent change, and an appropriate p-value descibing the probability that the change is due to change alone.
+First, for two time periods of interest, trends were estimated as the difference in means from the GAM results between the periods, an estimate of percent change, and an estimate of statistical certainty of the comparison.
 @Murphy19 defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.
-It was further suggested the the base and test periods include a minimum of two years if evaluating years at the beginning and end of a time series to account for instability of GAM estimates at the edges of the independent variable space.
 The differences are estimated as:
@@
 
-The second hypothesis test evaluates changes in seasonal estimates between years using results from a fitted GAM.
-This test provides both an estimate of a seasonal mean, its associated confidence interval, and if the rate of change in the seasonal means has varied over a specific period of interest.
+The second trend test evaluated changes in seasonal estimates between years using results from a fitted GAM.
+This test provided both an estimate of a seasonal mean, its associated confidence interval, and if the rate of change in the seasonal means has varied over a specific period of interest.
 For example, changes that occur during late fall or early spring can be evaluated between years to determine if the seasonal averages differ and if the direction and magnitude of change is significant.
@@
 
-Seasonal averages and confidence intervals were estimated from fitted GAMs by first creating a matrix of the predictors (i.e., `year`, `doy`) covering the seasonal periods across the time series.
+Seasonal averages and confidence intervals were estimated from fitted GAMs by first creating a matrix of the predictors (i.e., `decYear`, `doy`) covering defined seasons in the time series.
 For example, seasonal changes from late summer could be obtained from a prediction matrix covering all dates from Aug 1st to Sep.
@@
 The prediction matrix is then used to obtain fitted values from a GAM that describe the estimated values of chlorophyll that occurred during the periods of interest.
-Rather than the estimated values for the predictions, a linear predictor matrix is returned that provides a parameter vecotr at the supplied covariate values.
-This predictor matrix is then used to estimate the mean seasonal value within a year and associated variance using the sum of predictions from the matrix.
-As a result, each year has a mean value of chlorophyll with an associated confidence interval that describes the modelled expectation for the seasonal period.
-
-Although the seasonal averages obtained from GAMs describe the expected values for each year, describing the direction and magnitude of the seasonal change across years may also be of interest.
-Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected value of chlorophyll and the predictor is year as a numeric value.
-Regression analysis could be used to evaluate changes in the seasonal averages described above, although the fit from the regression would under-estimate the uncertainty in the model parameters by not accounting for the variance in the seasonal averages.
-To better describe the trends in the seasonal averages, we use mixed effects regression meta-analysis to evaluate seasonal changes and their uncertainty over a finite set of years.
-This analysis is an extension of conventional regression by accounting for different soures of uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:
+Rather than the estimated values for the predictions, a linear predictor matrix was obtained that provided a parameter vector for the supplied covariate values.
+This predictor matrix was then used to estimate the mean seasonal value within a year and associated variance using the sum of predictions from the matrix.
+As a result, each year had a mean value of chlorophyll with an associated confidence interval that describes the modelled expectation for the seasonal period.
+
+Although the seasonal averages obtained from GAMs described the expected values for each year, describing the direction and magnitude of the seasonal change across years was also of interest.
+Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected seasonal value of chlorophyll and the predictor is year as a numeric value.
+Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.
+To address this issue, we use mixed effects regression meta-analysis to evaluate seasonal changes and their uncertainty over a finite set of years.
+This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:
 
@@
 
-where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and devation of the seasonal mean from the true mean ($S_i$).
+where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and deviation of the seasonal mean from the true mean ($S_i$).
 Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.
 The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].
-Results appear similar to those from regression analysis (e.g., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.
+Results have a similar interpretation as those from regression analysis (i.e., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.
 
@@
 
-Each of the three GAM structures were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.
-For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoothers terms.
+Each of the three GAM structures (constant season, interactive season, and constant season*) were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.
+For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoother terms.
 Models were also evaluated using different transformations to determine which had improved fit for specific time series.
-Secondary analyses evaluating trends with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.
+Trend analyses with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.
 Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.
@@
 
-* start with description of model performance, which models had the best performance (Const. seas., Int. seas., Const. seas*, boxcox v log)
-* then describe figure 4
-
-Table \@ref(tab:modprftab), Table \@ref(tab:seastrndtab), Figure \@ref(fig:prddat), Figure \@ref(fig:trnddat), Figure \@ref(fig:trndmap)
+## Model performance and observed results
+
+```{r}
+aver2 <- round(100 * mean(modprf$R2), 0)
+minr2 <- round(100 * min(modprf$R2), 0)
+maxr2 <- round(100 * max(modprf$R2), 0)
+minst <- modprf$station[which.min(modprf$R2)]
+minstmod <- modprf$model[which.min(modprf$R2)]
+minsttrn <- modprf$trans[which.min(modprf$R2)]
+minsttrn <- ifelse(is.numeric(minsttrn), 'Box-Cox model', 'log-model')
+maxst <- modprf$station[which.max(modprf$R2)]
+maxstmod <- modprf$model[which.max(modprf$R2)]
+maxsttrn <- modprf$trans[which.max(modprf$R2)]
+maxsttrn <- ifelse(is.numeric(maxsttrn), 'Box-Cox model', 'log-model')
+avecsr2 <- modprf %>% filter(model %in% 'Const seas') %>% pull(R2) %>% mean %>% '*'(100) %>% round(0)
+logr2 <- modprf %>% filter(grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0)
+bxcxr2 <- modprf %>% filter(!grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0)
+```
+
+A total of six models (three GAM structures, two link functions) were fit to each of nine stations, producing summary statistics for 18 different models (Table \@ref(tab:modprftab)).
+Chlorophyll trends were generally well-explained across all models and stations, with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (`r minstmod` `r minsttrn` at station `r minst`) to `r maxr2`% (`r maxstmod` `r maxsttrn` at station `r maxst`).
+The constant season model had the worst performance for all stations (average R$^2$ `r avecsr2`%), with either the interactive season or constant season* model having improved measures of model fit.
+For all but stations 18 and 22, the constant season* model provided significantly lower AIC and GCV scores and higher R-squared values compared to the interactive season models.
+For station 18, the interactive season and constant season* models provided only a marginal improvement over the constant season model.
+For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.
+Moel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance.
+Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.
+
+GAM predictions from north to south on the longitidunal axis of South Bay provided different descriptions of annual and seasonal changes in chlorophyll.
+Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.
+All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record.
+Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location.
+However, the magnitude of the fall peak at more southern stations shifted over time depending on the model.
+In particular, the interactive season model indicates a shift in the magnitude of the fall peak at stations 24, 27, and 32 with maximum concentrations observed around the mid-early 2000s.
+The annual trends for all models suggested similar trends over time, i.e., an increase followed by a recent decrease, whereas the constant season* models with an unconstrained smoother on the `decYear` predictor varied considerably at station 24, 27, and 32.
+Annual trends at these stations were difficult to generalize and had substantial overlap with the seasonal trends.
+Conversely, station 18 had minimal visual differences in annual and seasonal trends across all models, which is reflected by similarities in the model performance statistics in Table \@ref(tab:modprftab).
+
+## Trend comparisons
+
+Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).
+This figure illustrates the narrative decriptions that can be obtained for each test and the differences that can be observed depending on season.
+Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.
+However, the change from 2000 to 2010 was statistically significant with an increase of 78%.
+The period from 2010 to 2017 showed a significant decrease of 39.8%.
+An assessment of the annual differences indicaates if changes overall were observe, but provides no information on seasonal shifts that may have contributed to these differences.
+The results in figure \@ref(fig:trnddat))d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences.
+Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.
+Similarly, rows g-i show the seasonal averages and trend estimates from August to December.
+The seasonal trend results suggest that increase in annual averages from 2000 to 2010 was likely causd by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).
+Seasonal changes from 2010 to 2017 showed that neither the January to July or August to December changes had significant changes, suggesting an overall decrease in chlorophyll regardless of season may have recently occurred.
+
+Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).
+From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.
+The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significiant.
+From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).
+From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.
+The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).
+For the January to July period, significant changes were only observed at the southern stations with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.
+No significant changes were observed for the August to December period.
+
+A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).
+Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).
+However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.
+Trends at staiton 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.
+Station 34 had signifificant increases in the JFM period from 1990-2010 and all but the OND (October, November, December) period for 2000 to 2010.
+Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.
+Station 36 had significant chlorophyll increases in all seasons except OND for the 1990 to 2000 period.
+Accordingly, table \@ref(tab:seastrndtab) demosntrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the alternative seasonal blocks shown in Figure \@ref(fig:trndmap)b.
 
@@
 A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
-Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
+Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.', eval = F}
 totab <- modprf %>% 

### 2020-08-07 R/dat_proc.R

@@ -7,12 +7,12 @@ library(officer)
 source('R/funcs.R')
 
 # model levels and labels
-modlvs <- c('gam0', 'gam1', 'gam2', 'gam6')
-modlbs <- c('gam0', 'Const seas', 'Int seas', 'Const seas*')
+modlvs <- c('gam1', 'gam2', 'gam6')
+modlbs <- c('Const seas', 'Int seas', 'Const seas*')
 names(modlvs) <- modlbs
 
 # best mod
-modbst <- modlbs[4]
+modbst <- modlbs[3]
 
 # # save all models with different naming convention ------------------------
 # 
@@ -29,8 +29,7 @@ modbst <- modlbs[4]
 #   dat <- dat %>% 
 #     filter(model != 'gam0') %>% 
 #     mutate(
-#       model = factor(model, levels = modlvs, labels = modlbs), 
-#       model = fct_drop(model)
+#       model = factor(model, levels = modlvs, labels = modlbs)
 #     )
 #   
 #   assign(nm, dat)
@@ -78,7 +77,8 @@ modprf <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names =
   mutate(
     value = gsub('^data/mods\\_chl|^data/modslog\\_chl|\\.RData$', '', value), 
     p.value = p_ast(p.value), 
-    p.value = ifelse(is.na(p.value), '-', p.value)
+    p.value = ifelse(is.na(p.value), '-', p.value), 
+    model = factor(model, levels = modlbs)
   ) %>% 
   rename(
     station = value, 

### 2020-08-11 manu_draft.Rmd
added trend method comparison data object and text results

@@
 data(chgtrnd)
+data(cmptrnd)
 data(mods_chl18)
@@
 
-Nitrogen and phospohorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.
+Nitrogen and phosphorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.
 However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.
@@
 Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
-Discrete chlorophyll concentrations at each station were determined by fluoremetric analysis with 90% acetone pigment extraction on GFF filters.
+Discrete chlorophyll concentrations at each station were determined by fluorimetric analysis with 90% acetone pigment extraction on GFF filters.
 Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.
@@
 where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an intercept term $s_0$.
-The smoothing functons are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.
+The smoothing functions are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.
 Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.
@@
 The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable.
-However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisying assumptions for GAMs.
+However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisfying assumptions for GAMs.
 The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation [@Box64].
@@
 
-where $g()$ is the transformation function for the appropriate GAM at the esitmated $\lambda$ value.
+where $g()$ is the transformation function for the appropriate GAM at the estimated $\lambda$ value.
 Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.
@@
 All models were compared using standard summary statistics describing overall model fit to the observed data.
-In addition to the Generalized Cross-Validatoin score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
+In addition to the Generalized Cross-Validation score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
 Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station.
@@
 Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.
-To address this issue, we use mixed effects regression meta-analysis to evaluate seasonal changes and their uncertainty over a finite set of years.
+To address this issue, we use mixed effects regression meta-analysis (mixed-meta hereafter) to evaluate seasonal changes and their uncertainty over a finite set of years.
 This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:
@@
 where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and deviation of the seasonal mean from the true mean ($S_i$).
-Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chloropyll averages.
+Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chlorophyll averages.
 The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].
@@
 Trend analyses with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.
-Temporal and spatial patterns within and between stations were also idenfied to better quanity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.
+Temporal and spatial patterns within and between stations were also identified to better quantity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.
 
@@
 For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.
-Moel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance.
+Moeel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance.
 Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.
 
-GAM predictions from north to south on the longitidunal axis of South Bay provided different descriptions of annual and seasonal changes in chlorophyll.
+GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll.
 Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.
@@
 
-## Trend comparisons
+## Trend estimates
 
 Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).
-This figure illustrates the narrative decriptions that can be obtained for each test and the differences that can be observed depending on season.
+This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
 Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.
@@
 The period from 2010 to 2017 showed a significant decrease of 39.8%.
-An assessment of the annual differences indicaates if changes overall were observe, but provides no information on seasonal shifts that may have contributed to these differences.
-The results in figure \@ref(fig:trnddat))d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences.
+An assessment of the annual differences indicated if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
+The results in figure \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences.
 Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.
 Similarly, rows g-i show the seasonal averages and trend estimates from August to December.
-The seasonal trend results suggest that increase in annual averages from 2000 to 2010 was likely causd by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).
-Seasonal changes from 2010 to 2017 showed that neither the January to July or August to December changes had significant changes, suggesting an overall decrease in chlorophyll regardless of season may have recently occurred.
+The seasonal trend results suggested that the increase in annual averages from 2000 to 2010 was likely caused by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).
+Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.
 
@@
 From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.
-The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significiant.
+The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.
 From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).
@@
 The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).
-For the January to July period, significant changes were only observed at the southern stations with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.
+For the January to July period, significant changes were only observed at the southern stations, with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.
 No significant changes were observed for the August to December period.
@@
 However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.
-Trends at staiton 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.
-Station 34 had signifificant increases in the JFM period from 1990-2010 and all but the OND (October, November, December) period for 2000 to 2010.
+Trends at station 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.
+Station 34 had significant increases in the JFM period from 1990-2010 and all other periods the OND (October, November, December) period for 2000 to 2010.
 Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.
-Station 36 had significant chlorophyll increases in all seasons except OND for the 1990 to 2000 period.
-Accordingly, table \@ref(tab:seastrndtab) demosntrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the alternative seasonal blocks shown in Figure \@ref(fig:trndmap)b.
+Station 36 had significant chlorophyll increases in all seasons, except OND for the 1990 to 2000 period.
+Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.
+
+## Trend comparisons
+
+Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.
+Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.
+In addition to mixed-meta analysis, two alternative trend methods were used where the first shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a linear regression model on annual averages derived from the GAM seasonal estimates.
+The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.
+
+The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
+The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
+Slope estimates from the linear model applied to the averages from the GAMs as compared ot the mixed-meta results were identical at almost all stations, excluding station 32 in the January-July period from 1990-2000 when the former showed a negative slope and the latter showed a positive slope.
+The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.
+Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.
 
@@
 
+```{r, results = 'hide', message = F}
+toplo <- cmptrnd
+p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = mod, colour = pval)) + 
+  facet_grid(seas ~ yrs, scales = 'free_x') + 
+  geom_point(size = 3, position = position_dodge(width = 0.5), alpha = 0.6) + 
+  scale_shape('Trend model') + 
+  scale_colour_discrete_diverging(name = 'Trend', palette = 'BlueRed2') +
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  theme_minimal() + 
+  theme(
+    strip.background = element_blank(),
+    panel.background = element_rect(fill = 'gray98', colour = NA), 
+    panel.border = element_blank(), 
+    legend.position = 'top'
+    ) +
+  labs(
+    y = 'Station (north to south)', 
+    x = expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+  )
+
+png('figs/trndcmp.png', height = 5, width = 7.5, family = 'serif', units = 'in', res = 400)
+p
+dev.off()
+```
+```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station.
+The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM constant season* model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM constant season* model.
+Values for each model are the slope esimates as annual change per year within each season, with color denoting significant trends.'}
+knitr::include_graphics('figs/trndcmp.png')
+```
+
 # Tables
 
-```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salnity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.
+```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salinity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.
 Rows are arranged from north to south.'}
@@
 ```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
-Results are for models of chlorophyll-a using BoxCox and logarathmic transformations.
+Results are for models of chlorophyll-a using BoxCox and logarithmic transformations.
 Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `Const seas`),  $\\Delta$ GCV, and R-squared values.

### 2020-08-11 R/dat_proc.R

@@ -283,3 +283,115 @@ chgtrnd <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>%
 save(chgtrnd, file = 'data/chgtrnd.RData', compress = 'xz')
 
 
+# trend model comparisons -------------------------------------------------
+
+cmptrnd <- list.files('data', pattern = '^mods\\_chl', full.names = T) %>% 
+  crossing(
+    fl = ., 
+    tibble(
+      # doystr = c(1, 91, 182, 274), 
+      # doyend = c(90, 181, 273, 364)
+      doystr = c(41, 214), 
+      doyend = c(213, 338)
+    ), 
+    tibble(
+      yrstr = c(1990, 2000, 2010),
+      yrend = c(1999, 2009, 2017)
+    )
+  ) %>% 
+  group_by(fl) %>% 
+  nest() %>% 
+  mutate(
+    mod = purrr::map(fl, function(x){
+      
+      load(file = x)
+      
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+      
+      out <- get(nm) %>% 
+        filter(model == !!modbst) %>% 
+        select(model, modi) %>% 
+        deframe()
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest(c('data')) %>% 
+  group_by(doystr, doyend, mod, fl) %>% 
+  nest() %>% 
+  mutate(
+    avgseas = purrr::pmap(list(mods = mod, doystr, doyend), anlz_avgseason)
+  ) %>% 
+  unnest('data') %>% 
+  mutate(
+    yrstr = purrr::pmap(list(yrstr, avgseas), function(yrstr, avgseas){
+      max(yrstr, min(avgseas$yr))
+    }), 
+    yrend = purrr::pmap(list(yrend, avgseas), function(yrend, avgseas){
+      min(yrend, max(avgseas$yr))
+    }),
+    metatrnd = purrr::pmap(list(avgseas, yrstr, yrend), anlz_mixmeta),
+    lmtrnd = purrr::pmap(list(avgseas, yrstr, yrend), function(avgseas, yrstr, yrend){
+      out <- avgseas %>% 
+        filter(yr >= yrstr & yr <= yrend) %>% 
+        lm(predicted ~ yr, .)
+      return(out)
+    }), 
[…]

### 2020-09-22 manu_draft.Rmd
adding some intro text building off of PDV text

@@
 # Introduction
+
+Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.
+The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.
+At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that are not of direct interest for longer-term trends.
+For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends.
+However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.
+Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.
+
+In this paper we propose analysis methods that separate the signal-extraction of fine-scale data from the trend analysis at longer time-scales, with full propagation of statistical uncertainties between the two analysis stages.
+For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines.
+While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data, alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines.
+Analysis questions or management goals that require information about about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.
+For the trend-analysis stage, error propagation in summary statistics can be addressed using mixed-effects meta-analysis, enabling the application of established tools that incorporate uncertainties in model estimates of summary statistics.
+This two-stage approach is useful because across-year time-scales of interest can be imposed or estimated explicitly, which contrasts with implicit absorption of time-scales into spline formulations in other methods.
+For example, across-year smoothing methods can be applied to seasonal averages from the signal-extraction stage while accurately propagating uncertainties.
+On a practical level, the signal-extraction stage can be viewed as creating a "data product" that can be used in second-stage analyses without the difficulty of embedding every long-term trend question in a GAM modeling framework to model the fine-scale fluctuations.
+
+The smoothing methods under discussion here are distinct from older methods of trend detection, such as seasonal Kendall tests or other non-parametric approaches.
+The former has been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years.
+The limitations are such that only the magnitude and direction of a change between selected a priori years can be identified, nor are any quantitative descriptions of changes occurring at different temporal scales explicitly provided.
+Alternative approaches that decompose a time series into additive or multiplicative components have also been used, such as the seasonal time series composition that isolates a trend, seasonal, and residual component from an observed time series.
+Although computationally efficient, such approaches may over-simplify trends into absolute components that do not smoothly vary over time, nor do they allow for explicit hypothesis testing.
+More recently, moving window regression approaches have been applied to dynamically estimate parameters that influence temporal changes in water quality time series.
+The Weighted Regression on Time, Discharge, and Season (WRTDS) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.
+The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable.
+
+GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.
+The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.
+Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity.
+Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic or management questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).
+Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.
+
+Although our approach is not the first to apply GAMs to environmental monitoring data, our approach is distinct from previous applications in several ways.
+First, some approaches attempt to estimate a parameter such as a seasonal average with the GAM itself [q1].
+This represents a mismatch between time scales and does not make sense as an estimation method if there is really no seasonal treatment factor, for example.
+Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the GAM fitting machinery [q2].
+There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be hard to interpret.
+Perhaps more important, it risks obscuring the different time-scales implicit in choices such as number of spline knots, leaving the rendering results harder to interpret.
+Third…[q3]...
+We illustrate each of these points in more detail below.
+
+We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.
+For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.
+Measurements from water samples include chlorophyll a, [others].
+Analysis is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data could be misleading.
+Examples of long-term trend questions include: Are there significant trends in spring mean chl-a at four-year (or other) time-scales?
+At what across-year scale do within-year summer-fall mean chl-a levels change?
+Are there significant across-year trends in within-year timing of the spring bloom in chl-a or in off-bloom baseline levels?
+We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.
+The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.
+For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evlauating these trends between year.
+Practically, this approach answers if a seasonal change has occurred across years of interest by providing an estimate of the change and its associated confidence as propagated through the GAM.
 

### 2020-09-23 manu_draft.Rmd
editing intro for completeness

@@
 The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.
-At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that are not of direct interest for longer-term trends.
+At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that may not be of direct interest for longer-term trends.
 For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends.
@@
 The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.
-Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity.
-Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic or management questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).
+Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity, including modeling additive annual and seasonal changes and the interaction between the two.
+Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).
 Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.
 
-Although our approach is not the first to apply GAMs to environmental monitoring data, our approach is distinct from previous applications in several ways.
-First, some approaches attempt to estimate a parameter such as a seasonal average with the GAM itself [q1].
-This represents a mismatch between time scales and does not make sense as an estimation method if there is really no seasonal treatment factor, for example.
-Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the GAM fitting machinery [q2].
-There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be hard to interpret.
-Perhaps more important, it risks obscuring the different time-scales implicit in choices such as number of spline knots, leaving the rendering results harder to interpret.
-Third…[q3]...
-We illustrate each of these points in more detail below.
+Although our approach is not the first to apply GAMs for trend analysis environmental monitoring data, our approach is distinct from previous applications in several ways.
+First, some approaches use the GAM smoothing functions to approximate components of a time series to estimate average conditions for a finite period of time.
+This can represent a mismatch between processes occurring at different time scales and does not make sense as an estimation method if, for example, a seasonal treatment factor is not an explicit component of the time series.
+Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the boundaries of the GAM estimation methods.
+There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be confounded.
+Perhaps more important, this approach risks obscuring the different time-scales implicit in choices such as an appropriate number of spline knots, leaving the rendering results more difficult to interpret.
+Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including propogation to secondary analyses, can lead to misinformed decisions.
+GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental component of model application to decision-making, particularly when wrong decisions can lead to expensive consequences.
 
@@
 For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.
-Measurements from water samples include chlorophyll a, [others].
-Analysis is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data could be misleading.
-Examples of long-term trend questions include: Are there significant trends in spring mean chl-a at four-year (or other) time-scales?
-At what across-year scale do within-year summer-fall mean chl-a levels change?
-Are there significant across-year trends in within-year timing of the spring bloom in chl-a or in off-bloom baseline levels?
+Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.
+Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?
+At what across-year scale do within-year summer-fall mean chloropyll levels change?
+Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?
 We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.
 The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.
-For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evlauating these trends between year.
-Practically, this approach answers if a seasonal change has occurred across years of interest by providing an estimate of the change and its associated confidence as propagated through the GAM.
+For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between year.
 

### 2020-09-23 manu_draft.Rmd
adding some refs to intro

@@
 # # extract bib entries from online
-# bib_scrp('manu_draft.Rmd', 'refs.bib')
+bib_scrp('manu_draft.Rmd', 'refs.bib')
 ```
@@
 In this paper we propose analysis methods that separate the signal-extraction of fine-scale data from the trend analysis at longer time-scales, with full propagation of statistical uncertainties between the two analysis stages.
-For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines.
-While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data, alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines.
+For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines [@Wood03;@Wood17].
+While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data [@He06;@Pearce10;@Haraguchi15;@Murphy19], alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines.
 Analysis questions or management goals that require information about about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.
@@
 More recently, moving window regression approaches have been applied to dynamically estimate parameters that influence temporal changes in water quality time series.
-The Weighted Regression on Time, Discharge, and Season (WRTDS) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.
-The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable.
+The Weighted Regression on Time, Discharge, and Season (WRTDS, @Hirsch10;@Beck15) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.
+The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable [@Beck17].
 
 GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.
-The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.
+The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators [@Murphy19].
 Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity, including modeling additive annual and seasonal changes and the interaction between the two.
 Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).
-Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.
+Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias [@Yang20], addressing serial correlation in high resolution data [@Morton08;@Yang20], or addressing different time lags in describing relationships between response and predictor variables [@Lefcheck17;@Testa18].
+Many applications have also evaluated GAMs or other non-linear smoothing methods to account for confounding effects of flow or hydrology changes on trend analysis [@Hirsch10;@Haraguchi15;@Beck17;@Wan17]
 
@@
 Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the boundaries of the GAM estimation methods.
-There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be confounded.
+There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of @Wood17 that the degrees of smoothness attributed to multiple splines can be confounded.
 Perhaps more important, this approach risks obscuring the different time-scales implicit in choices such as an appropriate number of spline knots, leaving the rendering results more difficult to interpret.
-Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including propogation to secondary analyses, can lead to misinformed decisions.
+Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including prorogation to secondary analyses, can lead to misinformed decisions.
 GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental component of model application to decision-making, particularly when wrong decisions can lead to expensive consequences.

### 2020-09-24 manu_draft.Rmd
more intro editing and refs added

@@
 # # extract bib entries from online
-bib_scrp('manu_draft.Rmd', 'refs.bib')
+# bib_scrp('manu_draft.Rmd', 'refs.bib')
 ```
@@
 Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.
-The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.
+The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18].
 At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that may not be of direct interest for longer-term trends.
-For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends.
-However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.
-Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.
+General trends may be easily described by simple smoothing or averaging over time, whereas similar analyses that aggregate or generalize observations can create mismatches between questions of trends at finer scales and what is returned by the summarized results.
+Further, many analysis questions are not easily addressed by simple smoothing methods, particularly when smoothing parameters must be chosen subjectively depending on temporal scales of interest [@Enting87;@Wan17].
 
@@
 For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines [@Wood03;@Wood17].
-While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data [@He06;@Pearce10;@Haraguchi15;@Murphy19], alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines.
-Analysis questions or management goals that require information about about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.
-For the trend-analysis stage, error propagation in summary statistics can be addressed using mixed-effects meta-analysis, enabling the application of established tools that incorporate uncertainties in model estimates of summary statistics.
+While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19], alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines.
+Analysis questions or management goals that require information about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.
+For the trend-analysis stage, error propagation in summary statistics can be addressed using mixed-effects meta-analysis [@Gasparrini12;@Sera19], enabling the application of established tools that incorporate uncertainties in model estimates of summary statistics.
 This two-stage approach is useful because across-year time-scales of interest can be imposed or estimated explicitly, which contrasts with implicit absorption of time-scales into spline formulations in other methods.
@@
 
-The smoothing methods under discussion here are distinct from older methods of trend detection, such as seasonal Kendall tests or other non-parametric approaches.
-The former has been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years.
+The smoothing methods under discussion here are distinct from older and more conventional methods of trend detection, such as seasonal Kendall tests or other non-parametric approaches [see @Helsel20].
+Kendall tests been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].
 The limitations are such that only the magnitude and direction of a change between selected a priori years can be identified, nor are any quantitative descriptions of changes occurring at different temporal scales explicitly provided.
-Alternative approaches that decompose a time series into additive or multiplicative components have also been used, such as the seasonal time series composition that isolates a trend, seasonal, and residual component from an observed time series.
-Although computationally efficient, such approaches may over-simplify trends into absolute components that do not smoothly vary over time, nor do they allow for explicit hypothesis testing.
+Alternative approaches that decompose a time series into additive or multiplicative components have also been used, such as the seasonal time series composition that isolates a trend, seasonal, and residual component from an observed time series [@Cleveland90;@Cloern10;@Stow15].
+Although computationally efficient, such approaches may over-simplify trends into absolute components that do not smoothly vary over time, nor do they allow explicit hypothesis testing.
 More recently, moving window regression approaches have been applied to dynamically estimate parameters that influence temporal changes in water quality time series.
-The Weighted Regression on Time, Discharge, and Season (WRTDS, @Hirsch10;@Beck15) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.
+The Weighted Regression on Time, Discharge, and Season [WRTDS, @Hirsch10;@Beck15] method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.
 The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable [@Beck17].
@@
 
-Although our approach is not the first to apply GAMs for trend analysis environmental monitoring data, our approach is distinct from previous applications in several ways.
+Although our approach is not the first to apply GAMs for trend analysis of environmental monitoring data, our approach is distinct from previous applications in several ways.
 First, some approaches use the GAM smoothing functions to approximate components of a time series to estimate average conditions for a finite period of time.
@@
 Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including prorogation to secondary analyses, can lead to misinformed decisions.
-GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental component of model application to decision-making, particularly when wrong decisions can lead to expensive consequences.
+GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental requirement for decision-making, which can lead to wrong decisions with expensive consequences.
 
@@
 Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.
-To address this issue, we use mixed effects regression meta-analysis (mixed-meta hereafter) to evaluate seasonal changes and their uncertainty over a finite set of years.
+To address this issue, we use mixed-effects regression meta-analysis (mixed-meta hereafter) to evaluate seasonal changes and their uncertainty over a finite set of years.
 This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:

### 2020-10-24 manu_draft.Rmd
intro replaced with Perry edits

--- 2020-09-23 manu_draft.Rmd
@@
 ---
-title: "Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study"
+title: "Multi-scale trend analysis of water quality using error propagation of generalized additive models"
 output: 
@@
 
-Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.
-The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.
-At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that may not be of direct interest for longer-term trends.
-For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends.
-However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.
-Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.
-
-In this paper we propose analysis methods that separate the signal-extraction of fine-scale data from the trend analysis at longer time-scales, with full propagation of statistical uncertainties between the two analysis stages.
-For the signal-extraction stage, we use smoothing splines, or the larger framework of generalized additive models (GAMs) that encompasses smoothing splines.
-While GAMs are a common statistical method and have been used for trend-analysis of environmental monitoring data, alternative and more commonly used approaches can entangle the trend-analysis goal with the signal-extraction goal in the formulation of splines.
-Analysis questions or management goals that require information about about non-trivial temporal features, such as across-year trends in timing of within-year seasonal events or baseline levels, may not be amenable to a single GAM formulation.
-For the trend-analysis stage, error propagation in summary statistics can be addressed using mixed-effects meta-analysis, enabling the application of established tools that incorporate uncertainties in model estimates of summary statistics.
-This two-stage approach is useful because across-year time-scales of interest can be imposed or estimated explicitly, which contrasts with implicit absorption of time-scales into spline formulations in other methods.
-For example, across-year smoothing methods can be applied to seasonal averages from the signal-extraction stage while accurately propagating uncertainties.
-On a practical level, the signal-extraction stage can be viewed as creating a "data product" that can be used in second-stage analyses without the difficulty of embedding every long-term trend question in a GAM modeling framework to model the fine-scale fluctuations.
-
-The smoothing methods under discussion here are distinct from older methods of trend detection, such as seasonal Kendall tests or other non-parametric approaches.
-The former has been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years.
-The limitations are such that only the magnitude and direction of a change between selected a priori years can be identified, nor are any quantitative descriptions of changes occurring at different temporal scales explicitly provided.
-Alternative approaches that decompose a time series into additive or multiplicative components have also been used, such as the seasonal time series composition that isolates a trend, seasonal, and residual component from an observed time series.
-Although computationally efficient, such approaches may over-simplify trends into absolute components that do not smoothly vary over time, nor do they allow for explicit hypothesis testing.
-More recently, moving window regression approaches have been applied to dynamically estimate parameters that influence temporal changes in water quality time series.
-The Weighted Regression on Time, Discharge, and Season (WRTDS) method can describe multi-decadal changes by estimating effects of the predictors that are specific to a given season, year, and flow regime.
-The approach provides results that are comparable to non-linear, spline-based methods, although at the expense of computation time and uncertainty estimates that are not easily obtainable.
-
-GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.
-The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.
-Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity, including modeling additive annual and seasonal changes and the interaction between the two.
-Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).
-Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.
-
-Although our approach is not the first to apply GAMs for trend analysis environmental monitoring data, our approach is distinct from previous applications in several ways.
-First, some approaches use the GAM smoothing functions to approximate components of a time series to estimate average conditions for a finite period of time.
-This can represent a mismatch between processes occurring at different time scales and does not make sense as an estimation method if, for example, a seasonal treatment factor is not an explicit component of the time series.
-Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the boundaries of the GAM estimation methods.
-There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be confounded.
-Perhaps more important, this approach risks obscuring the different time-scales implicit in choices such as an appropriate number of spline knots, leaving the rendering results more difficult to interpret.
-Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including propogation to secondary analyses, can lead to misinformed decisions.
-GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental component of model application to decision-making, particularly when wrong decisions can lead to expensive consequences.
+Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes.
+The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18] [Marcus will go over refs].
+At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).
+Such fluctuations may not be of direct interest for longer-term trends or may not be well-suited to multi-scale smoothing methods.
+However, aggregate features of seasonal patterns that integrate or summarize short-term fluctuations may be of interest.
+In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.
+
+Previous methods for water quality trend analysis can be generalized into four basic approaches.
+Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].
+A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.
+They do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data [is this true?
+I would think so.], do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.
+Thus, while these non-parametric approaches have some degree of robustness, they apply only to narrow goals.
+
+The seasonal trend decomposition using loess (STL) decomposes a time series into additive components of a long-term trend, a seasonal pattern, and residuals [@Cleveland90;@Cloern10;@Stow15].
+While useful and widely applied, this method also has important limitations.
+STL decomposition does not incorporate explanatory variables, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually include standard errors to allow hypothesis testing [but see @Hafen10].
+Conventional STL approaches may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.
+This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].
+
+The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15].
+Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.
+This yields parameters that are specific to season, year, and flow regime.
+Conceptually, the approach is similar to localized multi-polynomial smoothing methods, although the application was developed specifically for describing long-term water quality trends.
+Standard error estimates of predictions are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15].
+Although a useful addition to the original method in @Hirsch10, the approach requires extensive resampling as a post-hoc application to a previously fitted model.
+
+The final and most recent approach is to use smoothing splines to separate fluctuations on different time scales and do so within the larger framework of generalized additive models (GAMs) [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].
+These may be seen as generalizing the concepts behind STL and WRTDS.
+In statistics, the evolution of non-parametric regression methods has largely converged on GAMs rather than more generalized kernel smoothing methods used by both STL and WRTDS.
+Kernel smoothing and spline smoothing are closely related, and a key challenge for each is to determine the appropriate degree of smoothing.
+For example, WRTDS can potentially give results similar to the spline-based smoothing methods described next, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17].
+User input is also required to specify an acceptable degree of smoothing used by the "windows" that define the localized fit of WRTDS at each point in the time series.
+These windows are conceptually similar to the kernel (or bandwidth) that used in more conventional smoothing methods.
+There is no simple rule to guide the choice of defining an appropriate size and a tradeoff between over- and under-smoothing is a hallmark of these approaches.
+
+Compared to kernel smoothing methods, GAMs have various advantages.
+They are formulated using "basis functions", and these can be customized for needs such as cyclic splines (e.g., for an annual pattern) and low-dimensional interactions.
+They can naturally include both parametric (e.g.
+linear or quadratic) components and non-parametric (spline) components.
+Importantly, multiple approaches to automatically determine the optimal degree of smoothness have been developed, based on likelihood and/or optimizing out-of-sample prediction error.
+They have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].
+For these reasons, GAMs are widely used for non-parametric regression smoothing in many fields.
+
+GAMs have recently been applied to trend analysis of water quality time-series, particularly from long-term monitoring programs [@Haraguchi15;@Murphy19], but with different formulations and goals than given here.
+For example, in the US, the Chesapeake Bay Program uses GAMs to decompose time-series into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.
+That is related to the methods here, but this paper gives more general methods for analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.
+Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and different time lags in describing relationships between response and predictor variables [@Lefcheck17;@Testa18].
+
+Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point.
+Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as Chl-a.
+However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages.
+STL and/or WRTDS could potentially separate seasonal from long-term trends, but doing so is not necessary to determine seasonal averages.
+What is needed is an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, allowing for irregular sampling and the non-independence inherent in time-series.
+This can be done with GAMs, but we develop an application that is distinct from previous studies.
+Even if estimates of seasonal averages and their standard errors are available, none of these methods are designed to understand interannual trends in those averages.
+A Kendall test would not incorporate the standard errors or reveal useful long-term patterns beyond a significance test.
+Similarly, STL and WRTDS are not designed for this goal.
 
@@
 Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?
-At what across-year scale do within-year summer-fall mean chloropyll levels change?
+At what across-year scale do within-year summer-fall mean chlorophyll levels change?
 Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?
-We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.
-The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.
-For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between year.
+We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] following signal extraction with GAMs that is new to environmental trend-detection problems.
+For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between years.
 
@@
 ## GAM application
+
+### General approach
+
+The methods proposed here involve three stages.
+In the first stage, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.
+In the second stage, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty.
+In the examples here, the feature is a seasonal average.
+Other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
+In the third stage, a mixed effects meta-analysis is used to estimate trends and/or test hypotheses about the change in seasonal averages across years.
+While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
+They use response data that come with standard errors (uncertainties), which is exactly the need here.
+We give two kinds of mixed effects meta-analysis steps: simple comparison of whether seasonal features differ across years, and estimation of short-term linear trends on time-scales chosen by the analyst.
+These allow analysis of new kinds of questions for long-term water quality monitoring data.
+
+The three-stage approach is motivated in several ways.
+In the first stage, the GAM serves the role of signal-extraction at the scale of the raw data.
+We explain the relationships among various GAM specifications and how some previous methods have entangled the trend-analysis goal with the signal-extraction goal in the formulation of splines, making it harder to interpret either one.
+The second stage harnesses the capability to propagate uncertainty from estimated GAM parameters to features of interest such as seasonal averages.
+Such features are usefully viewed as functions of the fitted GAM rather than parameters to be estimated within the GAM.
+On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they incorporate the uncertainty associated with each point estimate.
+This offers an important practical benefit for visualization and management purposes.
+For the third stage (long-term trend-analysis), we harness existing meta-analysis methods that are well-suited to the task at hand.
 
@@
 
-```{r sumtab, tab.cap = 'Station locations, sample sizes, and median values for salinity, dissolved inorganic nitrogen (DIN, sum of ammonia, nitrate, and nitrite), and chlorophyll.
-Rows are arranged from north to south.'}
+```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chlorophyll.  Rows are arranged from north to south.'}
 totab <- rawdat %>% 
-  filter(param %in% c('sal', 'din', 'chl')) %>% 
+  filter(param %in% c('chl')) %>% 
   group_by(station, param) %>% 
@@
     n = n(), 
-    medv = median(value, na.rm = T)
+    medv = median(value, na.rm = T), 
+    minv = min(value, na.rm = T), 
+    maxv = max(value, na.rm = T)
   ) %>% 
   ungroup() %>% 
-  gather('var', 'val', n, medv) %>% 
+  gather('var', 'val', n, medv, minv, maxv) %>% 
   unite('var', param, var) %>% 
-  mutate(var = factor(var, levels = c('sal_n', 'sal_medv', 'din_n', 'din_medv', 'chl_n', 'chl_medv'))) %>% 
+  mutate(var = factor(var, levels = c('chl_n', 'chl_medv', 'chl_minv', 'chl_maxv'))) %>% 
   left_join(locs, by = 'station') %>% 
@@
     lon = 'Longitude', 
-    sal_n = 'n', 
-    sal_medv = 'Med.', 
-    din_n = 'n', 
-    din_medv = 'Med.', 
     chl_n = 'n', 
-    chl_medv = 'Med.'
-  ) %>% 
-  add_header_row(values = c('', '', '', 'Salinity (psu)', '', 'DIN (mg/L)', '', 'Chl-a (ug/L)', '')) %>% 
-  merge_at(i = 1, j = 4:5, part = 'header') %>% 
-  merge_at(i = 1, j = 6:7, part = 'header') %>% 
-  merge_at(i = 1, j = 8:9, part = 'header') %>% 
-  colformat_num(j = c(5, 7, 9), digits = 1) %>%
+    chl_medv = 'Med.', 
+    chl_minv = 'Min.', 
+    chl_maxv = 'Max.'
+  ) %>% 
+  # add_header_row(values = c('', '', '', 'Chl-a (ug/L)', '', '', '')) %>% 
+  # merge_at(i = 1, j = 4:7, part = 'header') %>% 
+  colformat_num(j = c(5, 6, 7), digits = 1) %>%
   colformat_num(j = c(2, 3), digits = 3) %>% 

### 2020-10-24 manu_draft.Rmd
workingin PErry methods

@@
 
-### General approach
-
 The methods proposed here involve three stages.
-In the first stage, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.
-In the second stage, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty.
-In the examples here, the feature is a seasonal average.
-Other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
-In the third stage, a mixed effects meta-analysis is used to estimate trends and/or test hypotheses about the change in seasonal averages across years.
+First, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.
+Second, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty.
+For this example, the seasonal averages are extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
+Third, a mixed effects meta-analysis is used to estimate trends and test hypotheses about the change in seasonal averages across years.
 While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
-They use response data that come with standard errors (uncertainties), which is exactly the need here.
-We give two kinds of mixed effects meta-analysis steps: simple comparison of whether seasonal features differ across years, and estimation of short-term linear trends on time-scales chosen by the analyst.
-These allow analysis of new kinds of questions for long-term water quality monitoring data.
+Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions.
+Two mixed effects meta-analysis approaches are developed, 1) simple comparison of whether seasonal features differ across years, and 2) estimation of short-term linear trends on time-scales chosen by the analyst.
 
 The three-stage approach is motivated in several ways.
-In the first stage, the GAM serves the role of signal-extraction at the scale of the raw data.
-We explain the relationships among various GAM specifications and how some previous methods have entangled the trend-analysis goal with the signal-extraction goal in the formulation of splines, making it harder to interpret either one.
-The second stage harnesses the capability to propagate uncertainty from estimated GAM parameters to features of interest such as seasonal averages.
-Such features are usefully viewed as functions of the fitted GAM rather than parameters to be estimated within the GAM.
-On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they incorporate the uncertainty associated with each point estimate.
+In the first stage, the GAM is used for signal-extraction at the scale of the raw data and we explain the relationships among various GAM specifications to emphasis concordance between the trend-analysis goal and signal-extraction goal in the formulation of splines.
+The second stage uses mixed-meta analysis to propagate uncertainty from estimated GAM parameters to features of interest, i.e., seasonal averages.
+Such features can be practically evaluated as functions of the fitted GAM, rather than as parameters to be estimated within the GAM.
+On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they include the uncertainty associated with each point estimate.
 This offers an important practical benefit for visualization and management purposes.
@@
 
-The basic structural form of a Generalized Additive Model is [@Hastie90]:
+### First-stage analysis: GAM estimation
+
+Generalized additive models use sums of parametric terms (e.g., linear) and non-parametric terms (e.g., smoothing splines) to predict response data.
+In our approach, GAMs are used to estimate signal and uncertainty from the time-series data.
+Results are subsequently used in the second-stage analysis, described below.
+
+In this section we describe four GAM formulations that can achieve the same or similar fits to data in slightly different ways.
+We present these to clarify relationships that may otherwise be confusing and to emphasize that results can be similarly achieved if a model unconstrained to have sufficient capability to extract a signal from raw data.
+Models will be written in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17].
+
+The fundamental goal is to smooth the raw data across time in order to separate the signal from the noise. The simplest GAM for this purpose can be expressed as:
+
+Model S: `y ~ s(cont_year, k = num_knots_Y)`
+
+Here `y` is the time-series of interest, such as Chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.
+
+For smoothing in `mgcv`, appropriate smoothness is achieved not by choice of knots but by a penalty on the net curvature of the spline [@Wood04].
+Smoothing was determined using generalized cross-validation (GCV, as implemented in `mgcv`), which approximately minimizes out-of-sample prediction error.
+To allow GCV (or other alternatives) to work as intended, the number of knots must be sufficiently large.
+Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.
+Therefore, we choose a sufficiently large number of knots for `num_knots_Y` as 12 times the number of years in the time series that was modeled.
+This created the potential to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.
+In some cases, this upper limit was too large given the observed data and the multiplier was reduced by one until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).
+
+The next three spline formulations differ in how spline terms compose a model to smooth the raw data.
+For convenience, a linear trend in `cont_year` can specified explicitly as a separate component.
+This would be expressed as:
+
+Model SY: `y ~ cont_year + s(cont_year, k = num_knots_Y)`
+
+This model is effectively mathematically equivalent to model S.
+The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
+When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.
+With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.
+With model SY, it is available more conveniently.
+Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.
+For our approach, this option is not used and all models specify `select = FALSE`.
+Details in the supplement explain this justification.
+
+Next, an average within-year cyclic pattern as a separate spline can be separted in the model formulation.  This can be expressed as:
+
+Model SYD: `y ~ cont_year + s(cont_year, k = num_knots_Y) + s(doy, bs = 'cc', k = num_knots_D)`
+
+Here `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.
+While model SYD is not mathematically equivalent to models S and SY, it should be functionally similar.
+In this model, the `doy` spline gives the average pattern of fluctuations within each year.
+This changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.
+
+Models S, SY, and SYD can all potentially extract a similar signal from the raw data.
+What differs is the allocation of penalties for curvature used to determine smoothness for each spline.
+In model SYD, there are separate penalties for the two splines.
+Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17].
+For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations.
+In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.
+
+Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.
+The model will impute an average seasonal pattern where data are limited.
+However, a conceptual representation of what an average within-year pattern represents is challenging.
+For example, suppose in every year includes a spring chl-a peak.
+If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.
+However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.
+The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.
+
+More importantly, the average within-year pattern (and its uncertainty) can be obtained from estimated models S and SY, as well as from model SYD, but using a different approach.
+In models S and SY, the average within-year pattern is simply the average across years of the smoothed within-year patterns.
+From this perspective, model SYD, like model SY, is a convenience, estimating the average within-year pattern directly as a separate estimated spline.
+The average pattern extracted from model S or SY may not be identical to the `doy` spline estimated in model SYD, again due to different smoothing formulations, but in practice they should be similar.
+
+Finally, the raw data can be smoothed using a bivariate spline representing an interaction between `cont_year` and `doy`. This can be expressed as:
+
+Model SYDI: `y ~ cont_year + s(cont_year) + s(doy, bs = "cc") + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knits_D_ti))`
+
+Here, `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.
+The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.
+In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knits_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both to avoid over-parameterization.
+
+Following similar rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.
+Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.
+The splines in `dec_time` and `doy` will likely not capture as much variation as above because of fewer knots.
+The `ti` term represents an interaction by allowing the pattern in `dec_time` to vary by `doy` and vice-versa.
+The interaction term in model SYDI provides an appearance that this model is fundamentally different form those provided by the other models.
+However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.
+Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually.
+By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.
+However, this outcome is difficult to achieve for reasons explained below.
+
+The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly across years.
+For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially.
+For the raw San Francisco Bay data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.
+If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.
+However, if that is the case, the conceptual motivation for model SYDI may not be justified.
+In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.
+If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare.
+
+All of the models require a sufficiently large number of knots in one or more of the spline terms to accommodate the fluctuations in the data and use the GAM estimation methods in the manner for which they were designed.
+In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.
+In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.
+Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.
+In practice, allocation of knots between the different splines can be chosen by the analyst so long it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.
+
+@Murphy19 used a related set of spline formulations but used them differently than we do here.
+@Murphy19 have a "`gam0`" with only a `s(doy)` term, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
+Their `gam0` is *a priori* not of interest for our data and was not explored further.
+Compared to our uses of SYD and SYDI, their uses differ in choice of knots and interpretation of results.
+For SYD, they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we use 12 times the number of years, roughly one per month.
+Thus, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate signal from noise at a scale informed by the data.
+In the SYDI model, @Murphy19 do not explicitly consider number of knots in the interaction spline.
+In both cases, @Murphy19 in effect use choice of number of knots in different spline components to serve as ad hoc allocation of variation in the data to different components while not necessarily allowing any component to fully estimate the signal in the data.
+@Murphy19 acknowledge that incomplete modeling of fluctuations in the data may lead to inflated Type I error rates, which would apply to their later comparisons of changes across time, but they leave that problem for future work.
+For the methods presented here, we seek to avoid inflated Type I error rates arising in this way.
+Finally, @Murphy19 present large AIC differences between their spline formulations.
+We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits.
+In our example, such large differences in AIC would only reflect inadequate choice of knots in one or more splines.
+
+### Log and Box-Cox transformations
+
+Following previous authors, we consider base 10 log-transformed chl-a levels for analysis.
+To generalize this choice, we also consider Box-Cox transformations, which scale from log-transformation to no (i.e., identity) transformation as a parameter $\lambda$ scales from 0 to 1.
+
+### Second-stage analysis: Uncertainty propagation from estimated GAMs to seasonal features
+
+\newcommand{\hm}{\hat{\mu}}
+\newcommand{\hmt}{\hat{\mu}_t}
+\newcommand{\mt}{\mu_t}
+\newcommand{\hmr}{\hat{\mu}_r}
+\newcommand{\hshmt}{\hat{\sigma}_{\hat{\mu}, t}}
+\newcommand{\hshmr}{\hat{\sigma}_{\hat{\mu}, r}}
+\newcommand{\hshm}{\hat{\sigma}_{\hat{\mu}}}
+\newcommand{\hsshm}{\hat{\sigma}^2_{\hat{\mu}}}
+\newcommand{\hsshmt}{\hat{\sigma}^2_{\hat{\mu}, t}}
+\newcommand{\bX}{\mathbf{X}}
+\newcommand{\hby}{\hat{\mathbf{y}}}
+\newcommand{\hbb}{\hat{\boldsymbol{\beta}}}
+\newcommand{\hShb}{\hat{\Sigma}_{\hat{\beta}}}
+\newcommand{\hShy}{\hat{\Sigma}_{\hat{\mathbf{y}}}}
+\newcommand{\sbs}{\sigma_b^2}
+
+Next we consider the goal of estimating a seasonal average, such as the mean spring Chl-A level, and its uncertainty in each year.
+Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
+The season includes $n$ days.
+In what follows, subscript $t$ will be omitted.
+
+Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of estimate parameters and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.
+Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
+For example, using model SY, if a point estimate for Chl-A is needed for a single day, given as `dec_year` = $r$, $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
+The number of spline basis functions is related to the number of knots.
+Note that $r$ can be any time; it need not be the time of an observation.
+
+To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.
+The seasonal averages used in our examples are calculated at the resolution of days.
+The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.
+The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$.
+Since parameter estimates are correlated, it is important to use the full $\hShb$ to determine the covariance of $\hby$.
+This is given by $\hShy = \bX \hShb \bX^T$.
+
+Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.
+The variance of this is $\hsshm = A^T \hShy A$.
+The standard error of the seasonal average is $\hshm$.
+These are all standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.
+@Murphy19 gave a similar summary for the purpose of comparing multi-year averages of month-scale spline values for sets of years at the beginning vs.
+end of a time-series.
+[I note that equations 2-4 of Murphy et al.
+appear to be wrong, giving standard least-squares rather than penalized least-squares.
+I do not necessarily feel we should distract readers with this correction, but we could do so if desired.]
+
+[Here are more detailed notes on @Murphy19.
+I think this is more detail than necessary to include in a paper, but I note it here in case it is helpful for clarity among our group.
+I'm fairly sure we are both doing the same thing, just with slightly different notation.
+@Murphy19 used the above relationships to set up a test of the null hypothesis that `y` is the same at two times (each potentialy an average, but not shown here).
+To do so, $\bX$ would have two rows, one for each time.
+The vector $A^T$ would [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in Chl-A between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.
+One can then determine if the difference is significantly different from 0.
+For me, the most confusing statement of the explanation in @Murphy19 was to call $Z_p$ the "prediction vector".
+If I follow their notation, it is the vector of spline basis function evaluations, so it is used to predict `y` but is not itself a set of predictions.]
+
+### Third-stage analysis: Trend analysis of seasonal features with uncertainties
+
+In this section we explain how to use the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, to investigate across-year trends of within-year features.
+In worked examples, we use as features two seasonal averages of Chl-A, one for January-July and the other for August-December.
+These periods are chosen for their relevance to California's climate.
+(In the *Discussion* we consider other features of interest that may be obtained from estimated GAMs.)  For any feature of interest, as long as its estimates come with associated standard errors, we give meta-analysis methods to address two kinds of questions:
+
+1. Do two years, or a group of years, differ significantly in a seasonal average?
+2.
+Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?
+This question can be posed in a moving-window manner across a time-series.
+
+[A third question would be "3.
+Are there significant trends or non-linear fluctuations across years in a seasonal average, where the time-scale of trends is chosen by statistical criteria?"  This needs further work.
+My current instinct is that this is too much for this paper, but it should be discussed.]
+
+For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.
+
+### Differences among years in seasonal averages
+
+To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.
+To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test.
+The latter readily includes consideration of multiple testing.
+For example, this is provided in R package `multcomp`.
+While this package typically uses the results of a linear or generalized linear model fit as input, it is easy to set up the values of $\hmt$ and $\hshmt$ as inputs.
+Details are given in the Appendix.
+
+### Linear trends across years at a chosen time-scale of interest
+
+The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19].
 
 \begin{equation}
-E\left(Y|X_1, X_2, ..., X_p\right) = s_0 + \sum_{j=1}^{p} s_j\left(X_j\right) 
-(\#eq:gengam)
-\end{equation}
-
-where the expected value of a dependent variable $Y$ conditional on predictors X$_1$ through X$_p$ is the sum of $s_j\left(.\right)$ smoothing functions for each predictor plus an intercept term $s_0$.
-The smoothing functions are standardized and have expected values equal to zero so that $Es_j\left(X_j\right) = 0$.
-Because GAMs are an extension of Generalized Linear Models, they can also include linear predictors in addition to the smoothed functions.
-All methods herein use functions provided by the *mgcv* R package to fit GAMs using multiple parameter smoothing estimation methods [@Wood17].
-
-To evaluate GAMs with SFE chlorophyll data, we followed an approach that built on past work by the Chesapeake Bay Program [@Murphy19].
-We specifically build our models using the `gam1` and `gam2` structures in @Murphy19, but present a different naming convention to better distinguish the functional components of the models.
-The `gam1` and `gam2` models are shown in equations \@ref(eq:gamone) and \@ref(eq:gamtwo), denoted as "Constant season" and "Interactive season", respectively.
-We also describe the "Constant season*" model below as a slight modification of the former.
-Each model has a similar functional form for time in relation to chlorophyll with slight differences in the smoothers that describe time.
-
-\begin{equation}
-\boldsymbol{Constant\ season:} g\left(chl\right) = s_0 + \beta_1 decYear + s_1\left(doy\right) + s_2\left(decYear\right) + \epsilon 
-(\#eq:gamone)
-\end{equation}
-
-\begin{equation}
-\boldsymbol{Interactive\ season:} g\left(chl\right) = s_0 + \beta_1 decYear + s_1\left(doy\right) + s_2\left(decYear\right) + ti_1(doy, decYear) + \epsilon
-(\#eq:gamtwo)
-\end{equation}
-
-For each GAM, chlorophyll was modelled using a link function $g()$ that relates the response variable to the independent variables $decYear$ and day of year ($doy$) (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
-The $decYear$ variable is a continuous measure for date in decimal time, where date is expressed as a continuous numeric value with year on the left of the decimal and date within the year to the right of the decimal (e.g., July 1st 2019 would be 2019.5).
-The $doy$ variable is an integer value for the Julian day in each year (from 1 to 366) that represents the seasonal component of the time series.
-The estimate $s_0$ is an intercept value and $\beta_1$ is a slope estimate for a linear year effect included in all models.
-The $s_1()$ and $s_2()$ functions are different smoothers for $decYear$ and $doy$ where the former uses a thin-plate spline and the latter uses a cyclic cubic spline to capture the periodic seasonal component [@Wood03].
-The $ti_1()$ function is a tensor product interaction between $decYear$ and $doy$ that allows the seasonal component to change between years.
-Within the $ti_1()$ function are the two separate smoothers for year as a thin-plate spline and $doy$ as a cyclic cubic spline.
-As with conventional linear models, the remaining variance in the response not captured by the functional form of the model is expressed as an error term $\epsilon$ that is normally-distributed having mean equal to zero and variance as $\sigma^2$.
-The residual variance is used to back-transform model estimates for trend testing, described below.
-
-A challenge for optimizing GAMs to time series data with significant intra- and interannual variation is choosing an appropriate functional form (i.e., equation \@ref(eq:gamone) or \@ref(eq:gamtwo)) and determining how much variation could be explained by the smoothers in each function.
-The two models above provide tradeoffs in the functional forms that balance descriptive and computational efficiency (Table \@ref(tab:modsumtab), Figure \@ref(fig:modsumfig)).
-However, the individual smoothing functions (i.e, $s_1()$, $s_2()$, and $ti_1()$) require user input on the potential upper limit for how much variation could be explained by the predictors during model fitting.
-Each smoother is approximated by multiple, localized spline functions that are connected by knots (*k*-values) to create a continuous function.
-Increasing the number of knots creates a smoother fit of the response variable against the predictor, whereas choosing a small number of knots creates a more rigid function that may underfit the data.
-Initial testing by @Murphy19 showed that the default *k* values in the mgcv package were insufficient for describing the interannual variation in chlorophyll data from Chesapeake Bay.
-The upper limit for the number of knots for the $s_2(decYear)$ smoother was modified from the default value to the maximum between 10 or  $2 / 3 \times$ the number of years in the time series.
-We follow the same approach herein both models.
-Further, the default *k*-values for $s_2()$ and $ti_1()$ were considered appropriate and not tested further.
-For the application to SFE chlorophyll data, an extension of the constant season model was also added as "Constant season*", where the total number of knots for the $decYear$ smoother, $s_2^*()$, was increased to 12 times the total number of years in the time series.
-This approach potentially allows the smoother for $decYear$ to be unconstrained in how much variance is fit within and between years.
-
-The link functions $g()$ for each model were chosen as either logarithmic transformations (base 10) or best estimates of Box-Cox power transformations for chlorophyll.
-A comparison of GAM performance between transformation methods has not yet been explored in the application of GAMs to long-term water quality monitoring data.
-The logarithmic transformation is commonly used for chlorophyll to approximate a log-normal distribution of the response variable.
-However, alternative transformations could provide improved model fits and reduced uncertainty estimates in the fitted parameters by better satisfying assumptions for GAMs.
-The Box-Cox method was also used to transform chlorophyll as a comparison to the logarithmic transformation [@Box64].
-The Box-Cox method requires an estimate of the parameter $\lambda$, whereby the the optimal value is based on a minimization function of a log-likelihood profile vector.
-The profile vector for each chlorophyll time series was obtained using the *boxcox* function from the MASS package [@Venables02] to evaluate log-likelihood estimates of $\lambda$ across the range -4 to 4.
-Once the optimal $\lambda$ value was identified, chlorophyll was transformed using the following power transformation [@Box64]:
-
-\begin{equation}
-g\left(chl;\lambda\right) = \begin{cases} \frac{chl ^ {\lambda} - 1}{\lambda} & \lambda \ne 0 \\ \log\left(chl\right) & \lambda = 0 \end{cases}
-(\#eq:boxcox)
-\end{equation}
-
-where $g()$ is the transformation function for the appropriate GAM at the estimated $\lambda$ value.
-Note that in some cases the Box-Cox transformation is a logarithmic transformation if $\lambda$ is estimated as zero.
-The *BoxCox* function from the forecast package [@Hyndman20] in R was used to transform the chlorophyll time series with the optimal $\lambda$ value.
-Equation \@ref(eq:boxcox) is only appropriate for positive values of a response variable, as for chlorophyll.
-
-All models were compared using standard summary statistics describing overall model fit to the observed data.
-In addition to the Generalized Cross-Validation score provided by the mgcv package, models were compared using the Akaike Information Criterion (AIC) and $R^2$ summmary statistics.
-Comparisons were made between the different GAMs and for the transformations used to define the $g()$ function for chlorophyll at each station.
-Model predictions were based on the standard model output from the *predict()* function from the mgcv package.
-A long-term trend independent of seasonal variation was also estimated for each model by subtracting the seasonal terms (i.e., $s_1(doy)$ and $ti_1(doy, year)$) from the predictions.
-Back-transformation of results were obtained by exponentiation if the log-transformed variable was used or using the *InvBoxCox()* function from the forecast package if the Box-Cox transformation was used.
-For hypothesis tests that required estimates of central tendency and confidence intervals for mean values, back-transformations were based on an exponentiation or inverse function of the transformed response variable as before, with the addition of a correction factor.
-This bias-correction factor was simply the variance of the model residuals divided by two for variables in log-space and correction with a forecast variance estimate as specified in the *InvBoxCox()* function from the forecast package.
-
-## Using GAMs to evaluate trends
-
-Each of the GAM structures provided different approaches to quantify long-term variation in annual and seasonal changes.
-Although useful as a general approach to view functional forms for different time series, additional information on whether an observed change is statistically significant can be useful to inform environmental management decisions.
-We present two methods for performing trend tests on model results that address different changes that may be of management interest.
-First, a method for evaluating the percent change in annual averages between two time periods is briefly summarized [@Murphy19].
-Second, a method for evaluating means and confidence intervals from GAM results for seasonal periods is presented, including an estimate of direction and rate of seasonal change over time.
-
-First, for two time periods of interest, trends were estimated as the difference in means from the GAM results between the periods, an estimate of percent change, and an estimate of statistical certainty of the comparison.
-@Murphy19 defined the two time periods as "base" and "test" periods, where the former represents a baseline condition earlier in the time series as a basis of comparison to the more recent "test" period.
-The differences are estimated as:
-
-\begin{equation}
-\text{Difference} = Z_d\hat{\beta}
-(\#eq:diffest)
-\end{equation}
-
-where $Z_d$ is a point estimate of the differences of the predictions between the base and test years and $\hat{\beta}$ is the estimated parameter vector from the fitted GAM.
-The standard error of this difference is then:
-
-\begin{equation}
-\text{se}_{\text{diff}} = \sqrt{Z_d\hat{\Sigma}_\beta Z_d^T}
-(\#eq:sediffeest)
-\end{equation}
-
-defined as the square root of the product of $Z_d$, the estimated variance-covariance matrix of parameter vector for the fitted GAM $\hat{\Sigma}_\beta$, and the transposition of $Z_d$.
-The trend is considered significant if the appropriate confidence interval at a specified $\alpha$ does not include zero.
-
-The second trend test evaluated changes in seasonal estimates between years using results from a fitted GAM.
-This test provided both an estimate of a seasonal mean, its associated confidence interval, and if the rate of change in the seasonal means has varied over a specific period of interest.
-For example, changes that occur during late fall or early spring can be evaluated between years to determine if the seasonal averages differ and if the direction and magnitude of change is significant.
-This has implications for understanding potential changes in seasonal bloom phenology resulting from shifting system responses to nutrient inputs or changes in the latter.
-
-Seasonal averages and confidence intervals were estimated from fitted GAMs by first creating a matrix of the predictors (i.e., `decYear`, `doy`) covering defined seasons in the time series.
-For example, seasonal changes from late summer could be obtained from a prediction matrix covering all dates from Aug 1st to Sep.
-30th for the period from 2000 to 2010.
-The prediction matrix is then used to obtain fitted values from a GAM that describe the estimated values of chlorophyll that occurred during the periods of interest.
-Rather than the estimated values for the predictions, a linear predictor matrix was obtained that provided a parameter vector for the supplied covariate values.
-This predictor matrix was then used to estimate the mean seasonal value within a year and associated variance using the sum of predictions from the matrix.
-As a result, each year had a mean value of chlorophyll with an associated confidence interval that describes the modelled expectation for the seasonal period.
-
-Although the seasonal averages obtained from GAMs described the expected values for each year, describing the direction and magnitude of the seasonal change across years was also of interest.
-Conventionally, trend analysis of changes in a response variable could be achieved with ordinary least squares regression, where the response variable is the expected seasonal value of chlorophyll and the predictor is year as a numeric value.
-Although this approach could be used to estimate a seasonal rate of change, the linear fit through the seasonal means does not account for the variance in the seasonal averages.
-To address this issue, we use mixed-effects regression meta-analysis (mixed-meta hereafter) to evaluate seasonal changes and their uncertainty over a finite set of years.
-This analysis is an extension of conventional regression by accounting for uncertainty attributed to both variation of the true seasonal mean around the trend and variation of the estimated seasonal mean around the true seasonal mean:
-
-\begin{equation}
-chl_i = \beta_0 + \beta_1 year_i + b_i + \epsilon_i
+\hmt = \beta_0 + \beta_t t + b_t + \epsilon_t
 (\#eq:mixmet)
@@
 
-where the estimated value of chlorophyll at year $i$ for the predetermined season is a function of the predicted seasonal mean $\beta_0 +\beta_1 year_i$ plus deviation of the seasonal mean from the trend ($b_i$) and deviation of the seasonal mean from the true mean ($S_i$).
-Both the error terms are assumed to be normally distributed with mean zero and variance for $b_i$ equal to an unknown term and variance for $\epsilon_i$ equal to the squared standard error from the seasonal chlorophyll averages.
-The R package *mixmeta* was used to estimate trends in the seasonal averages over time [@Sera19].
-Results have a similar interpretation as those from regression analysis (i.e., slope and intercept estimates), although parameter uncertainty is more accurately estimated with mixed meta-analysis.
-
-## Model application and interpretation
-
-Each of the three GAM structures (constant season, interactive season, and constant season*) were used to fit models to the chlorophyll time series at each of the nine stations from central to lower South Bay.
-For each model and station, the fit was assessed using the standard summary statistics described above (i.e., GCV, AIC, and $R^2$) and relative significance of the individual smoother terms.
-Models were also evaluated using different transformations to determine which had improved fit for specific time series.
-Trend analyses with fitted GAM predictions were also compared between stations and model types, particularly to determine if and when significant results varied with alternative methods.
-Temporal and spatial patterns within and between stations were also identified to better quantity changes in chlorophyll production throughout the period of record using GAMs to demonstrate improved descriptions relative to interpretations from the observed data.
+where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$.
+Here the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.
+The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.
+This is assumed to be normally disributed with mean 0 and variance $\hsshmt$, which is assumed known from the calculations above.
+The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$, what would be considered the "residual" in the sense of unexplained variation not due to the estimation error.
+This is assumed normal with mean 0 and unknown variance, $\sbs$, to be estimated.
+
+The R package *mixmeta* was used to estimate this model [@Sera19].
+Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
+Following meta-analysis theory, $mixmeta$ handles this model as a linear mixed effects model with some variance components fixed.
+The default estimation method is restricted maximum likelihood (REML).
+
+We apply this method for windows of a chosen width of years.
+In this way, one can ask whether there is a significant linear trend in Chl-A over a chosen series of years, taking into account all the data in those years.
+Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error.
+The meta-analysis incorporates such differences in uncertainty magnitudes.
 
@@
 
+When `select = TRUE` is included, the comparison between models S and SY changes.
+This option tells `mgcv` to penalize the linear term's coefficient in the spline.
+This would be appropriate if `cont_year` was an explanatory variable subject to variable selection -- which it is not here -- but it would be a strange choice if one is explicitly including a linear term for `cont_year`.
+If `select = TRUE` is used, models S and SY would still be effectively equivalent, but it would appear by AIC selection that one model is superior.
+That would be an artifact of the strange choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
+A link to an RMarkdown file demonstrating this issue is provided.
+
 ```{r modprfboxtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).

### 2020-11-02 manu_draft.Rmd
methods from PDV added to manu

@@
 Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point.
-Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as Chl-a.
+Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.
 However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages.
@@
 
-The fundamental goal is to smooth the raw data across time in order to separate the signal from the noise. The simplest GAM for this purpose can be expressed as:
+The fundamental goal is to smooth the raw data across time to separate a signal of response variable from noise. The simplest GAM for this purpose can be expressed as:
 
@@
 
-Here `y` is the time-series of interest, such as Chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.
+Here `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.
+Following previous authors [@Murphy19;@Yang20], we consider log-transformed (base 10) chlorophyll levels for analysis.
 
@@
 
-This model is effectively mathematically equivalent to model S.
+This model is effectively equivalent to model S.
 The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
@@
 However, a conceptual representation of what an average within-year pattern represents is challenging.
-For example, suppose in every year includes a spring chl-a peak.
+For example, suppose in every year includes a spring chlorophyll peak.
 If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.
@@
 Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.
-In practice, allocation of knots between the different splines can be chosen by the analyst so long it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.
+In practice, allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.
 
@@
 In our example, such large differences in AIC would only reflect inadequate choice of knots in one or more splines.
-
-### Log and Box-Cox transformations
-
-Following previous authors, we consider base 10 log-transformed chl-a levels for analysis.
-To generalize this choice, we also consider Box-Cox transformations, which scale from log-transformation to no (i.e., identity) transformation as a parameter $\lambda$ scales from 0 to 1.
 
@@
 
-Next we consider the goal of estimating a seasonal average, such as the mean spring Chl-A level, and its uncertainty in each year.
+Next we consider the goal of estimating a seasonal average, such as the mean spring chlorophyll concentrations and the estimated uncertainty in each year.
 Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
 The season includes $n$ days.
-In what follows, subscript $t$ will be omitted.
+For simplicity, the following text omits subscript $t$.
 
@@
 Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
-For example, using model SY, if a point estimate for Chl-A is needed for a single day, given as `dec_year` = $r$, $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
+For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
 The number of spline basis functions is related to the number of knots.
-Note that $r$ can be any time; it need not be the time of an observation.
+Note that $r$ can be any time, not necessarily the time of an observation.
 
 To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.
-The seasonal averages used in our examples are calculated at the resolution of days.
+Here, the seasonal averages used in our examples are calculated at the resolution of days.
 The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.
@@
 Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.
-The variance of this is $\hsshm = A^T \hShy A$.
-The standard error of the seasonal average is $\hshm$.
-These are all standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.
-@Murphy19 gave a similar summary for the purpose of comparing multi-year averages of month-scale spline values for sets of years at the beginning vs.
-end of a time-series.
-[I note that equations 2-4 of Murphy et al.
-appear to be wrong, giving standard least-squares rather than penalized least-squares.
-I do not necessarily feel we should distract readers with this correction, but we could do so if desired.]
-
-[Here are more detailed notes on @Murphy19.
-I think this is more detail than necessary to include in a paper, but I note it here in case it is helpful for clarity among our group.
-I'm fairly sure we are both doing the same thing, just with slightly different notation.
-@Murphy19 used the above relationships to set up a test of the null hypothesis that `y` is the same at two times (each potentialy an average, but not shown here).
-To do so, $\bX$ would have two rows, one for each time.
-The vector $A^T$ would [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in Chl-A between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.
-One can then determine if the difference is significantly different from 0.
-For me, the most confusing statement of the explanation in @Murphy19 was to call $Z_p$ the "prediction vector".
-If I follow their notation, it is the vector of spline basis function evaluations, so it is used to predict `y` but is not itself a set of predictions.]
+The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.
+Each of these estimates are standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.
+A similar summary for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs.
+end of a time-series is provided by @Murphy19.
+For example, the above estimates can be used for a test of the null hypothesis that `y` is the same at two times.
+To do so, $\bX$ would have two rows for each time to compare.
+The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.
+This difference can then be used to determine if the two estimates in time are significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates.
 
@@
 
-In this section we explain how to use the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, to investigate across-year trends of within-year features.
-In worked examples, we use as features two seasonal averages of Chl-A, one for January-July and the other for August-December.
-These periods are chosen for their relevance to California's climate.
-(In the *Discussion* we consider other features of interest that may be obtained from estimated GAMs.)  For any feature of interest, as long as its estimates come with associated standard errors, we give meta-analysis methods to address two kinds of questions:
+This section explains how the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, are used to investigate across-year trends of within-year features.
+In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.
+These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the San Francisco Bay region.
+For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:
 
@@
 
-[A third question would be "3.
-Are there significant trends or non-linear fluctuations across years in a seasonal average, where the time-scale of trends is chosen by statistical criteria?"  This needs further work.
-My current instinct is that this is too much for this paper, but it should be discussed.]
-
 For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.
@@
 To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test.
-The latter readily includes consideration of multiple testing.
-For example, this is provided in R package `multcomp`.
-While this package typically uses the results of a linear or generalized linear model fit as input, it is easy to set up the values of $\hmt$ and $\hshmt$ as inputs.
+The latter readily includes consideration of multiple testing, which is provided in the R package `multcomp` [@Hothorn08].
+While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.
 Details are given in the Appendix.
@@
 
-The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19].
+The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19]:
 
@@
 The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.
-This is assumed to be normally disributed with mean 0 and variance $\hsshmt$, which is assumed known from the calculations above.
-The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$, what would be considered the "residual" in the sense of unexplained variation not due to the estimation error.
-This is assumed normal with mean 0 and unknown variance, $\sbs$, to be estimated.
-
-The R package *mixmeta* was used to estimate this model [@Sera19].
+The residuls is assumed to be normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.
+The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.
+The random effect is assumed a normal distribution with mean 0 and unknown variance, $\sbs$, to be estimated.
+
+The R package *mixmeta* was used to estimate the model in equation \@ref{eqn:mixmet} [@Sera19].
 Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
-Following meta-analysis theory, $mixmeta$ handles this model as a linear mixed effects model with some variance components fixed.
-The default estimation method is restricted maximum likelihood (REML).
+Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed, whereas others are estimated ($\sbs$).
+The default estimation method for meta-analysis models is restricted maximum likelihood (REML).
 
 We apply this method for windows of a chosen width of years.
-In this way, one can ask whether there is a significant linear trend in Chl-A over a chosen series of years, taking into account all the data in those years.
+This approach provide evaluates whether there is a significant linear trend in chlorophyll over a chosen series of years, by providing a more comprehensive assessment of all the data in those years (i.e., the seasonal averages, plus their uncertainty).
 Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error.
-The meta-analysis incorporates such differences in uncertainty magnitudes.
+The meta-analysis explicitly includes these differences in uncertainty magnitudes.
 

### 2020-11-02 manu_draft.Rmd
fix to figure one decades in facets

@@
   scale_y_log10(ylab) +
+  facet_wrap(~yrcat, ncol = 3) + 
   theme_minimal(base_family = 'serif') +
@@
     legend.title = element_blank(), 
-    axis.title.x = element_blank()
+    legend.position = 'none', 
+    axis.title.x = element_blank(), 
+    axis.text.x = element_text(angle = 40, hjust = 1), 
+    strip.background = element_blank()
   ) +
@@
 p <- p1 + p2 + plot_layout(ncol = 1)
-png('figs/obsdat.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
+png('figs/obsdat.png', height = 7, width = 8, family = 'serif', units = 'in', res = 400)
 print(p)

### 2020-11-05 manu_draft.Rmd
reprocessed all models and summary data with model S

@@
 data(cmptrnd)
-data(mods_chl18)
-data(mods_chl24)
-data(mods_chl27)
-data(mods_chl32)
-data(mods_chl34)
+data(modslog_chl18)
+data(modslog_chl24)
+data(modslog_chl27)
+data(modslog_chl32)
+data(modslog_chl34)
 
@@
 
-This model is effectively equivalent to model S.
+This model is effectively and mathematically equivalent to model S.
 The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
@@
 
-Next, an average within-year cyclic pattern as a separate spline can be separted in the model formulation.  This can be expressed as:
+Next, an average within-year cyclic pattern as a separate spline can be separated in the model formulation.  This can be expressed as:
 
@@
 Models S, SY, and SYD can all potentially extract a similar signal from the raw data.
-What differs is the allocation of penalties for curvature used to determine smoothness for each spline.
-In model SYD, there are separate penalties for the two splines.
+What differs between the models is the allocation of penalties for curvature used to determine smoothness for each spline.
+In model SYD, there are separate penalties for the two splines, as compared to S and SY that include penalties only for the `cont_year` spline.
 Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17].
@@
 However, a conceptual representation of what an average within-year pattern represents is challenging.
-For example, suppose in every year includes a spring chlorophyll peak.
-If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude and appeas as a "typical" year.
+For example, suppose a spring chlorophyll peak is a notable feature in every year.
+If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude thta appears as a "typical" year.
 However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.
@@
 Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.
-The splines in `dec_time` and `doy` will likely not capture as much variation as above because of fewer knots.
-The `ti` term represents an interaction by allowing the pattern in `dec_time` to vary by `doy` and vice-versa.
-The interaction term in model SYDI provides an appearance that this model is fundamentally different form those provided by the other models.
+The splines in `cont_year` and `doy` will likely not capture as much variation as above because of fewer knots.
+The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.
+The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.
 However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.
@@
     axis.title.x = element_blank(), 
-    axis.text.x = element_text(angle = 40, hjust = 1), 
+    axis.text.x = element_text(angle = 45, hjust = 1), 
     strip.background = element_blank()

### 2020-11-05 R/dat_proc.R

@@ -1,46 +1,62 @@
 # libraries
 library(tidyverse)
 library(wqtrends)
+library(mgcv)
 library(flextable)
 library(officer)
+library(lubridate)
 
 source('R/funcs.R')
 
-# model levels and labels
-modlvs <- c('gam1', 'gam2', 'gam6')
-modlbs <- c('Const seas', 'Int seas', 'Const seas*')
-names(modlvs) <- modlbs
+# chlorophyll models, model S only ----------------------------------------
 
-# best mod
-modbst <- modlbs[3]
+modssta <- rawdat %>% 
+  filter(param == 'chl') %>% 
+  group_by(station) %>% 
+  nest() %>% 
+  mutate(
+    model = purrr::pmap(list(station, data), function(station, data){
+      
+      cat(station, '\n')
+      moddat <- anlz_trans(data, trans = 'log10')
+      out <- gam(value ~ s(cont_year, k = 360),
+        knots = list(doy = c(1, 366)),
+        data = moddat,
+        na.action = na.exclude,
+        select = F)
+      
+      out$trans <- 'log10'
+      
+      return(out)
+      
+    })
+  )
 
-# # save all models with different naming convention ------------------------
-# 
-# mods <- list.files('data', pattern = '^mods\\_chl|modslog\\_chl', full.names = T)
-# 
-# for(mod in mods){
-#   
-#   cat(mod, '\t')
-#   
-#   load(file = mod)
-#   nm <- gsub('\\.RData$', '', basename(mod))
-#   dat <- get(nm)
-#   
-#   dat <- dat %>% 
-#     filter(model != 'gam0') %>% 
-#     mutate(
-#       model = factor(model, levels = modlvs, labels = modlbs)
-#     )
-#   
-#   assign(nm, dat)
-#   
-#   save(list = nm, file = mod, compress = 'xz')
-#     
-# }
+# separate models into diff files by stations
+tosv <- modssta %>% 
+  pull(station) %>% 
+  unique
+
+for(i in seq_along(tosv)){
+  
+  cat(i, 'of', length(tosv), '\n')
+  
+  sta <- tosv[i]
+  
+  fl <- modssta %>% 
+    filter(station %in% !!sta) 
+  
+  flnm <- paste0('modslog_chl', sta)
+  
+  assign(flnm, fl)
+  
+  save(list = flnm, file = paste0('data/', flnm, '.RData'), compress = 'xz')
+  
+}
 
[…]

### 2020-11-05 manu_draft.Rmd
analyses extended to 2019

@@
 data(cmptrnd)
+data(modstr)
 data(modslog_chl18)
@@
 source('R/funcs.R')
-
-# model levels and labels
-modlbs <- c('Const seas', 'Int seas', 'Const seas*')
-modbst <- modlbs[3]
 
@@
 Discrete chlorophyll concentrations at each station were determined by fluorimetric analysis with 90% acetone pigment extraction on GFF filters.
-Data collected between 1990-2017 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.
+Data collected between 1990-2019 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.
 While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).
@@
 
-## Model performance and observed results
-
-```{r}
-aver2 <- round(100 * mean(modprf$R2), 0)
-minr2 <- round(100 * min(modprf$R2), 0)
-maxr2 <- round(100 * max(modprf$R2), 0)
-minst <- modprf$station[which.min(modprf$R2)]
-minstmod <- modprf$model[which.min(modprf$R2)]
-minsttrn <- modprf$trans[which.min(modprf$R2)]
-minsttrn <- ifelse(is.numeric(minsttrn), 'Box-Cox model', 'log-model')
-maxst <- modprf$station[which.max(modprf$R2)]
-maxstmod <- modprf$model[which.max(modprf$R2)]
-maxsttrn <- modprf$trans[which.max(modprf$R2)]
-maxsttrn <- ifelse(is.numeric(maxsttrn), 'Box-Cox model', 'log-model')
-avecsr2 <- modprf %>% filter(model %in% 'Const seas') %>% pull(R2) %>% mean %>% '*'(100) %>% round(0)
-logr2 <- modprf %>% filter(grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0)
-bxcxr2 <- modprf %>% filter(!grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0)
-```
-
-A total of six models (three GAM structures, two link functions) were fit to each of nine stations, producing summary statistics for 18 different models (Table \@ref(tab:modprftab)).
-Chlorophyll trends were generally well-explained across all models and stations, with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (`r minstmod` `r minsttrn` at station `r minst`) to `r maxr2`% (`r maxstmod` `r maxsttrn` at station `r maxst`).
-The constant season model had the worst performance for all stations (average R$^2$ `r avecsr2`%), with either the interactive season or constant season* model having improved measures of model fit.
-For all but stations 18 and 22, the constant season* model provided significantly lower AIC and GCV scores and higher R-squared values compared to the interactive season models.
-For station 18, the interactive season and constant season* models provided only a marginal improvement over the constant season model.
-For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.
-Moeel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance.
-Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.
-
-GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll.
-Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.
-All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record.
-Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location.
-However, the magnitude of the fall peak at more southern stations shifted over time depending on the model.
-In particular, the interactive season model indicates a shift in the magnitude of the fall peak at stations 24, 27, and 32 with maximum concentrations observed around the mid-early 2000s.
-The annual trends for all models suggested similar trends over time, i.e., an increase followed by a recent decrease, whereas the constant season* models with an unconstrained smoother on the `decYear` predictor varied considerably at station 24, 27, and 32.
-Annual trends at these stations were difficult to generalize and had substantial overlap with the seasonal trends.
-Conversely, station 18 had minimal visual differences in annual and seasonal trends across all models, which is reflected by similarities in the model performance statistics in Table \@ref(tab:modprftab).
-
-## Trend estimates
-
-Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).
-This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
-Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.
-However, the change from 2000 to 2010 was statistically significant with an increase of 78%.
-The period from 2010 to 2017 showed a significant decrease of 39.8%.
-An assessment of the annual differences indicated if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
-The results in figure \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences.
-Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.
-Similarly, rows g-i show the seasonal averages and trend estimates from August to December.
-The seasonal trend results suggested that the increase in annual averages from 2000 to 2010 was likely caused by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).
-Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.
-
-Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).
-From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.
-The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.
-From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).
-From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.
-The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).
-For the January to July period, significant changes were only observed at the southern stations, with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.
-No significant changes were observed for the August to December period.
-
-A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).
-Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).
-However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.
-Trends at station 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.
-Station 34 had significant increases in the JFM period from 1990-2010 and all other periods the OND (October, November, December) period for 2000 to 2010.
-Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.
-Station 36 had significant chlorophyll increases in all seasons, except OND for the 1990 to 2000 period.
-Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.
-
-## Trend comparisons
-
-Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.
-Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.
-In addition to mixed-meta analysis, two alternative trend methods were used where the first shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a linear regression model on annual averages derived from the GAM seasonal estimates.
-The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.
-
-The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
-The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
-Slope estimates from the linear model applied to the averages from the GAMs as compared ot the mixed-meta results were identical at almost all stations, excluding station 32 in the January-July period from 1990-2000 when the former showed a negative slope and the latter showed a positive slope.
-The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.
-Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.
-
 # Discussion
@@
       yr >= 2000 & yr < 2010 ~ '2000-2010', 
-      yr >= 2010 ~ '2010-2017'
+      yr >= 2010 ~ '2010-2019'
     ), 
-    yrcat = factor(yrcat, levels = c('1990-2000', '2000-2010', '2010-2017'))
+    yrcat = factor(yrcat, levels = c('1990-2000', '2000-2010', '2010-2019'))
   )
@@
 
-```{r, results = 'hide'}
+```{r, results = 'hide', eval = F}
 # get predictions for all three gams
@@
 ```{r modsumfig, fig.cap = "General output for each of the GAM structures for constant season, interactive season, and constant season*.
-Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration."}
+Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration.", eval = F}
 knitr::include_graphics('figs/modsumfig.png')
@@
 
-```{r, results = 'hide', message = F}
+```{r, results = 'hide', message = F, eval = F}
 # get predictions to plot
-prdplo <- bind_rows(mods_chl18, mods_chl24, mods_chl27, mods_chl32) %>% 
-  mutate(prddat = pmap(list(data, modi), function(data, modi){
+prdplo <- bind_rows(modslog_chl18, modslog_chl24, modslog_chl27, modslog_chl32) %>% 
+  mutate(prddat = pmap(list(data, model), function(data, model){
 
@@
 
-    prd <- predict(modi, newdata = prddat) %>% 
+    prd <- predict(model, newdata = prddat) %>% 
       as.numeric
 
-    annprd <- predict(modi, newdata = prddat, type = 'terms')
+    annprd <- predict(model, newdata = prddat, type = 'terms')
     int <- attr(annprd, 'constant')
@@
 
-    trans <- unique(modi$trans)
+    trans <- unique(model$trans)
 
@@
   })) %>% 
-  select(station, modi = model, prddat) %>% 
+  select(station, model = model, prddat) %>% 
   unnest(prddat)
@@
   ) + 
-  facet_grid(modi ~ station) +
+  facet_grid(model ~ station) +
   scale_y_log10(limits = c(0.74, 83)) + 
@@
   guides(colour = guide_colourbar(barheight = 10, barwidth = 1)) +
-  facet_grid(modi ~ station) +
+  facet_grid(model ~ station) +
   scale_y_log10() + 
@@
 Observed data in (a) are shown with the gray points.
-Station locations are in Figure \\@ref(fig:sitemap)."}
+Station locations are in Figure \\@ref(fig:sitemap).", eval = F}
 knitr::include_graphics('figs/prddat.png')
@@
 ```{r, results = 'hide', message = F}
-toplo <- mods_chl34 %>% 
-  filter(model %in% !!modbst) %>% 
-  pull(modi)
+toplo <- modslog_chl34 %>%
+  pull(model)
 
@@
   )
-p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2017, ylab = NULL) +
-  labs(
-    subtitle = '(c) 2010, 2017, Base: 10.1, Test: 6.1,\nChange: -39.8%, p < 0.01', 
+p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2019, ylab = NULL) +
+  labs(
+    subtitle = '(c) 2010, 2019, Base: 10.1, Test: 6.1,\nChange: -39.8%, p < 0.01', 
     title = NULL
@@
   )
-p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2016, ylab = NULL) +
+p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
   scale_y_continuous(limits = ylim) + 
   labs(
-    subtitle = '(f) 2010-2017, Jan-Jul, slope -0.4, ns', 
+    subtitle = '(f) 2010-2019, Jan-Jul, slope -0.4, ns', 
     title = NULL
@@
   )
-p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2017, ylab = NULL) +
+p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
   scale_y_continuous(limits = ylim) + 
   labs(
-    subtitle = '(i) 2010-2017, Aug-Dec, slope -0.52, ns', 
+    subtitle = '(i) 2010-2019, Aug-Dec, slope -0.52, ns', 
     title = NULL
@@
     ) %>% 
-  left_join(locs, by = 'station') %>% 
-  filter(model %in% !!modbst)
+  left_join(locs, by = 'station')
 
@@
 
-```{r modsumtab}
+```{r modsumtab, eval = T}
 tab <- tibble(
-  GAM = c('`Constant season`', '`Interactive season`', '`Constant season*`'),
-  Summary = c(
-    'Nonlinear trends seasonality (constrained knots)', 
-    'Nonlinear trends wiht seasonality, plus interaction',
-    'Nonlinear trends with seasonality (unconstrained knots)'
+  GAM = c('`S`', '`SY`', '`SYD`', '`SYDI`'),
+  `Additive components` = c(
+    '`s(cont_year)`', 
+    '`cont_year + s(cont_year)`',
+    '`cont_year + s(cont_year) + s(doy)`', 
+    '`cont_year + s(cont_year) + s(doy) + ti(cont_year, doy)`'
   ),
   Details = c(
-    'Treats year and doy with separate smoothers with a smooth non-linear trend through time, such that the seasonal peaks are similar across years, but interannual change is evident', 
-    'Treats cyear and doy same as `Const seas`; and Within-year seasonal fluctuations that vary across years but in a constrained pattern compared to `Const seas*` (`ti(decYear, doy, bs = c("tp", "cc"))`)',
-    'Variant of `Const seas` with much higher values of k, allowing the splines to follow greater fluctuations both within and across years.')
+    'A single smoother over a continuous year variable', 
+    'A linear continuous year variable and a single smoother over a continuous year variable', 
+    'A linear continuous year variable, a smoother over a continuous year variable, and smoother over a day of year variable',
+    'A linear continuous year variable, a smoother over a continuous year variable, smoother over a day of year variable, and an interactin smoother across continuous year and day of year variables')
   )
@@
 # table stuff
-cap.val <- 'Summary and details for each of the GAM structures.'
+cap.val <- 'Summary and details for each of the GAM structures.
+In practive, a sufficiently large number of knots provided to the additive terms will produce identical or comparable estimates for a response variable.
+The models differ in the allocation of penalties for the smoothness of each spline (`s()`).'
 
@@
 
-```{r modprftab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
-Results are for models of chlorophyll-a using BoxCox and logarithmic transformations.
-Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `Const seas`),  $\\Delta$ GCV, and R-squared values.
-Probability values ("P-val") show the results of Analysis of Variance comparisons between all BoxCox or logarithmic models at a station.
-The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`Const seas` is the base model and does not have a p-value).
-A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
-ns: not significant at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
-
+```{r modstrtab}
+totab <- modstr %>% 
+  enframe() %>% 
+  mutate(value = purrr::map(value, function(x){
+
+    GCV <- as.numeric(x$gcv.ubre)
+    R2 <- summary(x)$r.sq 
+    smooths <- summary(x)$s.table %>% data.frame %>% tibble::rownames_to_column('smoother')
+
+    out <- data.frame(GCV = GCV, R2 = R2, smooths)
+
+    return(out)
+
+  })) %>% 
+  unnest('value') %>% 
+  select(-Ref.df) %>% 
+  mutate(p.value = p_ast(p.value)) %>% 
+  mutate_if(is.numeric, round, 2) %>% 
+  mutate(
+    GCV = ifelse(duplicated(name), '', GCV),
+    R2 = ifelse(duplicated(name), '', R2),
+    name = ifelse(duplicated(name), '', name)
+  ) %>% 
+  rename(
+    model = name, 
+    `p-val` = p.value
+  )
+
+# table stuff
+cap.val <- 'Comparison of the four model structures (S, SY, SYD, SYDI) described in the first stage analysis of GAM estimation.
+The four models provide either identical or comparable ability to describe chlorophyll trends at an example station (32) in South San Francisco Bay.
+The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are comparable.
+GCV: generalized cross-validation score, R2: r-squared values for predictions, edf: effective degrees of freedom, F: F-statistic, p-val: probability value, ** p < 0.001'
+
+# table
+knitr::kable(totab, booktabs = T, caption = cap.val)
+```
+
+```{r modprftab}
 totab <- modprf %>% 
-  gather('stat', 'val', AIC, GCV, R2, `P-value`) %>% 
-  mutate(trans = case_when(
-    trans != 'log10' ~ 'BoxCox', 
-    T ~ trans)
-  ) %>% 
-  unite('stat', trans, stat) %>% 
-  mutate(
-    stat = factor(stat, levels = c('BoxCox_AIC', 'log10_AIC', 'BoxCox_GCV', 'log10_GCV', 'BoxCox_R2', 'log10_R2', 'BoxCox_P-value', 'log10_P-value'))
-  ) %>% 
-  spread(stat, val) %>% 
-  mutate_at(c('BoxCox_AIC', 'log10_AIC', 'BoxCox_GCV', 'log10_GCV', 'BoxCox_R2', 'log10_R2'), as.numeric) %>% 
-  group_by(station) %>%
-  mutate(
-    BoxCox_AIC = BoxCox_AIC - max(BoxCox_AIC),
-    log10_AIC = log10_AIC - max(log10_AIC),
-    BoxCox_GCV = BoxCox_GCV - max(BoxCox_GCV),
-    log10_GCV = log10_GCV - max(log10_GCV)#,
-    # minaic = AIC == min(AIC), 
-    # mingcv = GCV == min(GCV),
-    # maxr2 = R2 == max(R2)
-  ) %>% 
-  ungroup
-
-# boldaic <- as.numeric(rownames(totab)[totab$minaic])
-# boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
-# boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
-
-totab <- totab %>% 
-  # select(-minaic, -mingcv, -maxr2) %>% 
-  mutate(
-    station = ifelse(duplicated(station), '', station)
-  )
-
-tab <- flextable(totab) %>% 
-  set_header_labels(
-    station = '', 
-    model = '', 
-    BoxCox_AIC = 'Delta AIC',
-    log10_AIC = 'Delta AIC',
-    BoxCox_GCV = 'Delta GCV',
-    log10_GCV = 'Delta GCV',
-    BoxCox_R2 = 'R-squared',
-    log10_R2 = 'R-squared',
-    `BoxCox_P-value` = 'P-val',
-    `log10_P-value` = 'P-val'
-  ) %>% 
-  merge_at(i = 1, j = 3:4, part = "header") %>% 
-  merge_at(i = 1, j = 5:6, part = "header") %>% 
-  merge_at(i = 1, j = 7:8, part = "header") %>% 
-  merge_at(i = 1, j = 9:10, part = "header") %>% 
-  add_header_row(values = c('Station', 'GAM', 'BoxCox', 'Log', 'BoxCox', 'Log', 'BoxCox', 'Log', 'BoxCox', 'Log'), top = F) %>% 
-  align(j = c(3:10), align = 'center', part = 'all') %>% 
-  colformat_num(j = c(3, 4), digits = 1) %>%
-  colformat_num(j = c(5, 6), digits = 3) %>% 
-  colformat_num(j = c(7, 8), digits = 2) %>% 
-  # bold(i = boldaic, j = 'AIC') %>% 
-  # bold(i = boldgcv, j = 'GCV') %>% 
-  # bold(i = boldr2, j = 'R2') %>% 
-  border_remove() %>% 
-  hline_top(border = fp_border(color = 'black')) %>% 
-  font(fontname = 'Times', part = 'all') %>%
-  font(j = 2, fontname = 'Consolas') %>% 
-  fontsize(size = 8, part = 'all') %>% 
-  autofit()
-
-tab
+  mutate_if(is.numeric, round, 2) %>% 
+  rename(`R-squared` = R2)
+
+cap <- 'Model performance statistics for each station as generalized cross-validation scores (GCV) and r-squared values.'
+
+knitr::kable(totab, booktabs = T, caption = cap)
 ```
@@
 
-```{r modprfboxtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
-Results are for models of chlorophyll-a using BoxCox transformations.
-Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC, GCV, and R-squared values.
-Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
-Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
-The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
-A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
-Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.', eval = F}
-totab <- modprf %>% 
-  filter(trans != 'log10') %>% 
-  select(-trans) %>% 
-  group_by(station) %>% 
-  mutate(
-    AIC = AIC - max(AIC),
-    minaic = AIC == min(AIC), 
-    mingcv = GCV == min(GCV),
-    maxr2 = R2 == max(R2)
-  ) %>% 
-  ungroup
-
-boldaic <- as.numeric(rownames(totab)[totab$minaic])
-boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
-boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
-
-totab <- totab %>% 
-  select(-minaic, -mingcv, -maxr2) %>% 
-  mutate(
-    station = ifelse(duplicated(station), '', station)
-  )
-
-tab <- flextable(totab) %>% 
-  set_header_labels(
-    station = 'Station', 
-    model = 'GAM', 
-    AIC = 'Delta AIC',
-    R2 = 'R-squared'
-  ) %>% 
-  colformat_num(j = 3, digits = 1) %>%
-  colformat_num(j = c(4, 5), digits = 2) %>% 
-  bold(i = boldaic, j = 'AIC') %>% 
-  bold(i = boldgcv, j = 'GCV') %>% 
-  bold(i = boldr2, j = 'R2') %>% 
-  border_remove() %>% 
-  hline_top(border = fp_border(color = 'black')) %>% 
-  font(fontname = 'Times', part = 'all') %>%
-  font(j = 2, fontname = 'Consolas') %>% 
-  fontsize(size = 12, part = 'all') %>% 
-  autofit()
-
-tab
-```
-
-```{r modprflogtab, tab.cap = 'Model performance statistics for each station and GAM structure (Table \\@ref(tab:modsumtab)).
-Results are for models of chlorophyll-a using logarithmic (base-10) ransformations.
-Three GAMs were fit to each station and performance is summarized as $\\Delta$ AIC (from `gam0`), GCV, and R-squared values.
-Models with the lowest AIC, GCV and highest R-squared values for each station are emphasized.
-Probability values ("P-value") show the results of Analysis of Variance comparisons between all models at a station.
-The p-value for a row refers to the comparison of the model in the current row to the one preceeding (`gam0` is the base model and does not have a p-value).
-A significant p-value indicates the model in the current row provides a statistically significant difference in explaining chlorophyll variance from the model in the preceding row.
-Values indicate ns for not significantly different at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.', eval = F}
-totab <- modprf %>% 
-  filter(trans == 'log10') %>% 
-  select(-trans) %>% 
-  group_by(station) %>% 
-  mutate(
-    AIC = AIC - max(AIC),
-    minaic = AIC == min(AIC), 
-    mingcv = GCV == min(GCV),
-    maxr2 = R2 == max(R2)
-  ) %>% 
-  ungroup
-
-boldaic <- as.numeric(rownames(totab)[totab$minaic])
-boldgcv <- as.numeric(rownames(totab)[totab$mingcv])
-boldr2 <- as.numeric(rownames(totab)[totab$maxr2])
-
-totab <- totab %>% 
-  select(-minaic, -mingcv, -maxr2) %>% 
-  mutate(
-    station = ifelse(duplicated(station), '', station)
-  )
-
-tab <- flextable(totab) %>% 
-  set_header_labels(
-    station = 'Station', 
-    model = 'GAM', 
-    AIC = 'Delta AIC',
-    R2 = 'R-squared'
-  ) %>% 
-  colformat_num(j = 3, digits = 1) %>%
-  colformat_num(j = c(4, 5), digits = 2) %>% 
-  bold(i = boldaic, j = 'AIC') %>% 
-  bold(i = boldgcv, j = 'GCV') %>% 
-  bold(i = boldr2, j = 'R2') %>% 
-  border_remove() %>% 
-  hline_top(border = fp_border(color = 'black')) %>% 
-  font(fontname = 'Times', part = 'all') %>%
-  font(j = 2, fontname = 'Consolas') %>% 
-  fontsize(size = 12, part = 'all') %>% 
-  autofit()
-
-tab
-```
-
 # References

### 2020-11-05 R/dat_proc.R

@@ -5,6 +5,7 @@ library(mgcv)
 library(flextable)
 library(officer)
 library(lubridate)
+library(mgcv)
 
 source('R/funcs.R')
 
@@ -99,7 +100,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     ), 
     tibble(
       yrstr = c(1990, 2000, 2010),
-      yrend = c(1999, 2009, 2017)
+      yrend = c(1999, 2009, 2019)
     )
   ) %>% 
   group_by(fl) %>% 
@@ -150,7 +151,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     yrs = case_when(
       yrs < 1995 ~ '1990-2000', 
       yrs >=1995 & yrs < 2005 ~ '2000-2010', 
-      yrs >= 2005 ~ '2010-2016'
+      yrs >= 2005 ~ '2010-2019'
     ), 
     yrs = factor(yrs)
   ) %>% 
@@ -169,7 +170,7 @@ seastrnd2 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     ), 
     tibble(
       yrstr = c(1991, 2000, 2010),
-      yrend = c(1999, 2009, 2016)
+      yrend = c(1999, 2009, 2019)
     )
   ) %>% 
   group_by(fl) %>% 
@@ -220,7 +221,7 @@ seastrnd2 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     yrs = case_when(
       yrs < 1995 ~ '1990-2000', 
       yrs >=1995 & yrs < 2005 ~ '2000-2010', 
-      yrs >= 2005 ~ '2010-2016'
+      yrs >= 2005 ~ '2010-2019'
     ), 
     yrs = factor(yrs)
   ) %>% 
@@ -235,7 +236,7 @@ chgtrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     fl = ., 
     tibble(
       yrstr = c(1990, 2000, 2010),
-      yrend = c(1999, 2009, 2017)
+      yrend = c(1999, 2009, 2019)
     )
   ) %>% 
   group_by(fl) %>% 
@@ -268,7 +269,7 @@ chgtrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     yrs = case_when(
[…]

### 2020-11-05 manu_draft.Rmd
adding updated figures

@@
 
-```{r, results = 'hide', eval = F}
-# get predictions for all three gams
-tomod <- rawdat %>%
-  filter(station %in% 22) %>%
-  filter(param %in% 'chl')
-
+```{r, results = 'hide'}
 # use previously fitted list of models
-trans <- 'boxcox'
-mods <- list(
-  `Const seas` = anlz_gam(tomod, mod = 'gam1', trans = trans),
-  `Int seas` = anlz_gam(tomod, mod = 'gam2', trans = trans),
-  `Const seas*` = anlz_gam(tomod, mod = 'gam6', trans = trans)
-)
-
-p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8, faclev = modlbs) + 
-  theme(legend.position = 'right') + 
+mods <- modstr[2:4]
+
+thm <-theme_minimal() + 
+  theme(
+    legend.title = element_blank(),
+    legend.position = 'right'
+  )
+lncol <- 'tomato1'
+alph <- 0.5
+
+p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8) + 
   guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
   scale_colour_gradient(low = "white", high = "black") + 
-  theme_minimal() + 
-  theme(
-    legend.title = element_blank()
-  )
-
-png('figs/modsumfig.png', height = 2.5, width = 6, units = 'in', family = 'serif', res = 400)
-print(p1)
+  thm + 
+  labs(
+    title = '(a) Model predictions by day of year'
+  )
+
+prds <- anlz_prd(mods = mods, trans = 'log10')
+
+toplo2 <- prds %>% 
+  select(model, date, value) %>% 
+  spread(model, value)
+
+toplo3 <- prds %>% 
+  select(model, date, annvalue) %>% 
+  spread(model, annvalue)
+
+p2a <- ggplot(toplo2, aes(x = SY, y = SYD)) + 
+  geom_point(alpha = alph) + 
+  geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  thm + 
+  labs(
+    title = '(b) Estimated chlorophyll (log-space) between models'
+  )
+p2b <- ggplot(toplo2, aes(x = SYD, y = SYDI)) + 
+  geom_point(alpha = alph) + 
+  geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  thm
+p2c <- ggplot(toplo2, aes(x = SY, y = SYDI)) + 
+  geom_point(alpha = alph) + 
+  geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  thm
+
+p2 <- p2a + p2b + p2c + plot_layout(ncol = 3)
+
+p3a <- ggplot(toplo3, aes(x = SY, y = SYD)) + 
+  geom_point(alpha = alph) + 
+  geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  thm +
+  labs(
+    title = '(c) Estimated s(cont_year) between models'
+    )
+p3b <- ggplot(toplo3, aes(x = SYD, y = SYDI)) + 
+  geom_point(alpha = alph) + 
+  geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  thm
+p3c <- ggplot(toplo3, aes(x = SY, y = SYDI)) + 
+  geom_point(alpha = alph) + 
+  geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  thm
+
+p3 <- p3a + p3b + p3c + plot_layout(ncol = 3)
+
+out <- p1 / p2 / p3
+png('figs/modsumfig.png', height = 7, width = 8, units = 'in', family = 'serif', res = 400)
+print(out)
 dev.off()
 ```
-```{r modsumfig, fig.cap = "General output for each of the GAM structures for constant season, interactive season, and constant season*.
-Day of year (Julian) is on the x-axis and each curve represents a different fitted year for chlorophyll concentration.", eval = F}
+```{r modsumfig, fig.cap = "GAM output of estimated chlorophyll for models SY, SYD, and SYDI.
+Model S is identical to SY and is not shown.
+Plots in (a) show model predictions by day of year with separate lines for each year.
+Plots in (b) show pairwise comparisons of predicted chlorophyll between the models and plots in (c) show the same comparisons as in (b) but only for results from the esimated smoother for the `cont_year` variable.
+The plots demonstrate that results between the models are comparable except for a few observations at extreme values(a), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (b).
+The 1:1 lines are in red to faciliate comparisons."}
 knitr::include_graphics('figs/modsumfig.png')
@@
 
-```{r, results = 'hide', message = F, eval = F}
+```{r, results = 'hide', message = F}
 # get predictions to plot
-prdplo <- bind_rows(modslog_chl18, modslog_chl24, modslog_chl27, modslog_chl32) %>% 
-  mutate(prddat = pmap(list(data, model), function(data, model){
-
-    prddat <- data.frame(
-        dec_time = seq(min(data$dec_time), max(data$dec_time), length = 1000)
+modprds <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
+  enframe() %>% 
+  group_by(value) %>% 
+  nest %>% 
+  mutate(
+    prd = purrr::map(value, function(x){
+
+      load(file = x)
+
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+
+      mod <- get(nm) %>% 
+        pull(model)
+
+      prd <- anlz_prd(mods = mod) %>% 
+        anlz_backtrans()
+
+      return(prd)
+
+    })
   ) %>% 
+  ungroup %>% 
+  select(station = value, prd) %>% 
   mutate(
-        date = date_decimal(dec_time),
-        date = as.Date(date),
-        mo = month(date, label = TRUE),
-        doy = yday(date),
-        yr = year(date)
-      )
-
-    prd <- predict(model, newdata = prddat) %>% 
-      as.numeric
-
-    annprd <- predict(model, newdata = prddat, type = 'terms')
-    int <- attr(annprd, 'constant')
-    value <- rowSums(annprd) + int
-    annvalue <- rowSums(annprd[, !grepl('doy', colnames(annprd)), drop = FALSE]) + int
-    annvalue <- as.numeric(annvalue)
-
-    trans <- unique(model$trans)
-
-    if(!inherits(trans, 'numeric'))
-      stop('log used')
-
-    prddat <- prddat %>% 
-      mutate(
-        chl = prd,
-        chl = forecast::InvBoxCox(chl, trans), 
-        annchl = forecast::InvBoxCox(annvalue, trans)
-        )
-
-    return(prddat)
-
-  })) %>% 
-  select(station, model = model, prddat) %>% 
-  unnest(prddat)
-
-ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))
-
-toplo <- prdplo %>% 
-  filter(yr >= 1991)
-
-obsdat <- rawdat %>% 
-  filter(station %in% c(18, 24, 27, 32)) %>% 
-  filter(param == 'chl') %>% 
-  filter(yr >= 1991) %>% 
-  rename(chl = value)
-
-colpal <- "PuBuGn"
-
-p1 <- ggplot(toplo, aes(x = date, y = chl)) + 
-  geom_point(data = obsdat, color = 'grey', size = 0.15) + 
-  geom_line(size = 0.5) +
-  geom_line(aes(y = annchl), color = 'tomato1', size = 0.5, alpha = 0.9) +
-  theme_minimal(base_family = 'serif') + 
-  theme(
+    station = gsub('^data/modslog\\_chl|\\.RData$', '', station)
+  ) %>% 
+  unnest('prd')
+
+modobs <- rawdat %>% 
+  filter(param == 'chl')
+ylab <- 'Chlorophyll-a (ug/L)'
+
+p1 <- ggplot2::ggplot(modprds, ggplot2::aes(x = date)) + 
+  ggplot2::geom_point(data = modobs, ggplot2::aes(y = value), size = 0.5, colour = 'grey') +
+  ggplot2::geom_line(ggplot2::aes(y = value), size = 0.75, alpha = 0.7) + 
+  ggplot2::facet_grid(station~.) +
+  ggplot2::scale_color_viridis_d() + 
+  ggplot2::theme_minimal(base_family = 'serif', base_size = 14) + 
+  ggplot2::scale_y_log10() +
+  ggplot2::theme(
+    legend.position = 'top', 
+    legend.title = ggplot2::element_blank(),
+    strip.background = ggplot2::element_blank()
+  ) + 
+  ggplot2::labs(
+    y = ylab,
+    x = 'Continuous year', 
+    subtitle = '(a) Continous predictions'
+  )
+
+ p2 <- ggplot2::ggplot(modprds, ggplot2::aes(x = doy, group = factor(yr), colour = yr)) + 
+    ggplot2::geom_line(ggplot2::aes(y = value), size = 0.5, alpha = 1) + 
+    ggplot2::theme_minimal(base_family = 'serif', base_size = 14) + 
+    ggplot2::theme(
       legend.position = 'right', 
-    legend.title = element_blank(), 
-    # axis.text.x= element_text(size = 8)
-    strip.background = element_blank()
+      legend.title = ggplot2::element_blank(), 
+      strip.background = ggplot2::element_blank(), 
+      strip.text = ggplot2::element_blank(), 
+      axis.title.y = ggplot2::element_blank()
     ) + 
-  facet_grid(model ~ station) +
-  scale_y_log10(limits = c(0.74, 83)) + 
-  labs(x = 'Date', subtitle = '(a) Model predictions with annual estimates', y = ylab)
-
-p2 <- ggplot(toplo, aes(x = doy, colour = yr, group = factor(yr))) + 
-  geom_line(aes(y = chl)) + 
-  theme_minimal(base_family = 'serif') + 
-  theme(
-    legend.position = 'right', 
-    legend.title = element_blank(), 
-    # axis.text.x= element_text(size = 8)
-    strip.background = element_blank()
-  ) + 
-  scale_colour_continuous_sequential(palette = colpal, rev = T, begin = 0.2) +
-  guides(colour = guide_colourbar(barheight = 10, barwidth = 1)) +
-  facet_grid(model ~ station) +
-  scale_y_log10() + 
-  labs(x = 'Day of year', subtitle = '(b) Model predictions by day of year', y = ylab)
-
-p <- p1 + p2 + plot_layout(ncol = 1)
+    ggplot2::scale_colour_gradient(low = "white", high = "black") + 
+    ggplot2::facet_grid(station ~ .) +
+    ggplot2::scale_y_log10() +
+    ggplot2::guides(colour = ggplot2::guide_colourbar(barheight = 20, barwidth = 1)) +
+    ggplot2::labs(
+      x = "Day of year", 
+      subtitle = '(b) Day of year\npredictions'
+    )
+p <- p1 + p2 + plot_layout(ncol = 2, width = c(1, 0.4))
 
@@
 ```
-```{r prddat, fig.cap = "GAM predictions for select stations from north to south (18, 24, 27, 32) for each of the GAM structures for constant season, interactive season, and constant season*.
-The results show (a) predictions across the time series with annual estimates (red line) and (b) predictions by day of year with one smooth per year.
+```{r prddat, fig.cap = "GAM predictions for all stations from north to south for model S.
+The results show (a) predictions across the time series and (b) predictions by day of year.
 Observed data in (a) are shown with the gray points.
-Station locations are in Figure \\@ref(fig:sitemap).", eval = F}
+Station locations are in Figure \\@ref(fig:sitemap)."}
 knitr::include_graphics('figs/prddat.png')
@@
 Trends with $p<0.05$ are marked with an asterisk.
-All results are from `Const seas*`."}
+All results are from Model `S`."}
 knitr::include_graphics('figs/trndmap.png')

### 2020-11-05 manu_draft.Rmd
text about moving window

@@
 We apply this method for windows of a chosen width of years.
-This approach provide evaluates whether there is a significant linear trend in chlorophyll over a chosen series of years, by providing a more comprehensive assessment of all the data in those years (i.e., the seasonal averages, plus their uncertainty).
+This approach evaluates whether there is a significant linear trend in chlorophyll over a chosen series of years, by providing a more comprehensive assessment of all the data in those years (i.e., the seasonal averages, plus their uncertainty).
 Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error.
 The meta-analysis explicitly includes these differences in uncertainty magnitudes.
+
+As a final analysis, the chosen windows for evaluating a trend for a selected seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not.
+Although the initial window width and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.
+The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.
+We choose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
+Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.
 

### 2020-11-08 manu_draft.Rmd
window change figure added

@@
 data(seastrnd2)
+data(seastrnd3)
 data(chgtrnd)
@@
 
-As a final analysis, the chosen windows for evaluating a trend for a selected seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not.
+As a final analysis, the chosen windows for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not.
 Although the initial window width and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.
 The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.
-We choose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
+We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
 Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.
@@
 # Results
+
+<!-- ## Model performance and observed results -->
+
+<!-- ```{r} -->
+<!-- aver2 <- round(100 * mean(modprf$R2), 0) -->
+<!-- minr2 <- round(100 * min(modprf$R2), 0) -->
+<!-- maxr2 <- round(100 * max(modprf$R2), 0) -->
+<!-- minst <- modprf$station[which.min(modprf$R2)] -->
+<!-- minstmod <- modprf$model[which.min(modprf$R2)] -->
+<!-- minsttrn <- modprf$trans[which.min(modprf$R2)] -->
+<!-- minsttrn <- ifelse(is.numeric(minsttrn), 'Box-Cox model', 'log-model') -->
+<!-- maxst <- modprf$station[which.max(modprf$R2)] -->
+<!-- maxstmod <- modprf$model[which.max(modprf$R2)] -->
+<!-- maxsttrn <- modprf$trans[which.max(modprf$R2)] -->
+<!-- maxsttrn <- ifelse(is.numeric(maxsttrn), 'Box-Cox model', 'log-model') -->
+<!-- avecsr2 <- modprf %>% filter(model %in% 'Const seas') %>% pull(R2) %>% mean %>% '*'(100) %>% round(0) -->
+<!-- logr2 <- modprf %>% filter(grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0) -->
+<!-- bxcxr2 <- modprf %>% filter(!grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0) -->
+<!-- ``` -->
+
+<!-- A total of six models (three GAM structures, two link functions) were fit to each of nine stations, producing summary statistics for 18 different models (Table \@ref(tab:modprftab)).
+Chlorophyll trends were generally well-explained across all models and stations, with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (`r minstmod` `r minsttrn` at station `r minst`) to `r maxr2`% (`r maxstmod` `r maxsttrn` at station `r maxst`).
+The constant season model had the worst performance for all stations (average R$^2$ `r avecsr2`%), with either the interactive season or constant season* model having improved measures of model fit.
+For all but stations 18 and 22, the constant season* model provided significantly lower AIC and GCV scores and higher R-squared values compared to the interactive season models.
+For station 18, the interactive season and constant season* models provided only a marginal improvement over the constant season model.
+For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.
+Moeel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance.
+Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.
+-->
+
+<!-- GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll.
+Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.
+All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record.
+Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location.
+However, the magnitude of the fall peak at more southern stations shifted over time depending on the model.
+In particular, the interactive season model indicates a shift in the magnitude of the fall peak at stations 24, 27, and 32 with maximum concentrations observed around the mid-early 2000s.
+The annual trends for all models suggested similar trends over time, i.e., an increase followed by a recent decrease, whereas the constant season* models with an unconstrained smoother on the `decYear` predictor varied considerably at station 24, 27, and 32.
+Annual trends at these stations were difficult to generalize and had substantial overlap with the seasonal trends.
+Conversely, station 18 had minimal visual differences in annual and seasonal trends across all models, which is reflected by similarities in the model performance statistics in Table \@ref(tab:modprftab).
+-->
+
+<!-- ## Trend estimates -->
+
+<!-- Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).
+This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
+Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.
+However, the change from 2000 to 2010 was statistically significant with an increase of 78%.
+The period from 2010 to 2017 showed a significant decrease of 39.8%.
+An assessment of the annual differences indicated if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
+The results in figure \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences.
+Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.
+Similarly, rows g-i show the seasonal averages and trend estimates from August to December.
+The seasonal trend results suggested that the increase in annual averages from 2000 to 2010 was likely caused by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).
+Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.
+-->
+
+<!-- Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).
+From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.
+The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.
+From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).
+From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.
+The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).
+For the January to July period, significant changes were only observed at the southern stations, with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.
+No significant changes were observed for the August to December period.
+-->
+
+<!-- A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).
+Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).
+However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.
+Trends at station 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.
+Station 34 had significant increases in the JFM period from 1990-2010 and all other periods the OND (October, November, December) period for 2000 to 2010.
+Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.
+Station 36 had significant chlorophyll increases in all seasons, except OND for the 1990 to 2000 period.
+Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.
+-->
+
+<!-- ## Trend comparisons -->
+
+<!-- Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.
+Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.
+In addition to mixed-meta analysis, two alternative trend methods were used where the first shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a linear regression model on annual averages derived from the GAM seasonal estimates.
+The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.
+-->
+
+<!-- The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
+The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
+Slope estimates from the linear model applied to the averages from the GAMs as compared ot the mixed-meta results were identical at almost all stations, excluding station 32 in the January-July period from 1990-2000 when the former showed a negative slope and the latter showed a positive slope.
+The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.
+Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.
+-->
 
@@
 knitr::include_graphics('figs/trndcmp.png')
+```
+
+```{r, results = 'hide', message = F}
+
+toplo <- seastrnd3 %>% 
+  na.omit() %>% 
+  mutate(
+    sta = gsub('^data/modslog_chl|\\.RData', '', fl), 
+    subttl = ifelse(doystr < 150, '(a) Jan-Jun', '(b) Jul-Dec'),
+    pval = dplyr::case_when(
+      pval < 0.05 ~ 'p < 0.05', 
+      T ~ 'ns'
+    ), 
+    pval = factor(pval, levels = c('ns', 'p < 0.05'))
+  ) %>% 
+  dplyr::filter(yr <= 2016 & yr >= 1993)
+
+p <- ggplot(data = toplo, aes(x = yr, y = yrcoef, fill = pval)) + 
+  geom_point(shape = 21, size = 3) +
+  scale_x_continuous(limits = c(1993, 2016), breaks = seq(1993, 2016)) +
+  scale_y_continuous(limits = c(-2, 2)) +
+  scale_fill_manual(values = c(NA, 'tomato1'), drop = FALSE) +
+  geom_hline(yintercept = 0) + 
+  facet_grid(sta ~ subttl) + 
+  theme_minimal(base_family = 'serif') + 
+  theme(
+    axis.title.x = element_blank(), 
+    legend.position = 'right', 
+    legend.title = element_blank(),
+    strip.background = element_blank(), 
+    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
+    panel.grid.major.x = element_blank(),
+    panel.grid.minor.x = element_blank(), 
+    panel.grid.minor.y = element_blank(), 
+    panel.grid.major.y = element_blank(), 
+    strip.text.x = element_text(size = 12, hjust = 0), 
+    axis.ticks = element_line(), 
+    panel.spacing.y = grid::unit(1, "lines")
+  ) +
+  labs(
+    y = expression(paste('Chlorophyll change ( ', mu, 'g ', L^-1, yr^-1, ')'))
+  )
+
+png('figs/winchg.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
+print(p)
+dev.off()
+```
+```{r winchg, fig.cap='Estimates of chlorophyll change per year from applying the mixed-meta analysis across the time series for each station.
+Stations are arranged top to bottom from north to south.
+Plots in (a) show estimates for seasonal averages from January to June and plots in (b) show estimates for seasonal averages from July to December.
+Estimates in a year that are significant are shown in red.'}
+knitr::include_graphics('figs/winchg.png')
 ```

### 2020-11-08 R/dat_proc.R

@@ -229,6 +229,53 @@ seastrnd2 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
 
 save(seastrnd2, file = 'data/seastrnd2.RData', compress = 'xz')
 
+
+# moving window seasonal changes ------------------------------------------
+
+seastrnd3 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
+  crossing(
+    fl = ., 
+    tibble(
+      doystr = c(41, 214), 
+      doyend = c(213, 338)
+    )
+  ) %>% 
+  group_by(fl) %>% 
+  nest() %>% 
+  mutate(
+    mod = purrr::map(fl, function(x){
+      
+      load(file = x)
+      
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+      
+      out <- get(nm) %>% 
+        pull(model)
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest(c('data')) %>% 
+  group_by(doystr, doyend, mod, fl) %>% 
+  nest() %>% 
+  mutate(
+    res = purrr::pmap(list(fl, mods = mod, doystr, doyend), function(fl, mods, doystr, doyend){
+      
+      # get slope trends
+      out <- anlz_trndseason(mods = mod[[1]], doystr = doystr, doyend = doyend, justify = 'center', win = 5)
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest(res) %>%
+  ungroup() %>% 
+  select(-mod, -data)
+
+save(seastrnd3, file = 'data/seastrnd3.RData', compress = 'xz')
+
 # decadal percent changes -------------------------------------------------
 
 chgtrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 

### 2020-11-08 manu_draft.Rmd
working on updated results

@@
 
-In this section we describe four GAM formulations that can achieve the same or similar fits to data in slightly different ways.
+In this section we describe four GAM formulations that can achieve the same or similar fits to data in slightly different ways (Table \@ref(tab:modsumtab)).
 We present these to clarify relationships that may otherwise be confusing and to emphasize that results can be similarly achieved if a model unconstrained to have sufficient capability to extract a signal from raw data.
@@
 
-This model is effectively and mathematically equivalent to model S.
+This model is effectively and mathematically equivalent to model S (Table \@ref(tab:modstrtab)).
 The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
@@
 
-Models S, SY, and SYD can all potentially extract a similar signal from the raw data.
+Models S, SY, and SYD can all potentially extract a similar signal from the raw data (Table \@ref(tab:modstrtab)).
 What differs between the models is the allocation of penalties for curvature used to determine smoothness for each spline.
@@
 
-Model SYDI: `y ~ cont_year + s(cont_year) + s(doy, bs = "cc") + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knits_D_ti))`
+Model SYDI: `y ~ cont_year + s(cont_year) + s(doy, bs = "cc") + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knots_D_ti))`
 
@@
 The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.
-In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knits_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both to avoid over-parameterization.
+In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both to avoid over-parameterization.
 
@@
 In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.
-If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare.
+If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare (Table \@ref(tab:modstrtab)).
 
@@
 Finally, @Murphy19 present large AIC differences between their spline formulations.
-We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits.
+We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)).
 In our example, such large differences in AIC would only reflect inadequate choice of knots in one or more splines.
+
+We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).
+Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
+Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).
+However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an eexample), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to te basis functions.
+These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab).
+Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series but the model predictions are very similar.
 
@@
 
-<!-- ## Model performance and observed results -->
-
-<!-- ```{r} -->
-<!-- aver2 <- round(100 * mean(modprf$R2), 0) -->
-<!-- minr2 <- round(100 * min(modprf$R2), 0) -->
-<!-- maxr2 <- round(100 * max(modprf$R2), 0) -->
-<!-- minst <- modprf$station[which.min(modprf$R2)] -->
-<!-- minstmod <- modprf$model[which.min(modprf$R2)] -->
-<!-- minsttrn <- modprf$trans[which.min(modprf$R2)] -->
-<!-- minsttrn <- ifelse(is.numeric(minsttrn), 'Box-Cox model', 'log-model') -->
-<!-- maxst <- modprf$station[which.max(modprf$R2)] -->
-<!-- maxstmod <- modprf$model[which.max(modprf$R2)] -->
-<!-- maxsttrn <- modprf$trans[which.max(modprf$R2)] -->
-<!-- maxsttrn <- ifelse(is.numeric(maxsttrn), 'Box-Cox model', 'log-model') -->
-<!-- avecsr2 <- modprf %>% filter(model %in% 'Const seas') %>% pull(R2) %>% mean %>% '*'(100) %>% round(0) -->
-<!-- logr2 <- modprf %>% filter(grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0) -->
-<!-- bxcxr2 <- modprf %>% filter(!grepl('^log10$', trans)) %>% pull(R2) %>% mean %>% '*'(100) %>% round(0) -->
-<!-- ``` -->
-
-<!-- A total of six models (three GAM structures, two link functions) were fit to each of nine stations, producing summary statistics for 18 different models (Table \@ref(tab:modprftab)).
-Chlorophyll trends were generally well-explained across all models and stations, with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (`r minstmod` `r minsttrn` at station `r minst`) to `r maxr2`% (`r maxstmod` `r maxsttrn` at station `r maxst`).
-The constant season model had the worst performance for all stations (average R$^2$ `r avecsr2`%), with either the interactive season or constant season* model having improved measures of model fit.
-For all but stations 18 and 22, the constant season* model provided significantly lower AIC and GCV scores and higher R-squared values compared to the interactive season models.
-For station 18, the interactive season and constant season* models provided only a marginal improvement over the constant season model.
-For station 22, the interactive season and constant season* models provided similar but much improved fit to the data as compared to the constant season model.
-Moeel performance comparing either Box-Cox or logarithmic link functions were similar (average R$^2$ values of `r bxcxr2`% and `r logr2`% respectively), although the logarithmic models had slightly improved model performance.
-Station 18 was the only location where the three GAM models were not significantly different from each other for the Box-Cox transformation.
--->
-
-<!-- GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll.
-Four representative stations are shown in Figure \@ref(fig:trnddat) that capture spatial variation in chlorophyll and different temporal trends described by the models.
+## Model performance and observed results
+
+```{r, eval = T}
+aver2 <- round(100 * mean(modprf$R2), 0)
+minr2 <- round(100 * min(modprf$R2), 0)
+maxr2 <- round(100 * max(modprf$R2), 0)
+minst <- modprf$station[which.min(modprf$R2)]
+maxst <- modprf$station[which.max(modprf$R2)]
+```
+
+For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South San Francisco Bay.
+This model was chosen among the others because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
+Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
+GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)).
 All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record.
-Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location.
-However, the magnitude of the fall peak at more southern stations shifted over time depending on the model.
-In particular, the interactive season model indicates a shift in the magnitude of the fall peak at stations 24, 27, and 32 with maximum concentrations observed around the mid-early 2000s.
-The annual trends for all models suggested similar trends over time, i.e., an increase followed by a recent decrease, whereas the constant season* models with an unconstrained smoother on the `decYear` predictor varied considerably at station 24, 27, and 32.
-Annual trends at these stations were difficult to generalize and had substantial overlap with the seasonal trends.
-Conversely, station 18 had minimal visual differences in annual and seasonal trends across all models, which is reflected by similarities in the model performance statistics in Table \@ref(tab:modprftab).
--->
-
-<!-- ## Trend estimates -->
-
-<!-- Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to July and August to December (Figure \@ref(fig:trnddat)).
+Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).
+
+## Trend estimates
+
+Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December (Figure \@ref(fig:trnddat)).
 This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
@@
 Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.
--->
-
-<!-- Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).
+
+Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).
 From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.
@@
 No significant changes were observed for the August to December period.
--->
-
-<!-- A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).
+
+A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).
 Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).
@@
 Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.
--->
-
-<!-- ## Trend comparisons -->
-
-<!-- Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.
+
+## Trend comparisons
+
+Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.
 Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.
@@
 The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.
--->
-
-<!-- The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
+
+The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
 The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
@@
 Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.
--->
 
@@
 
+chg1 <- anlz_perchg(mods = toplo, baseyr = 1991, testy = 2000)
+chg2 <- anlz_perchg(mods = toplo, baseyr = 2001, testy = 2010)
+chg3 <- anlz_perchg(mods = toplo, baseyr = 2011, testy = 2019)
+chg <- bind_rows(chg1, chg2, chg3) %>% 
+  mutate_if(is.numeric, round, 1) %>% 
+  mutate(
+    pval = p_ast(pval)
+  )
+
+seas <- tibble(
+  doystr = c(41, 41, 41, 213, 213, 213), 
+  doyend = c(213, 213, 213, 338, 338, 338), 
+  yrstr = c(1991, 2000, 2010, 1991, 2000, 2010), 
+  yrend = c(2000, 2010, 2019, 2000, 2010, 2019)
+  ) %>% 
+  mutate(
+    est = purrr::pmap(list(doystr, doyend, yrstr, yrend), function(doystr, doyend, yrstr, yrend){
+
+      out <- anlz_avgseason(mods = toplo, doystr = doystr, doyend = doyend) %>% 
+        anlz_mixmeta(yrstr = yrstr, yrend = yrend) %>% 
+        .[[1]] %>% 
+        summary %>% 
+        coef(.) %>% 
+        .[2, ] 
+      out <- data.frame(est = out[['Estimate']], pval = out[['Pr(>|z|)']])
+
+      return(out)
+
+    })
+  ) %>% 
+  unnest(est) %>% 
+  mutate(
+    lets = c('(d)', '(e)', '(f)', '(g)', '(h)', '(i)'), 
+    doystr = ifelse(doystr == 41, 'Jan', 'Jul'), 
+    doyend = ifelse(doyend == 213, 'Jun', 'Dec'), 
+    est = paste('slope', round(est, 2)),
+    pval = p_ast(pval)
+  ) %>% 
+  unite('seas', doystr, doyend, sep = '-') %>% 
+  unite('yrs', yrstr, yrend, sep = '-') %>% 
+  unite('lets', lets, yrs, sep = ' ') %>% 
+  unite('lets', lets, seas, est, pval, sep = ', ') %>% 
+  deframe
+
 ylim <- c(-1, 19)
@@
   labs(
-    subtitle = '(a) 1990, 2000, Base: 3.6, Test: 5.7,\nChange: 57.7%, ns', 
+    subtitle = paste0('(a) 1991: ', chg[[1, 2]],' ug/L, 2000: ', chg[[1, 3]], 'ug/L\nChange: ', chg[[1,4]], '%, ', chg[[1,5]]), 
     title = NULL
@@
   labs(
-    subtitle = '(b) 2000, 2010, Base: 5.7, Test: 10.1,\nChange: 77.5%, p < 0.001', 
+    subtitle = paste0('(a) 1991: ', chg[[2, 2]],' ug/L, 2000: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
     title = NULL
@@
   labs(
-    subtitle = '(c) 2010, 2019, Base: 10.1, Test: 6.1,\nChange: -39.8%, p < 0.01', 
+    subtitle = paste0('(a) 1991: ', chg[[3, 2]],' ug/L, 2000: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
     title = NULL
@@
   labs(
-    subtitle = '(d) 1990-2000, Jan-Jul, slope 0.6, p < 0.01', 
+    subtitle = seas[[1]],
     title = NULL
@@
   labs(
-    subtitle = '(e) 2000-2010, Jan-Jul, slope 0.5, ns', 
+    subtitle = seas[[2]], 
     title = NULL
@@
   labs(
-    subtitle = '(f) 2010-2019, Jan-Jul, slope -0.4, ns', 
+    subtitle = seas[[3]], 
     title = NULL
@@
   labs(
-    subtitle = '(g) 1990-2000, Aug-Dec, slope 0.26, ns', 
+    subtitle = seas[[4]], 
     title = NULL
@@
   labs(
-    subtitle = '(h) 2000-2010, Aug-Dec, slope 0.56, p < 0.05', 
+    subtitle = seas[[5]], 
     title = NULL
@@
   labs(
-    subtitle = '(i) 2010-2019, Aug-Dec, slope -0.52, ns', 
+    subtitle = seas[[6]],
     title = NULL
@@
 ```
-```{r trnddat, fig.cap = "Examples of trend tests using results of GAM predictions for station 34 (`Const seas*`).
-Plots (a), (b), and (c) show estimates of percent change as the difference of means between base and test years for chlorophyll.
+```{r trnddat, fig.cap = "Examples of trend tests using results of GAM predictions for station 34.
+Plots (a), (b), and (c) show estimates of percent change as the difference of means between year pairs for chlorophyll.
 Plots (d) through (i) show seasonal averages and trend estimates over time.
-Plot (d), (e), and (f) show trend estimates for January through July and (g), (h), and (i) show trend estimates for August through December.
+Plot (d), (e), and (f) show trend estimates for January through June and (g), (h), and (i) show trend estimates for July through December.
 The trend lines in (d) through (i) estimate the rate of change of chlorophyll per year, reported as the slope in the plot title.
-ns: not significant"}
+ns: not significant, * p < 0.05, ** p < 0.005"}
 knitr::include_graphics('figs/trnddat.png')

### 2020-11-09 manu_draft.Rmd
fix trnddat fig with coord_cartesian

@@
 
-ylim <- c(-1, 19)
+ylim <- c(0, 16)
 ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))
@@
 p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
-  scale_y_continuous(limits = ylim) + 
+  coord_cartesian(ylim = ylim) + 
   labs(
@@
 p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
-  scale_y_continuous(limits = ylim) + 
+  coord_cartesian(ylim = ylim) + 
   labs(
@@
 p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
-  scale_y_continuous(limits = ylim) + 
+  coord_cartesian(ylim = ylim) +  
   labs(
@@
 p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
-  scale_y_continuous(limits = ylim) + 
+  coord_cartesian(ylim = ylim) + 
   labs(
@@
 p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
-  scale_y_continuous(limits = ylim) + 
+  coord_cartesian(ylim = ylim) + 
   labs(
@@
 p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
-  scale_y_continuous(limits = ylim) + 
+  coord_cartesian(ylim = ylim) + 
   labs(

### 2020-11-09 manu_draft.Rmd
using argument names in list to pmap function

@@
 For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South San Francisco Bay.
-This model was chosen among the others because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
+This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
 Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
@@
 
-Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December (Figure \@ref(fig:trnddat)).
+Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December for the same year pairs (Figure \@ref(fig:trnddat)).
 This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
-Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.
-However, the change from 2000 to 2010 was statistically significant with an increase of 78%.
-The period from 2010 to 2017 showed a significant decrease of 39.8%.
-An assessment of the annual differences indicated if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
-The results in figure \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences.
-Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.
-Similarly, rows g-i show the seasonal averages and trend estimates from August to December.
-The seasonal trend results suggested that the increase in annual averages from 2000 to 2010 was likely caused by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).
-Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.
+Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268,4% increase, although the change is insignificant.
+Similarly, the change from 2000 to 2010 was an insignificant increase of 11%, followed by an insignificant decrease of 13.9% from 2010 to 2019.
+Although these trends were insignificant, the results are based on a priori decisions of which years to compare.
+Significant results may be observed for other year pairs.
+
+An assessment of the annual differences indicates if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
+Accordingly, figures \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to evaluate changes across the same year pairs but GAM results for different seasonal periods.
+Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to June.
+Similarly, rows g-i show the seasonal averages and trend estimates from July to December.
+The seasonal trend results suggested that a significant increase in chlorophyll was observed from 1991 to 2000 but only for the January to June period (slope 0.57 ug/L chlorophyll change per year).
+Similarly, A significant increase was observed from 2000 to 2010 but only for the July to December period (slope 0.47)  Finally, a significant decrease was observed from 2010 to 2019 but only for the July to December period (slope -0.29).
+As a result, an assessment of trend changes by season provides a more precise assessment of trends over time and accounts for full uncertainty in estimates across model results.
 
@@
 chg1 <- anlz_perchg(mods = toplo, baseyr = 1991, testy = 2000)
-chg2 <- anlz_perchg(mods = toplo, baseyr = 2001, testy = 2010)
-chg3 <- anlz_perchg(mods = toplo, baseyr = 2011, testy = 2019)
+chg2 <- anlz_perchg(mods = toplo, baseyr = 2000, testy = 2010)
+chg3 <- anlz_perchg(mods = toplo, baseyr = 2010, testy = 2019)
 chg <- bind_rows(chg1, chg2, chg3) %>% 
@@
   labs(
-    subtitle = paste0('(a) 1991: ', chg[[2, 2]],' ug/L, 2000: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
+    subtitle = paste0('(a) 2001: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
     title = NULL
@@
   labs(
-    subtitle = paste0('(a) 1991: ', chg[[3, 2]],' ug/L, 2000: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
+    subtitle = paste0('(a) 2011: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
     title = NULL

### 2020-11-09 R/dat_proc.R

@@ -99,8 +99,8 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
       doyend = c(90, 181, 273, 364)
     ), 
     tibble(
-      yrstr = c(1990, 2000, 2010),
-      yrend = c(1999, 2009, 2019)
+      yrstr = c(1991, 2000, 2010),
+      yrend = c(2000, 2010, 2019)
     )
   ) %>% 
   group_by(fl) %>% 
@@ -124,7 +124,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, fl, mod) %>%
   nest() %>%
   mutate(
-    avgseas = purrr::pmap(list(mods = mod, doystr, doyend), anlz_avgseason)
+    avgseas = purrr::pmap(list(mods = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
   ) %>% 
   unnest('data') %>% 
   mutate(
@@ -134,7 +134,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     yrend = purrr::pmap(list(yrend, avgseas), function(yrend, avgseas){
       min(yrend, max(avgseas$yr))
     }),
-    metatrnd = purrr::pmap(list(avgseas, yrstr, yrend), anlz_mixmeta)
+    metatrnd = purrr::pmap(list(avgseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta)
   ) %>% 
   mutate(
     yrcoef = purrr::map(metatrnd, function(x) x[[1]]$coefficients['yr']), 
@@ -159,18 +159,18 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
 
 save(seastrnd, file = 'data/seastrnd.RData', compress = 'xz')
 
-# seasonal trends by decade, jan-jul and aug-dec --------------------------
+# seasonal trends by decade, jan-jun and jul-dec --------------------------
 
 seastrnd2 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
   crossing(
     fl = ., 
     tibble(
-      doystr = c(41, 214), 
+      doystr = c(41, 213), 
       doyend = c(213, 338)
     ), 
     tibble(
[…]

### 2020-11-09 manu_draft.Rmd
results text updated

@@
 This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
-Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268,4% increase, although the change is insignificant.
+Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268.4% increase, although the change is insignificant because of large uncertainties in the chlorophyll averages in the first year.
 Similarly, the change from 2000 to 2010 was an insignificant increase of 11%, followed by an insignificant decrease of 13.9% from 2010 to 2019.
@@
 
-Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although many trends were not significant and varied by location (Figure \@ref(fig:trndmap)a).
-From 1990 to 2000, only station 18 and 22 in the northern part of South Bay had significant increasing trends.
+Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although not all trends were significant and varied by location (Figure \@ref(fig:trndmap)a).
+From 1990 to 2000, only station 18, 32, 34, and 36 had increasing trends that were not significant.
 The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.
-From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and two southern stations (32, 34).
-From 2010 to 2016, all four southern stations had significant decreases in annual chlorophyll.
-The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to July and August to December (Figure \@ref(fig:trndmap)b).
-For the January to July period, significant changes were only observed at the southern stations, with increases at station 34 and 36 from 1990-2000, an increase at station 34 from 2000-2010, and a decrease at station 32 form 2010 to 2016.
-No significant changes were observed for the August to December period.
-
-A final analysis of the seasonal trends evaluated more discrete monthly periods in three month blocks to provide finer resolution to explain trends from the GAM results (Table \@ref(tab:seastrndtab)).
-Many stations and time periods did not show any seasonal chlorophyll trends or only had one significant trend (e.g., stations 21 and 24, JFM period from 2000-2010).
-However, several trends were observed in the southern stations, with two significant trends at station 32, six at station 34, and three at station 36.
-Trends at station 32 indicated a positive increase in the JFM (January, February, March) period from 2000-2010, followed by a significant decrease for the same seasonal period from 2010-2016.
-Station 34 had significant increases in the JFM period from 1990-2010 and all other periods the OND (October, November, December) period for 2000 to 2010.
-Significant decreases were observed at station 34 in the JFM and AMJ (April, March, July) period from 2010 to 2016.
-Station 36 had significant chlorophyll increases in all seasons, except OND for the 1990 to 2000 period.
+From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and the three southern-most stations (32, 34, 36).
+From 2010 to 2019, the three southern-most stations had significant decreases in annual chlorophyll, whereas station 24 had a significant increase during this period.
+The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to June and July to December (Figure \@ref(fig:trndmap)b).
+For the January to June period, significant increases were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas a significant decrease was observed at station 32 from 2010 to 2019.
+For the July to December period, significant increases were observed at stations 27 and 30 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas significant decreases were observed at stations 32 and 34 from 2010 to 2019.
+
+More discrete monthly periods in three month blocks were evaluated to explain a finer resolution of trends from the GAM results (Table \@ref(tab:seastrndtab)).
+Many stations and time periods did not show any seasonal chlorophyll trends or only had one or two significant trends at this finer temporal scale.
+The JAS (July, August, September) and OND (October, November, December) periods had significant trends that varied by decadal periods in the more northern stations (e.g., 18-27), whereas the southern stations tended to have more significant trends in the AMJ (April, May, June) and JAS periods.
 Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.
 
+A final analysis evaluated a moving window comparison of seasonal trends to provide additional context of when significant changes are occurring at each station (Figure \@ref(fig:winchg)).
+Results were obtained using a centered five year window to estimate the slope and significance of a seasonal trend of GAM results from January to June and July to December for each five-year period at each station (note that estimates at the far left and far right of each time series cannot be calculated).
+As before, more significant trends were generally observed with the more southern stations, particularly in the July to December period.
+Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record.
+Most stations showed a shift around 2007 when trends switched from significantly increasing to significantly decreasing, particularly in the January to June period.
+
 ## Trend comparisons
 
-Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate different conclusions that can be derived depending on method.
+Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
+As stated above, mixed-meta regression allow for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
 Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.
-In addition to mixed-meta analysis, two alternative trend methods were used where the first shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a linear regression model on annual averages derived from the GAM seasonal estimates.
+In addition to mixed-meta analysis, two alternative and more general trend methods were used.
+the first method shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a simple linear regression model on annual averages derived from the GAM seasonal estimates.
 The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.
@@
 The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
-The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
-Slope estimates from the linear model applied to the averages from the GAMs as compared ot the mixed-meta results were identical at almost all stations, excluding station 32 in the January-July period from 1990-2000 when the former showed a negative slope and the latter showed a positive slope.
+The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the GAM average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
+Slope estimates from the linear model applied to the averages from the GAMs as compared to the mixed-meta results were more identical, excluding some of the slope estimates for the southern stations.
 The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.
@@
 # Discussion
+
+Explain increasing resolution of methods, e.g., test/base, to seasonal averages, to mixed-meta, to moving window approach
 
@@
   labs(
-    subtitle = paste0('(a) 2001: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
+    subtitle = paste0('(a) 2000: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
     title = NULL
@@
   labs(
-    subtitle = paste0('(a) 2011: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
+    subtitle = paste0('(a) 2010: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
     title = NULL
@@
 knitr::include_graphics('figs/trndmap.png')
-```
-
-```{r, results = 'hide', message = F}
-toplo <- cmptrnd
-p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = mod, colour = pval)) + 
-  facet_grid(seas ~ yrs, scales = 'free_x') + 
-  geom_point(size = 3, position = position_dodge(width = 0.5), alpha = 0.6) + 
-  scale_shape('Trend model') + 
-  scale_colour_discrete_diverging(name = 'Trend', palette = 'BlueRed2') +
-  geom_vline(xintercept = 0, linetype = 'dashed') + 
-  theme_minimal() + 
-  theme(
-    strip.background = element_blank(),
-    panel.background = element_rect(fill = 'gray98', colour = NA), 
-    panel.border = element_blank(), 
-    legend.position = 'top'
-    ) +
-  labs(
-    y = 'Station (north to south)', 
-    x = expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
-  )
-
-png('figs/trndcmp.png', height = 5, width = 7.5, family = 'serif', units = 'in', res = 400)
-p
-dev.off()
-```
-```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station.
-The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM constant season* model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM constant season* model.
-Values for each model are the slope esimates as annual change per year within each season, with color denoting significant trends.'}
-knitr::include_graphics('figs/trndcmp.png')
 ```
@@
 
+```{r, results = 'hide', message = F}
+toplo <- cmptrnd
+p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = mod, colour = pval)) + 
+  facet_grid(seas ~ yrs, scales = 'free_x') + 
+  geom_point(size = 3, position = position_dodge(width = 0.5), alpha = 0.6) + 
+  scale_shape('Trend model') + 
+  scale_colour_discrete_diverging(name = 'Trend', palette = 'BlueRed2') +
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  theme_minimal() + 
+  theme(
+    strip.background = element_blank(),
+    panel.background = element_rect(fill = 'gray98', colour = NA), 
+    panel.border = element_blank(), 
+    legend.position = 'top'
+    ) +
+  labs(
+    y = 'Station (north to south)', 
+    x = expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+  )
+
+png('figs/trndcmp.png', height = 5, width = 7.5, family = 'serif', units = 'in', res = 400)
+p
+dev.off()
+```
+```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station.
+The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM constant season* model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM constant season* model.
+Values for each model are the slope esimates as annual change per year within each season, with color denoting significant trends.'}
+knitr::include_graphics('figs/trndcmp.png')
+```
+
 # Tables

### 2020-11-12 R/dat_proc.R
remove reml typo from modstr

@@ -470,11 +470,10 @@ SYD <- gam(value ~ cont_year + s(cont_year, k = 360) + s(doy, bs = 'cc', k = 230
            na.action = na.exclude,
            select = F)
 SYD$trans <- 'log10'
-SYDI <- gam(value ~ cont_year + s(cont_year, k = 300) + s(doy, bs = 'cc', k = 100) + ti(cont_year, doy, bs = c('tp', 'cc'), k = 15),
+SYDI <- gam(value ~ cont_year + s(cont_year, k = 300) + s(doy, bs = 'cc', k = 105) + ti(cont_year, doy, bs = c('tp', 'cc'), k = 15),
             knots = list(doy = c(1, 366)),
             data = moddat,
             na.action = na.exclude,
-            method = 'REML',
             select = F)
 SYDI$trans <- 'log10'
 
@@ -486,6 +485,6 @@ modstr <- list(
 )
 
 
-# anlz_fit(mods = modstr)
+anlz_fit(mods = modstr)
 
 save(modstr, file = 'data/modstr.RData', compress = 'xz')

### 2020-11-13 R/dat_proc.R
dat_proc and data objects updated from wqtrends single model

@@ -17,16 +17,9 @@ modssta <- rawdat %>%
   nest() %>% 
   mutate(
     model = purrr::pmap(list(station, data), function(station, data){
-      
+
       cat(station, '\n')
-      moddat <- anlz_trans(data, trans = 'log10')
-      out <- gam(value ~ s(cont_year, k = 360),
-        knots = list(doy = c(1, 366)),
-        data = moddat,
-        na.action = na.exclude,
-        select = F)
-      
-      out$trans <- 'log10'
+      out <- anlz_gam(data, trans = 'log10')
       
       return(out)
       
@@ -68,12 +61,12 @@ modprf <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
 
       nm <- basename(x)
       nm <- gsub('\\.RData', '', nm)
-      
-      dat <- get(nm) %>% 
-        select(model) %>% 
+
+      mod <- get(nm) %>% 
+        pull(model) %>% 
         deframe()
       
-      out <- anlz_fit(mods = dat)
+      out <- anlz_fit(mod)
       
       return(out)
       
@@ -82,9 +75,12 @@ modprf <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   ungroup %>% 
   select(-data) %>% 
   unnest('prf') %>% 
-  select(-AIC, -k, -F, -p.value, -value) %>% 
+  select(-AIC) %>% 
+  mutate(
+    value = gsub('^data/modslog\\_chl|\\.RData$', '', value)
+  ) %>% 
   rename(
-    station = model
+    station = value
   ) 
 
 save(modprf, file = 'data/modprf.RData', compress = 'xz')
@@ -114,7 +110,8 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
       nm <- gsub('\\.RData', '', nm)
 
       out <- get(nm) %>%
-        pull(model)
+        pull(model) %>% 
+        deframe
       
       return(out)
       
@@ -124,7 +121,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, fl, mod) %>%
   nest() %>%
   mutate(
-    avgseas = purrr::pmap(list(mods = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
[…]

### 2020-11-13 manu_draft.Rmd
code chunk updated to wqtrends one model in manu draft

@@
 alph <- 0.5
-
-p1 <- show_prddoy(mods = mods, ylab = 'Chlorophyll-a (ug/L)', nfac = 3, size = 0.8, alpha = 0.8) + 
+ylm <- c(0.8, 100)
+
+p1a <- show_prddoy(mods[[1]], ylab = 'SY: Chl-a (ug/L)', size = 0.8, alpha = 0.8) + 
   guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
   scale_colour_gradient(low = "white", high = "black") + 
-  thm + 
+  coord_cartesian(ylim = ylm) +
+  theme_minimal() + 
+  theme(
+    legend.position = 'none', 
+    axis.title.x = element_blank()
+  ) +
   labs(
@@
   )
-
-prds <- anlz_prd(mods = mods, trans = 'log10')
+p1b <- show_prddoy(modstr[[2]], ylab = 'SYD: Chl-a (ug/L)', size = 0.8, alpha = 0.8) + 
+  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
+  scale_colour_gradient(low = "white", high = "black") + 
+  coord_cartesian(ylim = ylm) +
+  theme_minimal() + 
+  theme(
+    legend.position = 'none'
+  )
+p1c <- show_prddoy(mods[[3]], ylab = 'SYDI: Chl-a (ug/L)', size = 0.8, alpha = 0.8) + 
+  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
+  scale_colour_gradient(low = "white", high = "black") + 
+  coord_cartesian(ylim = ylm) +
+  theme_minimal() + 
+  theme(
+    legend.position = 'right', 
+    axis.title.x = element_blank()
+  )
+
+p1 <- p1a + p1b + p1c + plot_layout(ncol = 3, widths = c(1, 1, 1))
+prds <- lapply(modstr, anlz_prd) %>% 
+  enframe('model', 'dat') %>%
+  unnest('dat')
 
@@
       mod <- get(nm) %>% 
-        pull(model)
-
-      prd <- anlz_prd(mods = mod) %>% 
+        pull(model) %>% 
+        deframe()
+
+      prd <- anlz_prd(mod) %>% 
         anlz_backtrans()
@@
 ```{r, results = 'hide', message = F}
-toplo <- modslog_chl34 %>%
-  pull(model)
-
-chg1 <- anlz_perchg(mods = toplo, baseyr = 1991, testy = 2000)
-chg2 <- anlz_perchg(mods = toplo, baseyr = 2000, testy = 2010)
-chg3 <- anlz_perchg(mods = toplo, baseyr = 2010, testy = 2019)
+mod <- modslog_chl34 %>%
+  pull(model) %>% 
+  deframe()
+
+chg1 <- anlz_perchg(mod, baseyr = 1991, testy = 2000)
+chg2 <- anlz_perchg(mod, baseyr = 2000, testy = 2010)
+chg3 <- anlz_perchg(mod, baseyr = 2010, testy = 2019)
 chg <- bind_rows(chg1, chg2, chg3) %>% 
@@
 
-      out <- anlz_avgseason(mods = toplo, doystr = doystr, doyend = doyend) %>% 
+      out <- anlz_avgseason(mod, doystr = doystr, doyend = doyend) %>% 
         anlz_mixmeta(yrstr = yrstr, yrend = yrend) %>% 
-        .[[1]] %>% 
         summary %>% 
@@
 
-p1 <- show_perchg(mods = toplo, baseyr = 1991, testyr = 2000, ylab = ylab) +
-  labs(
-    subtitle = paste0('(a) 1991: ', chg[[1, 2]],' ug/L, 2000: ', chg[[1, 3]], 'ug/L\nChange: ', chg[[1,4]], '%, ', chg[[1,5]]), 
+p1 <- show_perchg(mod, baseyr = 1991, testyr = 2000, ylab = ylab) +
+  labs(
+    subtitle = paste0('(a) 1991: ', chg[[1, 1]],' ug/L, 2000: ', chg[[1, 2]], 'ug/L\nChange: ', chg[[1,3]], '%, ', chg[[1,4]]), 
     title = NULL
   )
-p2 <- show_perchg(mods = toplo, baseyr = 2000, testyr = 2010, ylab = NULL) +
-  labs(
-    subtitle = paste0('(a) 2000: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
+p2 <- show_perchg(mod, baseyr = 2000, testyr = 2010, ylab = NULL) +
+  labs(
+    subtitle = paste0('(a) 2000: ', chg[[2, 1]],' ug/L, 2010: ', chg[[2, 2]], 'ug/L\nChange: ', chg[[2, 3]], '%, ', chg[[2, 4]]), 
     title = NULL
   )
-p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2019, ylab = NULL) +
-  labs(
-    subtitle = paste0('(a) 2010: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
+p3 <- show_perchg(mod, baseyr = 2010, testyr = 2019, ylab = NULL) +
+  labs(
+    subtitle = paste0('(a) 2010: ', chg[[3, 1]],' ug/L, 2019: ', chg[[3,2]], 'ug/L\nChange: ', chg[[3, 3]], '%, ', chg[[3, 4]]), 
     title = NULL
   )
-p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
+p4 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
+p5 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
+p6 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
   coord_cartesian(ylim = ylim) +  
@@
   )
-p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
+p7 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
+p8 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
+p9 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
   coord_cartesian(ylim = ylim) + 
@@
   mutate(
-    sta = gsub('^data/modslog_chl|\\.RData', '', fl), 
     subttl = ifelse(doystr < 150, '(a) Jan-Jun', '(b) Jul-Dec'),
@@
   geom_hline(yintercept = 0) + 
-  facet_grid(sta ~ subttl) + 
+  facet_grid(station ~ subttl) + 
   theme_minimal(base_family = 'serif') + 

### 2020-11-15 manu_draft.Rmd
data objects, figures, tables changed from back-transformation fixes on wqtrends

@@
   geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  scale_y_log10() + 
+  scale_x_log10() + 
   thm + 
   labs(
-    title = '(b) Estimated chlorophyll (log-space) between models'
+    title = '(b) Estimated chlorophyll between models'
   )
@@
   geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  scale_y_log10() + 
+  scale_x_log10() + 
   thm
@@
   geom_abline(slope = 1, intercept = 0, color = lncol) + 
+  scale_y_log10() + 
+  scale_x_log10() + 
   thm +
@@
 
-      prd <- anlz_prd(mod) %>% 
-        anlz_backtrans()
+      prd <- anlz_prd(mod)
 
@@
 
-ylim <- c(0, 16)
+ylim <- c(0, 21)
 ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))
@@
 
-pthm <- theme_bw(base_family = 'serif', base_size = 14) +
+pthm <- theme_bw(base_family = 'serif', base_size = 12) +
   theme(
@@
     axis.title = element_blank(), 
-    axis.text = element_text(size = 7)
+    axis.text = element_text(size = 8)
   )
@@
   scale_size(range = c(1, 6), guide = F) +
-  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 6)) +
+  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 8)) +
   labs(

### 2020-11-15 R/dat_proc.R

@@ -490,7 +490,6 @@ modstr <- list(
   SYDI = SYDI
 )
 
-
-anlz_fit(mods = modstr)
+# lapply(modstr, anlz_fit)
 
 save(modstr, file = 'data/modstr.RData', compress = 'xz')

### 2020-11-18 manu_draft.Rmd
trend model comparison example plot for select stations

@@
 Explain increasing resolution of methods, e.g., test/base, to seasonal averages, to mixed-meta, to moving window approach
+
+Explain how choosing window widths and time periods is a user choice that is arbitrary, identifying an objective approach to identifying trends can pave the way for future work (parallels with mgcv?)
 
@@
 ```{r, results = 'hide', message = F}
+show_cmpplo <- function(fl, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
+                        ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")"))){
+
+  load(file = fl)
+  nm <- basename(fl)
+  nm <- gsub('\\.RData', '', nm)
+  mod <- get(nm) %>%
+    pull(model) %>%
+    deframe()
+
+  # data to model
+  tomod1 <- mod$model %>%
+    mutate(
+      yr = floor(cont_year),
+      doy = yday(date_decimal(cont_year))
+    ) %>%
+    filter(doy >= doystr & doy <= doyend) %>%
+    group_by(yr) %>%
+    summarise(
+      avg = mean(value, na.rm = T),
+      sdavg = sd(value, na.rm = T),
+      .groups = 'drop'
+      ) %>%
+    mutate(bt_avg = 10^(avg + (sdavg^2) / 2))
+  avgseas <- anlz_avgseason(mod = mod, doystr = doystr, doyend = doyend)
+  tomod2 <- avgseas %>%
+    filter(yr <= yrend & yr >= yrstr) %>%
+    mutate(S = se^2)
+
+  # model
+  obmod <- tomod1 %>%
+    filter(yr >= yrstr & yr <= yrend) %>%
+    lm(avg ~ yr, .)
+  lmmod <- lm(avg ~ yr, tomod2)
+  mxmod <- mixmeta::mixmeta(avg ~ yr, S = S, random = ~1|yr, data = tomod2, method = 'reml')
+
+  # pvalues
+  pval1 <- summary(obmod) %>% coefficients %>% .[2, 4]
+  pval2 <- summary(lmmod) %>% coefficients %>% .[2, 4]
+  pval3 <- summary(mxmod) %>% coefficients %>% .[2, 4]
+  pval1 <- paste0('p = ', round(pval1, 2), ', ', p_ast(pval1))
+  pval2 <- paste0('p = ', round(pval2, 2), ', ', p_ast(pval2))
+  pval3 <- paste0('p = ', round(pval3, 2), ', ', p_ast(pval3))
+
+  dispersion <- summary(mod)$dispersion
+  s2a <- (summary(obmod)$sigma)^2
+  s2b <- (summary(lmmod)$sigma)^2
+
+  toplo1 <- data.frame(
+      yr = seq(yrstr, yrend, length = 50)
+    ) %>%
+    dplyr::mutate(
+      avg = predict(obmod, newdata = data.frame(yr = yr)),
+      se = predict(obmod, newdata = data.frame(yr = yr), se.fit = T)$se.fit,
+      bt_lwr = 10^((avg - 1.96 * se) + log(10) * s2a / 2),
+      bt_upr = 10^((avg + 1.96 * se) + log(10) * s2a / 2),
+      bt_avg = 10^(avg + log(10) * s2a / 2)
+    )
+
+  toplo2 <- data.frame(
+    yr = seq(yrstr, yrend, length = 50)
+    ) %>%
+    dplyr::mutate(
+      avg = predict(lmmod, newdata = data.frame(yr = yr)),
+      se = predict(lmmod, newdata = data.frame(yr = yr), se.fit = T)$se.fit,
+      bt_lwr = 10^((avg - 1.96 * se) + log(10) * dispersion / 2),
+      bt_upr = 10^((avg + 1.96 * se) + log(10) * dispersion / 2),
+      bt_avg = 10^(avg + log(10) * dispersion / 2)
+    )
+
+  toplo3 <- data.frame(
+    yr = seq(yrstr, yrend, length = 50)
+  ) %>%
+    dplyr::mutate(
+      avg = predict(mxmod, newdata = data.frame(yr = yr)),
+      se = predict(mxmod, newdata = data.frame(yr = yr), se = T)[, 2],
+      bt_lwr = 10^((avg - 1.96 * se) + log(10) * dispersion / 2),
+      bt_upr = 10^((avg + 1.96 * se) + log(10) * dispersion / 2),
+      bt_avg = 10^(avg + log(10) * dispersion / 2)
+    )
+
+  # y axis limits
+  ylim <- c(0, ymax)
+  if(is.null(ymax))
+    ylim <- c(0, max(avgseas$bt_upr))
+
+  # plots
+  p1 <- ggplot(data = tomod1, aes(x = yr, y = bt_avg)) +
+    geom_point(colour = 'deepskyblue3') +
+    geom_ribbon(data = toplo1, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'grey', alpha = 0.4) +
+    geom_line(data = toplo1, color = 'grey') +
+    theme_minimal(base_family = 'serif') +
+    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
+    theme(
+      axis.title.x = element_blank(),
+      axis.text.x = element_blank()
+    ) +
+    labs(
+      title = ttl,
+      subtitle = paste('Obesrved regression on raw data,', pval1),
+      y = NULL
+    )
+
+  p2 <- ggplot(data = avgseas, aes(x = yr, y = bt_avg)) +
+    geom_point(colour = 'deepskyblue3') +
+    geom_ribbon(data = toplo2, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'lightgreen', alpha = 0.4) +
+    geom_line(data = toplo2, color = 'green') +
+    theme_minimal(base_family = 'serif') +
+    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
+    theme(
+      axis.title.x = element_blank(),
+      axis.text.x = element_blank()
+    ) +
+    labs(
+      subtitle = paste('Average regression on GAM estimates,', pval2),
+      y = ylab
+    )
+
+  p3 <- ggplot(data = avgseas, aes(x = yr, y = bt_avg)) +
+    geom_point(colour = 'deepskyblue3') +
+    geom_errorbar(aes(ymin = bt_lwr, ymax = bt_upr), colour = 'deepskyblue3') +
+    geom_ribbon(data = toplo3, aes(ymin = bt_lwr, ymax = bt_upr), fill = 'pink', alpha = 0.4) +
+    geom_line(data = toplo3, color = 'pink') +
+    theme_minimal(base_family = 'serif') +
+    coord_cartesian(ylim = ylim, xlim = c(1990, 2020)) +
+    theme(
+      axis.title.x = element_blank()
+    ) +
+    labs(
+      subtitle = paste('Mixed-meta regression on GAM estimates', pval3),
+      y = NULL
+    )
+
+  p <- p1 + p2 + p3 + plot_layout(ncol = 1)
+
+  return(p)
+
+}
+
+p1 <- show_cmpplo('data/modslog_chl36.RData', doystr = 91, doyend = 181, yrstr = 1991, yrend = 2000, 
+                  ttl =  '(a) Station 36, Apr-Jun chl. averages', ymax = NULL)
+p2 <- show_cmpplo('data/modslog_chl22.RData', doystr = 274, doyend = 364, yrstr = 2000, yrend = 2010, 
+                  ttl =  '(b) Station 22, Oct-Dec chl. averages', ymax = NULL, ylab = NULL)
+# p3 <- show_cmpplo('data/modslog_chl34.RData', doystr = 274, doyend = 364, yrstr = 2010, yrend = 2019, 
+#                   ttl =  '(c) Station 34, Oct-Dec chl. averages', ymax = 15, ylab = NULL)
+
+p <- (p1 |  p2) + plot_layout(ncol = 2)
+png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
+p
+dev.off()
+```
+```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at two example stations.
+The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
+Regressions in each plot are fit through the seasonal estimates indicated in the plot titles.'}
+knitr::include_graphics('figs/trndcmpex.png')
+```
+
+```{r, results = 'hide', message = F}
 toplo <- cmptrnd
-p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = mod, colour = pval)) + 
+p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = modtyp, colour = pval)) + 
   facet_grid(seas ~ yrs, scales = 'free_x') + 
@@
   geom_vline(xintercept = 0, linetype = 'dashed') + 
+  # scale_x_continuous(limits = c(-1, 1)) +
   theme_minimal() + 
@@
 
-png('figs/trndcmp.png', height = 5, width = 7.5, family = 'serif', units = 'in', res = 400)
+png('figs/trndcmp.png', height = 8, width = 7.5, family = 'serif', units = 'in', res = 400)
 p
@@
 ```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station.
-The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM constant season* model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM constant season* model.
+The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
 Values for each model are the slope esimates as annual change per year within each season, with color denoting significant trends.'}
@@
 
-```{r seastrndtab, tab.cap = 'Seasonal trends from `Const seas*` results by station and decade.
-Results show the annual slope estimate as chlorophyll change per year for each seasonal period across years (JFM: January, February, March; AMJ: April, May, June; JAS: July, August, September, OND: October, November, December).
-Significant trends based on mixed-meta regression analysis are indicated with asterisks at $\\alpha = 0.05$, * $p< 0.05$, and ** $p < 0.005$.'}
-totab <- seastrnd %>% 
-  mutate(
-    pval = p_ast(pval),
-    pval = gsub('ns', '', pval), 
-    yrcoef = round(yrcoef, 2)
-    ) %>% 
-  unite('yrcoef', yrcoef, pval, sep = '') %>% 
-  spread(seas, yrcoef) %>% 
-  mutate(
-    station = ifelse(duplicated(station), '', station)
-  )
-
-boldjfm <- grep('\\*', totab$JFM)
-boldamj <- grep('\\*', totab$AMJ)
-boldjas <- grep('\\*', totab$JAS)
-boldond <- grep('\\*', totab$OND)
-
-tab <- flextable(totab) %>% 
-  set_header_labels(
-    station = 'Station', 
-    yrs = 'Years'
-  ) %>% 
-  add_header_row(values = c('', '', 'Chlorophyll change (ug/l) per year', '', '', '')) %>% 
-  merge_at(i = 1, j = 3:6, part = 'header') %>% 
-  bold(i = boldjfm, j = 'JFM') %>% 
-  bold(i = boldamj, j = 'AMJ') %>% 
-  bold(i = boldjas, j = 'JAS') %>% 
-  bold(i = boldond, j = 'OND') %>% 
-  border_remove() %>% 
-  hline_top(border = fp_border(color = 'black')) %>% 
-  font(fontname = 'Times', part = 'all') %>%
-  fontsize(size = 12, part = 'all') %>%   
-  width(j = c(2), width = 1) 
-tab
-```
-
 # Supplement

### 2020-11-18 R/dat_proc.R

@@ -337,10 +337,10 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   crossing(
     fl = ., 
     tibble(
-      # doystr = c(1, 91, 182, 274), 
-      # doyend = c(90, 181, 273, 364)
-      doystr = c(41, 213), 
-      doyend = c(213, 338)
+      doystr = c(1, 91, 182, 274),
+      doyend = c(90, 181, 273, 364)
+      # doystr = c(41, 213), 
+      # doyend = c(213, 338)
     ), 
     tibble(
       yrstr = c(1991, 2000, 2010),
@@ -381,20 +381,20 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     }),
     metatrnd = purrr::pmap(list(avgseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta),
     lmtrnd = purrr::pmap(list(avgseas, yrstr, yrend), function(avgseas, yrstr, yrend){
-
+      
       out <- avgseas %>% 
         filter(yr >= yrstr & yr <= yrend) %>% 
         lm(avg ~ yr, .)
       
       return(out)
+      
     }), 
     obstrnd = purrr::pmap(list(doystr, doyend, mod, yrstr, yrend), function(doystr, doyend, mod, yrstr, yrend){
-
+      
       moddat <- mod$model %>% 
         mutate(
           yr = floor(cont_year), 
           doy = yday(date_decimal(cont_year)),
-          value = 10 ^ value
         ) %>% 
         filter(doy >= doystr & doy <= doyend) %>% 
         filter(yr >= yrstr & yr <= yrend) %>% 
@@ -407,12 +407,30 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     })
   ) %>% 
   ungroup %>% 
-  select(-mod, -avgseas) %>% 
-  gather('mod', 'est', metatrnd, lmtrnd, obstrnd) %>% 
+  select(-avgseas) %>% 
+  gather('modtyp', 'est', metatrnd, lmtrnd, obstrnd) %>% 
   mutate(
-    yrcoef = purrr::pmap(list(mod, est), function(mod, est){
+    yrcoef = purrr::pmap(list(mod, modtyp, est), function(mod, modtyp, est){
       
[…]

### 2020-11-20 manu_draft.Rmd
results match figures/tables, one model only

@@
 data(seastrnd2)
-data(seastrnd3)
 data(chgtrnd)
@@
 
-The R package *mixmeta* was used to estimate the model in equation \@ref{eqn:mixmet} [@Sera19].
+The R package *mixmeta* was used to estimate the model in equation \@ref(eqn:mixmet) [@Sera19].
 Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
@@
 GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)).
-All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record.
-Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).
+All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, followed by a decreasing trend until the end of the record.
+Seasonally, all models indicated a larger spring peak in chlorophyll, particularly for more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).
 
@@
 Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268.4% increase, although the change is insignificant because of large uncertainties in the chlorophyll averages in the first year.
-Similarly, the change from 2000 to 2010 was an insignificant increase of 11%, followed by an insignificant decrease of 13.9% from 2010 to 2019.
+However, the change from 2000 to 2010 was an significant increase of 64%, followed by a significant decrease of 35% from 2010 to 2019.
 Although these trends were insignificant, the results are based on a priori decisions of which years to compare.
-Significant results may be observed for other year pairs.
 
 An assessment of the annual differences indicates if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
-Accordingly, figures \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to evaluate changes across the same year pairs but GAM results for different seasonal periods.
+Accordingly, figures \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to evaluate changes across the same year pairs but using seasonal averages from GAM results.
 Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to June.
 Similarly, rows g-i show the seasonal averages and trend estimates from July to December.
-The seasonal trend results suggested that a significant increase in chlorophyll was observed from 1991 to 2000 but only for the January to June period (slope 0.57 ug/L chlorophyll change per year).
-Similarly, A significant increase was observed from 2000 to 2010 but only for the July to December period (slope 0.47)  Finally, a significant decrease was observed from 2010 to 2019 but only for the July to December period (slope -0.29).
+The seasonal trend results suggested that a significant increase in chlorophyll was observed from 1991 to 2000 but only for the January to June period (slope 0.58 ug/L chlorophyll change per year).
+Similarly, A significant increase was observed from 2000 to 2010 but only for the July to December period (slope 0.58)  Finally, a significant decrease was observed from 2010 to 2019 but only for the July to December period (slope -0.28).
 As a result, an assessment of trend changes by season provides a more precise assessment of trends over time and accounts for full uncertainty in estimates across model results.
@@
 Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although not all trends were significant and varied by location (Figure \@ref(fig:trndmap)a).
-From 1990 to 2000, only station 18, 32, 34, and 36 had increasing trends that were not significant.
+Comparing years 1990 and 2000, stations 21, 24, 27, and 30 had increasing significant trends.
 The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.
-From 2000 to 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and the three southern-most stations (32, 34, 36).
-From 2010 to 2019, the three southern-most stations had significant decreases in annual chlorophyll, whereas station 24 had a significant increase during this period.
+Comparing 2000 and 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and the three southern-most stations (32, 34, 36).
+Comparing 2010 and 2019, the three southern-most stations had significant decreases in annual chlorophyll, whereas station 24 had a significant increase during this period.
 The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to June and July to December (Figure \@ref(fig:trndmap)b).
-For the January to June period, significant increases were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas a significant decrease was observed at station 32 from 2010 to 2019.
-For the July to December period, significant increases were observed at stations 27 and 30 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas significant decreases were observed at stations 32 and 34 from 2010 to 2019.
-
-More discrete monthly periods in three month blocks were evaluated to explain a finer resolution of trends from the GAM results (Table \@ref(tab:seastrndtab)).
-Many stations and time periods did not show any seasonal chlorophyll trends or only had one or two significant trends at this finer temporal scale.
-The JAS (July, August, September) and OND (October, November, December) periods had significant trends that varied by decadal periods in the more northern stations (e.g., 18-27), whereas the southern stations tended to have more significant trends in the AMJ (April, May, June) and JAS periods.
-Accordingly, table \@ref(tab:seastrndtab) demonstrates the value of obtaining greater insight into seasonal changes by evaluating different blocks as compared to the larger seasonal blocks shown in Figure \@ref(fig:trndmap)b.
+For the January to June period, significant increases were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas a significant decrease was observed at stations 30 and 32 from 2010 to 2019.
+For the July to December period, significant increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas significant decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
 
@@
 Results were obtained using a centered five year window to estimate the slope and significance of a seasonal trend of GAM results from January to June and July to December for each five-year period at each station (note that estimates at the far left and far right of each time series cannot be calculated).
-As before, more significant trends were generally observed with the more southern stations, particularly in the July to December period.
-Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record.
+As before, larger trends were generally observed with the more southern stations.
+Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until recent years when trends were relatively stable.
 Most stations showed a shift around 2007 when trends switched from significantly increasing to significantly decreasing, particularly in the January to June period.
@@
 Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
-As stated above, mixed-meta regression allow for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
-Figure \@ref(fig:trndcmp) shows estimates of chlorophyll change per year for the same seasonal periods and annual groupings as in figure \@ref(fig:trndmap)b.
-In addition to mixed-meta analysis, two alternative and more general trend methods were used.
-the first method shows slope estimates from a linear regression on annual averages of observed chlorophyll for the specified seasonal period and the second shows slope estimates from a simple linear regression model on annual averages derived from the GAM seasonal estimates.
-The latter uses the same averages from the mixed-meta analysis, but is a simpler approach that does not account for the uncertainty in the average estimates.
-
-The different trend analysis methods provided conflicting information on the magnitude and direction of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
-The slope estimates from the linear model applied to the observed data were understandably more variable than the slope estimate from the GAM average and mixed-meta methods, with much larger slopes observed especially at the more southern stations.
-Slope estimates from the linear model applied to the averages from the GAMs as compared to the mixed-meta results were more identical, excluding some of the slope estimates for the southern stations.
-The mixed-meta analyses had less significant trends compared to the average results for the same GAM estimates, which reflects the ability of the former to account for uncertainty in the average estimates when considering trends.
-Importantly, completely opposite and/or varying significance of trends were observed between the methods, suggesting choice of analysis method can influence the certainty of conclusions.
+As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
+Moreoever, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
+Trends assessed on observed data may include substantial noise indendent of any "canonical" signal derived from GAMs.
+
+These distinctions are demonstrated in Figure \@ref(fig:trndcmpex) for two stations and two time periods, where the significance of trends are compared based on regression through seasonal averages (Figure \@ref(fig:trndcmpex) row 1), regression through GAM estimates (row 2), and mixed-meta regression through GAM estimates (row 3).
+The stations and time periods were selected to demonstrate examples where different conclusions can be obtained depending on the methods and characteristics of the data.
+For example, Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
+Only the mixed-meta regression results are significant in this example.
+The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a significant trend.
+Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages form 2000 to 2010.
+Unlike the first example, the top two figures show significant trend estimates, whereas the bottom plot for the mixed-meta regression analysis does not show a significant trend because of added uncertainty in the averages provided by the GAMs.
+In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.
+
+Figure \@ref(fig:trndcmp) expands the results in Figure \@ref(fig:trndcmpex) by applying the method comparisons to all stations.
+As before, the different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade.
+The slope estimates from the linear models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.
+Slope estimates from the linear models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.
+Differences in significance of trends between the linear models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
 
@@
 Explain how choosing window widths and time periods is a user choice that is arbitrary, identifying an objective approach to identifying trends can pave the way for future work (parallels with mgcv?)
+
+Benefits of mixed-meta to provide accurate results, reduces possibility of type I/II errors (e.g., a/b examples in trndcmpex)
 
@@
 p1c <- show_prddoy(mods[[3]], ylab = 'SYDI: Chl-a (ug/L)', size = 0.8, alpha = 0.8) + 
-  guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
-  scale_colour_gradient(low = "white", high = "black") + 
+  guides(colour = ggplot2::guide_colourbar(barheight = 8, barwidth = 1)) +
+  scale_colour_gradient('Year', low = "white", high = "black") + 
   coord_cartesian(ylim = ylm) +
@@
   geom_abline(slope = 1, intercept = 0, color = lncol) + 
-  scale_y_log10() + 
-  scale_x_log10() + 
+  scale_y_log10(limits = ylm) + 
+  scale_x_log10(limits = ylm) + 
   thm + 
@@
   geom_abline(slope = 1, intercept = 0, color = lncol) + 
-  scale_y_log10() + 
-  scale_x_log10() + 
+  scale_y_log10(limits = ylm) + 
+  scale_x_log10(limits = ylm) + 
   thm
@@
   geom_abline(slope = 1, intercept = 0, color = lncol) + 
-  scale_y_log10() + 
-  scale_x_log10() + 
+  scale_y_log10(limits = ylm) + 
+  scale_x_log10(limits = ylm) + 
   thm +
   labs(
-    title = '(c) Estimated s(cont_year) between models'
+    title = '(c) Estimated smoother for continuous year between models'
     )
@@
 
-      out <- anlz_avgseason(mod, doystr = doystr, doyend = doyend) %>% 
-        anlz_mixmeta(yrstr = yrstr, yrend = yrend) %>% 
-        summary %>% 
-        coef(.) %>% 
-        .[2, ] 
-      out <- data.frame(est = out[['Estimate']], pval = out[['Pr(>|z|)']])
+      mixmet <- anlz_avgseason(mod, doystr = doystr, doyend = doyend) %>% 
+        anlz_mixmeta(yrstr = yrstr, yrend = yrend)
+
+      dispersion <- summary(mod)$dispersion
+      bt_prd <- 10 ^ (predict(mixmet) + log(10) * dispersion / 2)
+      df <- data.frame(chl = bt_prd, yr = mixmet$model$yr)
+      slope <- lm(chl ~ yr, df) %>% summary %>% coefficients %>% .[2, 1]
+
+      pval <- mixmet %>% summary %>% coefficients %>% .[2, 4]
+
+      out <- data.frame(est = slope, pval = pval)
 
@@
 
-toplo2 <- seastrnd2 %>%
+toplo2 <- seastrnd %>%
   mutate(
@@
 
-toplo <- seastrnd3 %>% 
+toplo <- seastrnd2 %>% 
   na.omit() %>% 
@@
     strip.background = element_blank(), 
+    strip.text = element_text(size = 13),
     axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
@@
     panel.grid.minor.y = element_blank(),
-    panel.grid.major.y = element_blank(), 
+    # panel.grid.major.y = element_blank(), 
     strip.text.x = element_text(size = 12, hjust = 0), 

### 2020-11-20 R/dat_proc.R

@@ -85,80 +85,9 @@ modprf <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
 
 save(modprf, file = 'data/modprf.RData', compress = 'xz')
 
-# seasonal trends by decade -----------------------------------------------
-
-seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
-  crossing(
-    fl = ., 
-    tibble(
-      doystr = c(1, 91, 182, 274), 
-      doyend = c(90, 181, 273, 364)
-    ), 
-    tibble(
-      yrstr = c(1991, 2000, 2010),
-      yrend = c(2000, 2010, 2019)
-    )
-  ) %>% 
-  group_by(fl) %>% 
-  nest() %>% 
-  mutate(
-    mod = purrr::map(fl, function(x){
-      
-      load(file = x)
-      
-      nm <- basename(x)
-      nm <- gsub('\\.RData', '', nm)
-
-      out <- get(nm) %>%
-        pull(model) %>% 
-        deframe
-      
-      return(out)
-      
-    })
-  ) %>% 
-  unnest(c('data')) %>% 
-  group_by(doystr, doyend, fl, mod) %>%
-  nest() %>%
-  mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
-  ) %>% 
-  unnest('data') %>% 
-  mutate(
-    yrstr = purrr::pmap(list(yrstr, avgseas), function(yrstr, avgseas){
-      max(yrstr, min(avgseas$yr))
-    }), 
-    yrend = purrr::pmap(list(yrend, avgseas), function(yrend, avgseas){
-      min(yrend, max(avgseas$yr))
-    }),
-    metatrnd = purrr::pmap(list(avgseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta)
-  ) %>% 
-  mutate(
-    yrcoef = purrr::map(metatrnd, function(x) x$coefficients['yr']), 
-    pval = purrr::map(metatrnd, function(x){
-      coefficients(summary(x)) %>% data.frame %>% .[2, 4]
-    }
-    )
-  ) %>% 
-  ungroup %>% 
-  select(station = fl, seas = doystr, yrs = yrstr, yrcoef, pval) %>% 
-  mutate(
-    station = gsub('^data/modslog\\_chl|\\.RData$', '', station),
[…]

### 2020-11-22 manu_draft.Rmd
started discussion

@@
 Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).
-However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an eexample), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to te basis functions.
+However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to the basis functions.
 These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab).
@@
 
+The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.
+This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv package that uses generalized cross-validation to extract an underlying signal in the observed data.
+Importantly, specifying an upper limit for the number of knots that can potentially be used to fit the smoother is critical to this approach.
+A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates.
+We further show that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended.
+This comparison demonstrated that model structure was of less importance than allowing the model sufficient freedom to estimate the trends over time.
+Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses developed herein.
+
+The extension of the GAM results to assess trends at an individual station is a new approach that previous trend analysis methods cannot sufficiently address.
+Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs.
+In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design.
+As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data.
+These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations.
+Aggregation of raw data, e.g., averaging of observations within a year or season to comply with the requirements of Kendall tests, risks loss of information by removing variation between observations at smaller time scales.
+The logical outcome is increased likelihood of incorrect conclusions from test results.
+
+Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainy from raw observations across methods is unaccounted for in trend assessment.
+These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.
+Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying Kendall test.
+In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.
+This is particularly problematic if sampling effort varies substantially between years.
+Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only).
+The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
+Further and most importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.
+As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.
+
+The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.
+For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing.
+Until implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.
+Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are not subject ot process or observation error.
+However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
+Although there are undoubtedly many scenarios where years of interest can be chosen obectively (e.g., regulatory compliance periods), a more general question of "when" changes occur independent of user decisions is also important to address.
+For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?
+There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.
+
+The trend assessment approach also does not address issues of causality in the observed trends.
+In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time.
+However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.
+This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.
+An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends.
+In such cases, considerations of model structure have direct implications on conclusions regarding potential causality.
+As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.
+Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs.
+year), whereas using GAMs with different predictors assess potential causality is a different application.
+Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to asssess different questions.
+
 Explain increasing resolution of methods, e.g., test/base, to seasonal averages, to mixed-meta, to moving window approach
 
-Explain how choosing window widths and time periods is a user choice that is arbitrary, identifying an objective approach to identifying trends can pave the way for future work (parallels with mgcv?)
-
-Benefits of mixed-meta to provide accurate results, reduces possibility of type I/II errors (e.g., a/b examples in trndcmpex)
+Application to other water quality variables and opportunity for ecological interpretation (cue up next paper?)
 

### 2020-12-02 manu_draft.Rmd
abstract done

@@
 # Abstract
+
+Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.
+However, mismatches between time scales present in the monitoring data and scales of analysis or management questions can create challenges for accurately assessing trends with full propagation of uncertainty across methods.
+Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in South San Francisco Estuary.
+GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.
+These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.
+Our proposed methods address the specific goal of estimating seasonal averages in a response variable and using those averages plus uncertainty to assess inter-annual trends, neither of which are addressed by previous applications of GAMs to environmental monitoring data.
+The approach first demonstrates that nearly identical descriptions of chlorophyll as a function of time can be obtained using different smoothing splines for annual or seasonal components of the time series.
+We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.
+Finally, we demonstrate how trends can be assessed with mixed-meta regression analyses that evaluates changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
+Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.
 
@@
 The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.
-The residuls is assumed to be normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.
+The residuals is assumed to be normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.
 The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.
@@
 As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
-Moreoever, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
-Trends assessed on observed data may include substantial noise indendent of any "canonical" signal derived from GAMs.
+Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
+Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.
 
@@
 This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv package that uses generalized cross-validation to extract an underlying signal in the observed data.
-Importantly, specifying an upper limit for the number of knots that can potentially be used to fit the smoother is critical to this approach.
+Importantly, specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.
 A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates.
-We further show that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended.
-This comparison demonstrated that model structure was of less importance than allowing the model sufficient freedom to estimate the trends over time.
-Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses developed herein.
-
-The extension of the GAM results to assess trends at an individual station is a new approach that previous trend analysis methods cannot sufficiently address.
+We further show that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package.
+This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time.
+Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significant of trends developed herein.
+
+The use of the GAM results with second and third stage analyses to assess trends is a new approach that previous trend analysis methods cannot sufficiently address.
 Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs.
@@
 
-Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainy from raw observations across methods is unaccounted for in trend assessment.
+Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
 These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.
-Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying Kendall test.
+Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test.
 In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.
@@
 The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
-Further and most importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.
+More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.
 As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.
@@
 For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing.
-Until implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.
-Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are not subject ot process or observation error.
+Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.
+Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minmally influenced process or observation error present in the raw data.
 However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
-Although there are undoubtedly many scenarios where years of interest can be chosen obectively (e.g., regulatory compliance periods), a more general question of "when" changes occur independent of user decisions is also important to address.
+Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.
 For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?
@@
 
-The trend assessment approach also does not address issues of causality in the observed trends.
+The trend assessment approach also does not address causality in the observed trends.
 In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time.
@@
 Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs.
-year), whereas using GAMs with different predictors assess potential causality is a different application.
-Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to asssess different questions.
-
-Explain increasing resolution of methods, e.g., test/base, to seasonal averages, to mixed-meta, to moving window approach
-
-Application to other water quality variables and opportunity for ecological interpretation (cue up next paper?)
-
-# Conclusions
+year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.
+Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.
+
+Finally, the evaluation of alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.
+The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay.
+These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.
+Measurements of dissolved oxygen can also be used to estimates gross primary production to assess process rates that may be more indicative of system function.
+We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.
+
+# Supplement
+
+The wqtrends R package includes all methods described in this manuscript.
+The package is available for download at https://tbep-tech.github.io/wqtrends.
+A full vignette describing installation and use is also available in the link provided.
+Please cite this manuscript in applications using the R package.
+
+When `select = TRUE` is included, the comparison between models S and SY changes.
+This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.
+This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it would be a strange choice if one is explicitly including a linear term for `cont_year`.
+If `select = TRUE` is used, models S and SY would still be effectively equivalent, but it would appear by AIC selection that one model is superior.
+That would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
+A link to an RMarkdown file demonstrating this issue is provided.
 
 # Acknowledgments
+
+We thank the staff of the US Geological Survey that collect and maintain long-term monitoring data in San Francisco Bay.
+This work benefited from discussions with the San Francisco Bay Nutrient Technical Workgroup and Steering Committee.
 
@@
 Plots in (a) show model predictions by day of year with separate lines for each year.
-Plots in (b) show pairwise comparisons of predicted chlorophyll between the models and plots in (c) show the same comparisons as in (b) but only for results from the esimated smoother for the `cont_year` variable.
+Plots in (b) show pairwise comparisons of predicted chlorophyll between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.
 The plots demonstrate that results between the models are comparable except for a few observations at extreme values(a), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (b).
-The 1:1 lines are in red to faciliate comparisons."}
+The 1:1 lines are in red to facilitate comparisons."}
 knitr::include_graphics('figs/modsumfig.png')
@@
 
-# Supplement
-
-When `select = TRUE` is included, the comparison between models S and SY changes.
-This option tells `mgcv` to penalize the linear term's coefficient in the spline.
-This would be appropriate if `cont_year` was an explanatory variable subject to variable selection -- which it is not here -- but it would be a strange choice if one is explicitly including a linear term for `cont_year`.
-If `select = TRUE` is used, models S and SY would still be effectively equivalent, but it would appear by AIC selection that one model is superior.
-That would be an artifact of the strange choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
-A link to an RMarkdown file demonstrating this issue is provided.
-
 # References

### 2020-12-02 manu_draft.Rmd
adding some references to discussion

@@
 
-The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15].
+The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15;@Beck18b].
 Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.
@@
 The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.
-This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv package that uses generalized cross-validation to extract an underlying signal in the observed data.
+This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv R package that uses generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
 Importantly, specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.
 A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates.
-We further show that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package.
+We further demonstrated that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
 This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time.
@@
 Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs.
-In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design.
-As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data.
+In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design [@Junninen04;@Racault14].
+As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data [@Hirsch82;@Helsel20].
 These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations.
@@
 This is particularly problematic if sampling effort varies substantially between years.
-Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only).
+Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].
 The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.

### 2020-12-03 manu_draft.Rmd
additional refs to discussion

@@
 Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes.
-The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18] [Marcus will go over refs].
+The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18].
 At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).
@@
 A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.
-They do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data [is this true?
-I would think so.], do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.
-Thus, while these non-parametric approaches have some degree of robustness, they apply only to narrow goals.
+For descriptive decomposition of long-term monitoring data, they do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data, do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.
+Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.
 
@@
 
-We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.
+We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Estuary, California, USA.
 For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.
@@
 For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially.
-For the raw San Francisco Bay data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.
+For the raw SFE data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.
 If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.
@@
 Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).
-However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially (Figure \@ref(fig:modsumfig)) because of structural differences and differences in the penalties applied to the basis functions.
+However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially because of structural differences and differences in the penalties applied to the basis functions.
 These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab).
-Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series but the model predictions are very similar.
+Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series, but the model predictions are very similar.
 
@@
 In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.
-These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the San Francisco Bay region.
+These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the SFE region.
 For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:
@@
 While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.
-Details are given in the Appendix.
 
@@
 
-The R package *mixmeta* was used to estimate the model in equation \@ref(eqn:mixmet) [@Sera19].
+The R package *mixmeta* was used to estimate the model in equation \@ref(eq:mixmet) [@Sera19].
 Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
@@
 
-For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South San Francisco Bay.
+For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.
 This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
@@
 The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
-More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses.
+More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].
 As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.
@@
 The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.
-For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing.
+For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].
 Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.
-Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minmally influenced process or observation error present in the raw data.
+Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced process or observation error present in the raw data.
 However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
@@
 This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.
-An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends.
-In such cases, considerations of model structure have direct implications on conclusions regarding potential causality.
+An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09].
+In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.
 As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.
@@
 
-Finally, the evaluation of alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.
-The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay.
+Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.
+The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20].
 These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.
-Measurements of dissolved oxygen can also be used to estimates gross primary production to assess process rates that may be more indicative of system function.
+Chlorophyll measurements can also be used to estimates gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
 We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.
@@
 
-When `select = TRUE` is included, the comparison between models S and SY changes.
+A note about the `select` argument in `mgcv`: When `select = TRUE` is included, the comparison between models S and SY changes.
 This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.
-This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it would be a strange choice if one is explicitly including a linear term for `cont_year`.
-If `select = TRUE` is used, models S and SY would still be effectively equivalent, but it would appear by AIC selection that one model is superior.
-That would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
-A link to an RMarkdown file demonstrating this issue is provided.
+This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it is irrelevant if including both a linear and spline term for `cont_year`.
+If `select = TRUE` is used, models S and SY would still be effectively equivalent, but AIC selection would suggest that one model is superior.
+This result would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
 
@@
 cap.val <- 'Comparison of the four model structures (S, SY, SYD, SYDI) described in the first stage analysis of GAM estimation.
-The four models provide either identical or comparable ability to describe chlorophyll trends at an example station (32) in South San Francisco Bay.
+The four models provide either identical or comparable ability to describe chlorophyll trends at an example station (32) in South San Francisco Estuary.
 The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are comparable.

### 2020-12-04 manu_draft.Rmd
fixing typos

@@
 The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt), whereas the southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
-The South Bay embayment is heavily urbanized and includes thirty-seven wastewater treatments plants that serve 7.2 million people.
+The South Bay embayment is heavily urbanized and includes thirty-seven wastewater treatment plants that serve 7.2 million people.
 Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.
@@
 The three-stage approach is motivated in several ways.
-In the first stage, the GAM is used for signal-extraction at the scale of the raw data and we explain the relationships among various GAM specifications to emphasis concordance between the trend-analysis goal and signal-extraction goal in the formulation of splines.
+In the first stage, the GAM is used for signal-extraction at the scale of the raw data and we explain the relationships among various GAM specifications to emphasize concordance between the trend-analysis goal and signal-extraction goal in the formulation of splines.
 The second stage uses mixed-meta analysis to propagate uncertainty from estimated GAM parameters to features of interest, i.e., seasonal averages.
@@
 The next three spline formulations differ in how spline terms compose a model to smooth the raw data.
-For convenience, a linear trend in `cont_year` can specified explicitly as a separate component.
+For convenience, a linear trend in `cont_year` can be specified explicitly as a separate component.
 This would be expressed as:
@@
 Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268.4% increase, although the change is insignificant because of large uncertainties in the chlorophyll averages in the first year.
-However, the change from 2000 to 2010 was an significant increase of 64%, followed by a significant decrease of 35% from 2010 to 2019.
+However, the change from 2000 to 2010 was a significant increase of 64%, followed by a significant decrease of 35% from 2010 to 2019.
 Although these trends were insignificant, the results are based on a priori decisions of which years to compare.
@@
 The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a significant trend.
-Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages form 2000 to 2010.
+Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.
 Unlike the first example, the top two figures show significant trend estimates, whereas the bottom plot for the mixed-meta regression analysis does not show a significant trend because of added uncertainty in the averages provided by the GAMs.
@@
 This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time.
-Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significant of trends developed herein.
+Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significance of trends developed herein.
 
@@
 This is particularly problematic if sampling effort varies substantially between years.
-Averages may be skewed in a particular direction if an annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].
+Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].
 The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
@@
 Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.
-Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced process or observation error present in the raw data.
+Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.
 However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
@@
 These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.
-Chlorophyll measurements can also be used to estimates gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
+Chlorophyll measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
 We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.
@@
 ```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at two example stations.
-The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
+The "observed" trend model is based on a linear fit to the annual averages of chlorophyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
 Regressions in each plot are fit through the seasonal estimates indicated in the plot titles.'}
@@
 ```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station.
-The "observed" trend model is based on a linear fit to the annual averages of chloropyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
-Values for each model are the slope esimates as annual change per year within each season, with color denoting significant trends.'}
+The "observed" trend model is based on a linear fit to the annual averages of chlorophyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
+Values for each model are the slope estimates as annual change per year within each season, with color denoting significant trends.'}
 knitr::include_graphics('figs/trndcmp.png')
@@
     'A linear continuous year variable, a smoother over a continuous year variable, and smoother over a day of year variable',
-    'A linear continuous year variable, a smoother over a continuous year variable, smoother over a day of year variable, and an interactin smoother across continuous year and day of year variables')
+    'A linear continuous year variable, a smoother over a continuous year variable, smoother over a day of year variable, and an interaction smoother across continuous year and day of year variables')
   )
@@
 cap.val <- 'Summary and details for each of the GAM structures.
-In practive, a sufficiently large number of knots provided to the additive terms will produce identical or comparable estimates for a response variable.
+In practice, a sufficiently large number of knots provided to the additive terms will produce identical or comparable estimates for a response variable.
 The models differ in the allocation of penalties for the smoothness of each spline (`s()`).'

### 2021-01-19 manu_draft.Rmd
started incorporating JH edits, intro done

@@
 Accurate and flexible trend assessment methods are valuable tools for describing historical changes in environmental monitoring datasets.
-However, mismatches between time scales present in the monitoring data and scales of analysis or management questions can create challenges for accurately assessing trends with full propagation of uncertainty across methods.
+A key requirement is complete propagation of uncertainty through the analysis.
+However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods.
 Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in South San Francisco Estuary.
@@
 These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.
-Our proposed methods address the specific goal of estimating seasonal averages in a response variable and using those averages plus uncertainty to assess inter-annual trends, neither of which are addressed by previous applications of GAMs to environmental monitoring data.
-The approach first demonstrates that nearly identical descriptions of chlorophyll as a function of time can be obtained using different smoothing splines for annual or seasonal components of the time series.
+Our proposed methods estimate seasonal averages in a response variable with GAMs and then use those ancertainty with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.
+We first demonstrate that nearly identical descriptions of temporal changes in l can be obtained using different smoothing splines for annual or seasonal components of the time series.
 We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.
-Finally, we demonstrate how trends can be assessed with mixed-meta regression analyses that evaluates changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
+Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluates changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
 Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.
@@
 
-Many environmental monitoring programs generate fine-scale but potentially irregular time-series data to assess long-term trends for regulatory, management, or research purposes.
-The mismatch between the scales of monitoring versus analysis questions or management goals can lead to statistical challenges [@Urquhart98;@Cumming06;@Forbes18].
-At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events, management, or seasonal changes).
-Such fluctuations may not be of direct interest for longer-term trends or may not be well-suited to multi-scale smoothing methods.
-However, aggregate features of seasonal patterns that integrate or summarize short-term fluctuations may be of interest.
+Accurate quantification of trends must consider variation at different temporal scales regardless of the question of interest, such that ignoring variation at one scale could lead to incorrect conclusions about variation at another scale.
+Many environmental monitoring programs collect temporally resolved but irregular time-series data to quantify trends for regulatory, management, or research purposes.
+The mismatch between the scales of monitoring versus analysis questions or management goals can present statistical challenges [@Urquhart98;@Cumming06;@Forbes18].
+At short temporal scales typically less than a year, environmental systems exhibit variability caused by multiple factors (e.g., weather events, management, or seasonal changes).
+Such fluctuations may not be of direct interest for inter-annual trends or may not be well-suited to multi-scale smoothing methods.
 In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.
 
-Previous methods for water quality trend analysis can be generalized into four basic approaches.
-Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes that account for seasonal variation between years [@Hirsch82;@Helsel20].
-A literature survey of @Wan17 revealed non-parametric approaches to be the most commonly used methods in long-term water quality trend analysis, yet they have limited scope.
-For descriptive decomposition of long-term monitoring data, they do not account for changes occurring at different temporal scales, do not adequately evaluate irregularly spaced data, do not incorporate other explanatory variables, and do not estimate a model that could be useful for other purposes.
+Existing methods that address our objectives in water quality trend analysis can be generalized into four basic approaches: seasonal Kendall tests, seasonal trend decomposition (STL), weighted regression on time, discharge, and season (WRTDS), and generalized additive models (GAMs).
+Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes over several years while accounting for the predictable patterns among seasons [@Hirsch82;@Helsel20].
+@Wan17 showed that non-parametric approaches have been the most commonly used methods in long-term water quality trend analysis despite critical limitations.
+For descriptive decomposition of long-term monitoring data, they assume seasonal patterns within year do not change, require regularly spaced or balanced data, only use time as a descriptor for trends, and do not estimate a model that could be useful for other purposes.
 Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.
@@
 While useful and widely applied, this method also has important limitations.
-STL decomposition does not incorporate explanatory variables, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually include standard errors to allow hypothesis testing [but see @Hafen10].
-Conventional STL approaches may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.
+STL decomposition does not incorporate explanatory variables besides time, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually estimate standard errors to allow hypothesis testing [but see @Hafen10].
+STL methods may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.
 This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].
 
-The more recently developed method of weighted regression on time, discharge, and season (WRTDS) uses a more general local regression scheme than STL [@Hirsch10;@Beck15;@Beck18b].
-Designed for river data where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allows parameters to vary smoothly in relation to both time and discharge.
+The weighted regression on time, discharge, and season (WRTDS) method addresses the problem of inflexibility in STL by using a more general local regression scheme [@Hirsch10;@Beck15;@Beck18b].
+Designed for evaluating water quality in rivers where separating the effect of discharge on constituent concentration is important, WRTDS estimates a moving window regression model with components that allow parameters to vary smoothly in relation to both time and discharge.
 This yields parameters that are specific to season, year, and flow regime.
-Conceptually, the approach is similar to localized multi-polynomial smoothing methods, although the application was developed specifically for describing long-term water quality trends.
-Standard error estimates of predictions are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15].
-Although a useful addition to the original method in @Hirsch10, the approach requires extensive resampling as a post-hoc application to a previously fitted model.
-
-The final and most recent approach is to use smoothing splines to separate fluctuations on different time scales and do so within the larger framework of generalized additive models (GAMs) [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].
-These may be seen as generalizing the concepts behind STL and WRTDS.
-In statistics, the evolution of non-parametric regression methods has largely converged on GAMs rather than more generalized kernel smoothing methods used by both STL and WRTDS.
+The WRTDS approach is conceptually similar to localized multi-polynomial smoothing methods, with specific application to descriptive variables relevant for water quality constituents (i.e., season, year, and discharge).
+Standard error estimates of predictions from WRTDS are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15].
+Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling as a post-hoc application to a previously fitted model.
+Alternative methods that include standard error estimates simultaneously with model output may be preferred for intensive or more iterative applications
+
+Generalized additive models (GAMs) are central to this paper and form the basis of our fourth and last method to separate fluctuations on different time scales.
+GAMs combine one or more smoothing splines to model patterns in data and may be seen as generalizing the concepts behind STL and WRTDS [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].
+In statistics, the evolution of non-parametric regression methods has largely converged on GAMs, rather than kernel smoothing methods used by both STL and WRTDS.
 Kernel smoothing and spline smoothing are closely related, and a key challenge for each is to determine the appropriate degree of smoothing.
-For example, WRTDS can potentially give results similar to the spline-based smoothing methods described next, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17].
-User input is also required to specify an acceptable degree of smoothing used by the "windows" that define the localized fit of WRTDS at each point in the time series.
-These windows are conceptually similar to the kernel (or bandwidth) that used in more conventional smoothing methods.
-There is no simple rule to guide the choice of defining an appropriate size and a tradeoff between over- and under-smoothing is a hallmark of these approaches.
-
-Compared to kernel smoothing methods, GAMs have various advantages.
-They are formulated using "basis functions", and these can be customized for needs such as cyclic splines (e.g., for an annual pattern) and low-dimensional interactions.
-They can naturally include both parametric (e.g.
+For example, WRTDS can potentially give results similar to the spline-based smoothing methods used in GAMs, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17].
+WRTDS also requires specifying the desired degree of smoothing via the "windows" that define the local weighting at each point in the time series.
+The windows in WRTDS are conceptually similar to the kernel (or bandwidth) used in more conventional smoothing methods.
+A challenge of WRTDS is that there is no clear rule that can be applied to determine the best choice of window size.
+The associated tradeoff between over- and under-smoothing is a hallmark of these approaches.
+
+GAMs have various advantages compared to kernel smoothing methods and are therefore used for non-parametric regression smoothing in many fields.
+The "basis functions" used to formulate GAMs can be customized based on expected patterns in the data.
+Examples inlcude cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
+GAMs have added flexibility because they can include both parametric (e.g.
 linear or quadratic) components and non-parametric (spline) components.
-Importantly, multiple approaches to automatically determine the optimal degree of smoothness have been developed, based on likelihood and/or optimizing out-of-sample prediction error.
-They have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].
-For these reasons, GAMs are widely used for non-parametric regression smoothing in many fields.
-
-GAMs have recently been applied to trend analysis of water quality time-series, particularly from long-term monitoring programs [@Haraguchi15;@Murphy19], but with different formulations and goals than given here.
-For example, in the US, the Chesapeake Bay Program uses GAMs to decompose time-series into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.
-That is related to the methods here, but this paper gives more general methods for analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.
-Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and different time lags in describing relationships between response and predictor variables [@Lefcheck17;@Testa18].
-
-Our motivating problem has several needs that are not satisfied by previous methods, but can use GAMs as a starting point.
+Multiple approaches have been developed to determine the optimal degree of smoothness.
+These approaches are based on likelihood estimates or optimization of out-of-sample prediction error, which address a key concern around methods like WRTDS that do not have analogs for choosing optimal degrees of smoothing.
+Further, GAMs have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (i.e., generalized additive mixed models or GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].
+
+GAMs have been applied previously to evaluate trends in water quality time-series from long-term monitoring programs [@Haraguchi15;@Murphy19].
+For example, @Murphy19 used GAMs to decompose water quality time series from Chesapeake Bay into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.
+This study generalizes the approach to analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.
+Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17;@Testa18].
+
+Our motivating problem has several characteristics that are partially addressed by previous methods and can further build on  GAMs as a starting point.
 Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.
-However, the seasonal average within each year must account for different sampling times and intervals, and any trend analysis must incorporate the uncertainties in seasonal averages.
-STL and/or WRTDS could potentially separate seasonal from long-term trends, but doing so is not necessary to determine seasonal averages.
-What is needed is an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, allowing for irregular sampling and the non-independence inherent in time-series.
-This can be done with GAMs, but we develop an application that is distinct from previous studies.
-Even if estimates of seasonal averages and their standard errors are available, none of these methods are designed to understand interannual trends in those averages.
-A Kendall test would not incorporate the standard errors or reveal useful long-term patterns beyond a significance test.
-Similarly, STL and WRTDS are not designed for this goal.
-
-We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Estuary, California, USA.
-For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.
-Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.
-Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?
+However, the seasonal average within each year must be robust to inconsistent sampling times and intervals, and any trend analysis must consider the uncertainties in seasonal averages.
+The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common to time series data.
+This paper develops GAMs as a distinct application that can meet these requirements and simultaneously addressing key limitations of seasonal Kendall tests and the more complex STL and WRTDS methods.
+We do so with the explicit goal of quantifying interannual trends in seasonal averages.
+
+We describe and demonstre the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.
+Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.
+Analysis of these data is complicated by irregularities in timing and consistency of data collection which can generate artifacts affecting simple seasonal averages of the data.
+We were interested in questions such as: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?
 At what across-year scale do within-year summer-fall mean chlorophyll levels change?
-Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?
-We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] following signal extraction with GAMs that is new to environmental trend-detection problems.
-For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between years.
+Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or baseline levels of productivity between blooms?
+We provide examples illustrating how these questions can be addressed using methods with GAMs.
+We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] after signal extraction with GAMs.
+This approach is new to environmental trend-detection problems.
+For this step, we provide methods using meta-anlaysis for isolating seasonal trends with reasonalbe certainty from from GAM results and evaluating these trends between years.
 

### 2021-01-19 README.Rmd
trying gh action build for manu

---
output: github_document
---

# README

[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions): last manuscript build on `r Sys.Date()`

Materials for water quality trends assessment using GAMs

## Data

* `datprc.RData`: Formatted water quality data from <https://github.com/fawda123/SFbaytrends>

* `mods_chl18.RData` through `mods_chl36.RData` also from <https://github.com/fawda123/SFbaytrends>

## Docs

* Outline <https://tbep-tech.github.io/wqtrends-manu/outline.html>

* Literature <https://tbep-tech.github.io/wqtrends-manu/lit.html>

* Manuscript draft (rendered in Word) <https://tbep-tech.github.io/wqtrends-manu/manu_draft.docx>

### 2021-01-20 README.Rmd
working on methods revision from JH

@@
 
-[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions): last manuscript build on `r Sys.Date()`
+[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions) last manuscript build `r Sys.time()`
 

### 2021-01-20 manu_draft.Rmd
trigger build
trigger build again with maptools depends
working on methods revision from JH

@@
 
-`r paste('Compiled', Sys.time())`
+`r paste('Last manuscript build', Sys.time())`
 
@@
 The "basis functions" used to formulate GAMs can be customized based on expected patterns in the data.
-Examples inlcude cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
+Examples include cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
 GAMs have added flexibility because they can include both parametric (e.g.
@@
 
-The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America and drains an area of approximately 200 thousand km$^2$ in the US state of California.
+The San Francisco Estuary (SFE) is the largest estuary on the Pacific Coast of North America.
+Its watershed covers 200 thousand km$^2$ in the US state of California.
 Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.
-The northern subembayments are river-dominated (salinity ranging from 0 to 15 ppt), whereas the southern subembayments are marine-dominated with salinity ranging from 5 to 35 ppt depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
+Salinity ranges from 0 to 15 ppt in the northern subembayments and from 5 to 35 ppt in southern subembayments closer to the Pacific Ocean, depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
 The South Bay embayment is heavily urbanized and includes thirty-seven wastewater treatment plants that serve 7.2 million people.
-Secondary treatment occurs at a majority of the treatment plants and the remaining effluent is discharged into the SFE.
-Agricultural runoff from the upper watershed also contributes to nutrient loading in the SFE with the annual nutrient export estimated as approximately 30 thousand kg dy$^{-1}$ of nitrogen from the Delta.
-
-Nitrogen and phosphorus levels in SFE generally exceed concentrations that have been observed to promote excess primary production in other large estuarine systems.
-However, eutrophic conditions have not been regularly observed since routine monitoring began in the 1970s.
-Historical resistance of SFE to eutrophication has been attributed to several factors, including elevated suspended sediments that reduce light penetration in the water column, regular exchange and mixing with low-nutrient marine waters and export of estuarine nutrients to the Pacific Ocean, and benthic grazing by filter-feeding bivalves that reduce algal concentrations.
-Renewed interest in the potential for nutrient loading to negatively affect water quality has occurred recently, particularly in South Bay, where harmful algal blooms (HABs), increases in summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations beginning in 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
-Although visual changes in observed data are apparent, statistical analyses to quantify current status and to provide estimates of annual and seasonal trends with appropriate bounds on uncertainty have not been sufficiently developed, particularly on a seasonal basis.
-
-The analysis evaluated near-surface chlorophyll-a data collected biweekly to monthly along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
+An estimated 73.8 metric tons dy$^{-1}$ of inorganic nitrogen are discharged into the Bay, primarily from wastewater [@Novick14].
+Secondary treatment occurs at most of the wastewater treatment plants that discharge into the SFE.
+Agricultural runoff from the upper watershed contributes 30 metric tons dy$^{-1}$ of nitrogen to the SFE via the Delta.
+
+Nitrogen and phosphorus levels in SFE usually exceed concentrations that cause eutrophication in other estuaries.
+However, SFE has demonstrated resistance to eutrophication, which to high suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves.
+Interest in the potential for nutrient loading to negatively affect water quality has increased recently particularly for South Bay, where harmful algal blooms (HABs), elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations have occurred since 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
+Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient particularly with respect to seasonal differences between years.
+
+We evaluated near-surface chlorophyll-a data measured biweekly to monthly from 1990 to 2019 along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
 Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
-Discrete chlorophyll concentrations at each station were determined by fluorimetric analysis with 90% acetone pigment extraction on GFF filters.
-Data collected between 1990-2019 were selected for analysis because it represented a suitable balance among three factors relevant to testing the statistical approaches, including sufficient length of record, consistent biweekly-monthly sampling, and a diverse set of stations covering the salinity gradient across multiple subembayments.
-While sampling frequency varied somewhat over time or by station, all data were treated as unique time series within the statistical models (i.e., no spatial or temporal binning or averaging was done).
+Chlorophyll on GFF filters was analyzed fluorometrically after extraction in 90% acetone.
+Sampling frequency varied somewhat over time and by station.
+Every observation was included directly in the statistical models without spatial or temporal binning or averaging.
 
@@
 
-The methods proposed here involve three stages.
-First, a GAM is used to estimate a smooth pattern of variation in the raw data along with its uncertainty.
-Second, a feature of interest is calculated from the estimated GAM, along with its propagated uncertainty.
-For this example, the seasonal averages are extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
-Third, a mixed effects meta-analysis is used to estimate trends and test hypotheses about the change in seasonal averages across years.
+We implemented the GAMs in three stages.
+First, we used a GAM to estimate a smooth pattern of variation in the raw data along with its uncertainty.
+Second, we calculated a feature of interest from the estimated GAM, along with its propagated uncertainty.
+For this example, the seasonal averages were extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
+Third, we used a mixed effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
 While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
 Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions.
-Two mixed effects meta-analysis approaches are developed, 1) simple comparison of whether seasonal features differ across years, and 2) estimation of short-term linear trends on time-scales chosen by the analyst.
-
-The three-stage approach is motivated in several ways.
-In the first stage, the GAM is used for signal-extraction at the scale of the raw data and we explain the relationships among various GAM specifications to emphasize concordance between the trend-analysis goal and signal-extraction goal in the formulation of splines.
-The second stage uses mixed-meta analysis to propagate uncertainty from estimated GAM parameters to features of interest, i.e., seasonal averages.
-Such features can be practically evaluated as functions of the fitted GAM, rather than as parameters to be estimated within the GAM.
-On a practical level, the first two stages can be viewed as creating a "data product" that can be used in any subsequent analyses as long as they include the uncertainty associated with each point estimate.
-This offers an important practical benefit for visualization and management purposes.
-For the third stage (long-term trend-analysis), we harness existing meta-analysis methods that are well-suited to the task at hand.
+We developed two mixed effects meta-analysis approaches that provided 1) a simple comparison of whether seasonal features differ across years, and 2) estimation of short-term linear trends on time-scales chosen by the analyst.
 
@@
 
-Generalized additive models use sums of parametric terms (e.g., linear) and non-parametric terms (e.g., smoothing splines) to predict response data.
-In our approach, GAMs are used to estimate signal and uncertainty from the time-series data.
-Results are subsequently used in the second-stage analysis, described below.
-
-In this section we describe four GAM formulations that can achieve the same or similar fits to data in slightly different ways (Table \@ref(tab:modsumtab)).
-We present these to clarify relationships that may otherwise be confusing and to emphasize that results can be similarly achieved if a model unconstrained to have sufficient capability to extract a signal from raw data.
-Models will be written in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17].
-
-The fundamental goal is to smooth the raw data across time to separate a signal of response variable from noise. The simplest GAM for this purpose can be expressed as:
+We used four different GAMs to estimate the long-term trend and uncertainty from the time series data, each with the potential to achieve a similar fit to the data but with a slightly different model structure (Table \@ref(tab:modsumtab)).
+Models are shown in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17].
+
+GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor from noise inherent in the raw data.
+The simplest GAM for this purpose can be expressed as:
 
@@
 Here `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.
-Following previous authors [@Murphy19;@Yang20], we consider log-transformed (base 10) chlorophyll levels for analysis.
+Log$_{10}$-transformed chlorophyll was used for all analyses to meet assumptions of normally-distributed residuals.
 

### 2021-01-21 manu_draft.Rmd
more methods edits from JH

@@
 
-GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor from noise inherent in the raw data.
-The simplest GAM for this purpose can be expressed as:
+GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor, indepdent of noise inherent in the raw data.
+The simplest GAM for this purpose is expressed as:
 
@@
 
-Here `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.
+where `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
 Log$_{10}$-transformed chlorophyll was used for all analyses to meet assumptions of normally-distributed residuals.
 
-For smoothing in `mgcv`, appropriate smoothness is achieved not by choice of knots but by a penalty on the net curvature of the spline [@Wood04].
+The optimal level of smoothing in `mgcv` is determined by a penalty on the net curvature of the spline [@Wood04].
 Smoothing was determined using generalized cross-validation (GCV, as implemented in `mgcv`), which approximately minimizes out-of-sample prediction error.
-To allow GCV (or other alternatives) to work as intended, the number of knots must be sufficiently large.
+To allow GCV (or other alternatives) to work as intended, a potential upper limit on the number of knots determined by the analyst must be sufficiently large.
 Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.
-Therefore, we choose a sufficiently large number of knots for `num_knots_Y` as 12 times the number of years in the time series that was modeled.
-This created the potential to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.
-In some cases, this upper limit was too large given the observed data and the multiplier was reduced by one until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).
-
-The next three spline formulations differ in how spline terms compose a model to smooth the raw data.
-For convenience, a linear trend in `cont_year` can be specified explicitly as a separate component.
-This would be expressed as:
+We chose a sufficiently large number of knots for `num_knots_Y` (Model S) as 12 times the number of years in the time series that was modeled.
+This created the potential, as determined through GCV, to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.
+If the data were too sparse, the model will have insufficient degrees of freedom to fit 12 knots per year, in which case the number of knots was reduced by one knot per year until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).
+
+The next three spline formulations (Model SY, SYD, and SYDI) provide progressively increasing complexity in how spline terms compose a model to smooth the raw data.
+Model SY describes the time series using a linear trend plus a spline for `cont_year`:
 
@@
 
-This model is effectively and mathematically equivalent to model S (Table \@ref(tab:modstrtab)).
+This model is mathematically equivalent to model S (Table \@ref(tab:modstrtab)).
 The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
-When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.
-With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.
-With model SY, it is available more conveniently.
-Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.
+When `cont_year` is included explicitly as a linear term in model SY, `mgcv` adjusts the basis functions for the spline to exclude the linear term, thereby not over-parameterizing the model.
+Whereas an estimated linear trend in `cont_year` and its uncertainty can be extracted from the fitted spline in model S, model SY provides this trend directly, given the equivalent result.
+Further, package `mgcv` can penalize linear trends in splines to provide a method for variable selection (option `select = TRUE`), such as when numerous splines are included in the model formulation for variables that may or may not be important.
 For our approach, this option is not used and all models specify `select = FALSE`.
@@
 
-Next, an average within-year cyclic pattern as a separate spline can be separated in the model formulation.  This can be expressed as:
+MOdel SYD adds an average within-year cyclic pattern as a separate spline:
 
@@
 
-Here `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.
-While model SYD is not mathematically equivalent to models S and SY, it should be functionally similar.
-In this model, the `doy` spline gives the average pattern of fluctuations within each year.
-This changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.
+where `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates that the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.
+While model SYD is not mathematically equivalent to models S and SY, it should produce nearly identical results.
+The `doy` spline in model SYD gives the average seasonal pattern and changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.
 
@@
 In model SYD, there are separate penalties for the two splines, as compared to S and SY that include penalties only for the `cont_year` spline.
-Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17].
-For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations.
+This is important because variation in the response variable can be differently attributed to each spline depending on model, whereas the sum of components for each model produce comparable results between models.
+Our goal is to extract signals (i.e., seasonal estimates) from the fitted time series and different allocation of penalties among the splines in each model is unimportant for this task.
+However, interpreting differences in fit between model SYD and models S or SY can be problematic because the penalties for smoothing splines based on curvature are heuristic [@Wood17].
+For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations and does not imply one model fits better.
 In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.
 
-Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.
-The model will impute an average seasonal pattern where data are limited.
-However, a conceptual representation of what an average within-year pattern represents is challenging.
-For example, suppose a spring chlorophyll peak is a notable feature in every year.
-If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude thta appears as a "typical" year.
-However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.
-The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.
-
-More importantly, the average within-year pattern (and its uncertainty) can be obtained from estimated models S and SY, as well as from model SYD, but using a different approach.
-In models S and SY, the average within-year pattern is simply the average across years of the smoothed within-year patterns.
-From this perspective, model SYD, like model SY, is a convenience, estimating the average within-year pattern directly as a separate estimated spline.
-The average pattern extracted from model S or SY may not be identical to the `doy` spline estimated in model SYD, again due to different smoothing formulations, but in practice they should be similar.
+Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will consider data from the same period in other years by imputing an average seasonal pattern with the `doy` spline.
+However, an interpretation of these imputations is challenging.
+For example, the spring chlorophyll peak is a notable feature every year in south SFE.
+If the peak occurs at the same time every year but the magnitude varies, then the average within-year pattern can be interpreted as the average magnitude.
+However, if the magnitude is the same but the timing varies across years, then the magnitude of the average peak cannot be similarly interpreted and instead underestimates the magnitude that usually occurs.
+Moreover, the width or duration of the peak will be longer than typically occurs in a given year.
 
@@
 
-Here, `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.
+where `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.
 The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.
-In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both to avoid over-parameterization.
-
-Following similar rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.
-Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.
-The splines in `cont_year` and `doy` will likely not capture as much variation as above because of fewer knots.
+In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both given limits on the model degrees of freedom.
+
+Following the rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.
+Model SYDI differs formulaically from model S to a greater extent than model SYDI, but all of the splines use the same inputs to smooth the same data.
+The splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model SY given the fewer knots that are available to the former.
 The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.

### 2021-01-23 manu_draft.Rmd
through first attachment from JH

@@
 
-We describe and demonstre the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.
+We describe and demonstrate the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.
 Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.
@@
 
-The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly across years.
-For example, if one year has an early and large magnitude spring bloom, the years before and after would have similar patterns described by the model, perhaps different but not substantially.
-For the raw SFE data, this may not be the case, as one year might have an early, high spring bloom and the next year a late, low spring bloom, or other such large variation across years.
-If the interaction spline has sufficient knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.
-However, if that is the case, the conceptual motivation for model SYDI may not be justified.
-In summary, model SDYI may best be suited for systems where within-year fluctuations actually vary smoothly across years, in which case it may out-perform the other models.
-If that is not the case, it may perform similarly to the others, but the spline structure and penalty structure are more different, making performance harder to compare (Table \@ref(tab:modstrtab)).
-
-All of the models require a sufficiently large number of knots in one or more of the spline terms to accommodate the fluctuations in the data and use the GAM estimation methods in the manner for which they were designed.
+The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year.
+For example, if one year has an especially early and large magnitude spring bloom, the years before and after would be expected also to have earlier and larger than average spring blooms.
+For the raw SFE data, and chlorophyll-a dynamics in many estuaries, this is not expected to be the case.
+An early and large spring bloom could easily be followed the next year by a late and relatively small spring bloom.
+If the interaction spline has many knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.
+However, if that is the case, the conceptual motivation for model SYDI may not be supported by the data.
+Given that observed bloom dynamics for most systems do not vary smoothly across years, the potential benefits of using model SYDI could be outweighed by using simpler models.
+Moreover, the spline and penalty structure differ from the other models, making it difficult to evaluate their relative performance (Table \@ref(tab:modstrtab)).
+
+For models SYD and SYDI, the allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to related differences in allocation of variation among the splines.
 In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.
@@
 Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.
-In practice, allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to more variation being explained between the splines.
-
-@Murphy19 used a related set of spline formulations but used them differently than we do here.
+
+@Murphy19 used spline formulations related to those proposed here, but are insufficient for our needs of extracting and quantifying seasonal trends.
 @Murphy19 have a "`gam0`" with only a `s(doy)` term, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
-Their `gam0` is *a priori* not of interest for our data and was not explored further.
-Compared to our uses of SYD and SYDI, their uses differ in choice of knots and interpretation of results.
-For SYD, they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we use 12 times the number of years, roughly one per month.
-Thus, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate signal from noise at a scale informed by the data.
-In the SYDI model, @Murphy19 do not explicitly consider number of knots in the interaction spline.
-In both cases, @Murphy19 in effect use choice of number of knots in different spline components to serve as ad hoc allocation of variation in the data to different components while not necessarily allowing any component to fully estimate the signal in the data.
-@Murphy19 acknowledge that incomplete modeling of fluctuations in the data may lead to inflated Type I error rates, which would apply to their later comparisons of changes across time, but they leave that problem for future work.
-For the methods presented here, we seek to avoid inflated Type I error rates arising in this way.
+Their `gam0` is *a priori* not of interest for our data because of an assumption of constant seasonality and a linear annual trend.
+Compared to our application of SYD and SYDI, their implementation of "`gam1`" and "`gam2`" allocated knots differently, leading to a different interpretation of results.
+For "`gam1`" (our SYD), they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we adopt a different approach using a number of knots equal to 12 times the number of years, or one per month.
+Accordingly, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate intra-annual changes in season from noise at a finer temporal scale.
+In the "`gam2`" model (our SYDI model), @Murphy19 do not explicitly consider number of knots in the interaction spline.
+In both cases, they effectively use their choice of number of knots in different spline components as an ad hoc allocation of variation in the data to different components, while not allowing any component to fully estimate the signal in the data.
+@Murphy19 acknowledge that incomplete modeling of fluctuations in the data may inflate their Type I error rates for estimating temporal changes, but they leave that problem for future work.
+We seek to avoid inflating Type I error rates in this way.
 Finally, @Murphy19 present large AIC differences between their spline formulations.
-We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)).
-In our example, such large differences in AIC would only reflect inadequate choice of knots in one or more splines.
-
-We present a simple visual comparison of chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).
+We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)), making differences in AIC unimportant.
+In our example, large differences in AIC only reflect inadequate choice of knots in one or more splines, which should be avoided.
+
+We visually compare chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).
 Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
-Model predictions by day of year are visually similar between the models (Figure \@ref(fig:modsumfig)a) and are very similar when compared to a 1:1 line (Figure \@ref(fig:modsumfig)b).
-However, when comparing similarities among the models using only the continuous year smoother (`s(cont_year)`, as an example), the models differ substantially because of structural differences and differences in the penalties applied to the basis functions.
-These results are also demonstrated in differences between the effective degrees of freedom among the additive components of each model in Table \@ref(tab:modstrtab).
-Accordingly, the models may differ in which structural component describes variation in the chlorophyll time series, but the model predictions are very similar.
+Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).
+However, when contrasting the models using only the continuous year smoother (`s(cont_year)`, as an example), the model fits differ substantially because of structural differences and differences in the penalties applied to the basis functions.
+These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)).
+Accordingly, even though the models differ by which structural component describes variation in the chlorophyll time series, they provide similar predictions.
 

### 2021-01-24 manu_draft.Rmd
methods revision from JH done

@@
 
-Next we consider the goal of estimating a seasonal average, such as the mean spring chlorophyll concentrations and the estimated uncertainty in each year.
-Define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
+In the second-stage analysis, we estimated a seasonal average, such as the mean spring chlorophyll concentrations and the uncertainty in each year.
+Define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
 The season includes $n$ days.
@@
 
-Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of estimate parameters and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.
-Here $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
+Point estimates of response values for the fitted GAM take the form $\hby = \bX \hbb$, where $\hbb$ is the vector of parameter estimates and $\mathbf{X}$ is a model matrix of explanatory variables, including spline basis function values.
+Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
 For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
@@
 To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.
-Here, the seasonal averages used in our examples are calculated at the resolution of days.
+Here, the seasonal averages used in our examples were calculated at the resolution of days.
 The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.
-The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$.
-Since parameter estimates are correlated, it is important to use the full $\hShb$ to determine the covariance of $\hby$.
-This is given by $\hShy = \bX \hShb \bX^T$.
-
-Finally, the estimated seasonal average is $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.
+The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$, wheras the off-diagonal elements are the correlations among the elements of $\hbb$.
+Since parameter estimates are correlated, the covariance of $\hby$ is $\hShy = \bX \hShb \bX^T$.
+
+The estimated seasonal average was calculated from the vector of daily values for each of the $n$ days in the season of interest with $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.
 The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.
-Each of these estimates are standard results based on approximate multivariate normality of the sampling distribution of $\hbb$.
-A similar summary for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs.
-end of a time-series is provided by @Murphy19.
-For example, the above estimates can be used for a test of the null hypothesis that `y` is the same at two times.
+Each of these estimates are from on approximate multivariate normality of the sampling distribution of $\hbb$.
+@Murphy19 provided a similar approach for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs.
+end of a time-series.
+For example, the above estimates were used for a test of the null hypothesis that `y` is the same at two times.
 To do so, $\bX$ would have two rows for each time to compare.
 The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.
-This difference can then be used to determine if the two estimates in time are significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates.
+This difference was used to determine if the two estimates in time were significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates.
+An alternative interpretation of this test is a quantification of differences between two points in time as affected by the magnitude of the difference and range of likely values for the magnitude of change.
 
@@
 
-This section explains how the within-year features ($\hmt$) from the estimated GAMs, and their standard errors $\hshmt$, are used to investigate across-year trends of within-year features.
-In the examples, we use two seasonal averages of chlorophyll as features, one for January-July and the other for August-December.
-These periods are chosen for their relevance to phytoplankton bloom phenology and broad seasonal patterns that occur in the SFE region.
-For any feature of interest that includes estimates with associated standard errors, meta-analysis methods can be used to address two kinds of questions:
-
-1. Do two years, or a group of years, differ significantly in a seasonal average?
-2.
-Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?
-This question can be posed in a moving-window manner across a time-series.
-
-For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.  $N$ is the number of years of the study.
-
-### Differences among years in seasonal averages
-
-To test the null hypothesis that  $\mu_t = \mu_r$ for years $r$ and $t$, one can use simple normal distribution theory that $\hmt - \hat{\mu}_r$ has standard error $(\hshmt^2 + \hshmr^2)^{1/2}$.
-To test the null hypothesis that all means in a group of (>2) years are equal, one can use a simple case of the meta-analysis given in the next section or a general linear hypothesis test.
-The latter readily includes consideration of multiple testing, which is provided in the R package `multcomp` [@Hothorn08].
-While this package typically uses the results of a linear or generalized linear model fit as input, alternative values of $\hmt$ and $\hshmt$ could be provided as inputs.
-
-### Linear trends across years at a chosen time-scale of interest
-
-The standard meta-analysis mixed effects model to estimate linear trends is given as follows, using notation similar to [@Sera19]:
+In stage three of the analysis, we used a meta-analysis method to evaluate across-year trends of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
+In the examples, we considered two periods of interest for chlorophyll, January-July and August-December, which are relevant to phytoplankton bloom phenology in south SFE.
+
+This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?
+For example, is there a trend in the spring chlorophyll average from 1990 to 2000?
+This question can also be posed in a moving-window manner across a time-series (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
+For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.
+$N$ is the number of years of the study.
+
+### Linear trends across years at a chosen time-scale
+
+A meta-analysis mixed effects model is useful to estimate linear trends when each observation has a unique and quantified standard error, which is the case with our estimates $\hmt$ and $\hshmt$.
+Differences in standard errors, which may result from different monitoring effort between years, are explicitly considered in the analysis.
+The model can be expressed using notation similar to @Sera19:
 
@@
 where $\beta_0$ is the intercept, $t$ is the year, $\beta_t$ is the slope, $b_t$ is the random effect for year $t$, and $\epsilon_t$ is the residual for year $t$.
-Here the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.
+Accordingly, the seasonal average for year $t$ is $\mu_t = \beta_0 + \beta_t t + b_t$.
 The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.
-The residuals is assumed to be normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.
+The residuals are assumed to be independent and normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.
 The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.
@@
 
-The R package *mixmeta* was used to estimate the model in equation \@ref(eq:mixmet) [@Sera19].
-Results have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
-Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed, whereas others are estimated ($\sbs$).
+We estimated the model (equation \@ref(eq:mximet)) using the *mixmeta* package in R [@Sera19].
+Results from *mixmeta* have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
+Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed and others estimated ($\sbs$).
 The default estimation method for meta-analysis models is restricted maximum likelihood (REML).
 
-We apply this method for windows of a chosen width of years.
-This approach evaluates whether there is a significant linear trend in chlorophyll over a chosen series of years, by providing a more comprehensive assessment of all the data in those years (i.e., the seasonal averages, plus their uncertainty).
-Some years may have had more sampling, resulting in smaller standard error for $\hmt$, while others may have had less sampling, result in higher standard error.
+This method was applied to a chosen sequence or "window" of years for estimating the linear trend.
+This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chlorophyll over a chosen series of years.
+Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).
+Years with more data have smaller standard errors for $\hmt$ compared to years with less data.
 The meta-analysis explicitly includes these differences in uncertainty magnitudes.
 
-As a final analysis, the chosen windows for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not.
-Although the initial window width and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.
+As a final analysis, the chosen sequence of years for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
+Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.
 The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.
-We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
+We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year (e.g., an estimate for 1990 for a six year window assesses the trend from 1987 to 1993).
 Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.

### 2021-01-27 manu_draft.Rmd
back trans section added

@@
 Accordingly, even though the models differ by which structural component describes variation in the chlorophyll time series, they provide similar predictions.
+
+For all results, model S was used with enough knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.
+This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
 
@@
 
+### Back-transformation of model results
+
+Model results were back-transformed from log$_{10}$-space to aid in the interpretation of trends.
+Back-transformation was accomplished using a bias correction factor for model predictions, estimates of mean values, and endpoints of confidence intervals from GAM results.
+Using a bias correction factor correctly back-transforms the values to arithmetic space, whereas a simple exponentiation of the results produces under-estimates at the scale of the geometric mean [@Bradu70;@Duan83].
+The following equation was used for back-transformation, such that:
+
+\begin{equation}
+E[y] = 10^{\mu + \frac{\log(10) \sigma^2}{2}}
+(\#eq:backtrans)
+\end{equation}
+
+where the back-transformed, expected value of the response variable $y$ (chlorophyll) is a function of the mean value $\mu$ in log$_{10}$-space, corrected by a dispersion estimate $\sigma^2$.
+The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals, corrected by the model degrees of freedom.
+
+Slope estimates that were used to determine significance of a seasonal trend across years from the mixed-meta analysis were not back-transformed.
+While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.
+However, an "approximate" linear slope was estimated for the mixed-meta regression models as the linear fit across the seasonal averages that were back-transformed using equation \@ref(eq:backtrans).
+This was done solely for interpretation of relative change across years, whereas the significance of the slope parameter in log$_{10}$-space from the mixed-meta model was used to assess significance of the trend.
+
 # Results
@@
 
-For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.
-This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
-Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
-GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)).
-All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, followed by a decreasing trend until the end of the record.
-Seasonally, all models indicated a larger spring peak in chlorophyll, particularly for more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).
+Model predictions for chlorophyll trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
+GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)).
+All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at south Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.
+The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).
 

### 2021-01-28 manu_draft.Rmd
trnddat fig only shows seasonal changes

@@
     reference_docx: my_styles.docx
+    number_sections: false
 always_allow_html: true
@@
 data(seastrnd2)
-data(chgtrnd)
 data(cmptrnd)
@@
 
-Model results were back-transformed from log$_{10}$-space to aid in the interpretation of trends.
-Back-transformation was accomplished using a bias correction factor for model predictions, estimates of mean values, and endpoints of confidence intervals from GAM results.
-Using a bias correction factor correctly back-transforms the values to arithmetic space, whereas a simple exponentiation of the results produces under-estimates at the scale of the geometric mean [@Bradu70;@Duan83].
-The following equation was used for back-transformation, such that:
+Model results were back-transformed from log$_{10}$-space to aid in the interpretation of trends [@Bradu70;@Duan83].
+Back-transformation was accomplished using equation \@ref(eq:backtrans) for model predictions, estimates of mean values, and endpoints of confidence intervals from GAM results, such that:
 
@@
 
-where the back-transformed, expected value of the response variable $y$ (chlorophyll) is a function of the mean value $\mu$ in log$_{10}$-space, corrected by a dispersion estimate $\sigma^2$.
-The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals, corrected by the model degrees of freedom.
+where the back-transformed, expected value of the response variable $y$ (chlorophyll) is a function of the mean value $\mu$ in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.
+The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals divided by the model degrees of freedom.
 
@@
 While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.
-However, an "approximate" linear slope was estimated for the mixed-meta regression models as the linear fit across the seasonal averages that were back-transformed using equation \@ref(eq:backtrans).
-This was done solely for interpretation of relative change across years, whereas the significance of the slope parameter in log$_{10}$-space from the mixed-meta model was used to assess significance of the trend.
+As such, all results reporting slope estimates as rates of change in chlorophyll per year were reported at the scale of the model in log$_{10}$-space.
 
@@
 
-chg1 <- anlz_perchg(mod, baseyr = 1991, testy = 2000)
-chg2 <- anlz_perchg(mod, baseyr = 2000, testy = 2010)
-chg3 <- anlz_perchg(mod, baseyr = 2010, testy = 2019)
-chg <- bind_rows(chg1, chg2, chg3) %>% 
-  mutate_if(is.numeric, round, 1) %>% 
-  mutate(
-    pval = p_ast(pval)
-  )
-
 seas <- tibble(
@@
 
-      dispersion <- summary(mod)$dispersion
-      bt_prd <- 10 ^ (predict(mixmet) + log(10) * dispersion / 2)
-      df <- data.frame(chl = bt_prd, yr = mixmet$model$yr)
-      slope <- lm(chl ~ yr, df) %>% summary %>% coefficients %>% .[2, 1]
+      slope <- summary(mixmet)$coefficients[2, c(1, 5, 6)]
+      slope <- round(slope, 2)
+      slope <- paste0(slope[1], ' (', slope[2], ', ', slope[3], ')')
 
@@
   mutate(
-    lets = c('(d)', '(e)', '(f)', '(g)', '(h)', '(i)'), 
+    lets = c('(a)', '(b)', '(c)', '(d)', '(e)', '(f)'), 
     doystr = ifelse(doystr == 41, 'Jan', 'Jul'), 
     doyend = ifelse(doyend == 213, 'Jun', 'Dec'), 
-    est = paste('slope', round(est, 2)),
     pval = p_ast(pval)
@@
 
-p1 <- show_perchg(mod, baseyr = 1991, testyr = 2000, ylab = ylab) +
-  labs(
-    subtitle = paste0('(a) 1991: ', chg[[1, 1]],' ug/L, 2000: ', chg[[1, 2]], 'ug/L\nChange: ', chg[[1,3]], '%, ', chg[[1,4]]), 
-    title = NULL
-  )
-p2 <- show_perchg(mod, baseyr = 2000, testyr = 2010, ylab = NULL) +
-  labs(
-    subtitle = paste0('(a) 2000: ', chg[[2, 1]],' ug/L, 2010: ', chg[[2, 2]], 'ug/L\nChange: ', chg[[2, 3]], '%, ', chg[[2, 4]]), 
-    title = NULL
-  )
-p3 <- show_perchg(mod, baseyr = 2010, testyr = 2019, ylab = NULL) +
-  labs(
-    subtitle = paste0('(a) 2010: ', chg[[3, 1]],' ug/L, 2019: ', chg[[3,2]], 'ug/L\nChange: ', chg[[3, 3]], '%, ', chg[[3, 4]]), 
-    title = NULL
-  )
-p4 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
+p1 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p5 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
+p2 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p6 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
+p3 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
   coord_cartesian(ylim = ylim) +  
@@
   )
-p7 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
+p4 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p8 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
+p5 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p9 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
+p6 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
   coord_cartesian(ylim = ylim) + 
@@
 
-p <- p1 + p2 + p3 + p4 + p5 + p6 + p7 + p8 + p9 + plot_layout(nrow = 3) & 
+p <- p1 + p2 + p3 + p4 + p5 + p6 + plot_layout(nrow = 2) & 
   theme_minimal(base_family = 'serif', base_size = 12) &
@@
 
-png('figs/trnddat.png', height = 9, width = 11, family = 'serif', units = 'in', res = 400)
+png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
 print(p)
@@
 ```
-```{r trnddat, fig.cap = "Examples of trend tests using results of GAM predictions for station 34.
-Plots (a), (b), and (c) show estimates of percent change as the difference of means between year pairs for chlorophyll.
-Plots (d) through (i) show seasonal averages and trend estimates over time.
-Plot (d), (e), and (f) show trend estimates for January through June and (g), (h), and (i) show trend estimates for July through December.
-The trend lines in (d) through (i) estimate the rate of change of chlorophyll per year, reported as the slope in the plot title.
+```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates using results of GAM predictions for station 34.
+Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.
+The trend lines estimate the rate of change of chlorophyll per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval)in the plot title.
 ns: not significant, * p < 0.05, ** p < 0.005"}

### 2021-01-28 R/dat_proc.R
chgtrnd data object and workflow deleted

@@ -219,57 +219,6 @@ seastrnd2 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
 
 save(seastrnd2, file = 'data/seastrnd2.RData', compress = 'xz')
 
-# decadal percent changes -------------------------------------------------
-
-chgtrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
-  crossing(
-    fl = ., 
-    tibble(
-      yrstr = c(1991, 2000, 2010),
-      yrend = c(2000, 2010, 2019)
-    )
-  ) %>% 
-  group_by(fl) %>% 
-  nest() %>% 
-  mutate(
-    mod = purrr::map(fl, function(x){
-      
-      load(file = x)
-      
-      nm <- basename(x)
-      nm <- gsub('\\.RData', '', nm)
-
-      out <- get(nm) %>%  
-        pull(model) %>% 
-        deframe()
-      
-      return(out)
-      
-    })
-  ) %>% 
-  unnest(c('data')) %>% 
-  mutate(
-    perchg = purrr::pmap(list(mod = mod, baseyr = yrstr, testyr = yrend), anlz_perchg)
-  ) %>% 
-  select(-mod) %>% 
-  unnest('perchg') %>% 
-  ungroup %>% 
-  select(station = fl, yrs = yrstr, perchg, pval) %>% 
-  mutate(
-    station = gsub('^data/modslog\\_chl|\\.RData$', '', station),
-    yrs = case_when(
-      yrs < 1995 ~ '1990, 2000', 
-      yrs >=1995 & yrs < 2005 ~ '2000, 2010', 
-      yrs >= 2005 ~ '2010, 2019'
-    ), 
-    yrs = factor(yrs), 
-    persgn = sign(perchg), 
-    persgn = factor(persgn, levels = c('1', '-1'), labels = c('inc', 'dec')), 
-    pval = ifelse(pval < 0.05, 'sig', 'ns')
-  ) 
-
-save(chgtrnd, file = 'data/chgtrnd.RData', compress = 'xz')
-
 # trend model comparisons -------------------------------------------------
 
 cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 

### 2021-02-02 manu_draft.Rmd
chgtrnd data object and workflow deleted
do not cache

@@
 # figure path, chunk options
-knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = T, dev.args = list(family = 'serif'), dpi = 300, warning = F, cache.path = 'manu_draft_cache/',
+knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F,
   fig.process = function(x) {
@@
 
-leglab2 <- expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+leglab <- expression(paste(log[10], "Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
 
@@
 
-toplo1 <- chgtrnd %>% 
-  mutate(
-    station = as.numeric(station),
-    pval = ifelse(pval == 'sig', '*', '')
-    ) %>% 
-  left_join(locs, by = 'station')
-
-toplo2 <- seastrnd %>%
+toplo <- seastrnd %>%
   mutate(
@@
 
-p1 <- ggmap(bsmap1_transparent) +
-  geom_point(data = toplo1, aes(x = lon, y = lat, size = abs(perchg), shape = persgn, fill = perchg, colour = pval)) + #, color = 'black') +
-  geom_text(data = toplo1, aes(x = lon, y = lat, label = pval), nudge_x = 0.04) + 
-  facet_grid( ~ yrs) + 
-  scale_fill_gradient2('Chl-a % change', low = 'green', mid = 'white',  high = 'tomato1', midpoint = 0) +
+p <- ggmap(bsmap1_transparent) +
+  geom_point(data = toplo, aes(x = lon, y = lat, size = abs(yrcoef), shape = coefsgn, fill = yrcoef, colour = pval)) + #, color = 'black') +
+  geom_text(data = toplo, aes(x = lon, y = lat, label = pval), nudge_x = 0.04) + 
+  facet_grid(seas ~ yrs) + 
+  scale_fill_gradient2(leglab, low = 'green', mid = 'white',  high = 'tomato1', midpoint = 0) +
   scale_colour_manual(values = c('black', 'black'), guide = F) +
@@
   scale_size(range = c(1, 6), guide = F) +
-  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 8)) +
-  labs(
-    subtitle = '(a) Percent change between year pairs'
-  )
-
-p2 <- ggmap(bsmap1_transparent) +
-  geom_point(data = toplo2, aes(x = lon, y = lat, size = abs(yrcoef), shape = coefsgn, fill = yrcoef, colour = pval)) + #, color = 'black') +
-  geom_text(data = toplo2, aes(x = lon, y = lat, label = pval), nudge_x = 0.04) + 
-  facet_grid(seas ~ yrs) + 
-  scale_fill_gradient2(leglab2, low = 'green', mid = 'white',  high = 'tomato1', midpoint = 0) +
-  scale_colour_manual(values = c('black', 'black'), guide = F) +
-  coord_map() + 
-  scale_shape_manual('Trend', values = c(24, 25)) + 
-  pthm +
-  scale_size(range = c(1, 6), guide = F) +
-  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 8)) +
-  labs(
-    subtitle = '(b) Annual rates of change across seasons by decade'
-  )
-
-p <- p1 + p2 + plot_layout(ncol = 1, heights = c(0.5, 1))
-
-png('figs/trndmap.png', height = 10, width = 6.5, family = 'serif', units = 'in', res = 400)
+  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 8))
+
+png('figs/trndmap.png', height = 5, width = 6.5, family = 'serif', units = 'in', res = 400)
 p

### 2021-02-02 manu_draft.Rmd
figures and captions updated for log slope

@@
 Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.
-The trend lines estimate the rate of change of chlorophyll per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval)in the plot title.
+The trend lines estimate the rate of change of chlorophyll per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title.
 ns: not significant, * p < 0.05, ** p < 0.005"}
@@
 
-leglab <- expression(paste(log[10], "Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+leglab <- expression(paste(log[10], " Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
 
@@
     legend.position = 'top',
+    legend.box = 'vertical', 
     strip.background = element_blank(),
@@
 
-png('figs/trndmap.png', height = 5, width = 6.5, family = 'serif', units = 'in', res = 400)
+png('figs/trndmap.png', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 400)
 p
@@
 ```
-```{r trndmap, fig.cap= "Trend estimates (a) between year pairs and (b) across seasons by decade for chlorophyll at each station.
-The top plots show percent change comparing the first and last year within a decade and the bottom plots show seasonal estimates of change per year for chlorophyll concentrations for each decade.
-Point type, shape, and color represent the direction and magnitude of an estimated trend.
+```{r trndmap, fig.cap= "Trend estimates across seasons by decade for chlorophyll at each station.
+Point type, shape, and color represent the direction and magnitude of an estimated trend as the log$_{10}$ slope for chlorophyll concentration per year.
 Trends with $p<0.05$ are marked with an asterisk.
@@
 p <- ggplot(data = toplo, aes(x = yr, y = yrcoef, fill = pval)) + 
+  geom_errorbar(ggplot2::aes(ymin = yrcoef_lwr, ymax = yrcoef_upr, color = pval), width = 0) +
   geom_point(shape = 21, size = 3) +
   scale_x_continuous(limits = c(1993, 2016), breaks = seq(1993, 2016)) +
-  scale_y_continuous(limits = c(-2, 2)) +
-  scale_fill_manual(values = c(NA, 'tomato1'), drop = FALSE) +
+  # scale_y_continuous(limits = c(-2, 2)) +
+  scale_color_manual(values = c('black', 'tomato1'), drop = FALSE) +
+  scale_fill_manual(values = c('white', 'tomato1'), drop = FALSE) +
   geom_hline(yintercept = 0) + 
@@
   labs(
-    y = expression(paste('Chlorophyll change ( ', mu, 'g ', L^-1, yr^-1, ')'))
+    y = expression(paste(log[10], ' Chl-a change ( ', mu, 'g ', L^-1, yr^-1, ')'))
   )
@@
 ```
-```{r winchg, fig.cap='Estimates of chlorophyll change per year from applying the mixed-meta analysis across the time series for each station.
+```{r winchg, fig.cap='Estimates of log$_{10}$-chlorophyll change per year from applying the mixed-meta analysis across the time series for each station.
 Stations are arranged top to bottom from north to south.
@@
     y = 'Station (north to south)', 
-    x = expression(paste("Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+    x = expression(paste(log[10], " Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
   )
@@
 The "observed" trend model is based on a linear fit to the annual averages of chlorophyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
-Values for each model are the slope estimates as annual change per year within each season, with color denoting significant trends.'}
+Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting significant trends.'}
 knitr::include_graphics('figs/trndcmp.png')

### 2021-02-02 manu_draft.Rmd
methods revision from JH done

@@
 
-MOdel SYD adds an average within-year cyclic pattern as a separate spline:
+Model SYD adds an average within-year cyclic pattern as a separate spline:
 
@@
 Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
-For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
+For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
 The number of spline basis functions is related to the number of knots.
@@
 end of a time-series.
-For example, the above estimates were used for a test of the null hypothesis that `y` is the same at two times.
-To do so, $\bX$ would have two rows for each time to compare.
-The vector $A^T$ would be [-1, 1], so that $A^T \bX$ gives the point estimate of the difference in chlorophyll between those times, and $A^T \bX \hShb \bX^T A$ gives the squared standard error of that difference.
-This difference was used to determine if the two estimates in time were significantly different from 0, e.g., is Chla-a different now than it was before based on GAM estimates.
-An alternative interpretation of this test is a quantification of differences between two points in time as affected by the magnitude of the difference and range of likely values for the magnitude of change.
 
@@
 
-As a final analysis, the chosen sequence of years for evaluating a trend for a seasonal average can be applied across the time series to assess periods of time within which trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
+### Trend comparisons
+
+The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year blocks from 1991 to 2000, 2000 to 2010, and 2010 to 2019.
+As an extension of these methods, a chosen sequence of years was applied across the time series to assess periods of time within which seasonal trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
 Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.
 The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.
-We chose a centered window where the model estimates are based on results at equal half-window widths to the left and right of a given year (e.g., an estimate for 1990 for a six year window assesses the trend from 1987 to 1993).
+We chose a centered five-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
 Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.
+
+Finally, trend results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
+As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
+Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
+Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.
+Trend estimates from mixed-meta regression applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averaged.
+The comparisons were demonstrated with select examples where differences were pronounced and then applied to all stations.
 
@@
 GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)).
-All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at south Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.
+All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at South Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.
 The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).
@@
 
-Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December for the same year pairs (Figure \@ref(fig:trnddat)).
-This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.
-Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268.4% increase, although the change is insignificant because of large uncertainties in the chlorophyll averages in the first year.
-However, the change from 2000 to 2010 was a significant increase of 64%, followed by a significant decrease of 35% from 2010 to 2019.
-Although these trends were insignificant, the results are based on a priori decisions of which years to compare.
-
-An assessment of the annual differences indicates if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences.
-Accordingly, figures \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to evaluate changes across the same year pairs but using seasonal averages from GAM results.
-Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to June.
-Similarly, rows g-i show the seasonal averages and trend estimates from July to December.
-The seasonal trend results suggested that a significant increase in chlorophyll was observed from 1991 to 2000 but only for the January to June period (slope 0.58 ug/L chlorophyll change per year).
-Similarly, A significant increase was observed from 2000 to 2010 but only for the July to December period (slope 0.58)  Finally, a significant decrease was observed from 2010 to 2019 but only for the July to December period (slope -0.28).
-As a result, an assessment of trend changes by season provides a more precise assessment of trends over time and accounts for full uncertainty in estimates across model results.
-
-Spatial variation in trends suggested that most stations in South Bay had increasing chlorophyll from 1990 to 2010 following a decrease in recent years, although not all trends were significant and varied by location (Figure \@ref(fig:trndmap)a).
-Comparing years 1990 and 2000, stations 21, 24, 27, and 30 had increasing significant trends.
-The magnitude of the chlorophyll increase in the southernmost stations (34, 36) was higher during this same period, although the trends were not significant.
-Comparing 2000 and 2010, significant increases were observed at the three northern-most stations (18, 21, and 22) and the three southern-most stations (32, 34, 36).
-Comparing 2010 and 2019, the three southern-most stations had significant decreases in annual chlorophyll, whereas station 24 had a significant increase during this period.
-The spatial trends were further evaluated with mixed-meta regression analysis to provide additional context to the annual changes by evaluating trends from January to June and July to December (Figure \@ref(fig:trndmap)b).
-For the January to June period, significant increases were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas a significant decrease was observed at stations 30 and 32 from 2010 to 2019.
-For the July to December period, significant increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas significant decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
-
-A final analysis evaluated a moving window comparison of seasonal trends to provide additional context of when significant changes are occurring at each station (Figure \@ref(fig:winchg)).
-Results were obtained using a centered five year window to estimate the slope and significance of a seasonal trend of GAM results from January to June and July to December for each five-year period at each station (note that estimates at the far left and far right of each time series cannot be calculated).
-As before, larger trends were generally observed with the more southern stations.
-Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until recent years when trends were relatively stable.
-Most stations showed a shift around 2007 when trends switched from significantly increasing to significantly decreasing, particularly in the January to June period.
+To demonstrate the differences in results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
+All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.
+The seasonal trend analyses showed that January to June chlorophyll increased (log$_{10}-Chla slope 0.03 $\mu$g L$^{-1} yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.
+Chlorophyll also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
+Finally, chlorophyll decreased from 2010 to 2019 but only for July to December (log$_{10}-Chla slope -0.02, -0.04-0 95\% confidence interval).
+Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.
+
+Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.
+Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.
+Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chlorophyll decreased for at stations 30 and 32 from 2010 to 2019.
+For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
 
@@
 
-Trends results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
-As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
-Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
-Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.
-
-These distinctions are demonstrated in Figure \@ref(fig:trndcmpex) for two stations and two time periods, where the significance of trends are compared based on regression through seasonal averages (Figure \@ref(fig:trndcmpex) row 1), regression through GAM estimates (row 2), and mixed-meta regression through GAM estimates (row 3).
-The stations and time periods were selected to demonstrate examples where different conclusions can be obtained depending on the methods and characteristics of the data.
-For example, Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
-Only the mixed-meta regression results are significant in this example.
-The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a significant trend.
+Results from a five-year moving window comparison of seasonal trends provided additional context on when significant changes were occurring at each station (Figure \@ref(fig:winchg)).
+The largest trends occurred at southern stations where increases occurred in both halves of the year (i.e., January to June, July to December).
+Decreases were observed later in the record, until recent years when trends were  Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until chlorophyll was relatively stable across all stations and seasons.
+A change from increasing to decreasing chlorophyll occurred at most stations around 2007, particularly for January to June.
+
+Results showing trend estimates from mixed-meta regression on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2).
+Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
+Only the mixed-meta regression results shows a trend in this example (based on $p < 0.05$).
+The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a trend.
 Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.
-Unlike the first example, the top two figures show significant trend estimates, whereas the bottom plot for the mixed-meta regression analysis does not show a significant trend because of added uncertainty in the averages provided by the GAMs.
+Unlike the first example, the top two figures show trends, whereas the bottom plot for the mixed-meta regression analysis does not show a trend because of added uncertainty in the averages provided by the GAMs.
 In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.
 
-Figure \@ref(fig:trndcmp) expands the results in Figure \@ref(fig:trndcmpex) by applying the method comparisons to all stations.
-As before, the different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade.
-The slope estimates from the linear models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.
-Slope estimates from the linear models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.
-Differences in significance of trends between the linear models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
+Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
+The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.
+Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.
+Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
 

### 2021-02-03 manu_draft.Rmd
discussion revision from JH done

@@
 
+### wqtrends R package
+
+The wqtrends R package developed by the authors includes all methods described above.
+The package is available for download at https://tbep-tech.github.io/wqtrends.
+A full vignette describing installation and use is also available in the link provided.
+
 # Results
@@
 
-The approach developed here can be used to estimate seasonal averages with uncertainty in water quality parameters and evaluate trends in these seasonal changes across years using full propagation of error estimates across methods.
-This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an "optimal" level of smoothing with the mgcv R package that uses generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
-Importantly, specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.
-A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risk inflated Type I error rates.
-We further demonstrated that models with different structures in the smoothed components, such as separate smoothers for continuous year and day of year, can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
-This comparison demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time.
-Although models with different structures allocate variance components of the response variable differently, this is a trivial issue for the second and third stage of analyses to estimate significance of trends developed herein.
-
 The use of the GAM results with second and third stage analyses to assess trends is a new approach that previous trend analysis methods cannot sufficiently address.
@@
 
+Our results demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time.
+This is an important conclusion that can provide guidance on how GAMs should be used to model time series from long-term water quality monitoring data.
+Models with separate smoothers for continuous year and day of year can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
+This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing with the mgcv R package using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
+Specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.
+A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risks incorrect conclusions from using these models with the second and third stage of analyses.
+
 Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
-These comparisons are analogous to the example in the previous paragraph where a common approach prior to trend analysis is to aggregate raw observations.
-Our assessment of trends using linear regression applied to seasonal averages from the raw observations is effectively similar to averaging results within a year and applying a simple Kendall test.
+These comparisons demonstrate a common approach where raw observations may be aggregated prior to use of more conventional trend analysis methods.
+Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
 In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.
-This is particularly problematic if sampling effort varies substantially between years.
-Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months (e.g., summer only)[@Fouquet12].
+A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the mixed-meta regression results.
+Conversely, a false positive result (type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, whereas the mixed-meta results did not.
+The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).
+
+Variability of sampling effort among years is a common characteristic of environmental monitoring data, which further increases the potential for incorrect conclusions when existing methods have been used.
+Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months [e.g., summer only, @Fouquet12].
 The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
 More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].
-As a result, trend assessments from mixed-meta regression should be considered an accurate representation of the actual trend within the limits of the raw observations.
+As a result, trend assessments from mixed-meta regression are accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.
 
@@
 For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].
-Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, leading to potential under- or over-smoothing of fitted results to the raw observations.
+Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, producing fitted results that are arbitraily under- or over-smoothed to the raw observations.
+This is especially problematic when for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.
 Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.
-However, we acknowledge that the third stage analyses that require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
+
+## Future work
+
+Additional work could be conducted to further strengthen the validity of conclusions based on trends from mixed-meta regression analyses applied to the seasonal GAM averages.
+We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
 Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.
@@
 
-The trend assessment approach also does not address causality in the observed trends.
-In this regard, the approach is similar to other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time.
+The trend assessment approach does not address causality in the observed trends.
+In this regard, the approach is like other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time.
 However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.
@@
 
-Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a trivial but logical extension of the methods proposed in this study.
+Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a simple and logical extension of the methods proposed in this study.
 The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20].
@@
 Chlorophyll measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
-We provide an open-source R package to facilitate these analyses (see supplement) that includes all methods described in this manuscript.
+The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
 
 # Supplement
-
-The wqtrends R package includes all methods described in this manuscript.
-The package is available for download at https://tbep-tech.github.io/wqtrends.
-A full vignette describing installation and use is also available in the link provided.
-Please cite this manuscript in applications using the R package.
 
@@
 This work benefited from discussions with the San Francisco Bay Nutrient Technical Workgroup and Steering Committee.
+We thank James D.
+Hagy III for reviewing an earlier draft of this manuscript.
 
@@
       title = ttl,
-      subtitle = paste('Obesrved regression on raw data,', pval1),
+      subtitle = paste('OLS regression on raw data,', pval1),
       y = NULL
@@
     labs(
-      subtitle = paste('Average regression on GAM estimates,', pval2),
+      subtitle = paste('OLS regression on GAM estimates,', pval2),
       y = ylab
@@
 ```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at two example stations.
-The "observed" trend model is based on a linear fit to the annual averages of chlorophyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
-Regressions in each plot are fit through the seasonal estimates indicated in the plot titles.'}
+The first row shows OLS (ordinary least squares) regression applied to annual averages of chlorophyll from the observed data, the second row shows OLS regression applied to annual averages of chlorophyll from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chlorophyll from the GAM.
+Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
 knitr::include_graphics('figs/trndcmpex.png')

### 2021-02-03 manu_draft.Rmd
sitemap has better inset

@@
 library(USAboundaries)
+library(rworldmap)
 library(hrbrthemes)
@@
 ```{r, results = 'hide', message = F}
+
 # station locations
@@
   filter(name %in% 'California')
-locbuff <- st_buffer(sflocs, dist = 0.1)
+continent <- getMap(resolution = 'low') %>% 
+  st_as_sf %>% 
+  filter(continent %in% c('North America', 'South America')) %>% 
+  st_transform(crs = st_crs(states)) %>% 
+  st_buffer(dist = 0) %>% 
+  st_union()
+locbuff <- st_buffer(sflocs, dist = 10)
 locbuff2 <- st_buffer(sflocs, dist = 0.07)
@@
   st_as_sfc(crs = 4326)
+statebuff <- st_buffer(states, dist = 0.1)
+insetylim <- st_bbox(statebuff)[c('ymin', 'ymax')]
+insetxlim <- st_bbox(statebuff)[c('xmin', 'xmax')]
 
@@
 
-# inset
 p2 <- ggplot() + 
-  geom_sf(data = states, fill = 'white', colour = 'black') + 
-  geom_sf(data = castate, fill = 'grey', colour = 'black') + 
+  geom_sf(data = continent, fill = 'grey', colour = 'grey') + 
+  # geom_sf(data = states, fill = 'grey', colour = 'grey') + 
   geom_sf(data = insetbb, fill = NA, color = 'blue', size = 2) +
-  theme_void()
+  coord_sf(ylim = insetylim, xlim = insetxlim) + 
+  theme_void() +
+  theme( 
+    panel.background = element_rect(fill = '#FFFFFF99', colour = 'white'), 
+    panel.border = element_rect(colour = 'black', fill = 'transparent')
+  )
 
@@
 p <- p1 + 
-  inset(ggplotGrob(p2), xmin = -122.3, xmax = -121.97, ymin = 37.75, ymax = 37.95)
+  inset(ggplotGrob(p2), xmin = -122.2, xmax = -121.97, ymin = 37.75, ymax = 37.983)
+p
 

### 2021-02-03 manu_draft.Rmd
flextable hack for gh action build

@@
   left_join(locs, by = 'station') %>% 
-  spread(var, val)
+  spread(var, val) %>% 
+  mutate_at(c('chl_medv', 'chl_minv', 'chl_maxv'), round, 1) %>% 
+  mutate_at(c('lat', 'lon'), round, 3)
 
@@
   # merge_at(i = 1, j = 4:7, part = 'header') %>% 
-  colformat_num(j = c(5, 6, 7), digits = 1) %>%
-  colformat_num(j = c(2, 3), digits = 3) %>% 
+  # colformat_num(j = c(5, 6, 7), digits = 1) %>%
+  # colformat_num(j = c(2, 3), digits = 3) %>% 
   align(align = 'right', part = 'header') %>% 

### 2021-02-03 manu_draft.Rmd
fixing build maybe i hope

@@
   st_union()
-locbuff <- st_buffer(sflocs, dist = 10)
+locbuff <- st_buffer(sflocs, dist = 0.1)
 locbuff2 <- st_buffer(sflocs, dist = 0.07)

### 2021-02-03 manu_draft.Rmd
trying to fix build for orphans that arent orphans

--- 2021-02-03 manu_draft.Rmd
@@
 These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.
-Our proposed methods estimate seasonal averages in a response variable with GAMs and then use those ancertainty with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.
-We first demonstrate that nearly identical descriptions of temporal changes in l can be obtained using different smoothing splines for annual or seasonal components of the time series.
+Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.
+We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.
 We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.
-Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluates changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
+Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluate changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
 Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.
@@
 At short temporal scales typically less than a year, environmental systems exhibit variability caused by multiple factors (e.g., weather events, management, or seasonal changes).
-Such fluctuations may not be of direct interest for inter-annual trends or may not be well-suited to multi-scale smoothing methods.
-In this paper, we describe methods to estimate across-year trends of within-year features of interest such as a seasonal average, seasonal peak, or seasonal timing of events, while accounting for uncertainties across analysis steps.
-
-Existing methods that address our objectives in water quality trend analysis can be generalized into four basic approaches: seasonal Kendall tests, seasonal trend decomposition (STL), weighted regression on time, discharge, and season (WRTDS), and generalized additive models (GAMs).
+Such fluctuations may not be related to inter-annual trends or may not be well-suited to multi-scale smoothing methods.
+In this paper, we develop methods to estimate across-year trends of within-year features, such as a seasonal averages, while accounting for uncertainties across analysis steps.
+
+Existing methods that begin to address our objectives in water quality trend analysis can be generalized into four basic approaches: seasonal Kendall tests, seasonal trend decomposition (STL), weighted regression on time, discharge, and season (WRTDS), and generalized additive models (GAMs).
 Seasonal Kendall tests or related non-parametric approaches have been used for decades in water quality trend assessments to identify monotonic changes over several years while accounting for the predictable patterns among seasons [@Hirsch82;@Helsel20].
 @Wan17 showed that non-parametric approaches have been the most commonly used methods in long-term water quality trend analysis despite critical limitations.
-For descriptive decomposition of long-term monitoring data, they assume seasonal patterns within year do not change, require regularly spaced or balanced data, only use time as a descriptor for trends, and do not estimate a model that could be useful for other purposes.
+For descriptive decomposition of long-term monitoring data, they assume seasonal patterns within years do not change, require regularly spaced or balanced data, only use time as a descriptor for trends, and do not estimate a model that could be useful for other purposes.
 Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.
@@
 For example, @Murphy19 used GAMs to decompose water quality time series from Chesapeake Bay into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.
-This study generalizes the approach to analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when they are used as designed [@Wood03;@Wood17] rather than for ad hoc separation of time scales, and prioritizes full incorporation of uncertainty.
-Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17;@Testa18].
+The study herein generalizes the approach to analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when spline flexibility is allowed to vary [@Wood03;@Wood17] rather than being constrained *a priori* for different time scales, and prioritizes full incorporation of uncertainty.
+Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17].
 
@@
 The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common to time series data.
-This paper develops GAMs as a distinct application that can meet these requirements and simultaneously addressing key limitations of seasonal Kendall tests and the more complex STL and WRTDS methods.
+This paper develops GAMs as a distinct application that can meet these requirements while addressing key limitations of seasonal Kendall tests and the more complex STL and WRTDS methods.
 We do so with the explicit goal of quantifying interannual trends in seasonal averages.
 
-We describe and demonstrate the proposed methods by analyzing water quality monitoring data from the southern portion of San Francisco Estuary, California, USA.
+We describe and demonstrate the proposed methods by analyzing water quality monitoring data from the southern portion of the San Francisco Estuary, California, USA.
 Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.
 Analysis of these data is complicated by irregularities in timing and consistency of data collection which can generate artifacts affecting simple seasonal averages of the data.
-We were interested in questions such as: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?
-At what across-year scale do within-year summer-fall mean chlorophyll levels change?
-Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or baseline levels of productivity between blooms?
+We were interested in questions such as: Are there significant trends in spring mean chlorophyll at multi-year time-scales?
+At what across-year scale do summer-fall mean chlorophyll levels change?
+Are there significant across-year trends in the spring phytoplankton bloom or baseline chlorophyll concentrations during periods of low productivity between blooms?
 We provide examples illustrating how these questions can be addressed using methods with GAMs.
@@
 This approach is new to environmental trend-detection problems.
-For this step, we provide methods using meta-anlaysis for isolating seasonal trends with reasonalbe certainty from from GAM results and evaluating these trends between years.
+For this step, we provide methods using meta-analysis for isolating seasonal trends with reasonable certainty from GAM results and evaluating these trends between years.
 
@@
 Salinity ranges from 0 to 15 ppt in the northern subembayments and from 5 to 35 ppt in southern subembayments closer to the Pacific Ocean, depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
-The South Bay embayment is heavily urbanized and includes thirty-seven wastewater treatment plants that serve 7.2 million people.
 An estimated 73.8 metric tons dy$^{-1}$ of inorganic nitrogen are discharged into the Bay, primarily from wastewater [@Novick14].
-Secondary treatment occurs at most of the wastewater treatment plants that discharge into the SFE.
 Agricultural runoff from the upper watershed contributes 30 metric tons dy$^{-1}$ of nitrogen to the SFE via the Delta.
@@
 Nitrogen and phosphorus levels in SFE usually exceed concentrations that cause eutrophication in other estuaries.
-However, SFE has demonstrated resistance to eutrophication, which to high suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves.
-Interest in the potential for nutrient loading to negatively affect water quality has increased recently particularly for South Bay, where harmful algal blooms (HABs), elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations have occurred since 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
+However, SFE has demonstrated resistance to eutrophication, which has been attributed to high suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves [@Cole84;@Alpine88;@Jassby08;@Kimmerer14;@Lehman17].
+Renewed interest in understanding the potential for nutrient loading to negatively affect water quality has occurred for South Bay, where harmful algal blooms (HABs), elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
 Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient particularly with respect to seasonal differences between years.
 
-We evaluated near-surface chlorophyll-a data measured biweekly to monthly from 1990 to 2019 along the South Bay axis extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
+We evaluated near-surface chlorophyll-a data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
 Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
@@
 Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions.
-We developed two mixed effects meta-analysis approaches that provided 1) a simple comparison of whether seasonal features differ across years, and 2) estimation of short-term linear trends on time-scales chosen by the analyst.
+We developed two mixed effects meta-analysis approaches that provided 1) a simple comparison of whether seasonal features differ across years, and 2) an estimation of short-term linear trends on time scales chosen by the analyst.
 
@@
 
-GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor, indepdent of noise inherent in the raw data.
+GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor, independent of noise inherent in the raw data.
 The simplest GAM for this purpose is expressed as:
@@
 @Murphy19 used spline formulations related to those proposed here, but are insufficient for our needs of extracting and quantifying seasonal trends.
-@Murphy19 have a "`gam0`" with only a `s(doy)` term, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
+@Murphy19 have a "`gam0`" with only `s(doy)` and linear `cont_year` terms, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
 Their `gam0` is *a priori* not of interest for our data because of an assumption of constant seasonality and a linear annual trend.
@@
 Accordingly, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate intra-annual changes in season from noise at a finer temporal scale.
-In the "`gam2`" model (our SYDI model), @Murphy19 do not explicitly consider number of knots in the interaction spline.
-In both cases, they effectively use their choice of number of knots in different spline components as an ad hoc allocation of variation in the data to different components, while not allowing any component to fully estimate the signal in the data.
-@Murphy19 acknowledge that incomplete modeling of fluctuations in the data may inflate their Type I error rates for estimating temporal changes, but they leave that problem for future work.
-We seek to avoid inflating Type I error rates in this way.
+In the "`gam2`" model (our SYDI model), @Murphy19 do not explicitly consider the number of knots in the interaction spline.
+In both cases, they effectively use their choice of number of knots in different spline components as an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system, while noting that additional explanatory variables are necessary to explain the residual variability or autocorrelation in the data.
+@Murphy19 acknowledge that incomplete modeling of fluctuations in the data may inflate their Type I error rates for estimating temporal changes, but they account for that by incorporating additional hydrologic variables in a subsequent phase of their models in a step-wise manner.
+For the methods presented here, which focus first on fully identifying temporal change, we seek to avoid inflating Type I error rates by increasing the upper limit for the knots.
 Finally, @Murphy19 present large AIC differences between their spline formulations.
@@
 All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.
-The seasonal trend analyses showed that January to June chlorophyll increased (log$_{10}-Chla slope 0.03 $\mu$g L$^{-1} yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.
+The seasonal trend analyses showed that January to June chlorophyll increased (log$_{10}$ Chl-a slope 0.03 $\mu$g L$^{-1} yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.
 Chlorophyll also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
-Finally, chlorophyll decreased from 2010 to 2019 but only for July to December (log$_{10}-Chla slope -0.02, -0.04-0 95\% confidence interval).
+Finally, chlorophyll decreased from 2010 to 2019 but only for July to December (log$_{10} Chl-a slope -0.02, -0.04-0 95\% confidence interval).
 Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.
@@
 
-Our example in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
+Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
 These comparisons demonstrate a common approach where raw observations may be aggregated prior to use of more conventional trend analysis methods.
@@
 The trend assessment approach does not address causality in the observed trends.
-In this regard, the approach is like other trend assessment methods where the focus is understanding direction, magnitude, and confidence in changes in water quality variables over time.
+In this regard, the approach is like other trend assessment methods where the focus is on understanding direction, magnitude, and confidence in changes in water quality variables over time.
 However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.
@@
   st_as_sfc(crs = 4326)
-statebuff <- st_buffer(states, dist = 0.1)
+statebuff <- st_buffer(states, dist = 10)
 insetylim <- st_bbox(statebuff)[c('ymin', 'ymax')]
@@
   # geom_sf(data = states, fill = 'grey', colour = 'grey') + 
-  geom_sf(data = insetbb, fill = NA, color = 'blue', size = 2) +
+  geom_sf(data = insetbb, fill = NA, color = 'blue', size = 1.5) +
   coord_sf(ylim = insetylim, xlim = insetxlim) + 
@@
   inset(ggplotGrob(p2), xmin = -122.2, xmax = -121.97, ymin = 37.75, ymax = 37.983)
-p
 
@@
 The trend lines estimate the rate of change of chlorophyll per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title.
-ns: not significant, * p < 0.05, ** p < 0.005"}
+ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
 knitr::include_graphics('figs/trnddat.png')
@@
 ```
-```{r winchg, fig.cap='Estimates of log$_{10}$-chlorophyll change per year from applying the mixed-meta analysis across the time series for each station.
+```{r winchg, fig.cap='Estimates of log$_{10}$ chlorophyll change per year from applying the mixed-meta analysis across the time series for each station.
 Stations are arranged top to bottom from north to south.
@@
 
-```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chlorophyll.  Rows are arranged from north to south.'}
+```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chlorophyll ($\\mu$g L$^{-1}$) .  Rows are arranged from north to south.'}
 totab <- rawdat %>% 

### 2021-02-04 manu_draft.Rmd
working through methods for coauth comments

@@
 However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods.
-Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in South San Francisco Estuary.
+Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.
 GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.
@@
 
-We evaluated near-surface chlorophyll-a data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
+We evaluated near-surface chlorophyll (chl-a) data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
 Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
-Chlorophyll on GFF filters was analyzed fluorometrically after extraction in 90% acetone.
+Chl-a on GFF filters was analyzed fluorometrically after extraction in 90% acetone.
 Sampling frequency varied somewhat over time and by station.
@@
 
-where `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
-Log$_{10}$-transformed chlorophyll was used for all analyses to meet assumptions of normally-distributed residuals.
+where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
+Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals.
 
@@
 However, an interpretation of these imputations is challenging.
-For example, the spring chlorophyll peak is a notable feature every year in south SFE.
+For example, the spring chl-a peak is a notable feature every year in SFE.
 If the peak occurs at the same time every year but the magnitude varies, then the average within-year pattern can be interpreted as the average magnitude.
@@
 For example, if one year has an especially early and large magnitude spring bloom, the years before and after would be expected also to have earlier and larger than average spring blooms.
-For the raw SFE data, and chlorophyll-a dynamics in many estuaries, this is not expected to be the case.
+For the raw SFE data, and chl-a dynamics in many estuaries, this is not expected to be the case.
 An early and large spring bloom could easily be followed the next year by a late and relatively small spring bloom.
@@
 
-We visually compare chlorophyll estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).
-Models SY, SYD, and SYDI were fit to chlorophyll data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
+We visually compare chl-a estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).
+Models SY, SYD, and SYDI were fit to chl-a data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
 Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).
@@
 These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)).
-Accordingly, even though the models differ by which structural component describes variation in the chlorophyll time series, they provide similar predictions.
-
-For all results, model S was used with enough knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South SFE.
+Accordingly, even though the models differ by which structural component describes variation in the chl-a time series, they provide similar predictions.
+
+For all results, model S was used with enough knots in `num_knots_y` to evaluate chl-a trends across the monitoring stations in SFE.
 This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
@@
 
-In the second-stage analysis, we estimated a seasonal average, such as the mean spring chlorophyll concentrations and the uncertainty in each year.
-Define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
+In the second-stage analysis, we estimated a seasonal average, such as the mean spring chl-a concentrations and the uncertainty in each year.
+We define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
 The season includes $n$ days.
@@
 Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
-For example, using model SY, if a point estimate for chlorophyll is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ (and possibly its corresponding `doy` value) in the remaining columns.
+For example, using model SY, if a point estimate for chl-a is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ in the remaining columns.
 The number of spline basis functions is related to the number of knots.
@@
 
-In stage three of the analysis, we used a meta-analysis method to evaluate across-year trends of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
-In the examples, we considered two periods of interest for chlorophyll, January-July and August-December, which are relevant to phytoplankton bloom phenology in south SFE.
+In stage three of the analysis, we used a meta-analysis method to evaluate linear trends across years of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
+In the examples, we considered two periods of interest for chl-a, January-July and August-December, which are relevant to phytoplankton bloom phenology in SFE.
 
 This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?
-For example, is there a trend in the spring chlorophyll average from 1990 to 2000?
+For example, is there a trend in the spring chl-a average from 1990 to 2000?
 This question can also be posed in a moving-window manner across a time-series (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
@@
 $N$ is the number of years of the study.
-
-### Linear trends across years at a chosen time-scale
 
@@
 This method was applied to a chosen sequence or "window" of years for estimating the linear trend.
-This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chlorophyll over a chosen series of years.
+This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chl-a over a chosen series of years.
 Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).
@@
 The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.
-We chose a centered five-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
+We chose a centered ten-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
 Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.
@@
 
-where the back-transformed, expected value of the response variable $y$ (chlorophyll) is a function of the mean value $\mu$ in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.
+where the back-transformed, expected value of the response variable $y$ (chl-a) is a function of the mean value $\mu$ in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.
 The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals divided by the model degrees of freedom.
@@
 While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.
-As such, all results reporting slope estimates as rates of change in chlorophyll per year were reported at the scale of the model in log$_{10}$-space.
+As such, all results reporting slope estimates as rates of change in chl-a per year were reported at the scale of the model in log$_{10}$-space.
 
@@
 
-Model predictions for chlorophyll trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
-GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chlorophyll towards the more southern stations (Figure \@ref(fig:prddat)).
-All the models suggested 1) increasing chlorophyll from 1990 until 2005 to 2010, followed by decreasing chlorophyll until the end of the record in 2019, 2) a spring chlorophyll peak, particularly at South Bay stations, and 3) a fall chlorophyll peak that was smaller than the spring peak.
+Model predictions for chl-a trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
+GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figure \@ref(fig:prddat)).
+All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at South Bay stations, and 3) a fall chl-a peak that was smaller than the spring peak.
 The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).
@@
 All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.
-The seasonal trend analyses showed that January to June chlorophyll increased (log$_{10}$ Chl-a slope 0.03 $\mu$g L$^{-1} yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.
-Chlorophyll also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
-Finally, chlorophyll decreased from 2010 to 2019 but only for July to December (log$_{10} Chl-a slope -0.02, -0.04-0 95\% confidence interval).
+The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.
+Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
+Finally, chl-a decreased from 2010 to 2019 but only for July to December (log$_{10} chl-a slope -0.02, -0.04-0 95\% confidence interval).
 Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.
@@
 Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.
-Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chlorophyll decreased for at stations 30 and 32 from 2010 to 2019.
+Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased for at stations 30 and 32 from 2010 to 2019.
 For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
@@
 
-Results from a five-year moving window comparison of seasonal trends provided additional context on when significant changes were occurring at each station (Figure \@ref(fig:winchg)).
-The largest trends occurred at southern stations where increases occurred in both halves of the year (i.e., January to June, July to December).
-Decreases were observed later in the record, until recent years when trends were  Significant increases were observed for both January to June and July to December earlier in the period of record, whereas significant decreases were observed later in the record until chlorophyll was relatively stable across all stations and seasons.
-A change from increasing to decreasing chlorophyll occurred at most stations around 2007, particularly for January to June.
+Results from a ten-year moving window comparison of seasonal trends provided additional context on when significant changes were occurring at each station (Figure \@ref(fig:winchg)).
+Trends were observed at all stations that followed a general pattern of increases early in the record followed by decreases later in the record.
+Increases and decreases were observed in both the January to June and July to December seasonal periods, with some notable exceptions.
+In particular, the most southern stations (32, 34, 36) had more trends in the July to December period.
+Additionally, trends in recent years in both seasonal periods were not observed for the more northern stations.
+For both seasonal periods, a change from increasing to decreasing chl-a occurred at most stations around 2007.
 
@@
 
-Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chlorophyll changes in each decade (Figure \@ref(fig:trndcmp)).
+Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).
 The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.
@@
 As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.
-Our goal herein was to describe chlorophyll changes relative to time where the predictors were variations on a general theme (e.g., season vs.
+Our goal herein was to describe chl-a changes relative to time where the predictors were variations on a general theme (e.g., season vs.
 year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.
@@
 
-Finally, the evaluation of trends for alternative water quality variables in addition to chlorophyll is a simple and logical extension of the methods proposed in this study.
-The long-term monitoring program maintained by USGS includes multiple parameters in addition to chlorophyll that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20].
+Finally, the evaluation of trends for alternative water quality variables in addition to chl-a is a simple and logical extension of the methods proposed in this study.
+The long-term monitoring program maintained by USGS includes multiple parameters in addition to chl-a that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20].
 These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.
-Chlorophyll measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
+Chl-a measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
 The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
@@
 
-ylab <- expression(paste("Chlorophyll-a (", italic(mu), "g ", L^-1, ")"))
+ylab <- expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")"))
 
@@
 ```
-```{r obsdat, fig.cap = "Observed chlorophyll concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
+```{r obsdat, fig.cap = "Observed chl-a concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
 knitr::include_graphics('figs/obsdat.png')
@@
   labs(
-    title = '(b) Estimated chlorophyll between models'
+    title = '(b) Estimated chl-a between models'
   )
@@
 ```
-```{r modsumfig, fig.cap = "GAM output of estimated chlorophyll for models SY, SYD, and SYDI.
+```{r modsumfig, fig.cap = "GAM output of estimated chl-a for models SY, SYD, and SYDI.
 Model S is identical to SY and is not shown.
 Plots in (a) show model predictions by day of year with separate lines for each year.
-Plots in (b) show pairwise comparisons of predicted chlorophyll between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.
+Plots in (b) show pairwise comparisons of predicted chl-a between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.
 The plots demonstrate that results between the models are comparable except for a few observations at extreme values(a), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (b).
@@
   filter(param == 'chl')
-ylab <- 'Chlorophyll-a (ug/L)'
+ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))
 
@@
 ylim <- c(0, 21)
-ylab <- expression(paste("Chlorophyll-a (", mu, "g ", L^-1, ")"))
+ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))
 
@@
 Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.
-The trend lines estimate the rate of change of chlorophyll per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title.
+The trend lines estimate the rate of change of chl-a per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title.
 ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
@@
 
-leglab <- expression(paste(log[10], " Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+leglab <- expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
 
@@
 ```
-```{r trndmap, fig.cap= "Trend estimates across seasons by decade for chlorophyll at each station.
-Point type, shape, and color represent the direction and magnitude of an estimated trend as the log$_{10}$ slope for chlorophyll concentration per year.
+```{r trndmap, fig.cap= "Trend estimates across seasons by decade for chl-a at each station.
+Point type, shape, and color represent the direction and magnitude of an estimated trend as the log$_{10}$ slope for chl-a concentration per year.
 Trends with $p<0.05$ are marked with an asterisk.
@@
   ) %>% 
-  dplyr::filter(yr <= 2016 & yr >= 1993)
+  dplyr::filter(yr <= 2014 & yr >= 1996)
 
@@
   geom_point(shape = 21, size = 3) +
-  scale_x_continuous(limits = c(1993, 2016), breaks = seq(1993, 2016)) +
+  scale_x_continuous(limits = c(1996, 2014), breaks = seq(1996, 2014)) +
   # scale_y_continuous(limits = c(-2, 2)) +
@@
   labs(
-    y = expression(paste(log[10], ' Chl-a change ( ', mu, 'g ', L^-1, yr^-1, ')'))
+    y = expression(paste(log[10], ' chl-a change ( ', mu, 'g ', L^-1, yr^-1, ')'))
   )
@@
 ```
-```{r winchg, fig.cap='Estimates of log$_{10}$ chlorophyll change per year from applying the mixed-meta analysis across the time series for each station.
+```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year from applying the mixed-meta analysis across the time series for each station.
 Stations are arranged top to bottom from north to south.
 Plots in (a) show estimates for seasonal averages from January to June and plots in (b) show estimates for seasonal averages from July to December.
+Results are from a ten-year, centered moving window where each point shows a linear trend estimate from five years prior to after each year.
+Estimates prior to 1996 and after 2014 are not available becuase of an incomplete ten year record for estimating the trend.
 Estimates in a year that are significant are shown in red.'}
@@
 ```
-```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at two example stations.
-The first row shows OLS (ordinary least squares) regression applied to annual averages of chlorophyll from the observed data, the second row shows OLS regression applied to annual averages of chlorophyll from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chlorophyll from the GAM.
+```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at two example stations.
+The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chl-a from the GAM.
 Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
@@
     y = 'Station (north to south)', 
-    x = expression(paste(log[10], " Chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
   )
@@
 ```
-```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chlorophyll in different annual periods at each station.
-The "observed" trend model is based on a linear fit to the annual averages of chlorophyll from the observed data, the "average" trend model is based on a linear fit to the annual averages of chlorophyll from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chlorophyll from the GAM model.
+```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station.
+The "observed" trend model is based on a linear fit to the annual averages of chl-a from the observed data, the "average" trend model is based on a linear fit to the annual averages of chl-a from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chl-a from the GAM model.
 Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting significant trends.'}
@@
 
-```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chlorophyll ($\\mu$g L$^{-1}$) .  Rows are arranged from north to south.'}
+```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chl-a ($\\mu$g L$^{-1}$) .  Rows are arranged from north to south.'}
 totab <- rawdat %>% 
@@
 cap.val <- 'Comparison of the four model structures (S, SY, SYD, SYDI) described in the first stage analysis of GAM estimation.
-The four models provide either identical or comparable ability to describe chlorophyll trends at an example station (32) in South San Francisco Estuary.
+The four models provide either identical or comparable ability to describe chl-a trends at an example station (32) in the southern end of the San Francisco Estuary.
 The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are comparable.

### 2021-02-04 R/dat_proc.R

@@ -201,7 +201,7 @@ seastrnd2 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     res = purrr::pmap(list(fl, mod = mod, doystr, doyend), function(fl, mod, doystr, doyend){
       
       # get slope trends
-      out <- anlz_trndseason(mod = mod, doystr = doystr, doyend = doyend, justify = 'center', win = 5)
+      out <- anlz_trndseason(mod = mod, doystr = doystr, doyend = doyend, justify = 'center', win = 10)
       
       return(out)
       

### 2021-02-04 manu_draft.Rmd
results revision done from coauth

@@
 Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).
-The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to March period.
+The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to June period.
 Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.
@@
   geom_vline(xintercept = 0, linetype = 'dashed') + 
-  # scale_x_continuous(limits = c(-1, 1)) +
+  # scale_x_continuous(limits = c(-1.14, 1.14)) +
   theme_minimal() + 
@@
 
-png('figs/trndcmp.png', height = 8, width = 7.5, family = 'serif', units = 'in', res = 400)
+png('figs/trndcmp.png', height = 4.5, width = 7.5, family = 'serif', units = 'in', res = 400)
 p
@@
 ```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station.
-The "observed" trend model is based on a linear fit to the annual averages of chl-a from the observed data, the "average" trend model is based on a linear fit to the annual averages of chl-a from the GAM model, and the "mixed-meta" trend model is based on a mixed-meta regression model fit to the annual averages of chl-a from the GAM model.
-Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting significant trends.'}
+The "OLS observed" trend model is based on an ordinary least squares (OLS) regression fit to the annual averages of chl-a from the observed data, the "OLS GAM" trend model is based on an OLS regression fit to the annual averages of chl-a from the GAM model, and the "Mixed-meta GAM" trend model is based on a mixed-meta regression fit to the annual averages of chl-a from the GAM model.
+Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting trend significance.'}
 knitr::include_graphics('figs/trndcmp.png')

### 2021-02-04 R/dat_proc.R

@@ -225,10 +225,10 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   crossing(
     fl = ., 
     tibble(
-      doystr = c(1, 91, 182, 274),
-      doyend = c(90, 181, 273, 364)
-      # doystr = c(41, 213), 
-      # doyend = c(213, 338)
+      # doystr = c(1, 91, 182, 274),
+      # doyend = c(90, 181, 273, 364)
+      doystr = c(41, 213),
+      doyend = c(213, 338)
     ), 
     tibble(
       yrstr = c(1991, 2000, 2010),
@@ -333,14 +333,15 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   select(station = fl, seas = doystr, yrs = yrstr, modtyp, yrcoef, pval) %>% 
   mutate(
     station = gsub('^data/modslog\\_chl|\\.RData$', '', station),
-    seas = factor(seas, levels = c('1', '91', '182', '274'), labels = c('Jan-Mar', 'Apr-Jun', 'Jul-Sep', 'Oct-Dec')),
+    # seas = factor(seas, levels = c('1', '91', '182', '274'), labels = c('Jan-Mar', 'Apr-Jun', 'Jul-Sep', 'Oct-Dec')),
+    seas = factor(seas, levels = c('41', '213'), labels = c('Jan-Jun', 'Jul-Dec')),
     yrs = case_when(
       yrs < 1995 ~ '1990-2000', 
       yrs >=1995 & yrs < 2005 ~ '2000-2010', 
       yrs >= 2005 ~ '2010-2019'
     ), 
     yrs = factor(yrs), 
-    modtyp = factor(modtyp, levels = c('obstrnd', 'lmtrnd', 'metatrnd'), labels = c('Observed', 'Average', 'Mixed-meta')), 
+    modtyp = factor(modtyp, levels = c('obstrnd', 'lmtrnd', 'metatrnd'), labels = c('OLS observed', 'OLS GAM', 'Mixed-meta GAM')), 
     pval = ifelse(pval <= 0.05, 'p < 0.05', 'ns')
   ) %>% 
   unnest('yrcoef') %>% 

### 2021-02-04 manu_draft.Rmd
revisions done from coauth, some figs put in supp, separate supp doc

@@
 We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.
-We then extract seasonal averages for an a priori period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.
+We then extract seasonal averages for an *a priori* period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.
 Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluate changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
@@
 Standard error estimates of predictions from WRTDS are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15].
-Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling as a post-hoc application to a previously fitted model.
+Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling as a *post-hoc* application to a previously fitted model.
 Alternative methods that include standard error estimates simultaneously with model output may be preferred for intensive or more iterative applications
@@
 The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.
-The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$, wheras the off-diagonal elements are the correlations among the elements of $\hbb$.
+The scalar standard errors of $\hbb$ are the square roots of the diagonal elements of $\hShb$, whereas the off-diagonal elements are the correlations among the elements of $\hbb$.
 Since parameter estimates are correlated, the covariance of $\hby$ is $\hShy = \bX \hShb \bX^T$.
@@
 Model predictions for chl-a trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
-GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figure \@ref(fig:prddat)).
+GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figure S1-S9).
 All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at South Bay stations, and 3) a fall chl-a peak that was smaller than the spring peak.
-The magnitude of the fall peak did not vary noticeably by location (Figure \@ref(fig:prddat)).
+The magnitude of the fall peak did not vary noticeably by location (Figures S1-S9).
 
@@
 
-Our results demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends over time.
-This is an important conclusion that can provide guidance on how GAMs should be used to model time series from long-term water quality monitoring data.
+Our results demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends.
+This is an important conclusion that provides guidance on how GAMs could be used to model time series from long-term water quality monitoring data when the purpose is to identify and describe which of many possible changes over time have occurred.
 Models with separate smoothers for continuous year and day of year can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
@@
 A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the mixed-meta regression results.
-Conversely, a false positive result (type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, whereas the mixed-meta results did not.
+Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, whereas the mixed-meta results did not.
 The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).
@@
 More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].
-As a result, trend assessments from mixed-meta regression are accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.
+As a result, trend assessments from mixed-meta regression are more accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.
 
@@
 For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].
-Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, producing fitted results that are arbitraily under- or over-smoothed to the raw observations.
+Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, producing fitted results that are arbitrarily under- or over-smoothed to the raw observations.
 This is especially problematic when for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.
@@
 We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
-Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention, etc.), a more general question of "when" changes occur independent of user decisions is also important to address.
+Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.
 For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?
 There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.
+Assessing water quality changes beyond an evaluation of seasonal averages could also be possible with our approach, such as assessing changes in the timing or magnitude of a seasonal peak across years.
 
@@
 In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.
-As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although a distinction here is made in the objective of the analysis.
-Our goal herein was to describe chl-a changes relative to time where the predictors were variations on a general theme (e.g., season vs.
-year), whereas using GAMs with different predictors assess potential causality is a different application with alternative goals.
+As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although this will depend on the objective of the analysis.
+Our goal was to describe chl-a changes relative to time where the predictors were variations on a general theme (e.g., season vs.
+year), which is a different application from using GAMs with predictors selected to assess potential causality.
 Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.
@@
 Finally, the evaluation of trends for alternative water quality variables in addition to chl-a is a simple and logical extension of the methods proposed in this study.
-The long-term monitoring program maintained by USGS includes multiple parameters in addition to chl-a that can provide additional context into broader water quality trends in South Bay [@Cloern16;@Schraga20].
+The long-term monitoring program maintained by USGS includes multiple parameters in addition to chl-a that can provide additional context into broader water quality trends in the SFE [@Cloern16;@Schraga20].
 These parameters include salinity, temperature, light attenuation, dissolved oxygen, suspended particulate matter, and dissolved inorganic nutrients, which collectively can be used to provide a broader understanding of potential eutrophication patterns or ecosystem shifts at seasonal and multi-decadal scales.
@@
 The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
-
-# Supplement
-
-A note about the `select` argument in `mgcv`: When `select = TRUE` is included, the comparison between models S and SY changes.
-This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.
-This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it is irrelevant if including both a linear and spline term for `cont_year`.
-If `select = TRUE` is used, models S and SY would still be effectively equivalent, but AIC selection would suggest that one model is superior.
-This result would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
 
@@
 knitr::include_graphics('figs/modsumfig.png')
-```
-
-```{r, results = 'hide', message = F}
-# get predictions to plot
-modprds <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
-  enframe() %>% 
-  group_by(value) %>% 
-  nest %>% 
-  mutate(
-    prd = purrr::map(value, function(x){
-
-      load(file = x)
-
-      nm <- basename(x)
-      nm <- gsub('\\.RData', '', nm)
-
-      mod <- get(nm) %>% 
-        pull(model) %>% 
-        deframe()
-
-      prd <- anlz_prd(mod)
-
-      return(prd)
-
-    })
-  ) %>% 
-  ungroup %>% 
-  select(station = value, prd) %>% 
-  mutate(
-    station = gsub('^data/modslog\\_chl|\\.RData$', '', station)
-  ) %>% 
-  unnest('prd')
-
-modobs <- rawdat %>% 
-  filter(param == 'chl')
-ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))
-
-p1 <- ggplot2::ggplot(modprds, ggplot2::aes(x = date)) + 
-  ggplot2::geom_point(data = modobs, ggplot2::aes(y = value), size = 0.5, colour = 'grey') +
-  ggplot2::geom_line(ggplot2::aes(y = value), size = 0.75, alpha = 0.7) + 
-  ggplot2::facet_grid(station~.) +
-  ggplot2::scale_color_viridis_d() + 
-  ggplot2::theme_minimal(base_family = 'serif', base_size = 14) + 
-  ggplot2::scale_y_log10() +
-  ggplot2::theme(
-    legend.position = 'top', 
-    legend.title = ggplot2::element_blank(),
-    strip.background = ggplot2::element_blank()
-  ) + 
-  ggplot2::labs(
-    y = ylab,
-    x = 'Continuous year', 
-    subtitle = '(a) Continous predictions'
-  )
-
- p2 <- ggplot2::ggplot(modprds, ggplot2::aes(x = doy, group = factor(yr), colour = yr)) + 
-    ggplot2::geom_line(ggplot2::aes(y = value), size = 0.5, alpha = 1) + 
-    ggplot2::theme_minimal(base_family = 'serif', base_size = 14) + 
-    ggplot2::theme(
-      legend.position = 'right', 
-      legend.title = ggplot2::element_blank(), 
-      strip.background = ggplot2::element_blank(), 
-      strip.text = ggplot2::element_blank(), 
-      axis.title.y = ggplot2::element_blank()
-    ) + 
-    ggplot2::scale_colour_gradient(low = "white", high = "black") + 
-    ggplot2::facet_grid(station ~ .) +
-    ggplot2::scale_y_log10() +
-    ggplot2::guides(colour = ggplot2::guide_colourbar(barheight = 20, barwidth = 1)) +
-    ggplot2::labs(
-      x = "Day of year", 
-      subtitle = '(b) Day of year\npredictions'
-    )
-p <- p1 + p2 + plot_layout(ncol = 2, width = c(1, 0.4))
-
-png('figs/prddat.png', height = 10, width = 9, family = 'serif', units = 'in', res = 400)
-print(p)
-dev.off()
-```
-```{r prddat, fig.cap = "GAM predictions for all stations from north to south for model S.
-The results show (a) predictions across the time series and (b) predictions by day of year.
-Observed data in (a) are shown with the gray points.
-Station locations are in Figure \\@ref(fig:sitemap)."}
-knitr::include_graphics('figs/prddat.png')
 ```

### 2021-02-04 supplement.Rmd
revisions done from coauth, some figs put in supp, separate supp doc

---
title: "Supplement to Multi-scale trend analysis of water quality using error propagation of generalized additive models"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Perry de Valpine (pdevalpine@berkeley.edu), Rebecca Murphy (rmurphy@chesapeakebay.net), Ian Wren (ianw@sfei.org), Ariella Chelsky (ariellac@sfei.org), Melissa Foley (melissaf@sfei.org), David B. Senn (davids@sfei.org)"
urlcolor: blue
csl: ecology.csl
link-citations: true
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F,
  fig.process = function(x) {
  x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
  if (file.rename(x, x2)) x2 else x
  })

# libraries
library(Jabbrev)
library(tidyverse)
library(wqtrends)
library(flextable)
library(officer)
library(ggmap)
library(sf)
library(ggsn)
library(USAboundaries)
library(rworldmap)
library(hrbrthemes)
library(patchwork)
library(lubridate)
library(colorspace)

data(locs)
data(modprf)
data(seastrnd)
data(seastrnd2)
data(cmptrnd)
data(modstr)
data(modslog_chl18)
data(modslog_chl24)
data(modslog_chl27)
data(modslog_chl32)
data(modslog_chl34)

source('R/funcs.R')

locs <- locs %>% 
  rename(station = Station)

# # extract bib entries from online
# bib_scrp('manu_draft.Rmd', 'refs.bib')
```

`r paste('Last build', Sys.time())`

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files('manu_draft.Rmd')
```

A note about the `select` argument in `mgcv`: When `select = TRUE` is included, the comparison between models S and SY changes.  This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.  This would be appropriate if `cont_year` was an explanatory variable subject to variable selection, but it is irrelevant if including both a linear and spline term for `cont_year`.  If `select = TRUE` is used, models S and SY would still be effectively equivalent, but AIC selection would suggest that one model is superior.  This result would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.

# Figures

```{r, results = 'hide', message = F}
# get predictions to plot
modprds <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
  enframe() %>% 
  group_by(value) %>% 
  nest %>% 
  mutate(
    prd = purrr::map(value, function(x){

      load(file = x)

      nm <- basename(x)
      nm <- gsub('\\.RData', '', nm)

      mod <- get(nm) %>% 
        pull(model) %>% 
        deframe()

      prd <- anlz_prd(mod)

      return(prd)

    })
  ) %>% 
  ungroup %>% 
  select(station = value, prd) %>% 
  mutate(
    station = gsub('^data/modslog\\_chl|\\.RData$', '', station)
  ) %>% 
  unnest('prd')

modobs <- rawdat %>% 
  filter(param == 'chl')
ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))

stas <- unique(modprds$station)

for(sta in stas){

  toplo1 <- modprds %>% 
    filter(station %in% sta)
  toplo2 <- modobs %>% 
    filter(station %in% sta)

  p1 <- ggplot2::ggplot(toplo1, ggplot2::aes(x = date)) + 
    ggplot2::geom_point(data = toplo2, ggplot2::aes(y = value), size = 0.5, colour = 'grey') +
    ggplot2::geom_line(ggplot2::aes(y = value), size = 0.75, alpha = 0.7) + 
    ggplot2::scale_color_viridis_d() + 
    ggplot2::theme_minimal(base_family = 'serif', base_size = 14) + 
    ggplot2::scale_y_log10(limits = c(0.26, 207)) +
    ggplot2::theme(
      legend.position = 'top', 
      legend.title = ggplot2::element_blank(),
      strip.background = ggplot2::element_blank()
    ) + 
    ggplot2::labs(
      y = ylab,
      x = 'Continuous year', 
      subtitle = paste0('Station ', sta, ' continuous and day of year predictions')
    )

  p2 <- ggplot2::ggplot(toplo1, ggplot2::aes(x = doy, group = factor(yr), colour = yr)) + 
    ggplot2::geom_line(ggplot2::aes(y = value), size = 0.5, alpha = 1) + 
    ggplot2::theme_minimal(base_family = 'serif', base_size = 14) + 
    ggplot2::theme(
      legend.position = 'right', 
      legend.title = ggplot2::element_blank(), 
      strip.background = ggplot2::element_blank(), 
      strip.text = ggplot2::element_blank(), 
      axis.title.y = ggplot2::element_blank()
    ) + 
    ggplot2::scale_colour_gradient(low = "white", high = "black") + 
    ggplot2::scale_y_log10(limits = c(0.26, 207)) +
    ggplot2::guides(colour = ggplot2::guide_colourbar(barheight = 10, barwidth = 1)) +
    ggplot2::labs(
      x = "Day of year"
    )
  p <- p1 + p2 + plot_layout(ncol = 2, width = c(1, 0.5))

  png(paste0('figs/prddat', sta, '.png'), height = 3.5, width = 9, family = 'serif', units = 'in', res = 400)
  print(p)
  dev.off()

}
```

```{r prddat18, fig.cap = "GAM predictions for station 18 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat18.png')
```

```{r prddat21, fig.cap = "GAM predictions for station 21 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat21.png')
```

```{r prddat22, fig.cap = "GAM predictions for station 22 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat22.png')
```

```{r prddat24, fig.cap = "GAM predictions for station 24 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat24.png')
```

```{r prddat27, fig.cap = "GAM predictions for station 27 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat27.png')
```

```{r prddat30, fig.cap = "GAM predictions for station 30 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat30.png')
```

```{r prddat32, fig.cap = "GAM predictions for station 32 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat32.png')
```

```{r prddat34, fig.cap = "GAM predictions for station 34 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat34.png')
```

```{r prddat36, fig.cap = "GAM predictions for station 36 for model S. The results show predictions across the time series and predictions by day of year.  Observed data in are shown with the gray points."}
knitr::include_graphics('figs/prddat36.png')
```

### 2021-02-05 manu_draft.Rmd
edits after proofing

@@
 However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods.
-Here, we propose a novel application of Generalized Additive Models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.
+Here, we propose a novel application of generalized additive models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.
 GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.
@@
 Such fluctuations may not be related to inter-annual trends or may not be well-suited to multi-scale smoothing methods.
-In this paper, we develop methods to estimate across-year trends of within-year features, such as a seasonal averages, while accounting for uncertainties across analysis steps.
+In this paper, we develop methods to estimate across-year trends of within-year features, such as seasonal averages, while accounting for uncertainties across analysis steps.
 
@@
 STL decomposition does not incorporate explanatory variables besides time, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually estimate standard errors to allow hypothesis testing [but see @Hafen10].
-STL methods may also over-simplify trends into absolute components that do not change over time, e.g., a seasonal estimate that is constant across years.
+STL methods may also over-simplify trends into stationary components that do not change over time, e.g., a seasonal estimate that is constant across years.
 This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].
@@
 We provide examples illustrating how these questions can be addressed using methods with GAMs.
-We also provide an approach for using meta-analysis methods [@Gasparrini12;@Sera19] after signal extraction with GAMs.
+We also provide an approach using meta-analysis methods [@Gasparrini12;@Sera19] for isolating seasonal trends with reasonable certainty from GAM results and evaluating these trends between years.
 This approach is new to environmental trend-detection problems.
-For this step, we provide methods using meta-analysis for isolating seasonal trends with reasonable certainty from GAM results and evaluating these trends between years.
 
@@
 However, SFE has demonstrated resistance to eutrophication, which has been attributed to high suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves [@Cole84;@Alpine88;@Jassby08;@Kimmerer14;@Lehman17].
-Renewed interest in understanding the potential for nutrient loading to negatively affect water quality has occurred for South Bay, where harmful algal blooms (HABs), elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
+Renewed interest in understanding the potential for nutrient loading to negatively affect water quality has occurred for more southern areas of the SFE, where harmful algal blooms (HABs), elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
 Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient particularly with respect to seasonal differences between years.
@@
 We used four different GAMs to estimate the long-term trend and uncertainty from the time series data, each with the potential to achieve a similar fit to the data but with a slightly different model structure (Table \@ref(tab:modsumtab)).
-Models are shown in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17].
-
-GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor, independent of noise inherent in the raw data.
+Models are shown in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17;@RCT20].
+
+GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory variable, independent of noise inherent in the raw data.
 The simplest GAM for this purpose is expressed as:
@@
 
-where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
+where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
 Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals.
@@
 When `cont_year` is included explicitly as a linear term in model SY, `mgcv` adjusts the basis functions for the spline to exclude the linear term, thereby not over-parameterizing the model.
-Whereas an estimated linear trend in `cont_year` and its uncertainty can be extracted from the fitted spline in model S, model SY provides this trend directly, given the equivalent result.
+Whereas an estimated linear trend in `cont_year` and its uncertainty can be extracted from the fitted spline in model S, model SY provides this trend directly, giving the equivalent result.
 Further, package `mgcv` can penalize linear trends in splines to provide a method for variable selection (option `select = TRUE`), such as when numerous splines are included in the model formulation for variables that may or may not be important.
@@
 
-Model SYDI: `y ~ cont_year + s(cont_year) + s(doy, bs = "cc") + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knots_D_ti))`
+Model SYDI: `y ~ cont_year + s(cont_year, k = num_knots_Y) + s(doy, bs = "cc", k = num_knots_D) + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knots_D_ti))`
 
@@
 The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.
-In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both given limits on the model degrees of freedom.
+In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `num_knots_Y` and `num_knots_D`, but not both given limits on the model degrees of freedom.
 
@@
 In the "`gam2`" model (our SYDI model), @Murphy19 do not explicitly consider the number of knots in the interaction spline.
-In both cases, they effectively use their choice of number of knots in different spline components as an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system, while noting that additional explanatory variables are necessary to explain the residual variability or autocorrelation in the data.
+In both cases, they effectively use their choice of number of knots in different spline components as an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system.
 @Murphy19 acknowledge that incomplete modeling of fluctuations in the data may inflate their Type I error rates for estimating temporal changes, but they account for that by incorporating additional hydrologic variables in a subsequent phase of their models in a step-wise manner.
@@
 In stage three of the analysis, we used a meta-analysis method to evaluate linear trends across years of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
-In the examples, we considered two periods of interest for chl-a, January-July and August-December, which are relevant to phytoplankton bloom phenology in SFE.
-
 This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?
@@
 
-We estimated the model (equation \@ref(eq:mximet)) using the *mixmeta* package in R [@Sera19].
+We estimated the model (equation \@ref(eq:mixmet)) using the *mixmeta* package in R [@Sera19].
 Results from *mixmeta* have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
@@
 This method was applied to a chosen sequence or "window" of years for estimating the linear trend.
-This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (slope) in chl-a over a chosen series of years.
+This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (log$_{10}$ slope) in chl-a over a chosen series of years.
 Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).
-Years with more data have smaller standard errors for $\hmt$ compared to years with less data.
+Years with more samples or less variability have smaller standard errors for $\hmt$ compared to years with less samples or more variability.
 The meta-analysis explicitly includes these differences in uncertainty magnitudes.
@@
 The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year blocks from 1991 to 2000, 2000 to 2010, and 2010 to 2019.
+These temporal separations are relevant to phytoplankton bloom phenology in SFE [@Cloern20].
 As an extension of these methods, a chosen sequence of years was applied across the time series to assess periods of time within which seasonal trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
@@
 Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.
-Trend estimates from mixed-meta regression applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averaged.
-The comparisons were demonstrated with select examples where differences were pronounced and then applied to all stations.
+Trend estimates from mixed-meta regression applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages.
+Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods.
+This analysis was then applied to all stations.
 
@@
 Model predictions for chl-a trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
-GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figure S1-S9).
-All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at South Bay stations, and 3) a fall chl-a peak that was smaller than the spring peak.
+GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figures S1-S9).
+All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at southern stations, and 3) a fall chl-a peak that was smaller than the spring peak.
 The magnitude of the fall peak did not vary noticeably by location (Figures S1-S9).
@@
 
-To demonstrate the differences in results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
-All plots show trends in seasonal averages within ten-year windows, whereas rows a-c show estimates from January to June and rows d-f show estimates from July to December.
-The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) form 1991 to 2000, whereas a trend for the same period in July to December was not observed.
+To demonstrate the results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
+All plots show trends in seasonal averages within ten-year windows, whereas plots a-c show estimates from January to June and plots d-f show estimates from July to December.
+The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) from 1991 to 2000, whereas a trend for the same period in July to December was not observed.
 Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
-Finally, chl-a decreased from 2010 to 2019 but only for July to December (log$_{10} chl-a slope -0.02, -0.04-0 95\% confidence interval).
+Finally, chl-a decreased from 2010 to 2019 but only for July to December (log$_{10}$ chl-a slope -0.02, -0.04-0 95\% confidence interval).
 Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.
@@
 Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.
-Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased for at stations 30 and 32 from 2010 to 2019.
+Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased at stations 30 and 32 from 2010 to 2019.
 For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
@@
 Increases and decreases were observed in both the January to June and July to December seasonal periods, with some notable exceptions.
-In particular, the most southern stations (32, 34, 36) had more trends in the July to December period.
-Additionally, trends in recent years in both seasonal periods were not observed for the more northern stations.
-For both seasonal periods, a change from increasing to decreasing chl-a occurred at most stations around 2007.
+In particular, the most southern stations (32, 34, 36) had increasing trends prior to 2005 only in the July to December period.
+Additionally, chl-a at the more northern stations has not changed in recent years for both seasonal periods.
+For most stations and seasonal periods, a change from increasing to decreasing chl-a occurred around 2007.
 
@@
 Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
-Only the mixed-meta regression results shows a trend in this example (based on $p < 0.05$).
-The observed regression results (top plot) and average regression on GAM estimates (middle plot) did not identify a trend.
+Only the mixed-meta regression results show a trend in this example (based on $p < 0.05$).
+The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.
 Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.
-Unlike the first example, the top two figures show trends, whereas the bottom plot for the mixed-meta regression analysis does not show a trend because of added uncertainty in the averages provided by the GAMs.
+Unlike the first example, the top two figures show trends, whereas the bottom plot for the mixed-meta regression analysis does not show a trend.
 In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.
@@
 Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.
-Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the former to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
+Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
 
@@
 Our results demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends.
-This is an important conclusion that provides guidance on how GAMs could be used to model time series from long-term water quality monitoring data when the purpose is to identify and describe which of many possible changes over time have occurred.
+This is an important conclusion that provides guidance on how GAMs could be used to model time series from long-term water or other environmental monitoring programs when the purpose is to identify and describe which of many possible changes over time have occurred.
 Models with separate smoothers for continuous year and day of year can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
-This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing with the mgcv R package using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
+This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
 Specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.
@@
 A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the mixed-meta regression results.
-Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, whereas the mixed-meta results did not.
+Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, but the mixed-meta results did not.
 The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).
@@
 
-The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions required by an analyst that may be necessary for the implementation of alternative trend assessment methods.
+The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions that may be necessary for the implementation of alternative trend assessment methods.
 For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].
 Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, producing fitted results that are arbitrarily under- or over-smoothed to the raw observations.
-This is especially problematic when for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.
+This is especially problematic for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.
 Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.
@@
 
-Additional work could be conducted to further strengthen the validity of conclusions based on trends from mixed-meta regression analyses applied to the seasonal GAM averages.
-We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends partially impose arbitrary decisions that can influence interpretation of results.
-Although there are undoubtedly many scenarios where years of interest can be chosen objectively (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.
+Additional work could be conducted to further strengthen the validity of conclusions based on trends from mixed-meta regression analyses applied to the GAM seasonal averages.
+We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends.
+Although there are undoubtedly many scenarios where years of interest can be chosen objectively by the needs of an analysis (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.
 For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?
@@
 An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09].
-In such cases, considerations of model structure can have direct implications on conclusions regarding potential causality.
+In such cases, considerations of model structure can have direct implications on conclusions of potential causality.
 As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although this will depend on the objective of the analysis.
-Our goal was to describe chl-a changes relative to time where the predictors were variations on a general theme (e.g., season vs.
-year), which is a different application from using GAMs with predictors selected to assess potential causality.
+Our goal was to describe chl-a changes relative to time, where the predictors were variations on a general theme (e.g., season vs.
+year).
+This is a different application from using GAMs with predictors selected to assess potential causality.
 Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.
@@
 Chl-a measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
-The open-source R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
+The open-source wqtrends R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
 
@@
 Plots in (b) show pairwise comparisons of predicted chl-a between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.
-The plots demonstrate that results between the models are comparable except for a few observations at extreme values(a), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (b).
+The plots demonstrate that results between the models are comparable except for a few observations at extreme values (a, b), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (c).
 The 1:1 lines are in red to facilitate comparisons."}
@@
 ```
-```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates using results of GAM predictions for station 34.
+```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates in ten year blocks from mixed-meta analyses using results of GAM predictions for station 34.
 Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.
@@
 ```{r trndmap, fig.cap= "Trend estimates across seasons by decade for chl-a at each station.
-Point type, shape, and color represent the direction and magnitude of an estimated trend as the log$_{10}$ slope for chl-a concentration per year.
+Point type and color represent the direction and magnitude of an estimated trend as the log$_{10}$ slope for chl-a concentration per year.
 Trends with $p<0.05$ are marked with an asterisk.
@@
 ```
-```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year from applying the mixed-meta analysis across the time series for each station.
+```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year (+/- 95\\% confidence interval) from applying the mixed-meta analysis across the seasonal averages for each station.
 Stations are arranged top to bottom from north to south.
@@
 Results are from a ten-year, centered moving window where each point shows a linear trend estimate from five years prior to after each year.
-Estimates prior to 1996 and after 2014 are not available becuase of an incomplete ten year record for estimating the trend.
+Estimates prior to 1996 and after 2014 are not available because of an incomplete ten year record for estimating the trend.
 Estimates in a year that are significant are shown in red.'}
@@
 ```
-```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at two example stations.
+```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at  (a) station 36 and (b) station 22.
 The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chl-a from the GAM.

### 2021-02-05 manu_draft.Rmd
wqtrends cite added, sent to coauth

@@
 
-The wqtrends R package developed by the authors includes all methods described above.
+The wqtrends R package developed by the authors includes all methods described above [@Beck21].
 The package is available for download at https://tbep-tech.github.io/wqtrends.
@@
 Chl-a measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
-The open-source wqtrends R package developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
+The open-source wqtrends R package [@Beck21] developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
 

### 2021-03-10 manu_draft.Rmd
starting revisions from PDV, no more mixed-meta language

@@
 A key requirement is complete propagation of uncertainty through the analysis.
-However, this is difficult when there are mismatches between time scales of monitoring data and analysis methods.
+However, this is difficult when there are mismatches between time scales of monitoring data and trends of interest.
 Here, we propose a novel application of generalized additive models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.
-GAMs are a general method for modelling a response variable as the additive sum of one or more smoother splines fit to predictor variables.
-These models have shown promise in other applications to water quality monitoring data to separate long-term (i.e., annual or decadal) trends from seasonal variation.
-Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-meta regression analyses to quantify inter-annual trends that account for full propagation of error across methods.
+GAMs can model a response variable as the additive sum of one or more smoothing splines fit to predictor variables.
+These models have shown promise in water quality trend analysis to separate long-term (i.e., annual or decadal) trends from seasonal variation.
+Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-effects meta-analysis regression to quantify inter-annual trends that account for full propagation of error across methods.
 We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.
-We then extract seasonal averages for an *a priori* period of time within each year from the GAM results, including an accurate assessment of standard error within each year where the seasonal average is estimated.
-Finally, we demonstrate how trends can be quantified with mixed-meta regression analyses that evaluate changes in the seasonal averages between years with full propagation of uncertainty in the estimates.
-Overall, this approach leverages the ability of GAMs to identify an optimal amount of smoothing from the raw data and uses those results to accurately assess trends that minimize the likelihood of false positive or negative results, which is particularly important for time series with missing observations or varying sample effort over time.
+We then extract seasonal averages and their standard errors for an *a priori* time period within each year from the GAM results.
+Finally, we demonstrate how across-year trends in seasonal averages can be modeled with mixed-effects meta-analysis regression that propagates uncertainties from the GAM fits to the across-year analysis.
+Overall, this approach leverages GAMs to smooth data with missing observations or varying sample effort across years to estimate seasonal averages and meta-analysis to estimate trends across years.
 
@@
 
-Finally, trend results from the mixed-meta regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
-As stated above, mixed-meta regression allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
-Moreover, the mixed-meta analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
+Finally, trend results from the meta-analysis regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
+As stated above, meta-analysis allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
+Moreover, the meta-analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
 Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.
-Trend estimates from mixed-meta regression applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages.
+Trend estimates from meta-analysis applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages.
 Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods.
@@
 
-Slope estimates that were used to determine significance of a seasonal trend across years from the mixed-meta analysis were not back-transformed.
+Slope estimates that were used to determine significance of a seasonal trend across years from the meta-analysis were not back-transformed.
 While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.
@@
 Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.
-Mixed-meta regression analyses applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased at stations 30 and 32 from 2010 to 2019.
+Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased at stations 30 and 32 from 2010 to 2019.
 For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
@@
 
-Results showing trend estimates from mixed-meta regression on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2).
+Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2).
 Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
-Only the mixed-meta regression results show a trend in this example (based on $p < 0.05$).
+Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).
 The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.
 Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.
-Unlike the first example, the top two figures show trends, whereas the bottom plot for the mixed-meta regression analysis does not show a trend.
-In both cases, only the mixed-meta regression results provide accurate trend estimates because of full propagation of uncertainty across methods.
+Unlike the first example, the top two figures show trends, whereas the bottom plot for the meta-analysis regression does not show a trend.
+In both cases, only the meta-analysis results provide accurate trend estimates because of full propagation of uncertainty across methods.
 
 Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).
-The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and mixed-meta methods, with much larger slopes observed especially at the more southern stations in the January to June period.
-Slope estimates from the OLS models applied to the averages from the GAMs as compared to the mixed-meta results were more similar, excluding some of the slope estimates for the southern stations.
-Differences in significance of trends between the OLS models applied to the GAM averages and the mixed-meta analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
+The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and meta-analysis methods, with much larger slopes observed especially at the more southern stations in the January to June period.
+Slope estimates from the OLS models applied to the averages from the GAMs as compared to the meta-analysis results were more similar, excluding some of the slope estimates for the southern stations.
+Differences in significance of trends between the OLS models applied to the GAM averages and the meta-analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
 
@@
 In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.
-A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the mixed-meta regression results.
-Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, but the mixed-meta results did not.
+A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the meta-analysis regression results.
+Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, but the meta-analysis results did not.
 The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).
@@
 The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
-More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-meta regression analyses [@Sera19].
-As a result, trend assessments from mixed-meta regression are more accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.
+More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-effects meta-analyses [@Sera19].
+As a result, trend assessments from meta-analyses are more accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.
 
@@
 
-Additional work could be conducted to further strengthen the validity of conclusions based on trends from mixed-meta regression analyses applied to the GAM seasonal averages.
+Additional work could be conducted to further strengthen the validity of conclusions based on trends from meta-analysis regression applied to the GAM seasonal averages.
 We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends.
@@
 ```
-```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates in ten year blocks from mixed-meta analyses using results of GAM predictions for station 34.
+```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates in ten year blocks from meta-analyses using results of GAM predictions for station 34.
 Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.
@@
 ```
-```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year (+/- 95\\% confidence interval) from applying the mixed-meta analysis across the seasonal averages for each station.
+```{r winchg, fig.cap='Estimates of log$_{10}$ chl-a change per year (+/- 95\\% confidence interval) from applying the meta-analysis across the seasonal averages for each station.
 Stations are arranged top to bottom from north to south.
@@
     labs(
-      subtitle = paste('Mixed-meta regression on GAM estimates', pval3),
+      subtitle = paste('Meta-analysis regression on GAM estimates', pval3),
       y = NULL
@@
 ```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at  (a) station 36 and (b) station 22.
-The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows mixed-meta regression applied to the annual averages of chl-a from the GAM.
+The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows meta-analysis regression applied to the annual averages of chl-a from the GAM.
 Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
@@
 ```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station.
-The "OLS observed" trend model is based on an ordinary least squares (OLS) regression fit to the annual averages of chl-a from the observed data, the "OLS GAM" trend model is based on an OLS regression fit to the annual averages of chl-a from the GAM model, and the "Mixed-meta GAM" trend model is based on a mixed-meta regression fit to the annual averages of chl-a from the GAM model.
+The "OLS observed" trend model is based on an ordinary least squares (OLS) regression fit to the annual averages of chl-a from the observed data, the "OLS GAM" trend model is based on an OLS regression fit to the annual averages of chl-a from the GAM model, and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the annual averages of chl-a from the GAM model.
 Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting trend significance.'}

### 2021-03-10 R/dat_proc.R

@@ -341,7 +341,7 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
       yrs >= 2005 ~ '2010-2019'
     ), 
     yrs = factor(yrs), 
-    modtyp = factor(modtyp, levels = c('obstrnd', 'lmtrnd', 'metatrnd'), labels = c('OLS observed', 'OLS GAM', 'Mixed-meta GAM')), 
+    modtyp = factor(modtyp, levels = c('obstrnd', 'lmtrnd', 'metatrnd'), labels = c('OLS observed', 'OLS GAM', 'Meta-analysis GAM')), 
     pval = ifelse(pval <= 0.05, 'p < 0.05', 'ns')
   ) %>% 
   unnest('yrcoef') %>% 

### 2021-03-10 manu_draft.Rmd
intro methods revision done from PDV

@@
 Overall, this approach leverages GAMs to smooth data with missing observations or varying sample effort across years to estimate seasonal averages and meta-analysis to estimate trends across years.
+Methods are provided in the *wqtrends* R package.
 
@@
 This yields parameters that are specific to season, year, and flow regime.
-The WRTDS approach is conceptually similar to localized multi-polynomial smoothing methods, with specific application to descriptive variables relevant for water quality constituents (i.e., season, year, and discharge).
-Standard error estimates of predictions from WRTDS are available through a "block bootstrap" approach that uses Monte Carlo estimates of false positive rates from the model results [@Hirsch15].
+The WRTDS approach is conceptually similar to local kernel smoothing methods, with specific application to explanatory variables relevant for water quality constituents (i.e., season, year, and discharge).
+Standard error estimates of predictions from WRTDS are available through a block bootstrap approach applied to the model results [@Hirsch15].
 Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling as a *post-hoc* application to a previously fitted model.
@@
 GAMs combine one or more smoothing splines to model patterns in data and may be seen as generalizing the concepts behind STL and WRTDS [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].
-In statistics, the evolution of non-parametric regression methods has largely converged on GAMs, rather than kernel smoothing methods used by both STL and WRTDS.
-Kernel smoothing and spline smoothing are closely related, and a key challenge for each is to determine the appropriate degree of smoothing.
-For example, WRTDS can potentially give results similar to the spline-based smoothing methods used in GAMs, although at higher computational expense and with the limitation that uncertainty estimates are not readily obtainable from the original method [@Beck17].
-WRTDS also requires specifying the desired degree of smoothing via the "windows" that define the local weighting at each point in the time series.
-The windows in WRTDS are conceptually similar to the kernel (or bandwidth) used in more conventional smoothing methods.
-A challenge of WRTDS is that there is no clear rule that can be applied to determine the best choice of window size.
-The associated tradeoff between over- and under-smoothing is a hallmark of these approaches.
-
-GAMs have various advantages compared to kernel smoothing methods and are therefore used for non-parametric regression smoothing in many fields.
-The "basis functions" used to formulate GAMs can be customized based on expected patterns in the data.
+The basis functions used to formulate GAMs can be customized based on expected patterns in the data.
 Examples include cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
-GAMs have added flexibility because they can include both parametric (e.g.
-linear or quadratic) components and non-parametric (spline) components.
+GAMs have added flexibility because they can include both parametric (e.g., linear or quadratic) components and non-parametric (spline) components.
 Multiple approaches have been developed to determine the optimal degree of smoothness.
-These approaches are based on likelihood estimates or optimization of out-of-sample prediction error, which address a key concern around methods like WRTDS that do not have analogs for choosing optimal degrees of smoothing.
-Further, GAMs have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (i.e., generalized additive mixed models or GAMMs), and have computationally efficient implementations that can be optimized more quickly than other approaches [@Wood17].
+These approaches are based on optimization of out-of-sample prediction error, which addresses a key concern around methods like WRTDS that do not have analogs for choosing optimal degrees of smoothing.
+GAMs can also produce comparable results similar to those provided by WRTDS [@Beck17] and have readily obtainable uncertainty estimates that do not require *post-hoc* resampling of model results.
+Further, GAMs have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (i.e., generalized additive mixed models or GAMMs), and have computationally efficient implementations [@Wood17].
 
@@
 For example, @Murphy19 used GAMs to decompose water quality time series from Chesapeake Bay into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.
+Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17].
 The study herein generalizes the approach to analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when spline flexibility is allowed to vary [@Wood03;@Wood17] rather than being constrained *a priori* for different time scales, and prioritizes full incorporation of uncertainty.
-Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17].
-
-Our motivating problem has several characteristics that are partially addressed by previous methods and can further build on  GAMs as a starting point.
+
+Our motivating problem has several characteristics that are only partially addressed by previous methods and can further build on  GAMs as a starting point.
 Our general goal is to understand interannual changes in seasonally averaged water quality metrics, such as chlorophyll.
 However, the seasonal average within each year must be robust to inconsistent sampling times and intervals, and any trend analysis must consider the uncertainties in seasonal averages.
-The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common to time series data.
-This paper develops GAMs as a distinct application that can meet these requirements while addressing key limitations of seasonal Kendall tests and the more complex STL and WRTDS methods.
-We do so with the explicit goal of quantifying interannual trends in seasonal averages.
+The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common in time series data.
+This paper develops the use of GAMs with mixed-effects meta-analysis [@Gasparrini12;@Sera19] to address multi-scale trend analysis questions for which seasonal Kendall tests and the more complex STL and WRTDS methods are not well-suited.
 
@@
 Are there significant across-year trends in the spring phytoplankton bloom or baseline chlorophyll concentrations during periods of low productivity between blooms?
-We provide examples illustrating how these questions can be addressed using methods with GAMs.
-We also provide an approach using meta-analysis methods [@Gasparrini12;@Sera19] for isolating seasonal trends with reasonable certainty from GAM results and evaluating these trends between years.
-This approach is new to environmental trend-detection problems.
+We provide examples illustrating how these questions can be addressed using GAMs to estimate seasonal trends and evaluated between years using meta-analysis methods.
+This approach is new to environmental trend-detection problems and is provided in the *wqtrends* R package developed by the authors [@Beck21, available at https://tbep-tech.github.io/wqtrends].
 
@@
 
-Nitrogen and phosphorus levels in SFE usually exceed concentrations that cause eutrophication in other estuaries.
-However, SFE has demonstrated resistance to eutrophication, which has been attributed to high suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves [@Cole84;@Alpine88;@Jassby08;@Kimmerer14;@Lehman17].
-Renewed interest in understanding the potential for nutrient loading to negatively affect water quality has occurred for more southern areas of the SFE, where harmful algal blooms (HABs), elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
+Nitrogen and phosphorus levels in the SFE usually exceed concentrations that cause eutrophication in other estuaries.
+However, the SFE has demonstrated resistance to eutrophication, which has been attributed to high concentrations of suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves [@Cole84;@Alpine88;@Jassby08;@Kimmerer14;@Lehman17].
+The Regional Water Quality Control Board has showed renewed interest in understanding the potential for nutrient loading to negatively affect water quality for more southern areas of the SFE, where harmful algal blooms, elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
 Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient particularly with respect to seasonal differences between years.
@@
 Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
-Chl-a on GFF filters was analyzed fluorometrically after extraction in 90% acetone.
 Sampling frequency varied somewhat over time and by station.
 Every observation was included directly in the statistical models without spatial or temporal binning or averaging.
+Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals.
+Methods for back-transformation of model results are provided in the supplement.
 
@@
 
-We implemented the GAMs in three stages.
-First, we used a GAM to estimate a smooth pattern of variation in the raw data along with its uncertainty.
-Second, we calculated a feature of interest from the estimated GAM, along with its propagated uncertainty.
+We implemented our analysis in three stages.
+First, we use a GAM to estimate a smooth temporal pattern in the raw data along with its uncertainty.
+Second, we calculate a feature of interest from the estimated GAM, along with its propagated uncertainty.
 For this example, the seasonal averages were extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
-Third, we used a mixed effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
+Third, we use a mixed effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
 While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
 Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions.
-We developed two mixed effects meta-analysis approaches that provided 1) a simple comparison of whether seasonal features differ across years, and 2) an estimation of short-term linear trends on time scales chosen by the analyst.
 
@@
 
-We used four different GAMs to estimate the long-term trend and uncertainty from the time series data, each with the potential to achieve a similar fit to the data but with a slightly different model structure (Table \@ref(tab:modsumtab)).
+We considered four different GAMs to smooth the raw data across time.
+Although they can achieve similar fits, they do so by partitioning variation in the time-series differently (Table \@ref(tab:modsumtab)).
+We discuss all four to clarify their relationships and interpretations.
 Models are shown in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17;@RCT20].
 
-GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory variable, independent of noise inherent in the raw data.
 The simplest GAM for this purpose is expressed as:
@@
 
-where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
-Log$_{10}$-transformed chl-a was used for all analyses to meet assumptions of normally-distributed residuals.
-
-The optimal level of smoothing in `mgcv` is determined by a penalty on the net curvature of the spline [@Wood04].
+where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case as a function of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
+
 Smoothing was determined using generalized cross-validation (GCV, as implemented in `mgcv`), which approximately minimizes out-of-sample prediction error.
-To allow GCV (or other alternatives) to work as intended, a potential upper limit on the number of knots determined by the analyst must be sufficiently large.
+GCV works by penalizing the net curvature of a spline [@Wood04].
+To allow GCV (or other alternatives) to work as intended, the number of knots that determine the maximum degrees of freedom chosen by the analyst must be sufficiently large so that the curvature penalty, rather than the number of knots, determines smoothness.
 Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.
-We chose a sufficiently large number of knots for `num_knots_Y` (Model S) as 12 times the number of years in the time series that was modeled.
-This created the potential, as determined through GCV, to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.
-If the data were too sparse, the model will have insufficient degrees of freedom to fit 12 knots per year, in which case the number of knots was reduced by one knot per year until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).
+In the examples below, we chose the number of knots, `num_knots_Y`, as 12 times the number of years in the time series, i.e., one knot per month.
+If the data were too sparse to fit 12 knots per year, the number of knots was reduced by one knot per year until the model could be estimated (i.e., 12 * years, 11 * years, etc.).
 
@@
 
-where `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates that the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.
+where `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates that the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the number of knots for the `doy` spline.
 While model SYD is not mathematically equivalent to models S and SY, it should produce nearly identical results.
@@
 In model SYD, there are separate penalties for the two splines, as compared to S and SY that include penalties only for the `cont_year` spline.
-This is important because variation in the response variable can be differently attributed to each spline depending on model, whereas the sum of components for each model produce comparable results between models.
-Our goal is to extract signals (i.e., seasonal estimates) from the fitted time series and different allocation of penalties among the splines in each model is unimportant for this task.
-However, interpreting differences in fit between model SYD and models S or SY can be problematic because the penalties for smoothing splines based on curvature are heuristic [@Wood17].
-For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations and does not imply one model fits better.
+This is important because variation in the response variable can be differently attributed to each spline depending on the model, even while the sum of components for each model produces similar results between models.
+Our goal is to extract seasonal averages from the fitted time series, which is not sensitive to different allocation of penalties among the splines in each model.
+
+If the fits were to differ substantially between model SYD and models S or SY, an interpretation could be difficult because the penalties for smoothing splines based on curvature are heuristic [@Wood17].
+For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations and does not imply one model fit is better.
 In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.
 
-Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will consider data from the same period in other years by imputing an average seasonal pattern with the `doy` spline.
-However, an interpretation of these imputations is challenging.
-For example, the spring chl-a peak is a notable feature every year in SFE.
+Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will impute an average seasonal pattern with the `doy` spline, thereby considering data from the same period in other years in the prediction of the period with missing data.
+However, an interpretation of these imputations may be challenging.
+For example, the spring chl-a peak is a notable feature every year in the SFE.
 If the peak occurs at the same time every year but the magnitude varies, then the average within-year pattern can be interpreted as the average magnitude.
@@
 Following the rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.
-Model SYDI differs formulaically from model S to a greater extent than model SYDI, but all of the splines use the same inputs to smooth the same data.
-The splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model SY given the fewer knots that are available to the former.
+Model SYDI differs formulaically from model S to a greater extent than model SYD, but all of the splines use the same inputs to smooth the same data.
+The univariate splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model S given the fewer knots that are available to the former.
 The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.
@@
 By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.
-However, this outcome is difficult to achieve for reasons explained below.
-
-The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year.
-For example, if one year has an especially early and large magnitude spring bloom, the years before and after would be expected also to have earlier and larger than average spring blooms.
-For the raw SFE data, and chl-a dynamics in many estuaries, this is not expected to be the case.
-An early and large spring bloom could easily be followed the next year by a late and relatively small spring bloom.
-If the interaction spline has many knots, then the degree of smoothness across years can be estimated to be very low, allowing fluctuations within years to vary strongly across years.
-However, if that is the case, the conceptual motivation for model SYDI may not be supported by the data.
-Given that observed bloom dynamics for most systems do not vary smoothly across years, the potential benefits of using model SYDI could be outweighed by using simpler models.
-Moreover, the spline and penalty structure differ from the other models, making it difficult to evaluate their relative performance (Table \@ref(tab:modstrtab)).
-
-For models SYD and SYDI, the allocation of knots between the different splines can be chosen by the analyst so long as it is recognized that different choices will arbitrarily lead to related differences in allocation of variation among the splines.
-In models S, SY, and SYD, the number of knots in the `cont_year` term need to be sufficiently large.
-In model SYDI, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.
-Allowing the `cont_year` spline to include many knots would somewhat defeat the purpose of the interaction spline.
-
-@Murphy19 used spline formulations related to those proposed here, but are insufficient for our needs of extracting and quantifying seasonal trends.
-@Murphy19 have a "`gam0`" with only `s(doy)` and linear `cont_year` terms, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
-Their `gam0` is *a priori* not of interest for our data because of an assumption of constant seasonality and a linear annual trend.
-Compared to our application of SYD and SYDI, their implementation of "`gam1`" and "`gam2`" allocated knots differently, leading to a different interpretation of results.
-For "`gam1`" (our SYD), they set a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years, whereas we adopt a different approach using a number of knots equal to 12 times the number of years, or one per month.
-Accordingly, they interpret this spline as separating a long-term (or low-frequency) trend from other patterns, whereas we use it to separate intra-annual changes in season from noise at a finer temporal scale.
-In the "`gam2`" model (our SYDI model), @Murphy19 do not explicitly consider the number of knots in the interaction spline.
-In both cases, they effectively use their choice of number of knots in different spline components as an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system.
-@Murphy19 acknowledge that incomplete modeling of fluctuations in the data may inflate their Type I error rates for estimating temporal changes, but they account for that by incorporating additional hydrologic variables in a subsequent phase of their models in a step-wise manner.
-For the methods presented here, which focus first on fully identifying temporal change, we seek to avoid inflating Type I error rates by increasing the upper limit for the knots.
+However, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.
+The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year and that feature could be accounted for with this term, but for the SFE data and chl-a dynamics in many estuaries, bloom size typically varies between years.
+Thus, the conceptual motivation for model SYDI and its practical application are not supported for this analysis.
+
+@Murphy19 used spline formulations for Chesapeake Bay water quality related to those proposed here, but for different goals and with different handling of smoothness.
+They evaluated a "`gam0`" with only `s(doy)` and linear `cont_year` terms, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
+In application, only "`gam2`" was used, including the addition of splines as functions of hydrologic variables to account for finer-scale variation.
+Murphy et al.
+allowed a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years and do not explicitly consider the number of knots in the interaction spline, following an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system.
+Constraining splines with insufficient knots could inflate Type I error rates for temporal changes and we seek to lower this risk by increasing the upper limit for the knots for the `s(cont_year)` term.
 Finally, @Murphy19 present large AIC differences between their spline formulations.
-We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)), making differences in AIC unimportant.
-In our example, large differences in AIC only reflect inadequate choice of knots in one or more splines, which should be avoided.
+We instead emphasize that, given sufficient knots, the models represent alternative formulations of conceptually similar explanations for the data and yield similar fits (Table \@ref(tab:modstrtab)), resulting in near ties for AIC between models.
 
@@
 Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).
-However, when contrasting the models using only the continuous year smoother (`s(cont_year)`, as an example), the model fits differ substantially because of structural differences and differences in the penalties applied to the basis functions.
+However, when contrasting the estimates using only the continuous year smoother (`s(cont_year)`), the fits differ substantially because of how each model allocates variation to the splines.
 These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)).
@@
 
-For all results, model S was used with enough knots in `num_knots_y` to evaluate chl-a trends across the monitoring stations in SFE.
-This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
+For all results, model S was used with enough knots in `num_knots_y` to evaluate chl-a trends across the monitoring stations in the SFE.
+This model was chosen because of the relatively faster processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
 
@@
 In the second-stage analysis, we estimated a seasonal average, such as the mean spring chl-a concentrations and the uncertainty in each year.
-We define $\mu_t$ as the average for the period of interest (e.g., seasonal average) in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
+We define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
 The season includes $n$ days.
@@
 Each of these estimates are from on approximate multivariate normality of the sampling distribution of $\hbb$.
-@Murphy19 provided a similar approach for comparing multi-year averages of month-scale spline values for sets of years at the beginning vs.
-end of a time-series.
 
@@
 
-In stage three of the analysis, we used a meta-analysis method to evaluate linear trends across years of within-year features such as seasonal water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
-This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the trend is chosen by the investigator?
+In stage three of the analysis, we used a meta-analysis method to evaluate linear trends across years of seasonal-average water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
+This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the seasonal average and duration of the across-year comparison is chosen by the investigator?
 For example, is there a trend in the spring chl-a average from 1990 to 2000?
@@
 
-A meta-analysis mixed effects model is useful to estimate linear trends when each observation has a unique and quantified standard error, which is the case with our estimates $\hmt$ and $\hshmt$.
+A mixed effects meta-analysis model can estimate linear trends when each observation has an associated standard error, which is the case with our estimates $\hmt$ and $\hshmt$.
 Differences in standard errors, which may result from different monitoring effort between years, are explicitly considered in the analysis.
@@
 The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.
-The random effect is assumed a normal distribution with mean 0 and unknown variance, $\sbs$, to be estimated.
+The random effect follows a normal distribution with mean 0 and variance, $\sbs$, to be estimated.
 
@@
 Results from *mixmeta* have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
-Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with some variance components fixed and others estimated ($\sbs$).
-The default estimation method for meta-analysis models is restricted maximum likelihood (REML).
-
-This method was applied to a chosen sequence or "window" of years for estimating the linear trend.
+Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with fixed and estimated ($\sbs$) variance components.
+The default estimation method for *mixmeta*, restricted maximum likelihood (REML), was used.
+The meta-analysis models were applied to a chosen sequence or "window" of years for estimating the linear trend.
 This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (log$_{10}$ slope) in chl-a over a chosen series of years.
-Using the meta-analysis considered statistical properties derived from all the data in the years of interest (i.e., the seasonal averages and their uncertainty).
-Years with more samples or less variability have smaller standard errors for $\hmt$ compared to years with less samples or more variability.
-The meta-analysis explicitly includes these differences in uncertainty magnitudes.
 
@@
 
-The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year blocks from 1991 to 2000, 2000 to 2010, and 2010 to 2019.
-These temporal separations are relevant to phytoplankton bloom phenology in SFE [@Cloern20].
-As an extension of these methods, a chosen sequence of years was applied across the time series to assess periods of time within which seasonal trends could be significant or not (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
-Although the initial year sequence and seasonal period to evaluate is chosen by the analyst, applying the method in a moving window approach reduces some of the ambiguity around points in the time series when trends may be changing with reasonable certainty.
-The moving window approach applies the meta-analysis model from left to right in the time series across the seasonal averages to obtain the slope and significance of the estimated trend.
-We chose a centered ten-year window where the model estimates are based on results at equal half-window widths to the left and right of a given year.
-Although a left- or right-centered window could also be applied, we limit the analysis to a centered window to demonstrate the concept.
-
-Finally, trend results from the meta-analysis regression method for each season and different time periods were compared to alternative trend analyses to demonstrate how different and potentially misleading conclusions about trends can arise from methods that insufficiently account for propagation of uncertainty.
-As stated above, meta-analysis allows for full consideration of uncertainty in trend assessments by explicitly incorporating standard error of averages from results obtained from the GAMs and it is expected that more generalized methods that do not account for uncertainty may lead to different conclusions.
-Moreover, the meta-analyses depend on GAM predictions to describe an estimated long-term signal in the observed time series.
-Trends assessed on observed data may include substantial noise independent of any "canonical" signal derived from GAMs.
-Trend estimates from meta-analysis applied to GAM seasonal averages were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages.
+The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year moving windows from 1991 to 2019.
+The choice of water-year seasons are relevant to phytoplankton bloom phenology in the SFE [@Cloern20].
+The moving-window approach applied the meta-analysis to each decadal window (e.g., 1991-2001, 1992-2002, etc.), allowing changes in slope and its significance to be interpreted as the window is shifted one year at a time.
+We interpret the slope as representative for the central year for each block, but a predictive trend for the final year of the window could also be interpreted.
+For some results, we focus on the windows 1990-2000, 200-2010, and 2010-2019.
+
+Finally, trend results from the meta-analysis regression method for each season and different time periods were compared "naive" across-year regressions that do not propagate uncertainty to demonstrate how different and potentially misleading conclusions can be obtained.
+Trend estimates were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages.
 Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods.
 This analysis was then applied to all stations.
-
-### Back-transformation of model results
-
-Model results were back-transformed from log$_{10}$-space to aid in the interpretation of trends [@Bradu70;@Duan83].
-Back-transformation was accomplished using equation \@ref(eq:backtrans) for model predictions, estimates of mean values, and endpoints of confidence intervals from GAM results, such that:
-
-\begin{equation}
-E[y] = 10^{\mu + \frac{\log(10) \sigma^2}{2}}
-(\#eq:backtrans)
-\end{equation}
-
-where the back-transformed, expected value of the response variable $y$ (chl-a) is a function of the mean value $\mu$ in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.
-The dispersion is obtained directly from the fitted GAM object as an estimate of the variance of the residuals divided by the model degrees of freedom.
-
-Slope estimates that were used to determine significance of a seasonal trend across years from the meta-analysis were not back-transformed.
-While it is possible to back-transform model predictions, estimates of the mean values, and endpoints of confidence intervals, it is not possible for slope estimates in log$_{10}$-space that can change in arithmetic space across the values of the independent variable.
-As such, all results reporting slope estimates as rates of change in chl-a per year were reported at the scale of the model in log$_{10}$-space.
-
-### wqtrends R package
-
-The wqtrends R package developed by the authors includes all methods described above [@Beck21].
-The package is available for download at https://tbep-tech.github.io/wqtrends.
-A full vignette describing installation and use is also available in the link provided.
 
@@
 Chl-a measurements can also be used to estimate gross primary production to assess process rates that may be more indicative of system function [@Jassby02;@Cloern07].
-The open-source wqtrends R package [@Beck21] developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in SFE and other estuarine systems.
+The open-source wqtrends R package [@Beck21] developed for this manuscript can be used for these analyses to provide additional insight into potential drivers of water quality change in the SFE and other estuarine systems.
 

### 2021-03-10 supplement.Rmd
intro methods revision done from PDV

@@
 
-A note about the `select` argument in `mgcv`: When `select = TRUE` is included, the comparison between models S and SY changes.
+# Comments on the `select` argument in `mgcv
+
+ When `select = TRUE` is included, the comparison between models S and SY changes.
 This option tells `mgcv` to penalize the coefficient of the linear terms in the spline.
@@
 This result would be an artifact of the choice in model SY to include a linear trend in `cont_year` both as a separate term and as part of the spline, with the latter subject to penalization.
+
+# Back-transformation of model results
+
+Model results were back-transformed from log$_{10}$-space to aid in the interpretation of trends [@Bradu70;@Duan83].
+Back-transformation was accomplished using equation \@ref(eq:backtrans) for estimates of mean values and endpoints of confidence intervals from GAM results, such that:
+
+\begin{equation}
+E[y] = 10^{\mu + \frac{\log(10) \sigma^2}{2}}
+(\#eq:backtrans)
+\end{equation}
+
+where the back-transformed, expected value of the response variable $y$ (chl-a) is a function of the predicted value $\mu$ (mean or confidence interval endpoint) in log$_{10}$-space and a dispersion estimate $\sigma^2$ from the model.
+The dispersion is the residual variance estimated from the GAM fit.
 

### 2021-03-10 manu_draft.Rmd
halfway through discussion edits

@@
 
-## Model performance and observed results
+## Model performance and predictions
 
@@
 Model predictions for chl-a trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
-GAM predictions from north to south on the longitudinal axis showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figures S1-S9).
+GAM predictions from north to south on the showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figures S1-S9).
 All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at southern stations, and 3) a fall chl-a peak that was smaller than the spring peak.
@@
 
-## Trend estimates
-
-To demonstrate the results that can be obtained with the trend tests, estimates of linear trends across years for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
-All plots show trends in seasonal averages within ten-year windows, whereas plots a-c show estimates from January to June and plots d-f show estimates from July to December.
+## Inter-annual trend estimates
+
+Estimates of linear trends across years in seasonal averages across roughly ten-year windows for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
+Plots a-c show trends in January to June averages while plots d-f show trends in July to December averages.
 The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) from 1991 to 2000, whereas a trend for the same period in July to December was not observed.
@@
 Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.
-Figure \@ref(fig:trndmap) shows results from similar analyses as those in figure \@ref(fig:trnddat), but applied to all stations.
-Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010, whereas chl-a decreased at stations 30 and 32 from 2010 to 2019.
+Figure \@ref(fig:trndmap) shows results from similar analyses as those in Figure \@ref(fig:trnddat), but applied to all stations.
+Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010; decreases were observed at stations 30 and 32 from 2010 to 2019.
 For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
-
-## Trend comparisons
 
@@
 
-Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from observed data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates (row 2).
+## Importance of uncertainty propagation
+
+Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2).
 Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
@@
 In both cases, only the meta-analysis results provide accurate trend estimates because of full propagation of uncertainty across methods.
+Even in cases where the p-value threshold is not of interest, the confidence intervals from the alternative methods will be inaccurate.
 
@@
 
-The use of the GAM results with second and third stage analyses to assess trends is a new approach that previous trend analysis methods cannot sufficiently address.
-Although the development of our approach was motivated by specific questions regarding assessment of seasonal changes over time, the application has advantages over more conventional methods that inadequately account for time series characteristics of water quality data from long-term monitoring programs.
-In particular, missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design [@Junninen04;@Racault14].
+Propagation of uncertainty from within-year features of estimated GAMs to across-year trends using mixed-effects meta-analysis is a new approach that can address different questions than previous methods.
+Our approach has several advantages over more conventional approaches for water quality data from long-term monitoring programs.
+GAMs are capable of modelling data with missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design [@Junninen04;@Racault14].
 As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data [@Hirsch82;@Helsel20].
-These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations.
+These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations if the raw data are aggregated to meet test requirements.
 Aggregation of raw data, e.g., averaging of observations within a year or season to comply with the requirements of Kendall tests, risks loss of information by removing variation between observations at smaller time scales.
-The logical outcome is increased likelihood of incorrect conclusions from test results.
-
-Our results demonstrated that model structure (i.e., types of smoothers) was less important than allowing the model sufficient freedom to estimate the trends.
-This is an important conclusion that provides guidance on how GAMs could be used to model time series from long-term water or other environmental monitoring programs when the purpose is to identify and describe which of many possible changes over time have occurred.
+The logical outcome is increased risk of incorrect conclusions from test results.
+
+Our results demonstrated that GAM structure (i.e., choice of smoothing terms) was less important than allowing the model sufficient freedom to fit the data.
+This is an important conclusion that provides guidance on how GAMs could be used to model time series from long-term environmental monitoring programs.
 Models with separate smoothers for continuous year and day of year can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
 This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
-Specifying an upper limit for the number of knots that can potentially be used to fit different smoothers is critical to this approach.
-A smaller limit can lead to under-smoothing and an insufficient characterization of trends that risks incorrect conclusions from using these models with the second and third stage of analyses.
-
-Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate how different conclusions can be obtained if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
-These comparisons demonstrate a common approach where raw observations may be aggregated prior to use of more conventional trend analysis methods.
+
+Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate the risk of incorrect conclusions if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
+This may be a likely scenario when aggregating observations to meet the assumptions of more conventional trend analysis methods.
 Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
-In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data.
-A false negative result (Type II) was shown in Figure \@ref(fig:trndcmpex)a where a trend was not shown for the OLS regression results, but was observed for the meta-analysis regression results.
-Conversely, a false positive result (Type I) was shown in Figure \@ref(fig:trndcmpex)b where the OLS regression results showed a trend, but the meta-analysis results did not.
-The potential danger of inflated error rates was further demonstrated at a larger spatial scale across all stations in Figure \@ref(fig:trndcmp).
+In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
+Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.
+Incorrect conclusions can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01].
 

### 2021-03-12 manu_draft.Rmd
edits from pv, rm, mf done

@@
 Accurate quantification of trends must consider variation at different temporal scales regardless of the question of interest, such that ignoring variation at one scale could lead to incorrect conclusions about variation at another scale.
-Many environmental monitoring programs collect temporally resolved but irregular time-series data to quantify trends for regulatory, management, or research purposes.
+Many environmental monitoring programs collect temporally resolved but irregular time series data to quantify trends for regulatory, management, or research purposes.
 The mismatch between the scales of monitoring versus analysis questions or management goals can present statistical challenges [@Urquhart98;@Cumming06;@Forbes18].
@@
 
-GAMs have been applied previously to evaluate trends in water quality time-series from long-term monitoring programs [@Haraguchi15;@Murphy19].
+GAMs have been applied previously to evaluate trends in water quality time series from long-term monitoring programs [@Haraguchi15;@Murphy19].
 For example, @Murphy19 used GAMs to decompose water quality time series from Chesapeake Bay into long-term and seasonal trends [@Murphy19] and test trend hypotheses between two points in time.
-Other studies of environmental time-series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17].
+Other studies of environmental time series with GAMs have addressed the use of transformed response data [@Yang20], serial correlation in high resolution data [@Morton08;@Yang20], and quantifying time lags in relationships between response and predictor variables [@Lefcheck17].
 The study herein generalizes the approach to analyzing trends of seasonal spline features, describes the relationships among alternative spline formulations when spline flexibility is allowed to vary [@Wood03;@Wood17] rather than being constrained *a priori* for different time scales, and prioritizes full incorporation of uncertainty.
@@
 Approximately twice-monthly monitoring has been conducted for several decades at fixed locations (stations) on the longitudinal axis of the Bay.
-Analysis of these data is complicated by irregularities in timing and consistency of data collection which can generate artifacts affecting simple seasonal averages of the data.
+Analysis of these data is complicated by irregularities in timing and consistency of data collection, which can generate artifacts affecting simple seasonal averages of the data.
 We were interested in questions such as: Are there significant trends in spring mean chlorophyll at multi-year time-scales?
@@
 Its watershed covers 200 thousand km$^2$ in the US state of California.
-Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay, where the combined inflow from both rivers is approximately 28 km$^3$ per year.
-Salinity ranges from 0 to 15 ppt in the northern subembayments and from 5 to 35 ppt in southern subembayments closer to the Pacific Ocean, depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff.
+Major freshwater inputs enter the system through the Sacramento-San Joaquin Delta complex upstream of Suisun Bay.
+Salinity ranges from 0 to 15 ppt in the northern subembayments and from 5 to 35 ppt in southern subembayments closer to the Pacific Ocean, depending on the tidal cycle, effluent discharge from wastewater treatment plants, and stormwater runoff [@Cloern12].
 An estimated 73.8 metric tons dy$^{-1}$ of inorganic nitrogen are discharged into the Bay, primarily from wastewater [@Novick14].
@@
 Nitrogen and phosphorus levels in the SFE usually exceed concentrations that cause eutrophication in other estuaries.
-However, the SFE has demonstrated resistance to eutrophication, which has been attributed to high concentrations of suspended sediments that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves [@Cole84;@Alpine88;@Jassby08;@Kimmerer14;@Lehman17].
-The Regional Water Quality Control Board has showed renewed interest in understanding the potential for nutrient loading to negatively affect water quality for more southern areas of the SFE, where harmful algal blooms, elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
+However, the SFE has demonstrated resistance to eutrophication, which has been attributed to high concentrations of suspended sediment that reduce light penetration in the water column, low residence time caused by vigorous river flushing, and removal of primary producers by abundant suspension feeding bivalves [@Cole84;@Alpine88;@Jassby08;@Kimmerer14;@Lehman17].
+The Regional Water Quality Control Board has showed renewed interest in understanding the potential for nutrient loading to negatively affect water quality for more southern areas of the SFE where harmful algal blooms, elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
 Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient particularly with respect to seasonal differences between years.
@@
 For this example, the seasonal averages were extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
-Third, we use a mixed effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
+Third, we use a mixed-effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
 While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
@@
 We considered four different GAMs to smooth the raw data across time.
-Although they can achieve similar fits, they do so by partitioning variation in the time-series differently (Table \@ref(tab:modsumtab)).
+Although they can achieve similar fits, they do so by partitioning variation in the time series differently (Table \@ref(tab:modsumtab)).
 We discuss all four to clarify their relationships and interpretations.
@@
 
-where `y` is the time-series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case as a function of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
+where `y` is the time series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case as a function of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
 
@@
 The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.
-However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.
+However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time series.
 Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually.
@@
 The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.
-Each of these estimates are from on approximate multivariate normality of the sampling distribution of $\hbb$.
+Each of these estimates are from the approximate multivariate normality of the sampling distribution of $\hbb$.
 
@@
 For example, is there a trend in the spring chl-a average from 1990 to 2000?
-This question can also be posed in a moving-window manner across a time-series (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
+This question can also be posed in a moving-window manner across a time series (e.g., spring average trend from 1990-2000, 1991-2001, etc.).
 For all analyses, the response data of interest are $\hmt$, $t = 1, \ldots, N$, with their associated standard errors, $\hshmt$.
@@
 
-A mixed effects meta-analysis model can estimate linear trends when each observation has an associated standard error, which is the case with our estimates $\hmt$ and $\hshmt$.
+A mixed-effects meta-analysis model can estimate linear trends when each observation has an associated standard error, which is the case with our estimates $\hmt$ and $\hshmt$.
 Differences in standard errors, which may result from different monitoring effort between years, are explicitly considered in the analysis.
@@
 Results from *mixmeta* have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
-Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed effects model with fixed and estimated ($\sbs$) variance components.
+Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed-effects model with fixed and estimated ($\sbs$) variance components.
 The default estimation method for *mixmeta*, restricted maximum likelihood (REML), was used.
@@
 
-Finally, trend results from the meta-analysis regression method for each season and different time periods were compared "naive" across-year regressions that do not propagate uncertainty to demonstrate how different and potentially misleading conclusions can be obtained.
-Trend estimates were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from observed data and 2) trends from OLS regression applied to GAM seasonal averages.
+Finally, trend results from the meta-analysis regression method for each season and different time periods were compared to "naive" across-year regressions that do not propagate uncertainty to demonstrate how different and potentially misleading conclusions can be obtained.
+Trend estimates were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from the raw data and 2) trends from OLS regression applied to GAM seasonal averages.
 Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods.
@@
 Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2).
-Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for April to June averages from 1991 to 2000.
+Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for January to July averages from 1991 to 2000.
 Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).
 The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.
-Figure \@ref(fig:trndcmpex)b shows trend estimates for station 22 for October to December averages from 2000 to 2010.
-Unlike the first example, the top two figures show trends, whereas the bottom plot for the meta-analysis regression does not show a trend.
+Figure \@ref(fig:trndcmpex)b shows trend estimates for the same station for July to December averages from 2000 to 2010.
+Unlike the first example, only the top figure shows a trend, whereas the bottom two plots do not show trends.
 In both cases, only the meta-analysis results provide accurate trend estimates because of full propagation of uncertainty across methods.
@@
 Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).
-The slope estimates from the OLS models applied to the observed data were understandably more variable than the slope estimates from the GAM averages and meta-analysis methods, with much larger slopes observed especially at the more southern stations in the January to June period.
-Slope estimates from the OLS models applied to the averages from the GAMs as compared to the meta-analysis results were more similar, excluding some of the slope estimates for the southern stations.
-Differences in significance of trends between the OLS models applied to the GAM averages and the meta-analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
+In many cases, the slope estimates were similar in magnitude, with some exceptions at the more southern stations where the OLS estimates suggested a larger trend than the meta-analysis methods.
+More importantly, differences in the magnitude of the confidence intervals between the OLS models applied to the GAM averages and the meta-analyses were also observed, reflecting the ability of the latter to more accurately assess significance of trends by accounting for uncertainty in the average estimates.
 
@@
 
+Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01].
+Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
+This may be a likely scenario when aggregating observations to meet the assumptions of more conventional trend analysis methods.
+Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
+In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
+Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.
+
 Our results demonstrated that GAM structure (i.e., choice of smoothing terms) was less important than allowing the model sufficient freedom to fit the data.
@@
 
-Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate the risk of incorrect conclusions if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
-This may be a likely scenario when aggregating observations to meet the assumptions of more conventional trend analysis methods.
-Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
-In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
-Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.
-Incorrect conclusions can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01].
-
-Variability of sampling effort among years is a common characteristic of environmental monitoring data, which further increases the potential for incorrect conclusions when existing methods have been used.
-Averages may be skewed in a particular direction if annual estimates are based on a handful of observations from select months [e.g., summer only, @Fouquet12].
-The use of GAMs to fit the long-term trend will reduce the potential of limited observations in a particular year skewing estimates of annual or seasonal averages.
-More importantly, limited observations in a year will be reflected in the standard error estimates derived from the GAM, which has direct implications for how uncertainty is treated in the mixed-effects meta-analyses [@Sera19].
-As a result, trend assessments from meta-analyses are more accurate for data with unequal sampling effort where other methods may fail by providing estimates that are biased or have inaccurate or incomplete estimates of uncertainty.
-
 The underlying cross-validation methods used by GAMs in the mgcv package also reduce the decisions that may be necessary for the implementation of alternative trend assessment methods.
 For example, WRTDS and similar smoothing approaches (e.g., LOESS) require decisions on appropriate window widths or bandwidths to define the neighborhood of observations for smoothing [@Hirsch10;@Wan17].
-Prior to the implementation of generalized cross-validation in GAMs, these decisions were also required, producing fitted results that are arbitrarily under- or over-smoothed to the raw observations.
 This is especially problematic for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.
-Because these decisions are no longer needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.
+Because these decisions are not needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.
 
@@
 
-Additional work could be conducted to further strengthen the validity of conclusions based on trends from meta-analysis regression applied to the GAM seasonal averages.
-We acknowledge that the third stage analyses require explicit user inputs on year periods to define trends.
+Additional work could be conducted to further strengthen the conclusions based on trends from meta-analysis regression applied to the GAM seasonal averages.
+Our third stage analyses require *a priori* decisions on long-term time scales of interest and future work could generalize these choices.
 Although there are undoubtedly many scenarios where years of interest can be chosen objectively by the needs of an analysis (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.
-For example, trends could be assessed based on five or ten year moving windows, but which result should be used to inform decisions?
-There may be additional methods to develop using objective criteria to more accurately identify inflection points or other important periods where changes occur independent of a user choice.
+Additional methods could be developed using objective criteria to identify inflection points or other important periods where changes occur independent of a user choice.
 Assessing water quality changes beyond an evaluation of seasonal averages could also be possible with our approach, such as assessing changes in the timing or magnitude of a seasonal peak across years.
 
-The trend assessment approach does not address causality in the observed trends.
-In this regard, the approach is like other trend assessment methods where the focus is on understanding direction, magnitude, and confidence in changes in water quality variables over time.
-However, a logical follow-up question is "what" is driving the trend after the trend has been adequately described.
+Additional explanatory variables could be identified that may be associated with the trend after the trend has been adequately described.
 This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.
 An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09].
-In such cases, considerations of model structure can have direct implications on conclusions of potential causality.
-As noted above, model structure in describing the long-term trend component was irrelevant for describing long-term trends, although this will depend on the objective of the analysis.
+In such cases, considerations of model structure can have direct implications on conclusions.
 Our goal was to describe chl-a changes relative to time, where the predictors were variations on a general theme (e.g., season vs.
 year).
-This is a different application from using GAMs with predictors selected to assess potential causality.
-Therefore, using our approach to assess causality will require considerations of model structure given how GAMs could be used to assess different questions.
+This is a different application from using GAMs with predictors selected to explain thos changes over time.
+Therefore, using our approach to evaluate explanatory variables will require considerations of model structure given how GAMs could be used to assess different questions.
 
@@
 ```
-```{r modsumfig, fig.cap = "GAM output of estimated chl-a for models SY, SYD, and SYDI.
+```{r modsumfig, fig.cap = "GAM output of estimated chl-a at station 32 for models SY, SYD, and SYDI.
 Model S is identical to SY and is not shown.
@@
 Plots in (b) show pairwise comparisons of predicted chl-a between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.
-The plots demonstrate that results between the models are comparable except for a few observations at extreme values (a, b), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (c).
+The plots demonstrate that results between the models are similar except for a few observations at extreme values (a, b), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (c).
 The 1:1 lines are in red to facilitate comparisons."}
@@
 ```{r trnddat, fig.cap = "Examples of seasonal averages and trend estimates in ten year blocks from meta-analyses using results of GAM predictions for station 34.
-Plots (a), (b), and (c) show trend estimates for January through June and (d), (e), and (f) show trend estimates for July through December.
-The trend lines estimate the rate of change of chl-a per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the plot title.
-ns: not significant at $\\alpha$ = 0.05, * p < 0.05, ** p < 0.005"}
+Plots (a), (b), and (c) show trend estimates for January to June averages and (d), (e), and (f) show trend estimates for July to December averages.
+The trend lines estimate the rate of change of chl-a per year, reported as the log$_{10}$-slope (+/- 95 % confidence interval) in the sub-plot titles.
+ns: not significant at $\\alpha$ = 0.05, * p < 0.05"}
 knitr::include_graphics('figs/trnddat.png')
@@
 ```
-```{r trndmap, fig.cap= "Trend estimates across seasons by decade for chl-a at each station.
+```{r trndmap, fig.cap= "Interannual trend estimates of seasonal averages by decade for chl-a at each station.
 Point type and color represent the direction and magnitude of an estimated trend as the log$_{10}$ slope for chl-a concentration per year.
@@
 Plots in (a) show estimates for seasonal averages from January to June and plots in (b) show estimates for seasonal averages from July to December.
-Results are from a ten-year, centered moving window where each point shows a linear trend estimate from five years prior to after each year.
+Results are from a ten-year, centered moving window where each point shows a linear trend estimate from five years prior to five years after each year.
 Estimates prior to 1996 and after 2014 are not available because of an incomplete ten year record for estimating the trend.
-Estimates in a year that are significant are shown in red.'}
+Significant estimates are shown in red.'}
 knitr::include_graphics('figs/winchg.png')
@@
 
-p1 <- show_cmpplo('data/modslog_chl36.RData', doystr = 91, doyend = 181, yrstr = 1991, yrend = 2000, 
-                  ttl =  '(a) Station 36, Apr-Jun chl. averages', ymax = NULL)
-p2 <- show_cmpplo('data/modslog_chl22.RData', doystr = 274, doyend = 364, yrstr = 2000, yrend = 2010, 
-                  ttl =  '(b) Station 22, Oct-Dec chl. averages', ymax = NULL, ylab = NULL)
+p1 <- show_cmpplo('data/modslog_chl36.RData', doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
+                  ttl =  '(a) Station 36, Jan-Jul chl. averages', ymax = NULL)
+p2 <- show_cmpplo('data/modslog_chl36.RData', doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
+                  ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL)
 # p3 <- show_cmpplo('data/modslog_chl34.RData', doystr = 274, doyend = 364, yrstr = 2010, yrend = 2019, 
@@
 ```
-```{r trndcmpex, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at  (a) station 36 and (b) station 22.
-The first row shows OLS (ordinary least squares) regression applied to annual averages of chl-a from the observed data, the second row shows OLS regression applied to annual averages of chl-a from the GAM, and the third row shows meta-analysis regression applied to the annual averages of chl-a from the GAM.
-Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
+```{r trndcmpex, fig.cap = 'Trend estimate comparisons (arithmetic scale) for three models applied to seasonal averages of chl-a in different annual periods at station 36.
+The first row shows OLS (ordinary least squares) regression applied to seasonal averages of chl-a from the raw data, the second row shows OLS regression applied to seasonal averages of chl-a from the GAM (without error propagation), and the third row shows meta-analysis regression applied to the seasonal averages of chl-a from the GAM.
+Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.
+These examples on selected periods of time show that slope estimates can be similar, but the confidence intervals vary.'}
 knitr::include_graphics('figs/trndcmpex.png')
@@
 toplo <- cmptrnd
-p <- ggplot(toplo, aes(y = station, x = yrcoef, shape = modtyp, colour = pval)) + 
+p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
   facet_grid(seas ~ yrs, scales = 'free_x') + 
-  geom_point(size = 3, position = position_dodge(width = 0.5), alpha = 0.6) + 
-  scale_shape('Trend model') + 
-  scale_colour_discrete_diverging(name = 'Trend', palette = 'BlueRed2') +
+  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
+  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
+  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
+  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
+  scale_shape(name = 'Trend model') + 
   geom_vline(xintercept = 0, linetype = 'dashed') + 
   # scale_x_continuous(limits = c(-1.14, 1.14)) +
+  guides(
+    color = guide_legend(override.aes = list(linetype = 0))
+  ) +
   theme_minimal() + 
@@
 
-png('figs/trndcmp.png', height = 4.5, width = 7.5, family = 'serif', units = 'in', res = 400)
+png('figs/trndcmp.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
 p
@@
 ```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station.
-The "OLS observed" trend model is based on an ordinary least squares (OLS) regression fit to the annual averages of chl-a from the observed data, the "OLS GAM" trend model is based on an OLS regression fit to the annual averages of chl-a from the GAM model, and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the annual averages of chl-a from the GAM model.
-Values for each model are the log$_{10}$-slope estimates as annual change per year within each season, with color denoting trend significance.'}
+The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.
+Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with color denoting trend significance.'}
 knitr::include_graphics('figs/trndcmp.png')
@@
     'A linear continuous year variable and a single smoother over a continuous year variable', 
-    'A linear continuous year variable, a smoother over a continuous year variable, and smoother over a day of year variable',
-    'A linear continuous year variable, a smoother over a continuous year variable, smoother over a day of year variable, and an interaction smoother across continuous year and day of year variables')
+    'A linear continuous year variable, a smoother over a continuous year variable, and a smoother over a day of year variable',
+    'A linear continuous year variable, a smoother over a continuous year variable, a smoother over a day of year variable, and an interaction smoother across continuous year and day of year variables')
   )
@@
 
+    AIC <- AIC(x)
     GCV <- as.numeric(x$gcv.ubre)
@@
 
-    out <- data.frame(GCV = GCV, R2 = R2, smooths)
+    out <- data.frame(AIC = AIC, GCV = GCV, R2 = R2, smooths)
 
@@
   mutate(
+    AIC = ifelse(duplicated(name), '', AIC),
     GCV = ifelse(duplicated(name), '', GCV),
@@
 The four models provide either identical or comparable ability to describe chl-a trends at an example station (32) in the southern end of the San Francisco Estuary.
-The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are comparable.
-GCV: generalized cross-validation score, R2: r-squared values for predictions, edf: effective degrees of freedom, F: F-statistic, p-val: probability value, ** p < 0.001'
+The models differ in additive smoothers and the amount of effective degrees of freedom (edf) in the smoothers (measure of wiggliness in each component), but the overall model predictions are similar.
+AIC: Akaike Information Criterion, GCV: generalized cross-validation score, R2: r-squared values for predictions, edf: effective degrees of freedom, F: F-statistic, p-val: probability value, ** p < 0.001'
 
@@
 
-cap <- 'Model performance statistics for each station as generalized cross-validation scores (GCV) and r-squared values.'
+cap <- 'Model performance statistics for each station as Akaike Information Criterion scores (AIC), generalized cross-validation scores (GCV), and r-squared values.'
 

### 2021-03-12 R/dat_proc.R

@@ -75,7 +75,6 @@ modprf <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   ungroup %>% 
   select(-data) %>% 
   unnest('prf') %>% 
-  select(-AIC) %>% 
   mutate(
     value = gsub('^data/modslog\\_chl|\\.RData$', '', value)
   ) %>% 
@@ -298,28 +297,21 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   select(-avgseas) %>% 
   gather('modtyp', 'est', metatrnd, lmtrnd, obstrnd) %>% 
   mutate(
-    yrcoef = purrr::pmap(list(mod, modtyp, est), function(mod, modtyp, est){
+    coefs = purrr::map(est, function(x){
       
-      if(modtyp %in% c('metatrnd', 'lmtrnd')){
-        
-        dispersion <- summary(mod)$dispersion
-        bt_prd <- 10 ^ (predict(est) + log(10) * dispersion / 2)
-        df <- data.frame(chl = bt_prd, yr = est$model$yr)
-        slope <- lm(chl ~ yr, df) %>% summary %>% coefficients %>% .[2, 1]
-        
-      }
+      summary(x)$coefficients
       
-      if(modtyp %in% 'obstrnd'){
-        
-        s2 <- (summary(est)$sigma)^2
-        bt_prd <- 10 ^ (predict(est) + log(10) * s2 / 2)
-        df <- data.frame(chl = bt_prd, yr = est$model$yr)
-        slope <- lm(chl ~ yr, df) %>% summary %>% coefficients %>% .[2, 1]
-        
-      }
-      
-      return(slope)
-      
+    }),
+    yrcoef = purrr::map(coefs, function(x){
+      x[2, 1]
+    }),
+    yrcoefupr = purrr::pmap(list(coefs, yrcoef), function(coefs, yrcoef){
+      se <- coefs[2, 2]
+      yrcoef + 1.96 * se
+    }),
+    yrcoeflwr = purrr::pmap(list(coefs, yrcoef), function(coefs, yrcoef){
+      se <- coefs[2, 2]
+      yrcoef - 1.96 * se
     }),
     pval = purrr::map(est, function(x){
       
[…]

### 2021-03-23 manu_draft.Rmd
additional minor edits added from MF, RM, PV

@@
 Here, we propose a novel application of generalized additive models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.
-GAMs can model a response variable as the additive sum of one or more smoothing splines fit to predictor variables.
-These models have shown promise in water quality trend analysis to separate long-term (i.e., annual or decadal) trends from seasonal variation.
+GAMs have shown promise in water quality trend analysis to separate long-term (i.e., annual or decadal) trends from seasonal variation.
 Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-effects meta-analysis regression to quantify inter-annual trends that account for full propagation of error across methods.
@@
 
-Accurate quantification of trends must consider variation at different temporal scales regardless of the question of interest, such that ignoring variation at one scale could lead to incorrect conclusions about variation at another scale.
+Accurate quantification of trends must consider variation at different temporal scales when ignoring variation at one scale could lead to incorrect conclusions about variation at another scale.
 Many environmental monitoring programs collect temporally resolved but irregular time series data to quantify trends for regulatory, management, or research purposes.
@@
 @Wan17 showed that non-parametric approaches have been the most commonly used methods in long-term water quality trend analysis despite critical limitations.
-For descriptive decomposition of long-term monitoring data, they assume seasonal patterns within years do not change, require regularly spaced or balanced data, only use time as a descriptor for trends, and do not estimate a model that could be useful for other purposes.
+For descriptive decomposition of long-term monitoring data, they assume seasonal patterns within years do not change, require regularly spaced or balanced data, do not include additional predictors to explain variation, and do not estimate a model that could be useful for other purposes.
 Thus, while these non-parametric approaches have some degree of robustness to assess magnitude and direction of trends, they apply only to narrow goals.
@@
 Standard error estimates of predictions from WRTDS are available through a block bootstrap approach applied to the model results [@Hirsch15].
-Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling as a *post-hoc* application to a previously fitted model.
-Alternative methods that include standard error estimates simultaneously with model output may be preferred for intensive or more iterative applications
-
-Generalized additive models (GAMs) are central to this paper and form the basis of our fourth and last method to separate fluctuations on different time scales.
+Although a useful addition to the original method [@Hirsch10], the approach requires extensive resampling using a previously fitted model.
+Alternative methods that include standard error estimates simultaneously with model output may be preferred for intensive or more iterative applications.
+
+Generalized additive models (GAMs) are central to this paper and form the basis of the fourth method to separate fluctuations on different time scales.
 GAMs combine one or more smoothing splines to model patterns in data and may be seen as generalizing the concepts behind STL and WRTDS [@He06;@Morton08;@Pearce11;@Haraguchi15;@Murphy19].
 The basis functions used to formulate GAMs can be customized based on expected patterns in the data.
-Examples include cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
+One example includes cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
 GAMs have added flexibility because they can include both parametric (e.g., linear or quadratic) components and non-parametric (spline) components.
@@
 These approaches are based on optimization of out-of-sample prediction error, which addresses a key concern around methods like WRTDS that do not have analogs for choosing optimal degrees of smoothing.
-GAMs can also produce comparable results similar to those provided by WRTDS [@Beck17] and have readily obtainable uncertainty estimates that do not require *post-hoc* resampling of model results.
+GAMs can also produce comparable results similar to those provided by WRTDS [@Beck17] and have readily obtainable uncertainty estimates.
 Further, GAMs have natural frequentist and Bayesian interpretations, are naturally extensible to include random effects (i.e., generalized additive mixed models or GAMMs), and have computationally efficient implementations [@Wood17].
@@
 At what across-year scale do summer-fall mean chlorophyll levels change?
-Are there significant across-year trends in the spring phytoplankton bloom or baseline chlorophyll concentrations during periods of low productivity between blooms?
+Is there a spatial difference in chlorophyll trends?
 We provide examples illustrating how these questions can be addressed using GAMs to estimate seasonal trends and evaluated between years using meta-analysis methods.
@@
 
-## GAM application
+## GAMs with uncertainty propagation
 
 We implemented our analysis in three stages.
-First, we use a GAM to estimate a smooth temporal pattern in the raw data along with its uncertainty.
-Second, we calculate a feature of interest from the estimated GAM, along with its propagated uncertainty.
+First, we used a GAM to estimate a smooth temporal pattern in the raw data along with its uncertainty.
+Second, we calculated a feature of interest from the estimated GAM, along with its propagated uncertainty.
 For this example, the seasonal averages were extracted, whereas other features could be the timing or magnitude of a seasonal peak, but those are not developed here.
-Third, we use a mixed-effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
+Third, we used a mixed-effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
 While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
@@
 
-where `y` is the time series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case as a function of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.
+where `y` is the time series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case as a function of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline.
 
@@
 While model SYD is not mathematically equivalent to models S and SY, it should produce nearly identical results.
-The `doy` spline in model SYD gives the average seasonal pattern and changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.
+The `doy` spline in model SYD gives the average within-year pattern and changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.
 
@@
 Following the rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.
-Model SYDI differs formulaically from model S to a greater extent than model SYD, but all of the splines use the same inputs to smooth the same data.
+Model SYDI differs from model S to a greater extent than model SYD, but all of the splines use the same inputs to smooth the same data.
 The univariate splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model S given the fewer knots that are available to the former.
@@
 However, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.
-The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year and that feature could be accounted for with this term, but for the SFE data and chl-a dynamics in many estuaries, bloom size typically varies between years.
+The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year, which is unlikely for the SFE data and chl-a dynamics in many estuaries because bloom size typically varies between years.
 Thus, the conceptual motivation for model SYDI and its practical application are not supported for this analysis.
@@
 
-We visually compare chl-a estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), S is identical to SY and is not shown).
-Models SY, SYD, and SYDI were fit to chl-a data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
+We visually compare chl-a estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), SY is identical to S and is not shown).
+Models S, SYD, and SYDI were fit to chl-a data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
 Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).
@@
 
-### Second-stage analysis: Uncertainty propagation from estimated GAMs to seasonal features
+### Second-stage analysis: Seasonal features with uncertainties
 
@@
 
-In the second-stage analysis, we estimated a seasonal average, such as the mean spring chl-a concentrations and the uncertainty in each year.
-We define $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
+In the second-stage analysis, we estimated a seasonal average, such as the mean spring chl-a concentrations, with the associated uncertainty in each year.
+We defined $\mu_t$ as the seasonal average in year $t$, $\hmt$ as an estimate of $\mu_t$, and $\hshmt$ as the estimated standard error of $\hmt$.
 The season includes $n$ days.
@@
 Vector $\hbb$ includes both fixed effect parameters and spline parameters, and $\bX$ contains columns corresponding to each.
-For example, using model SY, if a point estimate for chl-a is needed for a single day, given as `dec_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ in the remaining columns.
+For example, using model SY, if a point estimate for chl-a is needed for a single day, given as `cont_year` = $r$, then $\bX$ would have a row with $1$ in the first column (for the intercept parameter), $r$ (for the linear time trend) in the second column, and an evaluation of each spline basis function at $r$ in the remaining columns.
 The number of spline basis functions is related to the number of knots.
@@
 In stage three of the analysis, we used a meta-analysis method to evaluate linear trends across years of seasonal-average water quality, characterized by the within-year means ($\hmt$) and their standard errors ($\hshmt$) that we estimated in stage two of the analysis.
-This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average, where the time-scale of the seasonal average and duration of the across-year comparison is chosen by the investigator?
+This analysis provided a direct answer to the question: Is there a significant linear trend across a group of years in a seasonal average?
 For example, is there a trend in the spring chl-a average from 1990 to 2000?
@@
 Results from *mixmeta* have a similar interpretation as those from regression analysis, but parameter estimates and their standard errors incorporate the known standard errors of the response values.
-Following meta-analysis theory, $mixmeta$ evaluates this model as a linear mixed-effects model with fixed and estimated ($\sbs$) variance components.
 The default estimation method for *mixmeta*, restricted maximum likelihood (REML), was used.
 The meta-analysis models were applied to a chosen sequence or "window" of years for estimating the linear trend.
-This approach evaluated whether there was a significant linear trend and provided an estimated rate of change (log$_{10}$ slope) in chl-a over a chosen series of years.
 
@@
 
-The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximate ten year moving windows from 1991 to 2019.
-The choice of water-year seasons are relevant to phytoplankton bloom phenology in the SFE [@Cloern20].
+The above methods were applied to each station by evaluating changes in seasonal averages from January to June and July to December for approximately ten year moving windows from 1991 to 2019.
+The choice of within-year seasons are relevant to phytoplankton bloom phenology in the SFE [@Cloern20].
 The moving-window approach applied the meta-analysis to each decadal window (e.g., 1991-2001, 1992-2002, etc.), allowing changes in slope and its significance to be interpreted as the window is shifted one year at a time.
 We interpret the slope as representative for the central year for each block, but a predictive trend for the final year of the window could also be interpreted.
-For some results, we focus on the windows 1990-2000, 200-2010, and 2010-2019.
+For some results, we focus on the windows 1991-2000, 2000-2010, and 2010-2019.
 
@@
 
-Model predictions for chl-a trends across all stations had an average R-squared value equal to `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
-GAM predictions from north to south on the showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figures S1-S9).
+Model predictions for chl-a trends across all stations had an average R-squared value of `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
+GAM predictions from north to south showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figures S1-S9).
 All the models suggested 1) increasing chl-a from 1990 until 2005 to 2010, followed by decreasing chl-a until the end of the record in 2019, 2) a spring chl-a peak, particularly at southern stations, and 3) a fall chl-a peak that was smaller than the spring peak.
@@
 
-Estimates of linear trends across years in seasonal averages across roughly ten-year windows for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
+Estimates of linear trends in seasonal averages across roughly ten-year windows for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
 Plots a-c show trends in January to June averages while plots d-f show trends in July to December averages.
-The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) from 1991 to 2000, whereas a trend for the same period in July to December was not observed.
+The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) from 1991 to 2000, whereas July to December chl-a did not change significantly.
 Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
@@
 
-Temporal changes varied among regions of the Bay and were in fact estimated to be moving in opposite directions.
+Temporal changes varied among regions of the Bay and were more pronounced at southern stations.
 Figure \@ref(fig:trndmap) shows results from similar analyses as those in Figure \@ref(fig:trnddat), but applied to all stations.
@@
 Unlike the first example, only the top figure shows a trend, whereas the bottom two plots do not show trends.
-In both cases, only the meta-analysis results provide accurate trend estimates because of full propagation of uncertainty across methods.
+In both cases, only the meta-analysis results give reliable conclusions because of full propagation of uncertainty across methods.
 Even in cases where the p-value threshold is not of interest, the confidence intervals from the alternative methods will be inaccurate.
@@
 Propagation of uncertainty from within-year features of estimated GAMs to across-year trends using mixed-effects meta-analysis is a new approach that can address different questions than previous methods.
-Our approach has several advantages over more conventional approaches for water quality data from long-term monitoring programs.
-GAMs are capable of modelling data with missing observations or irregular sampling can complicate trend assessment and comparison of trends between locations that may differ in sampling design [@Junninen04;@Racault14].
+Our approach has several advantages over more conventional approaches for analysis of water quality data from long-term monitoring programs.
+GAMs are capable of modelling time series with missing observations or irregular sampling which can complicate trend assessment and comparison of trends between locations [@Junninen04;@Racault14].
 As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data [@Hirsch82;@Helsel20].
@@
 Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
-This may be a likely scenario when aggregating observations to meet the assumptions of more conventional trend analysis methods.
 Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
@@
 
-Our results demonstrated that GAM structure (i.e., choice of smoothing terms) was less important than allowing the model sufficient freedom to fit the data.
+Results here also show that GAM structure (i.e., choice of smoothing terms) was less important than allowing the model sufficient freedom to fit the data.
 This is an important conclusion that provides guidance on how GAMs could be used to model time series from long-term environmental monitoring programs.
 Models with separate smoothers for continuous year and day of year can produce nearly identical results in the predicted trends if the knots are sufficiently high to allow the GAMs to be fit as intended by the methods in the mgcv package (Figure \@ref(fig:modsumfig)).
-This approach leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
+The approach presented here leverages the ability of GAMs to objectively estimate smoothed trends across years by identifying an optimal level of smoothing using generalized cross-validation to extract an underlying signal in the observed data [@Wood04;@Wood17].
 
@@
 # use previously fitted list of models
-mods <- modstr[2:4]
+mods <- modstr[c(1, 3, 4)]
 
@@
 
-p1a <- show_prddoy(mods[[1]], ylab = 'SY: Chl-a (ug/L)', size = 0.8, alpha = 0.8) + 
+p1a <- show_prddoy(mods[[1]], ylab = 'S: Chl-a (ug/L)', size = 0.8, alpha = 0.8) + 
   guides(colour = ggplot2::guide_colourbar(barheight = 15, barwidth = 1)) +
@@
 
-p2a <- ggplot(toplo2, aes(x = SY, y = SYD)) + 
+p2a <- ggplot(toplo2, aes(x = S, y = SYD)) + 
   geom_point(alpha = alph) + 
@@
   thm
-p2c <- ggplot(toplo2, aes(x = SY, y = SYDI)) + 
+p2c <- ggplot(toplo2, aes(x = S, y = SYDI)) + 
   geom_point(alpha = alph) + 
@@
 
-p3a <- ggplot(toplo3, aes(x = SY, y = SYD)) + 
+p3a <- ggplot(toplo3, aes(x = S, y = SYD)) + 
   geom_point(alpha = alph) + 
@@
   thm
-p3c <- ggplot(toplo3, aes(x = SY, y = SYDI)) + 
+p3c <- ggplot(toplo3, aes(x = S, y = SYDI)) + 
   geom_point(alpha = alph) + 
@@
 ```
-```{r modsumfig, fig.cap = "GAM output of estimated chl-a at station 32 for models SY, SYD, and SYDI.
-Model S is identical to SY and is not shown.
+```{r modsumfig, fig.cap = "GAM output of estimated chl-a at station 32 for models S, SYD, and SYDI.
+Model SY is identical to S and is not shown.
 Plots in (a) show model predictions by day of year with separate lines for each year.
 Plots in (b) show pairwise comparisons of predicted chl-a between the models and plots in (c) show the same comparisons as in (b) but only for results from the estimated smoother for the `cont_year` variable.
-The plots demonstrate that results between the models are similar except for a few observations at extreme values (a, b), but they vary in the penalties applied to the basis functions for any particular smoother depending on which additive components are included in each model (c).
+The plots demonstrate that results between the models are similar except for a few observations at extreme values (a, b), but they vary in how they allocate contributions to the predictions among different additive splines (c).
 The 1:1 lines are in red to facilitate comparisons."}
@@
 The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.
-Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with color denoting trend significance.'}
+Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
 knitr::include_graphics('figs/trndcmp.png')
@@
 
-```{r sumtab, tab.cap = 'Station locations, sample sizes, and summary values (median, minimum, maximum) for chl-a ($\\mu$g L$^{-1}$) .  Rows are arranged from north to south.'}
+```{r sumtab, tab.cap = 'Station locations, sample sizes (from 1991 to 2019), and summary values (median, minimum, maximum) for chl-a ($\\mu$g L$^{-1}$).  Stations are arranged from north to south.'}
 totab <- rawdat %>% 

### 2021-03-31 manu_draft.Rmd
edits from AC

@@
 While model SYD is not mathematically equivalent to models S and SY, it should produce nearly identical results.
-The `doy` spline in model SYD gives the average within-year pattern and changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.
+The `doy` spline in model SYD gives the average within-year pattern and changes the interpretation of the `cont_year` spline to represent smoothed deviations from that pattern.
 
@@
 Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
-In many cases the results may be similar, but loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
+In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
 Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.
@@
 An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09].
-In such cases, considerations of model structure can have direct implications on conclusions.
+In such cases, considerations of model structure can have direct implications on conclusions given how GAMs could be used to assess different questions.
 Our goal was to describe chl-a changes relative to time, where the predictors were variations on a general theme (e.g., season vs.
 year).
-This is a different application from using GAMs with predictors selected to explain thos changes over time.
-Therefore, using our approach to evaluate explanatory variables will require considerations of model structure given how GAMs could be used to assess different questions.
+This is a different application from using GAMs with predictors selected to explain those changes over time.
+Therefore, using our approach to evaluate explanatory variables will require testing of different model structures.
 

### 2021-04-09 manu_draft.Rmd
adding sentence to intro
cls for ecolapp, cover letter, all figs as tiff

@@
 urlcolor: blue
-csl: ecology.csl
+csl: ecological-applications.csl
 link-citations: true
@@
 Such fluctuations may not be related to inter-annual trends or may not be well-suited to multi-scale smoothing methods.
+Many trend analysis methods lack the flexibility to evaluate one to many independent variables in an extendable structure that accommodates hypothesis testing at different time scales of interest.
 In this paper, we develop methods to estimate across-year trends of within-year features, such as seasonal averages, while accounting for uncertainties across analysis steps.
@@
 p <- p1 + p2 + plot_layout(ncol = 1)
-png('figs/obsdat.png', height = 7, width = 8, family = 'serif', units = 'in', res = 400)
+tiff('figs/obsdat.tiff', height = 7, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 print(p)
@@
 ```{r obsdat, fig.cap = "Observed chl-a concentrations for all stations in central and south San Francisco Estuary (18-36, Figure \\@ref(fig:sitemap)), with (a) annual summer/fall concentrations (Aug - Dec) and (b) monthly concentrations by decade."}
-knitr::include_graphics('figs/obsdat.png')
+knitr::include_graphics('figs/obsdat.tiff')
 ```
@@
 
-# save as png
-png('figs/sitemap.png', height = 6.5, width = 5, units = 'in', family = 'serif', res = 300)
+# save as tiff
+tiff('figs/sitemap.tiff', height = 6.5, width = 5, units = 'in', family = 'serif', res = 300)
 print(p)
@@
 Full dataset described in @Schraga20.'}
-knitr::include_graphics('figs/sitemap.png')
+knitr::include_graphics('figs/sitemap.tiff')
 ```
@@
 out <- p1 / p2 / p3
-png('figs/modsumfig.png', height = 7, width = 8, units = 'in', family = 'serif', res = 400)
+tiff('figs/modsumfig.tiff', height = 7, width = 8, units = 'in', family = 'serif', res = 600, compression = 'lzw')
 print(out)
@@
 The 1:1 lines are in red to facilitate comparisons."}
-knitr::include_graphics('figs/modsumfig.png')
+knitr::include_graphics('figs/modsumfig.tiff')
 ```
@@
 
-png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
+tiff('figs/trnddat.tiff', height = 6, width = 11, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 print(p)
@@
 ns: not significant at $\\alpha$ = 0.05, * p < 0.05"}
-knitr::include_graphics('figs/trnddat.png')
+knitr::include_graphics('figs/trnddat.tiff')
 ```
@@
 
-png('figs/trndmap.png', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 400)
+tiff('figs/trndmap.tiff', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 p
@@
 All results are from Model `S`."}
-knitr::include_graphics('figs/trndmap.png')
+knitr::include_graphics('figs/trndmap.tiff')
 ```
@@
 
-png('figs/winchg.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
+tiff('figs/winchg.tiff', height = 8, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 print(p)
@@
 Significant estimates are shown in red.'}
-knitr::include_graphics('figs/winchg.png')
+knitr::include_graphics('figs/winchg.tiff')
 ```
@@
 p <- (p1 |  p2) + plot_layout(ncol = 2)
-png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
+tiff('figs/trndcmpex.tiff', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 p
@@
 These examples on selected periods of time show that slope estimates can be similar, but the confidence intervals vary.'}
-knitr::include_graphics('figs/trndcmpex.png')
+knitr::include_graphics('figs/trndcmpex.tiff')
 ```
@@
 
-png('figs/trndcmp.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
+tiff('figs/trndcmp.tiff', height = 6, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 p
@@
 Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
-knitr::include_graphics('figs/trndcmp.png')
+knitr::include_graphics('figs/trndcmp.tiff')
 ```

### 2021-04-09 supplement.Rmd
cls for ecolapp, cover letter, all figs as tiff

@@
 
-  png(paste0('figs/prddat', sta, '.png'), height = 3.5, width = 9, family = 'serif', units = 'in', res = 400)
+  tiff(paste0('figs/prddat', sta, '.tiff'), height = 3.5, width = 9, family = 'serif', units = 'in', res = 600, compression = 'lzw')
   print(p)
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat18.png')
+knitr::include_graphics('figs/prddat18.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat21.png')
+knitr::include_graphics('figs/prddat21.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat22.png')
+knitr::include_graphics('figs/prddat22.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat24.png')
+knitr::include_graphics('figs/prddat24.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat27.png')
+knitr::include_graphics('figs/prddat27.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat30.png')
+knitr::include_graphics('figs/prddat30.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat32.png')
+knitr::include_graphics('figs/prddat32.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat34.png')
+knitr::include_graphics('figs/prddat34.tiff')
 ```
@@
 Observed data in are shown with the gray points."}
-knitr::include_graphics('figs/prddat36.png')
+knitr::include_graphics('figs/prddat36.tiff')
 ```

### 2021-04-09 manu_draft.Rmd
figs as tiff for submission, png in draft

--- 2021-03-31 manu_draft.Rmd
@@
 p <- p1 + p2 + plot_layout(ncol = 1)
+
+# save as tiff
+tiff('submission/Fig1.tiff', height = 7, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+print(p)
+dev.off()
+
+# save as png
 png('figs/obsdat.png', height = 7, width = 8, family = 'serif', units = 'in', res = 400)
@@
 
+# save as tiff
+tiff('submission/Fig2.tiff', height = 6.5, width = 5, units = 'in', family = 'serif', res = 600, compression = 'lzw')
+print(p)
+dev.off()
+
 # save as png
-png('figs/sitemap.png', height = 6.5, width = 5, units = 'in', family = 'serif', res = 300)
+png('figs/sitemap.png', height = 6.5, width = 5, units = 'in', family = 'serif', res = 400)
 print(p)
@@
 out <- p1 / p2 / p3
+
+# save as tiff
+tiff('submission/Fig3.tiff', height = 7, width = 8, units = 'in', family = 'serif', res = 600, compression = 'lzw')
+print(out)
+dev.off()
+
+# save as png
 png('figs/modsumfig.png', height = 7, width = 8, units = 'in', family = 'serif', res = 400)
@@
 
+# save as tiff
+tiff('submission/Fig4.tiff', height = 6, width = 11, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+print(p)
+dev.off()
+
+# save as png
 png('figs/trnddat.png', height = 6, width = 11, family = 'serif', units = 'in', res = 400)
@@
 
+# save as tiff
+tiff('submission/Fig5.tiff', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+p
+dev.off()
+
+# save as png
 png('figs/trndmap.png', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 400)
@@
 
+# save as tiff
+tiff('submission/Fig6.tiff', height = 8, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+print(p)
+dev.off()
+
+# save as png
 png('figs/winchg.png', height = 8, width = 8, family = 'serif', units = 'in', res = 400)
@@
 p <- (p1 |  p2) + plot_layout(ncol = 2)
+
+# save as tif
+tiff('submission/Fig7.tiff', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+p
+dev.off()
+
+# save as png
 png('figs/trndcmpex.png', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 400)
@@
 
+# save as tiff
+tiff('submission/Fig8.tiff', height = 6, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+p
+dev.off()
+
+# save as png
 png('figs/trndcmp.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)

### 2021-04-09 supplement.Rmd
figs as tiff for submission, png in draft

--- 2021-03-10 supplement.Rmd
@@
 
+  # # save as tiff
+  # tiff(paste0('figs/prddat', sta, '.tiff'), height = 3.5, width = 9, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+  # print(p)
+  # dev.off()
+
+  # save as png
   png(paste0('figs/prddat', sta, '.png'), height = 3.5, width = 9, family = 'serif', units = 'in', res = 400)

### 2021-04-09 manu_draft.Rmd
submitted to ecolapp

@@
 library(wqtrends)
-library(flextable)
 library(officer)
@@
 
-```{r sumtab, tab.cap = 'Station locations, sample sizes (from 1991 to 2019), and summary values (median, minimum, maximum) for chl-a ($\\mu$g L$^{-1}$).  Stations are arranged from north to south.'}
+```{r sumtab}
 totab <- rawdat %>% 
@@
   mutate_at(c('chl_medv', 'chl_minv', 'chl_maxv'), round, 1) %>% 
-  mutate_at(c('lat', 'lon'), round, 3)
-
-tab <- flextable(totab) %>% 
-  set_header_labels(
-    station = 'Station', 
-    lat = 'Latitude', 
-    lon = 'Longitude', 
-    chl_n = 'n', 
-    chl_medv = 'Med.', 
-    chl_minv = 'Min.', 
-    chl_maxv = 'Max.'
-  ) %>% 
-  # add_header_row(values = c('', '', '', 'Chl-a (ug/L)', '', '', '')) %>% 
-  # merge_at(i = 1, j = 4:7, part = 'header') %>% 
-  # colformat_num(j = c(5, 6, 7), digits = 1) %>%
-  # colformat_num(j = c(2, 3), digits = 3) %>% 
-  align(align = 'right', part = 'header') %>% 
-  border_remove() %>% 
-  hline_top(border = fp_border(color = 'black')) %>% 
-  font(fontname = 'Times', part = 'all') %>%
-  fontsize(size = 12, part = 'all') %>% 
-  width(j = c(2, 3), width = 0.9) 
-tab
+  mutate_at(c('lat', 'lon'), round, 3) %>% 
+  select(
+    Station = station, 
+    Latitude = lat, 
+    Longitude = lon, 
+    n = chl_n, 
+    `Med.` = chl_medv, 
+    `Min.` = chl_minv,
+    `Max.` = chl_maxv
+  )
+
+# table stuff
+cap.val <- 'Station locations, sample sizes (from 1991 to 2019), and summary values (median, minimum, maximum) for chl-a ($\\mu$g L$^{-1}$).  Stations are arranged from north to south.'
+
+tab <- totab
+
+# table
+knitr::kable(tab, booktabs = T, caption = cap.val)
 ```

### 2021-04-09 README.Rmd

@@
 
-[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions) last manuscript build `r Sys.time()`
-
+[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions)
 Materials for water quality trends assessment using GAMs
+[![DOI](https://zenodo.org/badge/249504727.svg)](https://zenodo.org/badge/latestdoi/249504727)
 

### 2021-04-12 README.Rmd
resubmit with compressed folders for package and manu

--- 2021-01-20 README.Rmd
@@
 
-[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions) last manuscript build `r Sys.time()`
+[![build](https://github.com/tbep-tech/wqtrends-manu/workflows/build/badge.svg)](https://github.com/tbep-tech/wqtrends-manu/actions)
+[![DOI](https://zenodo.org/badge/249504727.svg)](https://zenodo.org/badge/latestdoi/249504727)
 

### 2021-05-05 manu_draft.Rmd
5 and 10 year trend comparisons added to supplement

@@
 data(seastrnd2)
-data(cmptrnd)
+data(cmptrnd10)
 data(modstr)
@@
 
-Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp)).
+Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp10), see supplement for the same analysis with five and fifteen year time periods).
 In many cases, the slope estimates were similar in magnitude, with some exceptions at the more southern stations where the OLS estimates suggested a larger trend than the meta-analysis methods.
@@
 Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks [@Smith01].
-Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
+Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp10) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
 Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
@@
 ```{r, results = 'hide', message = F}
-toplo <- cmptrnd
+toplo <- cmptrnd10
 p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
@@
 # save as png
-png('figs/trndcmp.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
+png('figs/trndcmp10.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
 p
@@
 ```
-```{r trndcmp, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station.
+```{r trndcmp10, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in different annual periods at each station.
 The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.
 Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
-knitr::include_graphics('figs/trndcmp.png')
+knitr::include_graphics('figs/trndcmp10.png')
 ```

### 2021-05-05 supplement.Rmd
5 and 10 year trend comparisons added to supplement

@@
 data(seastrnd2)
-data(cmptrnd)
+data(cmptrnd05)
+data(cmptrnd15)
 data(modstr)
@@
 ```
+
+```{r, results = 'hide', message = F}
+toplo <- cmptrnd05
+p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
+  facet_grid(seas ~ yrs, scales = 'free_x') + 
+  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
+  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
+  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
+  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
+  scale_shape(name = 'Trend model') + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  # scale_x_continuous(limits = c(-1.14, 1.14)) +
+  guides(
+    color = guide_legend(override.aes = list(linetype = 0))
+  ) +
+  theme_minimal() + 
+  theme(
+    strip.background = element_blank(),
+    panel.background = element_rect(fill = 'gray98', colour = NA), 
+    panel.border = element_blank(), 
+    legend.position = 'top', 
+    axis.text.x = element_text(angle = 45, hjust = 1)
+  ) +
+  labs(
+    y = 'Station (north to south)', 
+    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+  )
+
+# save as png
+png('figs/trndcmp05.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
+p
+dev.off()
+```
+```{r trndcmp05, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in 5 year periods at each station.
+The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.
+Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
+knitr::include_graphics('figs/trndcmp05.png')
+```
+
+```{r, results = 'hide', message = F}
+toplo <- cmptrnd15
+p <- ggplot(toplo, aes(y = station, x = yrcoef, linetype = pval, shape = modtyp, colour = modtyp)) + 
+  facet_grid(seas ~ yrs, scales = 'free_x') + 
+  geom_point(size = 2, position = position_dodge(width = 0.6), alpha = 0.8) + 
+  geom_errorbarh(aes(xmin = yrcoeflwr, xmax = yrcoefupr), position = position_dodge(width = 0.6), height = 0) +
+  scale_linetype_manual('Trend significance', values = c('dashed', 'solid')) + 
+  scale_colour_discrete_qualitative(name = 'Trend model', palette = 'Dark3') +
+  scale_shape(name = 'Trend model') + 
+  geom_vline(xintercept = 0, linetype = 'dashed') + 
+  # scale_x_continuous(limits = c(-1.14, 1.14)) +
+  guides(
+    color = guide_legend(override.aes = list(linetype = 0))
+  ) +
+  theme_minimal() + 
+  theme(
+    strip.background = element_blank(),
+    panel.background = element_rect(fill = 'gray98', colour = NA), 
+    panel.border = element_blank(), 
+    legend.position = 'top'
+  ) +
+  labs(
+    y = 'Station (north to south)', 
+    x = expression(paste(log[10], " chl-a change (", italic(mu), "g ", L^-1, yr^-1, ")"))
+  )
+
+# save as png
+png('figs/trndcmp15.png', height = 6, width = 7.5, family = 'serif', units = 'in', res = 400)
+p
+dev.off()
+```
+```{r trndcmp15, fig.cap = 'Trend estimate comparisons for three models applied to seasonal averages of chl-a in 15 year periods at each station.
+The "OLS raw" trend model is based on an ordinary least squares (OLS) regression fit to the seasonal averages of chl-a from the raw data, the "OLS GAM" trend model is based on an OLS regression fit to the seasonal averages of chl-a from the GAM model (without error propagation), and the "Meta-analysis GAM" trend model is based on a meta-analysis regression fit to the seasonal averages of chl-a from the GAM model.
+Values for each model are the log$_{10}$-slope estimates (+/- 95\\% confidence interval) as annual change per year within each season, with line style denoting trend significance.'}
+knitr::include_graphics('figs/trndcmp15.png')
+```
+
+# References

### 2021-05-05 R/dat_proc.R

@@ -229,7 +229,9 @@ save(seastrnd2, file = 'data/seastrnd2.RData', compress = 'xz')
 
 # trend model comparisons -------------------------------------------------
 
-cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
+## 10 year windows, main text ---------------------------------------------
+
+cmptrnd10 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
   crossing(
     fl = ., 
     tibble(
@@ -348,7 +350,255 @@ cmptrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   unnest(c('yrcoef', 'yrcoefupr', 'yrcoeflwr')) %>% 
   mutate(station = factor(station, levels = rev(unique(station))))
 
-save(cmptrnd, file = 'data/cmptrnd.RData', compress = 'xz')
+save(cmptrnd10, file = 'data/cmptrnd10.RData', compress = 'xz')
+
+## 5 year windows, supplement ---------------------------------------------
+
+cmptrnd05 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>% 
+  crossing(
+    fl = ., 
+    tibble(
+      # doystr = c(1, 91, 182, 274),
+      # doyend = c(90, 181, 273, 364)
+      doystr = c(41, 213),
+      doyend = c(213, 338)
+    ), 
+    tibble(
+      yrstr = c(1991, 1995, 2000, 2005, 2010, 2015),
+      yrend = c(1995, 2000, 2005, 2010, 2015, 2019)
+    )
+  ) %>% 
+  group_by(fl) %>% 
+  nest() %>% 
+  mutate(
+    mod = purrr::map(fl, function(x){
+      
+      load(file = x)
+      
+      nm <- basename(x)
+      nm <- gsub('\\.RData', '', nm)
+      
+      out <- get(nm) %>% 
+        pull(model) %>% 
+        deframe()
+      
+      return(out)
+      
+    })
+  ) %>% 
+  unnest(c('data')) %>% 
+  group_by(doystr, doyend, mod, fl) %>% 
+  nest() %>% 
+  mutate(
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
[…]

### 2021-05-09 manu_draft.Rmd
ste submitted

@@
 urlcolor: blue
-csl: ecological-applications.csl
+csl: science-of-the-total-environment.csl
 link-citations: true
@@
 
+*Key words*: chlorophyll, Generalized Additive Models, meta-analysis, San Francisco Estuary, Trend analysis
+
 # Introduction
@@
 We provide examples illustrating how these questions can be addressed using GAMs to estimate seasonal trends and evaluated between years using meta-analysis methods.
-This approach is new to environmental trend-detection problems and is provided in the *wqtrends* R package developed by the authors [@Beck21, available at https://tbep-tech.github.io/wqtrends].
+This approach is new to environmental trend-detection problems and is provided in the *wqtrends* R package developed by the authors [@Beck21, available at https://tbep-tech.github.io/wqtrends, including an online dashboard for viewing results at https://nutrient-data.sfei.org/apps/SFbaytrends/].
 
@@
 
-Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp10), see supplement for the same analysis with five and fifteen year time periods).
+Applying the same comparison to all stations showed that different trend analysis methods provided conflicting information on the magnitude and significance of the seasonal chl-a changes in each decade (Figure \@ref(fig:trndcmp10)).
 In many cases, the slope estimates were similar in magnitude, with some exceptions at the more southern stations where the OLS estimates suggested a larger trend than the meta-analysis methods.

### 2021-05-09 supplement.Rmd
ste submitted

--- 2021-04-09 supplement.Rmd
@@
 ```
+
+# Online dashboard
+
+An interactive, online dashboard for viewing results in this manuscript is available at https://nutrient-data.sfei.org/apps/SFbaytrends/.
 
@@
 ```
+
+# References

### 2021-05-09 R/graphabs.R

@@ -1,0 +1,221 @@
+library(wqtrends)
+library(tidyverse)
+
+data(rawdat)
+
+tomod <- rawdat %>% 
+  filter(station %in% 30) %>% 
+  filter(param %in% 'chl')
+
+mod <- anlz_gam(tomod, trans = 'log10')
+
+yrstr1 <- 1990
+yrend1 <- 2007
+yrstr2 <- 2007
+yrend2 <- 2019
+doystr <- 213
+doyend <- 304
+  
+ylab <- 'Chl-a (ug/L)'
+alpha <- 1
+
+# get predictions
+prds <- anlz_prd(mod)
+
+# get transformation
+trans <- unique(prds$trans)
+
+# raw data
+tobacktrans <- mod$model %>% 
+  dplyr::mutate(
+    trans = mod$trans
+  )
+
+moddat <- anlz_backtrans(tobacktrans) %>% 
+  dplyr::mutate(
+    date = lubridate::date_decimal(cont_year), 
+    date = as.Date(date)
+  )
+
+# transformation used
+trans <- mod$trans
+
+# get seasonal averages
+avgseason <- anlz_avgseason(mod, doystr = doystr, doyend = doyend) 
+
+# get mixmeta models
+mixmet1 <- anlz_mixmeta(avgseason, yrstr = yrstr1, yrend = yrend1)
+mixmet2 <- anlz_mixmeta(avgseason, yrstr = yrstr2, yrend = yrend2)
+
+# title
+dts <- as.Date(c(doystr, doyend), origin = as.Date("2000-12-31"))
+strt <- paste(lubridate::month(dts[1], label = T, abbr = T), lubridate::day(dts[1]))
+ends <- paste(lubridate::month(dts[2], label = T, abbr = T), lubridate::day(dts[2]))
+ttl <- paste0('Fitted averages with 95% confidence intervals: ', strt, '-',  ends)
+
+# plot objects
+toplo1 <- avgseason
+
+toplo2 <- data.frame(
+  yr = seq(yrstr1, yrend1, length = 50)
+  ) %>% 
+  dplyr::mutate( 
+    avg = predict(mixmet1, newdata = data.frame(yr = yr)), 
+    se = predict(mixmet1, newdata = data.frame(yr = yr), se = T)[, 2], 
+    bt_lwr = avg - 1.96 * se,
+    bt_upr = avg + 1.96 * se,
+    bt_avg = avg
+  )
+
+# subtitle info
[…]

### 2021-06-23 revision/review_response.Rmd
STE revisions started

---
output: 
  bookdown::word_document2:
    reference_docx: response_styles.docx
    number_sections: false
always_allow_html: true
urlcolor: blue
link-citations: true
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files('revision-response.Rmd')
```

We sincerely thank the reviewers for providing thoughtful comments on our manuscript.  We have provided a point-by-point response to these comments below.  We are confident that these additions and revisions improve the content of our manuscript for readers of Science of the Total Environment.

## Reviewer 1

This manuscript is a welcome addition in the world of trend analyses. The authors propose combinations of GAM and mixed-effects meta-analysis regression to account for uncertainty. The methodology is sound and results are convincing. My recommendation would be even more enthusiastic if the authors had formatted their methods sections in a more traditional manner. I am of the opinion that models should be presented using clear mathematical equations and not in R code semantics. This comments and a few other can be found in the attached document.

* __Response__: We really appreciate these comments that affirm our opinion on the value of this work for trend analysis of long-term water quality data.  We have changed our formulas and text to a conventional format, as opposed to R code notation.  Briefly, here are the updated equations for the models described in the methods.

\begin{equation}
\textrm{Model S}: y_i \sim\ \beta_0 + f_1\left(cont\_year_i\right) + \epsilon_i
(\#eq:Smod)
\end{equation}

\begin{equation}
\textrm{Model SY:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + \epsilon_i
(\#eq:SYmod)
\end{equation}

\begin{equation}
\textrm{Model SYD:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + \epsilon_i
(\#eq:SYDmod)
\end{equation}

\begin{equation}
\textrm{Model SYDI:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + f_3\left(cont\_year_i, doy_i\right) + \epsilon_i
(\#eq:SYDImod)
\end{equation}

## Reviewer 2

### Overview

This manuscript presents a new application of GAMs approaches to fit irregular time series data (gaps, changing frequency of measurements) to estimate trends. The paper presents a case study using Chl a data from San Francisco Bay.  The manuscript is presented as a general approach with applications to other similar environmental time series data.  A comparison of methods shows that the GAMs plus meta-analysis method provides some different trend conclusions from ordinary least squares and plain GAMs. The paper is generally well written and well-suited for the audience for STOTEN.  I have some clarifications that would help improve/clarify the utility and readability of the manuscript.  These are presented below.

* __Response__:

### Utility (1):

We applied the wqtrends code in R to a time series data set of X2 (salinity intrusion measure) over approximately in San Francisco Bay.  Some observations:

-the code did not work with the full time series of daily X2 data.  But we were able to make it work using a random sampling of 5 points per month.  Could the reason for this error be explored?  As time series data are concerned, we provided an input of ~100 years of daily data.  A sample result for the trends are shown in the figure below.  I think it would be helpful for readers/users to understand what the practical limitations of the code/algorithm are.

* __Response__:

### Utility (2):

From figure 7 it seems that there are very different conclusions for Station 36 depending on the method used.  From Figure 8, however, it seems that Station 36 is a bit of an anomaly in this regard, and that other 8 stations in the case study have findings on trends that are similar among the three methods—and that the results are not dramatically different.  Perhaps this should be the basis for added discussion in the final section—when is the added complexity of the GAMS and meta analysis method is appropriate in a real-world setting?  Alternatively, it would be helpful if the authors were to examine another data set from the region that they work with, to review the relative findings of trend significance and what this depends on.

* __Response__:

### Readability/Clarity:

The term meta-analysis seems to be an important part of the paper and is referred to multiple times, but it is never clearly explained in the manuscript.  For example, at line 164.  See the following sentences.

"Third, we used a mixed-effects meta-analysis to  estimate trends and test hypotheses about the change in seasonal averages across years. While 166 meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty (Gasparrini et al., 2012; Sera et al., 2019). Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions."

* __Response__:

It is not clear to me what exactly is meant by meta-analysis and what is being tested in the context of the time series data.  An explanation that is understandable by a general audience of potential users of this method would be helpful.  It does seem likely, as that sentence points out, that most readers would be more familiar with meta-analysis methods in the context of aggregating the results of multiple studies.  There are other helpful citations afterwards, but a more explicit connection between the familiar use case and the use case here of combining different seasonal features from a single regression model would have been helpful for reading this work.

* __Response__:

### 2021-06-23 manu_draft.Rmd
STE revisions started

@@
 
+nsims <- 1000
 # # extract bib entries from online
@@
 We discuss all four to clarify their relationships and interpretations.
-Models are shown in the notation of the `mgcv` R package as formulas for the `gam` function [@Wood17;@RCT20].
+All models were created using the `mgcv` R package [@Wood17;@RCT20], with utility functions included in the `wqtrends` package created by the authors [@Beck21].
 
@@
 
-Model S: `y ~ s(cont_year, k = num_knots_Y)`
-
-where `y` is the time series of interest, such as chl-a, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case as a function of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline.
+\begin{equation}
+\textrm{Model S:} y_i \sim\ \beta_0 + f_1\left(cont\_year_i\right) + \epsilon_i
+(\#eq:Smod)
+\end{equation}
+
+where `y` is the time series of interest, such as chl-a, for observation $i$ explained by an intercept-term $\beta_0$ and a smooth function $f_1()$ for `cont_year` as "continuous year" represented as a continuous numerical date (e.g., July 1st 2019 would be 2019.5).
+The $f_1()$ function is a smoothing spline composed of the sum of multiple "basis functions" describing the relationship of $cont\_year$ with $y$ [@Wood17].
+The smoothing splines includes a user-specified number of knots or "connections" along the spline.
+The $\epsilon$ term is for model errors having normal distribution with mean zero and constant variance.
 
@@
 Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.
-In the examples below, we chose the number of knots, `num_knots_Y`, as 12 times the number of years in the time series, i.e., one knot per month.
+In the examples below, we chose the number of knots for $f_1()$ as 12 times the number of years in the time series, i.e., one knot per month.
 If the data were too sparse to fit 12 knots per year, the number of knots was reduced by one knot per year until the model could be estimated (i.e., 12 * years, 11 * years, etc.).
@@
 
-Model SY: `y ~ cont_year + s(cont_year, k = num_knots_Y)`
+\begin{equation}
+\textrm{Model SY:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + \epsilon_i
+(\#eq:SYmod)
+\end{equation}
+
+where equation \@ref(eq:SYmod) is the same as equation \@ref(eq:Smod) with the addition of a linear term for $cont\_year$ related to $y_i$ by the $\beta_1$ slope parameter.
 
 This model is mathematically equivalent to model S (Table \@ref(tab:modstrtab)).
-The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
+The $f_1()$ spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.
 When `cont_year` is included explicitly as a linear term in model SY, `mgcv` adjusts the basis functions for the spline to exclude the linear term, thereby not over-parameterizing the model.
@@
 
-Model SYD: `y ~ cont_year + s(cont_year, k = num_knots_Y) + s(doy, bs = 'cc', k = num_knots_D)`
-
-where `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates that the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the number of knots for the `doy` spline.
+\begin{equation}
+\textrm{Model SYD:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + \epsilon_i
+(\#eq:SYDmod)
+\end{equation}
+
+where equation \@ref(eq:SYDmod) is the same as equation \@ref(eq:SYmod) with the addition of a smoothing spline for "day-of-year" ($doy$, i.e., Julian date, a count starting January 1 for each year).
+The $f_2()$ spline is cyclic (using `bs = 'cc'` in `mgcv`) to constrain the start and end at the same value.
+A user-specified number of knots is also included in $f_2()$.
 While model SYD is not mathematically equivalent to models S and SY, it should produce nearly identical results.
@@
 
-Model SYDI: `y ~ cont_year + s(cont_year, k = num_knots_Y) + s(doy, bs = "cc", k = num_knots_D) + ti(cont_year, doy, bs = c("tp", "cc"), k = c(num_knots_Y_ti, num_knots_D_ti))`
-
-where `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.
-The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.
-In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `num_knots_Y` and `num_knots_D`, but not both given limits on the model degrees of freedom.
+\begin{equation}
+\textrm{Model SYDI:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + f_3\left(cont\_year_i, doy_i\right) + \epsilon_i
+(\#eq:SYDImod)
+\end{equation}
+
+where equation \@ref(eq:SYDImod) is the same as equation \@ref(eq:SYDmod) with the addition of a tensor-product smoothing spline (`ti()` in `mgcv`) that varies smoothly as a function of both $cont\_year$ and $doy$.
+Both $cont\_year$ and $doy$ include their own number of knots, such that the total number of knots for the spline is the product of the two.
+The need for sufficient knots in SYDI can be satisfied either by sufficiently large values for $f_3()$ or a sufficiently large value for knots in $f_1()$ and $f_2()$, but not both given limits on the model degrees of freedom.
 
@@
 The univariate splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model S given the fewer knots that are available to the former.
-The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.
+The tensor-product spline represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.
 The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.
@@
 @Murphy19 used spline formulations for Chesapeake Bay water quality related to those proposed here, but for different goals and with different handling of smoothness.
-They evaluated a "`gam0`" with only `s(doy)` and linear `cont_year` terms, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
+They evaluated a "`gam0`" with only a cyclic spline for `doy` and linear `cont_year` terms, a "`gam1`" like our SYD, and a "`gam2`" like our SYDI.
 In application, only "`gam2`" was used, including the addition of splines as functions of hydrologic variables to account for finer-scale variation.
 Murphy et al.
-allowed a maximum number of knots in the `s(cont_year)` term of 2/3 times the number of years and do not explicitly consider the number of knots in the interaction spline, following an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system.
-Constraining splines with insufficient knots could inflate Type I error rates for temporal changes and we seek to lower this risk by increasing the upper limit for the knots for the `s(cont_year)` term.
+allowed a maximum number of knots for the `cont_year` spline ($f_1()$) of 2/3 times the number of years and do not explicitly consider the number of knots in the interaction spline, following an *ad hoc* allocation of variation in the data to different components based on previous interpretations of water quality dynamics in the system.
+Constraining splines with insufficient knots could inflate Type I error rates for temporal changes and we seek to lower this risk by increasing the upper limit for the knots for the $f_1()$ term.
 Finally, @Murphy19 present large AIC differences between their spline formulations.
@@
 We visually compare chl-a estimates from models SY, SYD, and SYDI to emphasize that similar fits can be achieved by all of the presented models (Figure \@ref(fig:modsumfig), SY is identical to S and is not shown).
-Models S, SYD, and SYDI were fit to chl-a data from station 34 using large k values for the arguments `num_knots_y`, `num_knots_D`, `num_knots_Y_ti`, and `num_knots_D_ti` for each model.
+Models S, SYD, and SYDI were fit to chl-a data from station 34 using a sufficiently high number of knots for the respective splines for each model.
 Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).
-However, when contrasting the estimates using only the continuous year smoother (`s(cont_year)`), the fits differ substantially because of how each model allocates variation to the splines.
+However, when contrasting the estimates using only the continuous year smoother ($f_1()$), the fits differ substantially because of how each model allocates variation to the splines.
 These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)).
@@
 
-For all results, model S was used with enough knots in `num_knots_y` to evaluate chl-a trends across the monitoring stations in the SFE.
+For all results, model S was used with enough knots in $f_1()$ to evaluate chl-a trends across the monitoring stations in the SFE.
 This model was chosen because of the relatively faster processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)).
@@
   st_transform(crs = st_crs(states)) %>% 
+  st_make_valid() %>% 
   st_buffer(dist = 0) %>% 
@@
 p1 <- p1a + p1b + p1c + plot_layout(ncol = 3, widths = c(1, 1, 1))
-prds <- lapply(modstr, anlz_prd) %>% 
+prds <- lapply(modstr, anlz_prd, annual = T) %>% 
   enframe('model', 'dat') %>%
@@
 
-      mixmet <- anlz_avgseason(mod, doystr = doystr, doyend = doyend) %>% 
+      mixmet <- anlz_metseason(mod, doystr = doystr, doyend = doyend, nsim = nsims) %>% 
         anlz_mixmeta(yrstr = yrstr, yrend = yrend)
@@
 
-p1 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
+p1 <- show_metseason(mod, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p2 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
+p2 <- show_metseason(mod, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p3 <- show_avgseason(mod, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
+p3 <- show_metseason(mod, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) +  
@@
   )
-p4 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
+p4 <- show_metseason(mod, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p5 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
+p5 <- show_metseason(mod, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p6 <- show_avgseason(mod, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
+p6 <- show_metseason(mod, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
 show_cmpplo <- function(fl, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
-                        ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")"))){
+                        ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")")), nsim){
 
@@
     mutate(bt_avg = 10^(avg + (sdavg^2) / 2))
-  avgseas <- anlz_avgseason(mod = mod, doystr = doystr, doyend = doyend)
+  avgseas <- anlz_metseason(mod = mod, doystr = doystr, doyend = doyend, nsim = nsims) %>% 
+    rename(
+      bt_avg = bt_met
+    )
   tomod2 <- avgseas %>%
     filter(yr <= yrend & yr >= yrstr) %>%
-    mutate(S = se^2)
+    mutate(S = se^2) %>% 
+    rename(
+      avg = met
+    )
 
@@
 p1 <- show_cmpplo('data/modslog_chl36.RData', doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
-                  ttl =  '(a) Station 36, Jan-Jul chl. averages', ymax = NULL)
+                  ttl =  '(a) Station 36, Jan-Jul chl. averages', ymax = NULL, nsim = nsims)
 p2 <- show_cmpplo('data/modslog_chl36.RData', doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
-                  ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL)
+                  ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL, nsim = nsims)
 # p3 <- show_cmpplo('data/modslog_chl34.RData', doystr = 274, doyend = 364, yrstr = 2010, yrend = 2019, 
-#                   ttl =  '(c) Station 34, Oct-Dec chl. averages', ymax = 15, ylab = NULL)
+#                   ttl =  '(c) Station 34, Oct-Dec chl. averages', ymax = 15, ylab = NULL, nsim = nsims)
 
@@
   `Additive components` = c(
-    '`s(cont_year)`', 
-    '`cont_year + s(cont_year)`',
-    '`cont_year + s(cont_year) + s(doy)`', 
-    '`cont_year + s(cont_year) + s(doy) + ti(cont_year, doy)`'
+    '`f_1(cont_year)`', 
+    '`\beta_0cont_year + f_1(cont_year)`',
+    '`\beta_0cont_year + f_1(cont_year) + f_2(doy)`', 
+    '`\beta_0cont_year + f_1(cont_year) + f_2(doy) + f_3(cont_year, doy)`'
   ),

### 2021-06-23 R/dat_proc.R

@@ -129,7 +129,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, mod, fl) %>% 
   nest() %>% 
   mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason)
   ) %>% 
   unnest('data') %>% 
   mutate(
@@ -267,7 +267,7 @@ cmptrnd10 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, mod, fl) %>% 
   nest() %>% 
   mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason)
   ) %>% 
   unnest('data') %>% 
   mutate(
@@ -390,7 +390,7 @@ cmptrnd05 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, mod, fl) %>% 
   nest() %>% 
   mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason)
   ) %>% 
   unnest('data') %>% 
   mutate(
@@ -516,7 +516,7 @@ cmptrnd15 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, mod, fl) %>% 
   nest() %>% 
   mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_avgseason)
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason)
   ) %>% 
   unnest('data') %>% 
   mutate(

### 2021-06-24 revision/review_response.Rmd
more STE revisions

@@
 
-* __Response__:
+* __Response__: Thank you for your comment on our manuscript.  We have addressed your concerns on the utility and readability of our manuscript below.
 
@@
 
-* __Response__:
+* __Response__: Very rarely do reviewers actually apply the proposed methods on novel datasets as a proof of concept.
+For that, we are very appreciative of your efforts in exploring the R package.
+These applications are important for testing both the method and software.
+As such, you exposed an important limitation of our method that was not clearly articulated in the draft.
+We provide two responses to this limitation.
+
+     First, the X2 dataset is collected daily, so the GAM smoothing methods proposed in our manuscript for "irregularly spaced or missing" monitoring data is unnecessary.
+The modelling approach is designed to provide both a continuous (e.g., daily) estimate of the long-term trend and a measure of uncertainty in the absence of more regularly collected values.
+So, this would be an unnecessary applications of GAMs.
+
+     Second, and more important, the practical limitations of applying a GAM with flexible smoothing to nearly 100 years of daily data is beyond the computational power most desktop computers.
+This is asking too much for the smoothing splines used in the mgcv package.
+Given this limitation, we do feel it is important to clearly state the intended use case for these models to avoid confusion in application.
+We have added clarification both in the manuscript and the package vignette so users are clear on the type of data that is appropriate for the methods.
+
+     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the the intended use case for the methods herein."
+
+     In the package vignette, first paragraph (https://tbep-tech.github.io/wqtrends/articles/introduction.html): "These models are appropriate for data typically collected for surface water quality monitoring programs at roughly monthly or biweekly collection intervals, covering at least a decade of observations (e.g., Cloern and Schraga 2016).
+Daily or continuous monitoring data are not appropriate for these methods, due to computational limitations and a goal of the analysis to estimate long-term, continuous trends from irregular or discontinuous sampling."
 
@@
 
-* __Response__:
+* __Response__: You are correct to note that many of the comparisons had similar results and we agree that it is useful to clarify that our approach is warranted relative to more conventional methods.
+If a simpler approach is used, there is no way to know for sure that a false positive or negative result is obtained without applying the proposed methods.
+So, we argue that using our approach is still the wiser option in all cases.
+Otherwise, a user cannot be certain the correct conclusions are obtained.
+We have added the following to the discussion:
+
+    Line xyz: "Moreover, use of methods that fully account for uncertainty in sampling design is recommended over more conventional analyses given the certainty of conclusions in the former.
+Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used.
+Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences."
 
@@
 
-* __Response__:
-
 It is not clear to me what exactly is meant by meta-analysis and what is being tested in the context of the time series data.
@@
 
-* __Response__:
+* __Response__: Thank you for point out this ambiguity.
+We agree that a distinction between the more common understanding of "meta-analysis" and the one applied here needs to be made early in the manuscript.
+We have added content in the introduction to clarify our methods.
+
+     Line xyz: "Herein, meta-analysis refers to the use of seasonal features and their uncertainty estimates for trend analysis, as compared to meta-analysis for the aggregation of results from multiple studies for analysis of effect sizes or other synthesis results (i.e., Lortie et al.
+2014)."

### 2021-06-24 manu_draft.Rmd
more STE revisions

@@
 This paper develops the use of GAMs with mixed-effects meta-analysis [@Gasparrini12;@Sera19] to address multi-scale trend analysis questions for which seasonal Kendall tests and the more complex STL and WRTDS methods are not well-suited.
+Herein, "meta-analysis" refers to the use of seasonal features and their uncertainty estimates for trend analysis, as compared to meta-analysis for the aggregation of results from multiple studies for analysis of effect sizes or other synthesis results [i.e., @Lortie14].
 
@@
 Sampling frequency varied somewhat over time and by station.
+Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the the intended use case for the methods herein.
 Every observation was included directly in the statistical models without spatial or temporal binning or averaging.
@@
 Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2).
-Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for January to July averages from 1991 to 2000.
+Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for January to June averages from 1991 to 2000.
 Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).
@@
 In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
-Our examples demonstrated the potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations.
+Moreover, use of methods that fully account for uncertainty in sampling design is recommended over more conventional analyses given the certainty of conclusions in the former.
+Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used.
+Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences.
 
@@
 ```{r, results = 'hide', message = F}
-show_cmpplo <- function(fl, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
+show_cmpplo <- function(datin, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
                         ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")")), nsim){
 
-  load(file = fl)
-  nm <- basename(fl)
-  nm <- gsub('\\.RData', '', nm)
-  mod <- get(nm) %>%
+  mod <- datin %>%
     pull(model) %>%
@@
 
-p1 <- show_cmpplo('data/modslog_chl36.RData', doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
-                  ttl =  '(a) Station 36, Jan-Jul chl. averages', ymax = NULL, nsim = nsims)
-p2 <- show_cmpplo('data/modslog_chl36.RData', doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
+load(file = 'data/modslog_chl32.RData')
+datin <- modslog_chl32
+
+p1 <- show_cmpplo(datin, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
+                  ttl =  '(a) Station 36, Jan-Jun chl. averages', ymax = NULL, nsim = nsims)
+p2 <- show_cmpplo(datin, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
                   ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL, nsim = nsims)

### 2021-06-24 R/dat_proc.R

@@ -267,7 +267,7 @@ cmptrnd10 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, mod, fl) %>% 
   nest() %>% 
   mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason)
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason, nsims = 1000)
   ) %>% 
   unnest('data') %>% 
   mutate(
@@ -277,12 +277,12 @@ cmptrnd10 <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     yrend = purrr::pmap(list(yrend, avgseas), function(yrend, avgseas){
       min(yrend, max(avgseas$yr))
     }),
-    metatrnd = purrr::pmap(list(avgseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta),
+    metatrnd = purrr::pmap(list(metseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta),
     lmtrnd = purrr::pmap(list(avgseas, yrstr, yrend), function(avgseas, yrstr, yrend){
       
       out <- avgseas %>% 
         filter(yr >= yrstr & yr <= yrend) %>% 
-        lm(avg ~ yr, .)
+        lm(met ~ yr, .)
       
       return(out)
       

### 2021-06-25 revision/review_response.Rmd
initial STE revision wihout formatting done, sent to coauthors

@@
 Briefly, here are the updated equations for the models described in the methods.
+All other instances in the main text referring to R code (except in a few locations) have been changed.
 
@@
 
-* __Response__: Thank you for your comment on our manuscript.  We have addressed your concerns on the utility and readability of our manuscript below.
+* __Response__: Thank you for your comments on our manuscript.  We have addressed your concerns on the utility and readability of our manuscript below.
 
@@
 The modelling approach is designed to provide both a continuous (e.g., daily) estimate of the long-term trend and a measure of uncertainty in the absence of more regularly collected values.
-So, this would be an unnecessary applications of GAMs.
+So, application of our method to datasets with sampling intervals longer than the daily scale would be more appropriate.
 
-     Second, and more important, the practical limitations of applying a GAM with flexible smoothing to nearly 100 years of daily data is beyond the computational power most desktop computers.
-This is asking too much for the smoothing splines used in the mgcv package.
+     Second, and more important, the practical limitations of applying a GAM with flexible smoothing to nearly 100 years of daily data is beyond the computational power of most desktop computers.
+This is asking too much of the smoothing splines used in the mgcv package.
 Given this limitation, we do feel it is important to clearly state the intended use case for these models to avoid confusion in application.
-We have added clarification both in the manuscript and the package vignette so users are clear on the type of data that is appropriate for the methods.
+We have added clarification both in the manuscript and the package vignette to clearly indicate the type of data that is appropriate for the methods.
 
-     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the the intended use case for the methods herein."
+     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the intended use case for the methods herein."
 
-     In the package vignette, first paragraph (https://tbep-tech.github.io/wqtrends/articles/introduction.html): "These models are appropriate for data typically collected for surface water quality monitoring programs at roughly monthly or biweekly collection intervals, covering at least a decade of observations (e.g., Cloern and Schraga 2016).
+     In the package vignette, first paragraph (https://tbep-tech.github.io/wqtrends/articles/introduction.html): "These models are appropriate for data typically from surface water quality monitoring programs at roughly monthly or biweekly collection intervals, covering at least a decade of observations (e.g., Cloern and Schraga 2016).
 Daily or continuous monitoring data are not appropriate for these methods, due to computational limitations and a goal of the analysis to estimate long-term, continuous trends from irregular or discontinuous sampling."
@@
 * __Response__: You are correct to note that many of the comparisons had similar results and we agree that it is useful to clarify that our approach is warranted relative to more conventional methods.
-If a simpler approach is used, there is no way to know for sure that a false positive or negative result is obtained without applying the proposed methods.
-So, we argue that using our approach is still the wiser option in all cases.
-Otherwise, a user cannot be certain the correct conclusions are obtained.
+If a simpler approach is used, there is no way to know for sure that a false positive or negative result is obtained without applying our proposed methods.
+As such, we argue that using our approach is still the wiser option in all cases.
+Otherwise, a user cannot be certain the correct conclusions are obtained using simpler methods.
 We have added the following to the discussion:
@@
 Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences."
+
+    We also want to mention that we recently discovered a minor bug in the seasonal feature extraction of wqtrends that prevented an estimate for years on the tail ends of the time series with incomplete data.
+Because of this, we have updated Figure 7 and 8, now using stations 30 and 34 as examples in the manuscript.
+None of the conclusions change in the manuscript.
 
@@
 
-* __Response__: Thank you for point out this ambiguity.
+* __Response__: Thank you for pointing out this ambiguity.
 We agree that a distinction between the more common understanding of "meta-analysis" and the one applied here needs to be made early in the manuscript.

### 2021-06-25 manu_draft.Rmd
initial STE revision wihout formatting done, sent to coauthors

@@
 
-nsims <- 1000
+nsims <- 1e4
+
 # # extract bib entries from online
@@
 Sampling frequency varied somewhat over time and by station.
-Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the the intended use case for the methods herein.
+Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the intended use case for the methods herein.
 Every observation was included directly in the statistical models without spatial or temporal binning or averaging.
@@
 Plots a-c show trends in January to June averages while plots d-f show trends in July to December averages.
-The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.06 95\% confidence interval) from 1991 to 2000, whereas July to December chl-a did not change significantly.
-Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0.01-0.05 95\% confidence interval).
+The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.05 95\% confidence interval) from 1991 to 2000, whereas July to December chl-a did not change significantly.
+Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0-0.05 95\% confidence interval).
 Finally, chl-a decreased from 2010 to 2019 but only for July to December (log$_{10}$ chl-a slope -0.02, -0.04-0 95\% confidence interval).
@@
 Figure \@ref(fig:trndmap) shows results from similar analyses as those in Figure \@ref(fig:trnddat), but applied to all stations.
-Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32, 34, and 36 from 1991 to 2000 and station 18 from 2000 to 2010; decreases were observed at stations 30 and 32 from 2010 to 2019.
-For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
+Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32 and 34 from 1991 to 2000; decreases were observed at stations 30 and 32 from 2010 to 2019.
+For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, 24, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
 
@@
 Increases and decreases were observed in both the January to June and July to December seasonal periods, with some notable exceptions.
-In particular, the most southern stations (32, 34, 36) had increasing trends prior to 2005 only in the July to December period.
+In particular, the most southern stations (32, 34, 36) had increasing trends prior to 2005 that were more often observed in the July to December period.
 Additionally, chl-a at the more northern stations has not changed in recent years for both seasonal periods.
@@
 Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2).
-Figure \@ref(fig:trndcmpex)a shows trend estimates for station 36 for January to June averages from 1991 to 2000.
+Figure \@ref(fig:trndcmpex)a shows trend estimates for station 30 for January to June averages from 2000 to 2010.
 Only the meta-analysis regression results show a trend in this example (based on $p < 0.05$).
 The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.
-Figure \@ref(fig:trndcmpex)b shows trend estimates for the same station for July to December averages from 2000 to 2010.
-Unlike the first example, only the top figure shows a trend, whereas the bottom two plots do not show trends.
+Figure \@ref(fig:trndcmpex)b shows trend estimates station 34 for July to December averages from 1991 to 2000.
+Unlike the first example, only the middle figure shows a trend, whereas the top and bottom plots do not show trends.
 In both cases, only the meta-analysis results give reliable conclusions because of full propagation of uncertainty across methods.
@@
 
-# station locations
+# for buffer, see https://github.com/r-spatial/sf/issues/1692
 sflocs <- locs %>% 
@@
   st_union()
-locbuff <- st_buffer(sflocs, dist = 0.1)
-locbuff2 <- st_buffer(sflocs, dist = 0.07)
-insetbb <- st_buffer(sflocs, dist = 1.5) %>% 
+locbuff <- st_buffer(sflocs, dist = units::set_units(0.1, degree))
+locbuff2 <- st_buffer(sflocs, dist = units::set_units(0.07, degree))
+insetbb <- st_buffer(sflocs, dist = units::set_units(1.5, degree)) %>% 
   st_bbox() %>% 
@@
     legend.title = element_blank(),
-    legend.position = 'right'
+    legend.position = 'right', 
+    panel.grid.minor = element_blank()
   )
@@
 alph <- 0.5
-ylm <- c(0.8, 100)
+ylm <- c(0.8, 110)
 
@@
   geom_point(alpha = alph) +
-  geom_abline(slope = 1, intercept = 0, color = lncol) + 
-  scale_y_log10(limits = ylm) + 
-  scale_x_log10(limits = ylm) + 
+  coord_trans(x = 'log10', y = 'log10', xlim = ylm, ylim = ylm) + 
+  scale_y_continuous(breaks = c(1, 10, 100))  +
+  scale_x_continuous(breaks = c(1, 10, 100)) +
+  geom_abline(intercept = 0, slope = 1, color = lncol) +
   thm + 
@@
   geom_point(alpha = alph) + 
-  geom_abline(slope = 1, intercept = 0, color = lncol) + 
-  scale_y_log10(limits = ylm) + 
-  scale_x_log10(limits = ylm) + 
+  coord_trans(x = 'log10', y = 'log10', xlim = ylm, ylim = ylm) + 
+  scale_y_continuous(breaks = c(1, 10, 100))  +
+  scale_x_continuous(breaks = c(1, 10, 100)) +
+  geom_abline(intercept = 0, slope = 1, color = lncol) +
   thm
@@
   geom_point(alpha = alph) + 
-  geom_abline(slope = 1, intercept = 0, color = lncol) + 
-  scale_y_log10(limits = ylm) + 
-  scale_x_log10(limits = ylm) + 
+  coord_trans(x = 'log10', y = 'log10', xlim = ylm, ylim = ylm) + 
+  scale_y_continuous(breaks = c(1, 10, 100))  +
+  scale_x_continuous(breaks = c(1, 10, 100)) +
+  geom_abline(intercept = 0, slope = 1, color = lncol) +
   thm +
@@
 seas <- tibble(
-  doystr = c(41, 41, 41, 213, 213, 213), 
-  doyend = c(213, 213, 213, 338, 338, 338), 
+  doystr = c(1, 1, 1, 182, 182, 182), 
+  doyend = c(182, 182, 182, 364, 364, 364), 
   yrstr = c(1991, 2000, 2010, 1991, 2000, 2010), 
@@
     lets = c('(a)', '(b)', '(c)', '(d)', '(e)', '(f)'), 
-    doystr = ifelse(doystr == 41, 'Jan', 'Jul'), 
-    doyend = ifelse(doyend == 213, 'Jun', 'Dec'), 
+    doystr = ifelse(doystr == 1, 'Jan', 'Jul'), 
+    doyend = ifelse(doyend == 182, 'Jun', 'Dec'), 
     pval = p_ast(pval)
@@
 
-p1 <- show_metseason(mod, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab, nsim = nsims) +
+p1 <- show_metseason(mod, doystr = 1, doyend = 182, yrstr = 1991, yrend = 2000, ylab = ylab, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p2 <- show_metseason(mod, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL, nsim = nsims) +
+p2 <- show_metseason(mod, doystr = 1, doyend = 182, yrstr = 2000, yrend = 2010, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p3 <- show_metseason(mod, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL, nsim = nsims) +
+p3 <- show_metseason(mod, doystr = 1, doyend = 182, yrstr = 2010, yrend = 2019, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) +  
@@
   )
-p4 <- show_metseason(mod, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab, nsim = nsims) +
+p4 <- show_metseason(mod, doystr = 182, doyend = 364, yrstr = 1991, yrend = 2000, ylab = ylab, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p5 <- show_metseason(mod, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL, nsim = nsims) +
+p5 <- show_metseason(mod, doystr = 182, doyend = 364, yrstr = 2000, yrend = 2010, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
   )
-p6 <- show_metseason(mod, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL, nsim = nsims) +
+p6 <- show_metseason(mod, doystr = 182, doyend = 364, yrstr = 2010, yrend = 2019, ylab = NULL, nsim = nsims) +
   coord_cartesian(ylim = ylim) + 
@@
 ```{r, results = 'hide', message = F}
-show_cmpplo <- function(datin, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
+show_cmpplo <- function(fl, doystr, doyend, yrstr, yrend, ttl, ymax = NULL, 
                         ylab = expression(paste("Chl-a (", italic(mu), "g ", L^-1, ")")), nsim){
 
-  mod <- datin %>%
+  load(file = fl)
+  nm <- basename(fl)
+  nm <- gsub('\\.RData', '', nm)
+  mod <- get(nm) %>%
     pull(model) %>%
@@
 
-load(file = 'data/modslog_chl32.RData')
-datin <- modslog_chl32
-
-p1 <- show_cmpplo(datin, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, 
-                  ttl =  '(a) Station 36, Jan-Jun chl. averages', ymax = NULL, nsim = nsims)
-p2 <- show_cmpplo(datin, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, 
-                  ttl =  '(b) Station 36, Jul-Dec chl. averages', ymax = NULL, ylab = NULL, nsim = nsims)
+p1 <- show_cmpplo('data/modslog_chl30.RData', doystr = 1, doyend = 182, yrstr = 2010, yrend = 2019, 
+                  ttl =  '(b) Station 30, Jan-Jun chl. averages', ymax = NULL, ylab = NULL, nsim = nsims)
+p2 <- show_cmpplo('data/modslog_chl34.RData', doystr = 182, doyend = 364, yrstr = 1991, yrend = 2000, 
+                  ttl =  '(a) Station 34, Jul-Dec chl. averages', ymax = 16, nsim = nsims)
 # p3 <- show_cmpplo('data/modslog_chl34.RData', doystr = 274, doyend = 364, yrstr = 2010, yrend = 2019, 
@@
 ```
-```{r trndcmpex, fig.cap = 'Trend estimate comparisons (arithmetic scale) for three models applied to seasonal averages of chl-a in different annual periods at station 36.
+```{r trndcmpex, fig.cap = 'Trend estimate comparisons (arithmetic scale) for three models applied to seasonal averages of chl-a in different annual periods at stations (a) 30 and (b) 34.
 The first row shows OLS (ordinary least squares) regression applied to seasonal averages of chl-a from the raw data, the second row shows OLS regression applied to seasonal averages of chl-a from the GAM (without error propagation), and the third row shows meta-analysis regression applied to the seasonal averages of chl-a from the GAM.
-Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.
-These examples on selected periods of time show that slope estimates can be similar, but the confidence intervals vary.'}
+Regressions in each plot are fit through the seasonal estimates indicated in the plot titles for a specified year range.'}
 knitr::include_graphics('figs/trndcmpex.png')

### 2021-06-25 R/dat_proc.R

@@ -5,9 +5,10 @@ library(mgcv)
 library(flextable)
 library(officer)
 library(lubridate)
-library(mgcv)
 
 source('R/funcs.R')
+
+nsims <- 1e4
 
 # station locations -------------------------------------------------------
 
@@ -28,6 +29,8 @@ modssta <- rawdat %>%
     model = purrr::pmap(list(station, data), function(station, data){
 
       cat(station, '\n')
+      
+      data <- ungroup(data)
       out <- anlz_gam(data, trans = 'log10')
       
       return(out)
@@ -99,8 +102,8 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   crossing(
     fl = ., 
     tibble(
-      doystr = c(41, 213), 
-      doyend = c(213, 338)
+      doystr = c(1, 182), 
+      doyend = c(182, 364)
     ), 
     tibble(
       yrstr = c(1991, 2000, 2010),
@@ -129,7 +132,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
   group_by(doystr, doyend, mod, fl) %>% 
   nest() %>% 
   mutate(
-    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason)
+    avgseas = purrr::pmap(list(mod = mod, doystr = doystr, doyend = doyend), anlz_metseason, nsim = nsims)
   ) %>% 
   unnest('data') %>% 
   mutate(
@@ -139,7 +142,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
     yrend = purrr::pmap(list(yrend, avgseas), function(yrend, avgseas){
       min(yrend, max(avgseas$yr))
     }),
-    metatrnd = purrr::pmap(list(avgseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta)
+    metatrnd = purrr::pmap(list(metseason = avgseas, yrstr = yrstr, yrend = yrend), anlz_mixmeta)
   ) %>% 
   mutate(
     yrcoef = purrr::pmap(list(mod, metatrnd), function(mod, metatrnd){
@@ -162,7 +165,7 @@ seastrnd <- list.files('data', pattern = '^modslog\\_chl', full.names = T) %>%
[…]

### 2021-07-01 revision/review_response.Rmd
included revision comments from AC

@@
 All other instances in the main text referring to R code (except in a few locations) have been changed.
+Also please note that the attached pdf that was included with the reviewer's response did not include any markup comments.
+This is contrary to what the reviewer indicated.
 
@@
 Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used.
-Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences."
+Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have major consequences for environmental management."
 

### 2021-07-01 manu_draft.Rmd
included revision comments from AC

@@
 Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used.
-Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have expensive consequences.
+Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have major consequences for environmental management.
 

### 2021-07-06 revision/review_response.Rmd
adding PDV response edits

@@
 We have provided a point-by-point response to these comments below.
-We are confident that these additions and revisions improve the content of our manuscript for readers of Science of the Total Environment.
+We are confident that these additions and revisions improve our manuscript for readers of Science of the Total Environment.
+
+In addition to making revisions that address points raised by the reviewers, we have also taken the opportunity to make minor improvements that we identified.
 
@@
 \begin{equation}
-\textrm{Model SY:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + \epsilon_i
+\textrm{Model SY:} y_i \sim\ \beta_0 + \beta_1 cont\_year_i + f_1\left(cont\_year_i\right) + \epsilon_i
 (\#eq:SYmod)
@@
 \begin{equation}
-\textrm{Model SYD:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + \epsilon_i
+\textrm{Model SYD:} y_i \sim\ \beta_0 + \beta_1 cont\_year_i + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + \epsilon_i
 (\#eq:SYDmod)
@@
 \begin{equation}
-\textrm{Model SYDI:} y_i \sim\ \beta_0 + \beta_1\left(cont\_year_i\right) + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + f_3\left(cont\_year_i, doy_i\right) + \epsilon_i
+\textrm{Model SYDI:} y_i \sim\ \beta_0 + \beta_1 cont\_year_i + f_1\left(cont\_year_i\right) + f_2\left(doy_i) + f_3\left(cont\_year_i, doy_i\right) + \epsilon_i
 (\#eq:SYDImod)
@@
 
-     First, the X2 dataset is collected daily, so the GAM smoothing methods proposed in our manuscript for "irregularly spaced or missing" monitoring data is unnecessary.
+     First, the X2 dataset is at the daily scale, so one aspect of the GAM smoothing methods proposed in our manuscript for "irregularly spaced or missing" monitoring data is unnecessary.
+The GAMs provide two benefits: smoothing irregularly spaced data and estimating the non-independence of sequential data so that propagated uncertainty does not incorrectly assume sequential data are independent.
 The modelling approach is designed to provide both a continuous (e.g., daily) estimate of the long-term trend and a measure of uncertainty in the absence of more regularly collected values.
-So, application of our method to datasets with sampling intervals longer than the daily scale would be more appropriate.
+So, application of our method to datasets with sampling intervals longer than the daily scale would be the most compelling use case, but they are nevertheless relevant for uncertainty propagation even with daily data.
+However, because some of the daily X2 values are in fact imputed, it would be appropriate to use the raw data only in our method, and possibly to enrich the GAM with whatever domain-specific model is used for smoothing X2 values, in order to fully propagate uncertainty.
 
@@
 We have added clarification both in the manuscript and the package vignette to clearly indicate the type of data that is appropriate for the methods.
+In practice, if one wants to analyze such a long series, one could break it down into multiple (possibly overlapping) shorter series for the first two stages (GAM estimation and seasonal averaging with uncertainty propagation).
+The only difference would be that the estimated optimal degree of smoothness might change across different time windows, but that may be appropriate and supported by having so much data.
 
-     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the intended use case for the methods herein."
+     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
 
      In the package vignette, first paragraph (https://tbep-tech.github.io/wqtrends/articles/introduction.html): "These models are appropriate for data typically from surface water quality monitoring programs at roughly monthly or biweekly collection intervals, covering at least a decade of observations (e.g., Cloern and Schraga 2016).
-Daily or continuous monitoring data are not appropriate for these methods, due to computational limitations and a goal of the analysis to estimate long-term, continuous trends from irregular or discontinuous sampling."
+Daily or continuous monitoring data covering many years are not appropriate for these methods, due to computational limitations and a goal of the analysis to estimate long-term, continuous trends from irregular or discontinuous sampling."
 

### 2021-07-06 revision/review_response.Rmd
pdv comments added

@@
 
+     We have also added a paragraph in the discussion regarding limitations of the approach, one of which is related to the issues above:
+
+     Lines xyz to xyz: "Several limitations of the proposed methods deserve mention.
+First, if sampling is so irregular that important fluctuations are missed entirely in some years, the GAM estimates and uncertainty propagation could become dubious in interpretation and usefulness.
+Second, estimation of GAMs for very long series can be computationally demanding.
+When this is an obstacle, one could do the first two analysis stages using temporal windows of the full data, with the only implication being that different degrees of smoothness may be estimated for different windows, which indeed might be justified by the data.
+Third, meta-analysis regression results for a very small number of years, particularly confidence intervals and associated p-values, may be inaccurate (e.g., in confidence interval coverage).
+In such cases, one could make alternative use of the GAM seasonal averages and standard errors, such as for pairwise comparisons among years."
+
 ### Utility (2):
@@
 * __Response__: You are correct to note that many of the comparisons had similar results and we agree that it is useful to clarify that our approach is warranted relative to more conventional methods.
-If a simpler approach is used, there is no way to know for sure that a false positive or negative result is obtained without applying our proposed methods.
+If a simpler approach is used, there is no way to know for sure that an invalid result is obtained without applying our proposed methods.
 As such, we argue that using our approach is still the wiser option in all cases.
 Otherwise, a user cannot be certain the correct conclusions are obtained using simpler methods.
-We have added the following to the discussion:
+The paragraph on lines xyz to xyz was revised to address these points:
 
-    Line xyz: "Moreover, use of methods that fully account for uncertainty in sampling design is recommended over more conventional analyses given the certainty of conclusions in the former.
-Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used.
-Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have major consequences for environmental management."
+    "Incorrect conclusions on trends can have dramatic consequences for regulated parties under existing water quality compliance frameworks (Smith et al., 2001).
+Our examples in Figures 7 and 8 demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
+The 'naïve' method using OLS regression applied to seasonal averages from the raw observations fails to propagate uncertainty, similarly to averaging results within a year and applying a simple Kendall test.
+In some cases the results may be similar to those from fully propagating uncertainty, but the loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation (Shabman and Smith, 2003).
+Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and, at larger spatial scales, across all stations if simpler trend analyses are used.
+Even though simpler methods may produce similar results in some cases, particularly with frequent sampling and similar effort between years, the only way to confirm such an outcome would be to compare results, relying on the method with full propagation of uncertainty to be the more robust method.
+Use of methods that fully account for uncertainty is recommended to obtain statistically valid results in a wider range of conditions."
 
@@
 
-     Line xyz: "Herein, meta-analysis refers to the use of seasonal features and their uncertainty estimates for trend analysis, as compared to meta-analysis for the aggregation of results from multiple studies for analysis of effect sizes or other synthesis results (i.e., Lortie et al.
-2014)."
+     Line xyz: "Meta-analysis regression incorporates a known (or estimated) standard error for each response datum.
+Usually meta-analysis is used when each response summarizes a dataset from a separate study, along with its standard error, and the meta-analysis looks for across-study patterns in effect size (i.e., Lortie et al.
+2014).
+Here each response summarizes one year or season of data, with standard error from the GAM, and the meta-analysis looks for patterns across years.
+Thus, while meta-analysis methods are most commonly associated with combining results from multiple studies into a larger analysis, their key modeling step is propagation of uncertainty (Gasparrini et al., 2012; Sera et al., 2019).
+To do this, meta-analysis makes use of a known (estimated) standard error for each response datum, which is exactly the same step required here to propagate standard errors from the GAM into a regression of seasonal averages."

### 2021-07-06 manu_draft.Rmd
pdv comments added

@@
 However, this is difficult when there are mismatches between time scales of monitoring data and trends of interest.
-Here, we propose a novel application of generalized additive models (GAMs) to model seasonal and multi-decadal changes in a long-term monitoring dataset of chlorophyll-a concentrations in the San Francisco Estuary.
+Here, we propose a novel application of generalized additive models (GAMs) to understand multi-decadal changes in chlorophyll-a concentrations in the San Francisco Estuary from roughly biweekly monitoring.
 GAMs have shown promise in water quality trend analysis to separate long-term (i.e., annual or decadal) trends from seasonal variation.
 Our proposed methods estimate seasonal averages in a response variable with GAMs and then use the uncertainty measures with mixed-effects meta-analysis regression to quantify inter-annual trends that account for full propagation of error across methods.
-We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing splines for annual or seasonal components of the time series.
+We first demonstrate that nearly identical descriptions of temporal changes can be obtained using different smoothing spline formulations of the original time series.
 We then extract seasonal averages and their standard errors for an *a priori* time period within each year from the GAM results.
@@
 Such fluctuations may not be related to inter-annual trends or may not be well-suited to multi-scale smoothing methods.
-Many trend analysis methods lack the flexibility to evaluate one to many independent variables in an extendable structure that accommodates hypothesis testing at different time scales of interest.
+Many trend analysis methods lack the flexibility to evaluate multiple independent variables in an extendable structure that accommodates hypothesis testing at different time scales of interest.
 In this paper, we develop methods to estimate across-year trends of within-year features, such as seasonal averages, while accounting for uncertainties across analysis steps.
@@
 STL decomposition does not incorporate explanatory variables besides time, it is defined more as an algorithm of statistical steps than as a coherent statistical model [e.g., @Wan17], and it does not usually estimate standard errors to allow hypothesis testing [but see @Hafen10].
-STL methods may also over-simplify trends into stationary components that do not change over time, e.g., a seasonal estimate that is constant across years.
+STL methods may also over-simplify trends into fixed components that do not change over time, e.g., a seasonal estimate that is constant across years.
 This limitation presents challenges when addressing questions relevant to long-term water quality data, such as timing of seasonal peaks that can suggest system response to changing  environmental conditions [@Cloern10;@Navarro12].
@@
 The basis functions used to formulate GAMs can be customized based on expected patterns in the data.
-One example includes cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
+Examples include cyclic splines, which can be used to model seasonal patterns, and low-dimensional interactions.
 GAMs have added flexibility because they can include both parametric (e.g., linear or quadratic) components and non-parametric (spline) components.
@@
 However, the seasonal average within each year must be robust to inconsistent sampling times and intervals, and any trend analysis must consider the uncertainties in seasonal averages.
-The critical need is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common in time series data.
+The critical need from GAMs is the ability to obtain an accurate estimate of uncertainty (e.g., a standard error) of seasonal averages, even with irregular sampling and serial correlation, which is common in time series data.
 This paper develops the use of GAMs with mixed-effects meta-analysis [@Gasparrini12;@Sera19] to address multi-scale trend analysis questions for which seasonal Kendall tests and the more complex STL and WRTDS methods are not well-suited.
-Herein, "meta-analysis" refers to the use of seasonal features and their uncertainty estimates for trend analysis, as compared to meta-analysis for the aggregation of results from multiple studies for analysis of effect sizes or other synthesis results [i.e., @Lortie14].
+
+Meta-analysis regression incorporates a known (or estimated) standard error for each response datum.
+Usually meta-analysis is used when each response summarizes a dataset from a separate study, along with its standard error, and the meta-analysis looks for across-study patterns in effect size [i.e., @Lortie14].
+Here each response summarizes one year or season of data, with standard error from the GAM, and the meta-analysis looks for patterns across years.
+Thus, while meta-analysis methods are most commonly associated with combining results from multiple studies into a larger analysis, their key modeling step is propagation of uncertainty [@Gasparrini12;@Sera19].
+To do this, meta-analysis makes use of a known (estimated) standard error for each response datum, which is exactly the same step required here to propagate standard errors from the GAM into a regression of seasonal averages.
 
@@
 Is there a spatial difference in chlorophyll trends?
-We provide examples illustrating how these questions can be addressed using GAMs to estimate seasonal trends and evaluated between years using meta-analysis methods.
+We provide examples illustrating how these questions can be addressed using GAMs to estimate seasonal patterns and use meta-analysis to evaluate trends between years.
 This approach is new to environmental trend-detection problems and is provided in the *wqtrends* R package developed by the authors [@Beck21, available at https://tbep-tech.github.io/wqtrends, including an online dashboard for viewing results at https://nutrient-data.sfei.org/apps/SFbaytrends/].
@@
 The Regional Water Quality Control Board has showed renewed interest in understanding the potential for nutrient loading to negatively affect water quality for more southern areas of the SFE where harmful algal blooms, elevated summer-fall chlorophyll concentrations, and low dissolved oxygen concentrations began around 1999 (Figure \@ref(fig:obsdat)) [@Cloern20].
-Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient particularly with respect to seasonal differences between years.
-
-We evaluated near-surface chlorophyll (chl-a) data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
+Although changes in the data are visually apparent, statistical analyses to quantify these changes have been insufficient with respect to seasonal differences between years.
+
+We used near-surface chlorophyll (chl-a) data measured biweekly to monthly from 1990 to 2019 along the longitudinal axis of the SFE extending from Central Bay (stations 18-23), South Bay (stations 24-32), and Lower South Bay (stations 34-36) (Table \@ref(tab:sumtab), Figure \@ref(fig:sitemap)).
 Monitoring data were obtained from the SFE Research Program of the US Geological Survey [@Cloern16;@Schraga20].
 Sampling frequency varied somewhat over time and by station.
-Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the intended use case for the methods herein.
+Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein.
 Every observation was included directly in the statistical models without spatial or temporal binning or averaging.
@@
 Third, we used a mixed-effects meta-analysis to estimate trends and test hypotheses about the change in seasonal averages across years.
-While meta-analysis methods arose from analyses of results from multiple studies, their distinguishing characteristic is propagation of uncertainty [@Gasparrini12;@Sera19].
-Meta-analysis uses response data that includes standard errors (uncertainties) as needed to address our questions.
 
@@
 
-where `y` is the time series of interest, such as chl-a, for observation $i$ explained by an intercept-term $\beta_0$ and a smooth function $f_1()$ for `cont_year` as "continuous year" represented as a continuous numerical date (e.g., July 1st 2019 would be 2019.5).
-The $f_1()$ function is a smoothing spline composed of the sum of multiple "basis functions" describing the relationship of $cont\_year$ with $y$ [@Wood17].
-The smoothing splines includes a user-specified number of knots or "connections" along the spline.
-The $\epsilon$ term is for model errors having normal distribution with mean zero and constant variance.
+where $y$ is measured chl-a, $\beta_0$ is an intercept, and $cont\_year$ is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5).
+The $f_1()$ function is a smoothing spline composed of the sum of multiple "basis functions" multiplied by coefficients.
+$f_1\left(cont\_year\right)$ describes the relationship of $cont\_year$ with $y$ in a way that smoothly follows the data [@Wood17].
+The basis functions involve user-specified knots, a grid of values on the $cont\_year$ axis that is discussed more below.
+The $\epsilon$ term is for residuals following a normal distribution with mean zero and constant variance.
 
@@
 GCV works by penalizing the net curvature of a spline [@Wood04].
-To allow GCV (or other alternatives) to work as intended, the number of knots that determine the maximum degrees of freedom chosen by the analyst must be sufficiently large so that the curvature penalty, rather than the number of knots, determines smoothness.
+To allow GCV (or other alternatives) to work as intended, the number of knots, which determine the maximum degrees of freedom, chosen by the analyst must be sufficiently large so that the curvature penalty, rather than the number of knots, determines smoothness.
 Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.
@@
 The next three spline formulations (Model SY, SYD, and SYDI) provide progressively increasing complexity in how spline terms compose a model to smooth the raw data.
-Model SY describes the time series using a linear trend plus a spline for `cont_year`:
+Model SY describes the time series using a linear trend plus a spline for $cont\_year$:
 
@@
 Both $cont\_year$ and $doy$ include their own number of knots, such that the total number of knots for the spline is the product of the two.
-The need for sufficient knots in SYDI can be satisfied either by sufficiently large values for $f_3()$ or a sufficiently large value for knots in $f_1()$ and $f_2()$, but not both given limits on the model degrees of freedom.
+The need for sufficient knots in SYDI can be satisfied either by sufficiently many knots for $f_3()$ or sufficiently many knots for $f_1()$ or $f_2()$, but not both given limits on the model degrees of freedom.
 
@@
 The univariate splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model S given the fewer knots that are available to the former.
-The tensor-product spline represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.
-The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.
-However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time series.
+The tensor-product spline represents an interaction by allowing the pattern in $cont\_year$ to vary by $doy$ and vice-versa.
+The interaction term provides an appearance that model SYDI is fundamentally different from those provided by the other models.
+However, models S, SY, and SYD also allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time series.
 Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually.
 By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.
-However, a very large number of knots in both the `cont_year` spline *and* in both dimensions of the interaction spline is impossible to achieve.
+However, a very large number of knots in both the $cont\_year$ spline *and* in both dimensions of the interaction spline is impossible to achieve.
 The distinct aspect of model SYDI is the anticipation that within-year fluctuations will vary smoothly from year to year, which is unlikely for the SFE data and chl-a dynamics in many estuaries because bloom size typically varies between years.
@@
 Predictions by day of year from each model are visually similar (Figure \@ref(fig:modsumfig)a) and closely follow the 1:1 line (Figure \@ref(fig:modsumfig)b).
-However, when contrasting the estimates using only the continuous year smoother ($f_1()$), the fits differ substantially because of how each model allocates variation to the splines.
+However, the component of predictions explained by the continuous year smoother ($f_1\left(\right)$) differs between models that include additional smoothers to absorb some patterns in the data.
 These results are also reflected in differences in the effective degrees of freedom among the additive components of each model (Table \@ref(tab:modstrtab)).
@@
 To obtain a vector, $\hby$, of fitted point estimates for every day in a season, $\bX$ would have one row for each day.
-Here, the seasonal averages used in our examples were calculated at the resolution of days.
+Here, the seasonal averages were calculated at the resolution of days.
 The estimated spline yields both $\hbb$ and $\hShb$, an estimate of the covariance matrix of the sampling distribution of $\hbb$.
@@
 The estimated seasonal average was calculated from the vector of daily values for each of the $n$ days in the season of interest with $\hm = A^T \hby$, where $A^T$ is a row vector with all values equal to $1/n$.
-The variance of $\hm$ is $\hsshm = A^T \hShy A$ and standard error is $\hshm$.
+The variance of $\hm$ is $\hsshm = A^T \hShy A$, and its standard error is $\hshm$.
 Each of these estimates are from the approximate multivariate normality of the sampling distribution of $\hbb$.
@@
 The "residual", $\epsilon_t$, represents estimation error in $\hmt$, namely $\hmt - \mt$.
-The residuals are assumed to be independent and normally distributed with mean 0 and variance $\hsshmt$, where the latter is estimated from the calculations above.
+The residuals are assumed to be independent and normally distributed with mean 0 and variances $\hsshmt$, where the latter is estimated from the calculations above.
 The random effect, $b_t$, is the difference between $\mt$ and $\beta_0 + \beta_t t$ and is considered the "residual" in the sense of unexplained variation not due to the estimation error.
@@
 Trend estimates were compared to 1) trends from ordinary least squares (OLS) regression applied to seasonal averages from the raw data and 2) trends from OLS regression applied to GAM seasonal averages.
-Select examples were used where differences were pronounced to illustrate false positive or negatives that may occur with alternative methods.
+Select examples where differences were pronounced were used for illustration.
 This analysis was then applied to all stations.
+The method that formally propagates uncertainty should have more robust statistical properties, such as accurate confidence interval coverage, than naïve methods.
+For this reason, even when results are similar across methods, the more robust method provides the best support for those results.
 
@@
 
-Model predictions for chl-a trends across all stations had an average R-squared value of `r aver2`%  (Table \@ref(tab:modprftab)) and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
+Model predictions for chl-a trends across all stations had an average R-squared value of `r aver2`%  (Table \@ref(tab:modprftab)) and range from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`).
 GAM predictions from north to south showed more pronounced annual and seasonal changes in chl-a towards the more southern stations (Figures S1-S9).
@@
 
-Estimates of linear trends in seasonal averages across roughly ten-year windows for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
+Estimates from the seasonal trend analyses (mixed-effects meta-analysis regressions) across roughly ten-year windows for different seasons are shown for station 34 (Figure \@ref(fig:trnddat)).
 Plots a-c show trends in January to June averages while plots d-f show trends in July to December averages.
-The seasonal trend analyses showed that January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.05 95\% confidence interval) from 1991 to 2000, whereas July to December chl-a did not change significantly.
+January to June chl-a increased (log$_{10}$ chl-a slope 0.03 $\mu$g L$^{-1}$ yr$^{-1}$, 0.01-0.05 95\% confidence interval) from 1991 to 2000, whereas July to December chl-a did not change significantly.
 Chl-a also increased from 2000 to 2010, but only for July to December (log$_{10}$ slope 0.03, 0-0.05 95\% confidence interval).
-Finally, chl-a decreased from 2010 to 2019 but only for July to December (log$_{10}$ chl-a slope -0.02, -0.04-0 95\% confidence interval).
+Finally, chl-a decreased from 2010 to 2019 but again only for July to December (log$_{10}$ chl-a slope -0.02, -0.04-0 95\% confidence interval).
 Because the trends were confined to certain times of the year, the seasonal estimates provide additional information beyond coarser estimates that cover the entire year.
@@
 Figure \@ref(fig:trndmap) shows results from similar analyses as those in Figure \@ref(fig:trnddat), but applied to all stations.
-Mixed-effects meta-analysis regressions applied to seasonal averages showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32 and 34 from 1991 to 2000; decreases were observed at stations 30 and 32 from 2010 to 2019.
+The seasonal trend analyses showed that increases (based on $p < 0.05$) for the January to June period were observed at stations 32 and 34 from 1991 to 2000; decreases were observed at stations 30 and 32 from 2010 to 2019.
 For the July to December period, increases were observed at stations 24, 27, 30, and 32 from 1991 to 2000 and stations 18, 21, 22, 24, and 34 from 2000 to 2010, whereas decreases were observed at stations 30, 32, and 34 from 2010 to 2019.
@@
 In particular, the most southern stations (32, 34, 36) had increasing trends prior to 2005 that were more often observed in the July to December period.
-Additionally, chl-a at the more northern stations has not changed in recent years for both seasonal periods.
+Additionally, chl-a at the more northern stations has not changed in recent years for either seasonal period.
 For most stations and seasonal periods, a change from increasing to decreasing chl-a occurred around 2007.
@@
 
-Results showing trend estimates from meta-analysis on GAM seasonal estimates provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM estimates without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2).
+Results showing trend estimates from meta-analysis on GAM seasonal averages provided different conclusions than those from either OLS regression through seasonal averages from raw data (Figure \@ref(fig:trndcmpex) row 1) or OLS regression through GAM seasonal averages without uncertainty propagation (Figure \@ref(fig:trndcmpex) row 2).
 Figure \@ref(fig:trndcmpex)a shows trend estimates for station 30 for January to June averages from 2000 to 2010.
@@
 The OLS regression on observed estimates (top plot) and OLS regression on GAM estimates (middle plot) did not identify trends.
-Figure \@ref(fig:trndcmpex)b shows trend estimates station 34 for July to December averages from 1991 to 2000.
+Figure \@ref(fig:trndcmpex)b shows trend estimates at station 34 for July to December averages from 1991 to 2000.
 Unlike the first example, only the middle figure shows a trend, whereas the top and bottom plots do not show trends.
@@
 As noted above, non-parametric approaches (i.e., seasonal Kendall tests) are by far the most common trend analysis methods applied to long-term water quality data [@Hirsch82;@Helsel20].
-These methods only assess the direction and significance of comparisons between year pairs, and importantly, do not account for full propagation of uncertainty inherent in raw observations if the raw data are aggregated to meet test requirements.
+These methods only assess the direction and significance of comparisons across years, and importantly, do not account for full propagation of uncertainty inherent in raw observations if the raw data are aggregated to meet test requirements.
 Aggregation of raw data, e.g., averaging of observations within a year or season to comply with the requirements of Kendall tests, risks loss of information by removing variation between observations at smaller time scales.
@@
 Our examples in Figures \@ref(fig:trndcmpex) and \@ref(fig:trndcmp10) demonstrate these risks if propagation of uncertainty from raw observations across methods is unaccounted for in trend assessment.
-Our assessment of trends using OLS regression applied to seasonal averages from the raw observations is effectively like averaging results within a year and applying a simple Kendall test.
-In many cases the results may be similar, but loss of information with averaging can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
-Moreover, use of methods that fully account for uncertainty in sampling design is recommended over more conventional analyses given the certainty of conclusions in the former.
-Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and at much larger spatial scales across all stations if simpler trend analyses are used.
-Even though simpler methods may produce the same result, particularly with frequent sampling and similar effort between years, decision-makers require certainty of conclusions when incorrect results can have major consequences for environmental management.
+The "naïve" method using OLS regression applied to seasonal averages from the raw observations fails to propagate uncertainty, similarly to averaging results within a year and applying a simple Kendall test.
+In some cases the results may be similar to those from fully propagating uncertainty, but the loss of information can lead to increased Type I or II error rates depending on characteristics of the raw data and the method used for their evaluation [@Shabman03].
+Our examples demonstrated the increased potential for incorrect conclusions at specific monitoring locations and, at larger spatial scales, across all stations if simpler trend analyses are used.
+Even though simpler methods may produce similar results in some cases, particularly with frequent sampling and similar effort between years, the only way to confirm such an outcome would be to compare results, relying on the method with full propagation of uncertainty to be the more robust method.
+Use of methods that fully account for uncertainty is recommended to obtain statistically valid results in a wider range of conditions.
 
@@
 This is especially problematic for policy analysis or regulatory decisions if the results change based on arbitrary decisions of the analyst.
-Because these decisions are not needed for GAMs, the results can be considered a more objective and potentially "true" signal of actual trends that are minimally influenced by process or observation error present in the raw data.
+Because these decisions are not needed for GAMs, the results can be considered a more objective and potentially accurate signal of actual trends that are minimally influenced by process or observation error present in the raw data.
+
+Several limitations of the proposed methods deserve mention.
+First, if sampling is so irregular that important fluctuations are missed entirely in some years, the GAM estimates and uncertainty propagation could become dubious in interpretation and usefulness.
+Second, estimation of GAMs for very long series can be computationally demanding.
+When this is an obstacle, one could do the first two analysis stages using temporal windows of the full data, with the only implication being that different degrees of smoothness may be estimated for different windows, which indeed might be justified by the data.
+Third, meta-analysis regression results for a very small number of years, particularly confidence intervals and associated p-values, may be inaccurate (e.g.
+in confidence interval coverage).
+In such cases, one could make alternative use of the GAM seasonal averages and standard errors, such as for pairwise comparisons among years.
 
@@
 Our third stage analyses require *a priori* decisions on long-term time scales of interest and future work could generalize these choices.
-Although there are undoubtedly many scenarios where years of interest can be chosen objectively by the needs of an analysis (e.g., regulatory compliance periods, time since management intervention), a more general question of "when" changes occur independent of user decisions is also important to address.
+Although there are undoubtedly many scenarios where years of interest can be chosen objectively by the needs of an analysis (e.g., regulatory compliance periods, time since management intervention), a more general question of when changes occur independent of user decisions is also important to address.
 Additional methods could be developed using objective criteria to identify inflection points or other important periods where changes occur independent of a user choice.
@@
 
-Additional explanatory variables could be identified that may be associated with the trend after the trend has been adequately described.
-This information has obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.
-An advantage of GAMs is their flexibility in including alternative predictors, such that the significance of a predictor or comparison of nested models with and without different predictors can provide evidence of which predictors are driving the observed trends [@Wood02;@Zuur09].
-In such cases, considerations of model structure can have direct implications on conclusions given how GAMs could be used to assess different questions.
-Our goal was to describe chl-a changes relative to time, where the predictors were variations on a general theme (e.g., season vs.
-year).
-This is a different application from using GAMs with predictors selected to explain those changes over time.
-Therefore, using our approach to evaluate explanatory variables will require testing of different model structures.
+Additional explanatory variables could be identified to explain trends in either the GAM stage or the meta-analysis stage of analysis.
+This information would have obvious implications for management decisions on factors that influence water quality changes, e.g., wastewater treatment upgrades, large-scale climatic factors, or flow regulation practices.
+Including alternative predictors in the GAMs [@Wood02;@Zuur09] could reduce the uncertainties of its estimates and, if relevant, allow the influence of those variables to be removed.
+Including alternative predictors in the meta-analysis could help in explaining long-term trends of seasonal averages or other metrics obtained from the GAMs.
+Our goal here was to describe chl-a changes relative to time, so the single predictor in both modeling stages was time.
 

### 2021-07-12 revision/review_response.Rmd
formatting revision for resubmission

@@
 
-     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
+     In the manuscript, lines 165 - 167: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
 
@@
 
-     Lines xyz to xyz: "Several limitations of the proposed methods deserve mention.
+     Lines 492 - 501: "Several limitations of the proposed methods deserve mention.
 First, if sampling is so irregular that important fluctuations are missed entirely in some years, the GAM estimates and uncertainty propagation could become dubious in interpretation and usefulness.
@@
 Otherwise, a user cannot be certain the correct conclusions are obtained using simpler methods.
-The paragraph on lines xyz to xyz was revised to address these points:
+The paragraph on lines 458 to 473 was revised to address these points:
 
@@
 
-     Line xyz: "Meta-analysis regression incorporates a known (or estimated) standard error for each response datum.
+     Lines 114 - 123: "Meta-analysis regression incorporates a known (or estimated) standard error for each response datum.
 Usually meta-analysis is used when each response summarizes a dataset from a separate study, along with its standard error, and the meta-analysis looks for across-study patterns in effect size (i.e., Lortie et al.

### 2021-07-12 manu_draft.Rmd
formatting revision for resubmission

@@
 We thank James D.
-Hagy III for reviewing an earlier draft of this manuscript.
+Hagy III and two anonymous reviewers for providing helpful comments on this manuscript.
 
@@
 # save as tiff
-tiff('submission/Fig1.tiff', height = 7, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+tiff('revision/Fig1.tiff', height = 7, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 print(p)
@@
 # save as tiff
-tiff('submission/Fig2.tiff', height = 6.5, width = 5, units = 'in', family = 'serif', res = 600, compression = 'lzw')
+tiff('revision/Fig2.tiff', height = 6.5, width = 5, units = 'in', family = 'serif', res = 600, compression = 'lzw')
 print(p)
@@
 # save as tiff
-tiff('submission/Fig3.tiff', height = 7, width = 8, units = 'in', family = 'serif', res = 600, compression = 'lzw')
+tiff('revision/Fig3.tiff', height = 7, width = 8, units = 'in', family = 'serif', res = 600, compression = 'lzw')
 print(out)
@@
 
-ylim <- c(0, 21)
+ylim <- c(0, 24)
 ylab <- expression(paste("Chl-a (", mu, "g ", L^-1, ")"))
@@
 # save as tiff
-tiff('submission/Fig4.tiff', height = 6, width = 11, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+tiff('revision/Fig4.tiff', height = 6, width = 11, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 print(p)
@@
 # save as tiff
-tiff('submission/Fig5.tiff', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+tiff('revision/Fig5.tiff', height = 6.5, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 p
@@
 # save as tiff
-tiff('submission/Fig6.tiff', height = 8, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+tiff('revision/Fig6.tiff', height = 8, width = 8, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 print(p)
@@
 # save as tif
-tiff('submission/Fig7.tiff', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+tiff('revision/Fig7.tiff', height = 6.5, width = 7.75, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 p
@@
 # save as tiff
-tiff('submission/Fig8.tiff', height = 6, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
+tiff('revision/Fig8.tiff', height = 6, width = 7.5, family = 'serif', units = 'in', res = 600, compression = 'lzw')
 p
@@
 In practice, a sufficiently large number of knots provided to the additive terms will produce identical or comparable estimates for a response variable.
-The models differ in the allocation of penalties for the smoothness of each spline (`s()`).'
+The models differ in the allocation of penalties for the smoothness of each spline.'
 

### 2021-07-28 revision/review_response.Rmd
ds edits included, proof all else, ready for resubmit

--- 2021-07-06 revision/review_response.Rmd
@@
 
-     In the manuscript, line xyx: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
+     In the manuscript, lines 189 - 191: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
 
@@
 
-     Lines xyz to xyz: "Several limitations of the proposed methods deserve mention.
+     Lines 527 - 536: "Several limitations of the proposed methods deserve mention.
 First, if sampling is so irregular that important fluctuations are missed entirely in some years, the GAM estimates and uncertainty propagation could become dubious in interpretation and usefulness.
@@
 Otherwise, a user cannot be certain the correct conclusions are obtained using simpler methods.
-The paragraph on lines xyz to xyz was revised to address these points:
+The paragraph on lines 493 to 508 was revised to address these points:
 
@@
 
-     Line xyz: "Meta-analysis regression incorporates a known (or estimated) standard error for each response datum.
+     Lines 125 - 135: "Meta-analysis regression incorporates a known (or estimated) standard error for each response datum.
 Usually meta-analysis is used when each response summarizes a dataset from a separate study, along with its standard error, and the meta-analysis looks for across-study patterns in effect size (i.e., Lortie et al.
@@
 Thus, while meta-analysis methods are most commonly associated with combining results from multiple studies into a larger analysis, their key modeling step is propagation of uncertainty (Gasparrini et al., 2012; Sera et al., 2019).
-To do this, meta-analysis makes use of a known (estimated) standard error for each response datum, which is exactly the same step required here to propagate standard errors from the GAM into a regression of seasonal averages."
+To do this, meta-analysis makes use of a known (estimated) standard error for each response datum, which is the required priority here to propagate standard errors from the GAM into a regression of seasonal averages."

### 2021-07-28 revision/review_response.Rmd
revision resubmitted

@@
 
-     In the manuscript, lines 189 - 191: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
+     In the manuscript, lines 194 - 196: "Approximate monthly or biweekly sampling with coverage of at least a decade is common for many long-term monitoring programs and is the motivating use case for the methods herein."
 
@@
 
-     Lines 527 - 536: "Several limitations of the proposed methods deserve mention.
+     Lines 532 - 541: "Several limitations of the proposed methods deserve mention.
 First, if sampling is so irregular that important fluctuations are missed entirely in some years, the GAM estimates and uncertainty propagation could become dubious in interpretation and usefulness.
@@
 Otherwise, a user cannot be certain the correct conclusions are obtained using simpler methods.
-The paragraph on lines 493 to 508 was revised to address these points:
+The paragraph on lines 498 to 513 was revised to address these points:
 
